### 2020-12-10 paper.tex
Add skeleton of the paper

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}

\title{Union and intersection contracts: theory vs practice}
\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}

\begin{document}

\maketitle

\begin{abstract}
 In this paper we do some cool stuff.
\end{abstract}

\section*{Introduction}
Intro

\section{First section}
Some text

\section{Second section}
Some text

\section{Third section}
Some text

\section*{Conclusion}
Conclusion

\end{document}

### 2021-02-15 paper.tex
WIP

@@
 
-\section{First section}
-Some text
+\section{Background on union types}
 
-\section{Second section}
-Some text
+\subsection{Nickel design space}
 
-\section{Third section}
-Some text
+Make clear the design of Nickel aims at being simple, interoperable with JSON,
+and to not type every possible untyped idioms, but rather the common typed one...
 
-\section*{Conclusion}
-Conclusion
+\subsection{Tagged versus untagged}
+
+Unions are fundamental and ubiquitous pieces of programming. While products
+represent conjunctions, which are several pieces of data put together, unions
+represent disjunctions, that are different alternatives values for the same
+data. Put together, they allow to model a large class of data types.
+
+\\
+EXAMPLES of union in configuration specification
+\\
+
+In statically typed languages, more specifically in functional programming
+languages, they are usually represented as algebraic data types [references],
+also called *tagged unions*. Tagged means that there is an additional field, or
+constructor, that allows to discriminate between each alternatives, usually at
+compile-time. Well-known programming languages with first-class tagged unions
+include Haskell, OCaml, Rust, Scala, PureScript, Elm, Swift and a even greater
+number of related and derived languages.
+
+\\
+EXAMPLES
+\\
+
+All of the aforementioned languages are statically typed. In this setting, the
+actual representation of a tagged union is an implementation detail: from the
+point of view of the programmer, a tagged union is a proper value 
+associated with a clearly defined semantics. However, in a gradually typed language like Nickel,
+the question of how tagged union are represented as untyped values arises. One
+problem with tagged union is that it is a fundamentally typed representation,
+which has no totally natural counterpart as an untyped data.
+
+One usual representation is as a structure with a tag and a value which actual
+shape depends on the tag:
+
+\\
+FIGURE OR C CODE EXAMPLE ILLUSTRATING THE USUAL REPRESENTATION
+\\
+
+But this is neither totally natural, nor the only possible solution: 
+. More importantly, this is not a natural representation for schemas, which
+favors simple, untagged union. Indeed, if one can determine the type at
+run-time, and there is no typy stuffy, the tag sounds rather useless.
+
+In an untyped language, it is thus natural to directly stores different data in a
+variable: actually, the runtime system is the one in charge of tags, which it
+makes accessible via functions such as JavaScript's
+\begin{verbabtim}typeof\end{verbatim}. Using so-called untagged unions has already
+been an idiom in practice for a long time, and being able to type and represent
+these unions is an important aspect of a type system applying to these
+languages.
+
+This is the reason why other gradually typed languages, which were either
+originally untyped or aim at backward compatibility with an untyped base
+language, seem to favor supporting union types which are the typed counter-part
+of these untagged union [example of gradually typed language with untagged
+union].
+
+In the case of Nickel, there is no pre-existing untyped language that we have to
+support, so we are freer of our choice. But are we? As a configuration language,
+Nickel does not evolve in the abstract. It aims at maximum interoperability with
+JSON. It is designed to generate data for other software to consume, and strives
+to be able to encode as much schemas as possible inside the language itself. If
+these schemas use untagged union, then this is a strong incentive to support
+them at the type level, or at least, at the contract level.
+
+\subsection{Union types}
+
+\subsection{Tagged as untagged}
+
+Note that together with the other existing features of the type system, one can
+encode tagged union in the following way:
+
+\\
+Encoding as: use singleton contract for tags, and then union.
+\begin{lstlisting}
+type sum = A of Num | B of Str
+\end{lstlisting}
+
+\begin{lstlisting}
+let Singleton = fun val l x => if val == x then x else blame l in
+let A = Singleton `A in
+let B = Singleton `B in
+{ tag : #A, val : Num} | { tag : #B, val : Str} 
+\end{lstlisting}
+
+This is just the encoding mentioned in the previous section. But it suffers a
+fundamental problem: the static type system is not able to deduce when the value
+is known to be a `Num` or a `Str`.
+
+\\
+EXAMPLE if x.tag == `A then x.val + 1 else if val.tag == `B then strings.fromNum
+x.val
+\\
+
+TypeScript is capable of emulating such union (see ??).
 

### 2021-02-18 paper.tex
WIP

--- 2020-12-10 paper.tex
@@
 \usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
+\usepackage{listings}
 \usepackage{xcolor}
@@
 
-\section{First section}
-Some text
+\section{Background on union types}
 
-\section{Second section}
-Some text
+\subsection{Nickel design space}
 
-\section{Third section}
-Some text
+\color{red}Goals of the section
 
-\section*{Conclusion}
-Conclusion
+Introduce Nickel, explain the choice of gradual
+typing and main design orientations, such as practice-oriented, lightweight,
+etc.\vspace{0.5cm}\color{black}
+
+This paper is centered around the design of the Nickel language. Nickel is a
+configuration programming language, meaning that a program eventually evaluates
+to a value that can be directly serialized to JSON, YAML or alike. Nickel aims at
+being lightweight, easy to write and easy to embed in a larger application. The
+focus is first and foremost practical: it originated from the language of the Nix
+package manager, used to describe the biggest set of software packages ever
+built [ ], and targets build systems and cloud deployment tools (Kubernetes,
+Terraform or NixOps) as well. While there's no hesitation in incorporating recent
+research when it serves a purpose, this is not a research language, as the
+design decisions are guided by the industrial use cases.
+
+Our aim is that writing simple Nickel should not be more complex than writing
+JSON or YAML.  Unlike JSON, though, Nickel anticipates large configurations by
+being both programmable and typed. There's a dilemma with respect to typing: if
+your program is only run once on fixed inputs, as are such programmable
+configurations, any type error will be reported at run-time anyway. Why bother
+with a static type system? On the other hand, more and more systems rely on
+complex configurations, like cloud infrastructure, leading the corresponding
+configurations to become increasingly complex, to the point where static types
+are beneficial. For reusable code — that is, library functions — static types
+brings all the usual, widely acknowledged benefits. This dilemma is naturally
+solved by an approach that can reunite static typing and dynamic typing: gradual
+typing [references]. Unlike language like TypeScript, we don't seek to provide a
+complex type system that tries hard to type most of the idioms arising naturally
+in dynamically typed code. Rather, we provide a reasonably solid base type
+system with good inference properties, that should make writing typed generic
+function operating on data easy. In fact, for more complex data validation,
+Nickel uses contracts rather static typing.
+
+\subsection{Unions types}
+\color{red}Goals of the section
+
+Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}
+
+In programming, unions are fundamental and ubiquitous building blocks of data.
+While products represent conjunctions, which are several different pieces of
+data stored all together in one value, unions represent disjunctions, which are
+several exclusive alternatives stored in one value. When nested together, unions
+and intersections are able to describe a large class of data shapes [ref,
+examples?].
+
+\begin{verbatim}
+EXAMPLES of union in configuration specification
+\end{verbatim}
+
+In statically typed languages, and more specifically in functional programming
+languages, unions are usually embodied by algebraic data types [references],
+also called \textbf{tagged unions}. Tagged means that there is an additional
+field, or constructor, that allows to discriminate between each alternative,
+usually via pattern matching. Well-known programming languages with first-class
+tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm, Swift and an
+even greater number of related and derived languages.
+
+\begin{verbatim}
+EXAMPLES
+\end{verbatim}
+
+\subsection{Tagged versus untagged}
+\color{red}Goals of the section
+
+Interpreted languages already need to tag their values, hence bare union are more natural.
+It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}
+
+In this statically typed setting, the actual representation of a tagged union is
+an implementation detail: from the point of view of the programmer, a tagged
+union is a fist-class data type associated with constructors, destructors and a
+well-defined semantics.  However, in a gradually typed language like Nickel, the
+question of the representation of tagged unions as untyped values arises.
+
+Unfortunately, it seems that tagged unions do not have a canonical counterpart
+as serializable untyped data. An intuitive representation is as a record with a
+tag together with a value which shape depends on the tag:
+
+\begin{lstlisting}
+{ tag: <A, B, C>, value: Dyn }
+\end{lstlisting}
+
+Another way of seeing it is a dependent $\Sigma$-type which first component is a
+tag and second component is the data type corresponding to this tag:
+\[
+    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
+\]
+
+Alas, choosing this representation as a default is not fully satisfying. As any
+representation, it does not correspond precisely to, say, a JSON type. In
+consequence, a lot of illegal combinations are structurally valid untyped (JSON)
+values. More importantly, this does not match with the configuration schemas one
+can find in the wild favors simple, untagged union [example of JSON
+schemas/protobuf/whatever with unions].
+
+In dynamically typed languages, which are interpreted for the vast majority
+[ref/examples?], it is more natural to directly stores different kind of data in
+one variable without further ceremony. The runtime system needs to attach type
+tags to values anyway, which usually are accessible via functions like the
+JavaScript's \verb+typeof+. Using so-called untagged unions has already been an
+idiom in dynamically typed language for a long time [ref/examples], and being
+able to type and handle these unions an important aspect of the migration to a
+gradual type system.
+
+Most of gradually typed languages either existed originally as dynamically typed
+language or aim at backward compatibility with a dynamically typed base
+language. This is the reason they tend to favor supporting union types, the
+typed model of the practice of simply storing different kind of data in one
+variable [example of gradually typed language with untagged union].
+
+In the case of Nickel, there is no pre-existing untyped language that we have to
+support, so we are freer of our choice. But are we? As a configuration language,
+Nickel does not evolve in the abstract. It aims at best interoperability with
+JSON and alike. It is designed to generate data for other softwares to consume,
+and tries as much as possible to be able to encode data schemas at the language
+level. If these schemas use untagged union, then this is a strong incentive to
+support them at the type level, or at least, at the contract level.
+
+\subsection{Union types}
+
+\subsection{Tagged as untagged}
+\color{red}Goals of the section
+
+Untagged unions, modulo a slighty smarter typechecker, can encode usual type-safe
+tagged unions. They let the user free of the representation though, and encode
+more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}
+
+Untagged union, together with the other existing features of the type system,
+are sufficient to implement the encoding mentioned in [cite previous section]:
+
+Encoding as: use singleton contract for tags, and then union.
+\begin{lstlisting}
+type sum = A of Num | B of Str
+\end{lstlisting}
+
+\begin{lstlisting}
+{ tag : <A>, val : Num} | { tag : <B>, val : Str}
+\end{lstlisting}
+
+As is, the static type system able to refine the type of the value though,
+making the following example rejected:
+
+\begin{lstlisting}
+switch x.tag {
+  A => x.val + 1,
+  B => then strings.fromNum x.val
+}
+\end{lstlisting}
+
+But TypeScript is capable of typing this kind of things [ ], and one can imagine
+doing the same in Nickel. Equipped with tagged union and a typechecker smart
+enough, classic tagged unions can be simulated. But the representation is not
+imposed by the language. Untagged unions empower the programmer to represent and
+handle a wider variety of unions appearing in the wild. Thus, adding union types
+(and their natural and unavoidable [ref??] counterpart, intersections) to Nickel
+seems very appealing.
+
+\subsection{Union and intersections}
+\color{red}Goals of the section
+
+Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}
 

### 2021-03-01 paper.tex
Skeleton and first few parts
WIP

--- 2020-12-10 paper.tex
@@
 
-\section{First section}
-Some text
+\section{Background on union types}
 
-\section{Second section}
-Some text
+\subsection{Nickel design space}
 
-\section{Third section}
-Some text
+\color{red}Goals of the section
 
-\section*{Conclusion}
-Conclusion
+Introduce Nickel, explain the choice of gradual
+typing and main design orientations, such as practice-oriented, lightweight,
+etc.\vspace{0.5cm}\color{black}
+
+This paper is centered around the design of the Nickel language. Nickel is a
+configuration programming language, meaning that a Nickel program eventually
+evaluates to a value that can be directly serialized to JSON, YAML or alike.
+Nickel aims at being lightweight, easy to write and easy to embed in a larger
+system, to serve either directly as a native language to specify configuration,
+or as a meta-language to generate configuration. The focus is first and foremost
+practical: Nickel originated as a rethink of the language of the Nix package
+manager, used to describe the biggest set of software packages ever built [ ],
+and targets build systems and cloud deployment tools (Kubernetes, Terraform or
+NixOps) as well.  While there's no hesitation in incorporating recent research
+when it serves a well-defined purpose, the design decisions are guided by the
+industrial use cases.
+
+Our aim is that writing simple Nickel should not be more complex than writing
+JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
+being both programmable and typed. In the configuration setting, there is a
+singular dilemma with respect to typing: if your program is only run once on
+fixed inputs, as are such programmable configurations, any type error will be
+reported at run-time anyway. Why bother with the complexity of a static type
+system? On the other hand, more and more software systems offload complexity to
+configurations, such as cloud infrastructure, build system, or operating
+systems, leading the corresponding configurations to become increasingly
+complex, to the point where static types are beneficial. For reusable code —
+that is, library functions — static types brings all the usual, widely
+acknowledged benefits.
+
+This dilemma is naturally solved by gradual typing [], an approach capable of
+reuniting static typing and dynamic typing. Unlike gradually-typed language like
+TypeScript though, we do not seek to build a complex type system that tries hard
+to type most of the idioms naturally arising in dynamically typed code.  Rather,
+we choose to provide a reasonably expressive type system with good inference
+properties, that should make writing typed generic function that operate on data
+relatively easy, but may require resorting to untyped code for more exotic expressions.
+However, this is not a surrender, as in fact Nickel provides an alternative and
+complementary mechanism for more complex data validation, contracts[references].
+
+\subsection{Unions types}
+\color{red}Goals of the section
+
+Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}
+
+In programming, unions are fundamental and ubiquitous building blocks of data,
+and Nickel is no exception. While products represent conjunctions, which are several pieces of
+data stored all together in the same value, unions represent disjunctions, which are
+several exclusive alternatives stored in one value. When nested together, unions
+and intersections are able to describe a large class of data shapes: lists,
+trees, structured errors, and so on [ref, examples?].
+
+\begin{verbatim}
+EXAMPLES of union in configuration specification
+\end{verbatim}
+
+In statically typed languages, and more specifically in functional programming
+languages, unions are usually embodied by algebraic data types [references],
+also called \textbf{tagged unions}. Tagged means that there is an additional
+field, or constructor, that allows to discriminate between each alternative.
+When an union is constructed, the programmer specifies which alternative is
+chosen by using the corresponding tag. When a union is consumed, usually
+via pattern matching, the code branches depending on the tag, and each branch is
+ensured to handle data of the right shape. Well-known programming languages with
+first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
+Swift and surely many more.
+
+\begin{verbatim}
+EXAMPLES
+\end{verbatim}
+
+\subsection{Tagged versus untagged}
+\color{red}Goals of the section
+
+Interpreted languages already need to tag their values, hence bare union are more natural.
+It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}
+
+In the statically typed setting, the actual representation of a tagged union is
+an implementation detail: from the point of view of the programmer, a tagged
+union is a fist-class data type associated with constructors and destructors all
+with well-defined semantics.  However, in a gradually typed language like
+Nickel, the question of the representation of tagged unions as untyped values
+arises.
+
+Unfortunately, it seems that tagged unions do not have a canonical counterpart
+as serializable untyped data. A natural representation is a record with a tag
+and a value which shape depends on the tag:
+
+\begin{lstlisting}
+{ tag: <A, B, C>, value: Dyn }
+\end{lstlisting}
+
+A better way to think of it is maybe as dependent $\Sigma$-type which first
+component is a tag and second component is the data type corresponding to this
+tag:
+\[
+    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
+\]
+
+Alas, choosing this representation as a default is not fully satisfying. As any
+representation, it does not map precisely to a JSON type. In consequence, a lot
+of illegal combinations are structurally valid as untyped JSON values. More
+importantly, this does not necessarily match with the configuration schemas one
+can find in the wild, that favor simple, untagged union [example of JSON
+schemas/protobuf/whatever with unions].
+
+In dynamically typed languages, which are interpreted for the vast majority
+[ref/examples?], it is customary to directly store different kind of data in one
+variable without further ceremony. The runtime system needs to attach type tags
+to values anyway, which can usually be observed using functions like the
+JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
+unions} has already been an idiom in dynamically typed language for a long time
+[ref/examples], and being able to type and handle these unions is an important
+aspect of the migration to a gradual type system. Indeed, most of gradually
+typed languages either existed originally as dynamically typed language or aim
+at backward compatibility with a dynamically typed base language. This is why
+they tend to favor supporting untagged union types, the typed counterpart of
+this practice [example of gradually typed language with untagged union].
+
+In the case of Nickel, there is no pre-existing untyped language that we have to
+support, so we are freer of our choice. But are we? As a configuration language,
+Nickel does not evolve in the abstract. It aims at the best interoperability
+possible with JSON. It is designed to generate data for other softwares to
+consume, and tries as much as possible to be expressive enough to encode data
+schemas at the language level. If these schemas use untagged union, then this is
+a strong incentive to support them at the type level, or at least, at the
+contract level.
+
+\subsection{Union types}
+
+\subsection{Tagged as untagged}
+\color{red}Goals of the section
+
+Untagged unions, modulo a slightly smarter typechecker, can encode usual type-safe
+tagged unions. They let the user free of the representation though, and encode
+more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}
+
+Untagged union, together with the other existing features of the type system,
+are sufficient to implement the encoding mentioned in [cite previous section]:
+
+Encoding as: use singleton contract for tags, and then union.
+\begin{lstlisting}
+type sum = A of Num | B of Str
+\end{lstlisting}
+
+\begin{lstlisting}
+{ tag : <A>, val : Num} | { tag : <B>, val : Str}
+\end{lstlisting}
+
+There is a twist, though. As it is, the static type system is unable to refine
+the type of the value, making the following example rejected:
+
+\begin{lstlisting}
+switch x.tag {
+  A => x.val + 1,
+  B => then strings.fromNum x.val
+}
+\end{lstlisting}
+
+Flow typing[ ], as implemented for example by TypeScript, is capable of solving
+this issue by correctly propagating the type information in each branch. One can
+imagine doing the same in Nickel. Thus equipped with untagged unions and a
+typechecker smart enough, classic tagged unions can be simulated, but with the
+benefit that the language is not imposing any representation. Untagged unions
+empower the programmer to represent and handle a wider variety of unions
+appearing in the wild. Thus, adding union types (and their natural and
+unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.
+
+\subsection{Union and intersections}
+\color{red}Goals of the section
+
+Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}
+
+\section{Contracts}
+\color{red}Goals of the section
+
+A primer on contracts, which in the end are the problematic bit when confronted
+with unions.\vspace{0.5cm}\color{black}
 

### 2021-03-23 paper.tex
Drafted WMW parts

--- 2021-02-15 paper.tex
@@
 \begin{abstract}
- In this paper we do some cool stuff.
+ In this paper we try to do some cool stuff,
+ but we mainly complain about uncool stuff.
 \end{abstract}
@@
 
-\section{Background on union types}
+\section{Challenges}
+\input{challenges}
 
-\subsection{Nickel design space}
-
-Make clear the design of Nickel aims at being simple, interoperable with JSON,
-and to not type every possible untyped idioms, but rather the common typed one...
-
-\subsection{Tagged versus untagged}
-
-Unions are fundamental and ubiquitous pieces of programming. While products
-represent conjunctions, which are several pieces of data put together, unions
-represent disjunctions, that are different alternatives values for the same
-data. Put together, they allow to model a large class of data types.
-
-\\
-EXAMPLES of union in configuration specification
-\\
-
-In statically typed languages, more specifically in functional programming
-languages, they are usually represented as algebraic data types [references],
-also called *tagged unions*. Tagged means that there is an additional field, or
-constructor, that allows to discriminate between each alternatives, usually at
-compile-time. Well-known programming languages with first-class tagged unions
-include Haskell, OCaml, Rust, Scala, PureScript, Elm, Swift and a even greater
-number of related and derived languages.
-
-\\
-EXAMPLES
-\\
-
-All of the aforementioned languages are statically typed. In this setting, the
-actual representation of a tagged union is an implementation detail: from the
-point of view of the programmer, a tagged union is a proper value 
-associated with a clearly defined semantics. However, in a gradually typed language like Nickel,
-the question of how tagged union are represented as untyped values arises. One
-problem with tagged union is that it is a fundamentally typed representation,
-which has no totally natural counterpart as an untyped data.
-
-One usual representation is as a structure with a tag and a value which actual
-shape depends on the tag:
-
-\\
-FIGURE OR C CODE EXAMPLE ILLUSTRATING THE USUAL REPRESENTATION
-\\
-
-But this is neither totally natural, nor the only possible solution: 
-. More importantly, this is not a natural representation for schemas, which
-favors simple, untagged union. Indeed, if one can determine the type at
-run-time, and there is no typy stuffy, the tag sounds rather useless.
-
-In an untyped language, it is thus natural to directly stores different data in a
-variable: actually, the runtime system is the one in charge of tags, which it
-makes accessible via functions such as JavaScript's
-\begin{verbabtim}typeof\end{verbatim}. Using so-called untagged unions has already
-been an idiom in practice for a long time, and being able to type and represent
-these unions is an important aspect of a type system applying to these
-languages.
-
-This is the reason why other gradually typed languages, which were either
-originally untyped or aim at backward compatibility with an untyped base
-language, seem to favor supporting union types which are the typed counter-part
-of these untagged union [example of gradually typed language with untagged
-union].
-
-In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. But are we? As a configuration language,
-Nickel does not evolve in the abstract. It aims at maximum interoperability with
-JSON. It is designed to generate data for other software to consume, and strives
-to be able to encode as much schemas as possible inside the language itself. If
-these schemas use untagged union, then this is a strong incentive to support
-them at the type level, or at least, at the contract level.
-
-\subsection{Union types}
-
-\subsection{Tagged as untagged}
-
-Note that together with the other existing features of the type system, one can
-encode tagged union in the following way:
-
-\\
-Encoding as: use singleton contract for tags, and then union.
-\begin{lstlisting}
-type sum = A of Num | B of Str
-\end{lstlisting}
-
-\begin{lstlisting}
-let Singleton = fun val l x => if val == x then x else blame l in
-let A = Singleton `A in
-let B = Singleton `B in
-{ tag : #A, val : Num} | { tag : #B, val : Str} 
-\end{lstlisting}
-
-This is just the encoding mentioned in the previous section. But it suffers a
-fundamental problem: the static type system is not able to deduce when the value
-is known to be a `Num` or a `Str`.
-
-\\
-EXAMPLE if x.tag == `A then x.val + 1 else if val.tag == `B then strings.fromNum
-x.val
-\\
-
-TypeScript is capable of emulating such union (see ??).
+\section*{Conclusion}
+Conclusion
 

### 2021-03-23 paper.tex
Add citations and subsection on contracts

--- 2021-03-01 paper.tex
@@
 \usepackage{tikz}
+\usepackage{url}
+
+\usepackage{biblatex}
+\addbibresource{nickel.bib}
 
@@
 
-This paper is centered around the design of the Nickel language. Nickel is a
-configuration programming language, meaning that a Nickel program eventually
-evaluates to a value that can be directly serialized to JSON, YAML or alike.
-Nickel aims at being lightweight, easy to write and easy to embed in a larger
-system, to serve either directly as a native language to specify configuration,
-or as a meta-language to generate configuration. The focus is first and foremost
-practical: Nickel originated as a rethink of the language of the Nix package
-manager, used to describe the biggest set of software packages ever built [ ],
-and targets build systems and cloud deployment tools (Kubernetes, Terraform or
+This paper is centered around the design of the Nickel
+language\cite{NickelRepo}. Nickel is a configuration programming language,
+meaning that a Nickel program eventually evaluates to a value that can be
+directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
+easy to write and easy to embed in a larger system, to serve either directly as
+a native language to specify configuration, or as a meta-language to generate
+configuration. The focus is first and foremost practical: Nickel originated as a
+rethink of the language of the Nix package manager, used to describe one of the
+biggest sets of software packages ever built\cite{repology}.
+Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform or
 NixOps) as well.  While there's no hesitation in incorporating recent research
@@
 singular dilemma with respect to typing: if your program is only run once on
-fixed inputs, as are such programmable configurations, any type error will be
-reported at run-time anyway. Why bother with the complexity of a static type
-system? On the other hand, more and more software systems offload complexity to
-configurations, such as cloud infrastructure, build system, or operating
-systems, leading the corresponding configurations to become increasingly
-complex, to the point where static types are beneficial. For reusable code —
-that is, library functions — static types brings all the usual, widely
-acknowledged benefits.
+fixed inputs and terminate, as are such programmable configurations, any type
+error will be reported at run-time anyway. Why bother with the complexity of a
+static type system? On the other hand, more and more software systems offload
+complexity to configurations, such as cloud infrastructure, build system, or
+operating systems, leading the corresponding configurations to become
+increasingly complex, to the point where static types are beneficial. For
+reusable code — that is, library functions — static types brings all the usual,
+widely acknowledged benefits.
 
-This dilemma is naturally solved by gradual typing [], an approach capable of
-reuniting static typing and dynamic typing. Unlike gradually-typed language like
-TypeScript though, we do not seek to build a complex type system that tries hard
-to type most of the idioms naturally arising in dynamically typed code.  Rather,
-we choose to provide a reasonably expressive type system with good inference
-properties, that should make writing typed generic function that operate on data
-relatively easy, but may require resorting to untyped code for more exotic expressions.
-However, this is not a surrender, as in fact Nickel provides an alternative and
-complementary mechanism for more complex data validation, contracts[references].
+This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}, an
+approach capable of reuniting static typing and dynamic typing. Unlike
+gradually-typed language like TypeScript, we do not seek to build a complex type
+system that tries hard to type most of the idioms naturally arising in
+dynamically typed code.  Rather, we choose to provide a reasonably expressive
+type system with good inference properties, that should make writing typed
+generic function that operate on data relatively easy, but may require resorting
+to untyped code for more exotic expressions.  However, this is not a surrender,
+as Nickel provides a complementary mechanism for more complex data validation,
+contracts.
+
+\subsection{Contracts}
+\color{red}Goals of the section
+
+A primer on contracts, which in the end are the problematic bit when confronted
+with unions.\vspace{0.5cm}
+
+Enforcing pre- and post-conditions at runtime is a widely established practice.
+In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
+Fellleisen introduces a principled approach to run-time predicate checking that
+nicely supports higher-order functions. It turns out these contracts are closely
+related to the type casts introduced by gradual typing. In
+\cite{WellTypedBlamed}, the authors precisely propose a system mixing gradual
+typing and contracts \textit{à la Findler \& Felleisen}. Nickel adopts a similar
+type system, with both statically typed terms, dynamically typed terms, and
+first-class contracts. In this setting, each construct of the type system -
+arrow types, record types, polymorphic types - must have a contract counterpart,
+that is able to check at run-time the membership of a value to the corresponding
+type.
+
+\color{black}
 
@@
 
-In programming, unions are fundamental and ubiquitous building blocks of data,
-and Nickel is no exception. While products represent conjunctions, which are several pieces of
-data stored all together in the same value, unions represent disjunctions, which are
-several exclusive alternatives stored in one value. When nested together, unions
-and intersections are able to describe a large class of data shapes: lists,
-trees, structured errors, and so on [ref, examples?].
+Unions are fundamental and ubiquitous building blocks of program data. Products
+represent conjunctions, that is several pieces of data whith potentially
+different shapes that are packed together in one value. Unions represent
+disjunctions, which are several exclusive alternatives stored in one value. When
+nested together, unions and intersections are able to represent aribrary
+tree-like structures, allowing to model a large class of data.
 
-\begin{verbatim}
-EXAMPLES of union in configuration specification
-\end{verbatim}
+Unions are also found in configuration, to represent nullable value in JSON,
+plain enumerations, or more general union types when several representations are
+supported.
 
@@
 When an union is constructed, the programmer specifies which alternative is
-chosen by using the corresponding tag. When a union is consumed, usually
-via pattern matching, the code branches depending on the tag, and each branch is
-ensured to handle data of the right shape. Well-known programming languages with
+chosen by using the corresponding tag. When a union is consumed, usually via
+pattern matching, code branches are tagged as well, and each branch is ensured
+to handle data of the right shape. Well-known programming languages with
 first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
@@
 
-\begin{verbatim}
-EXAMPLES
-\end{verbatim}
+\begin{lstlisting}[caption={Example of a binary tree as a tagged union in OCaml}, captionpos=b]
+type binary_tree =
+  | Node of 'a binary_tree * 'a binary_tree
+  | Leaf of int
+
+let sum =
+  let rec aux total = function
+  | Leaf val -> val
+  | Node(left,right) =>
+    let left = aux total left in
+    aux total right
+\end{lstlisting}
 
@@
 union is a fist-class data type associated with constructors and destructors all
-with well-defined semantics.  However, in a gradually typed language like
+with well-defined semantics. In contrast, in a gradually typed language like
 Nickel, the question of the representation of tagged unions as untyped values
@@
 
-A better way to think of it is maybe as dependent $\Sigma$-type which first
-component is a tag and second component is the data type corresponding to this
-tag:
+\lstinline+<A, B, C>+ is the type of a C-like enumeration that can take value
+$A$, $B$ or $C$, while \lstinline+Dyn+ is the unknown dynamic type.
+
+A better way to describe it would be as a dependent pair which first component
+is a tag and second component is the data type corresponding to this tag, but
+the language does not feature dependent types:
 \[
@@
 
-Alas, choosing this representation as a default is not fully satisfying. As any
-representation, it does not map precisely to a JSON type. In consequence, a lot
-of illegal combinations are structurally valid as untyped JSON values. More
-importantly, this does not necessarily match with the configuration schemas one
-can find in the wild, that favor simple, untagged union [example of JSON
-schemas/protobuf/whatever with unions].
+Alas, choosing the first representation as a default is not fully satisfying.
+First, it does not map precisely to a JSON type: in consequence, a lot of
+illegal combinations are structurally valid as untyped JSON values. But this is
+to be expected of any representation. More importantly, this does not
+necessarily match with the configuration schemas one can find in the wild, that
+favor simple, untagged union.
 
-In dynamically typed languages, which are interpreted for the vast majority
-[ref/examples?], it is customary to directly store different kind of data in one
-variable without further ceremony. The runtime system needs to attach type tags
-to values anyway, which can usually be observed using functions like the
-JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
+In dynamically typed languages, it is customary to directly store different kind
+of data in one variable without further ceremony. The runtime system needs to
+attach type tags to values anyway, which can usually be observed using functions
+like JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
 unions} has already been an idiom in dynamically typed language for a long time
 [ref/examples], and being able to type and handle these unions is an important
-aspect of the migration to a gradual type system. Indeed, most of gradually
+aspect of the migration to a gradual type system. Indeed, a lot of gradually
 typed languages either existed originally as dynamically typed language or aim
@@
 they tend to favor supporting untagged union types, the typed counterpart of
-this practice [example of gradually typed language with untagged union].
+this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
 
 In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. But are we? As a configuration language,
-Nickel does not evolve in the abstract. It aims at the best interoperability
-possible with JSON. It is designed to generate data for other softwares to
-consume, and tries as much as possible to be expressive enough to encode data
-schemas at the language level. If these schemas use untagged union, then this is
-a strong incentive to support them at the type level, or at least, at the
-contract level.
+support, so we are freer of our choice. But as a configuration language, Nickel
+does not evolve in the abstract. It aims at the best interoperability possible
+with JSON. It is designed to generate data for other softwares to consume, and
+tries as much as possible to be expressive enough to encode data schemas at the
+language level. If these schemas use untagged union, then this is a strong
+incentive to support them at the type level, or at least, at the contract level.
 
@@
 
-Encoding as: use singleton contract for tags, and then union.
 \begin{lstlisting}
@@
 benefit that the language is not imposing any representation. Untagged unions
-empower the programmer to represent and handle a wider variety of unions
+empower the programmer to represent and handle a wider variety of union idioms
 appearing in the wild. Thus, adding union types (and their natural and
@@
 
-\section{Contracts}
-\color{red}Goals of the section
-
-A primer on contracts, which in the end are the problematic bit when confronted
-with unions.\vspace{0.5cm}\color{black}
+\printbibliography
 

### 2021-03-23 paper.tex
Improve contract part

@@
 
-This paper is centered around the design of the Nickel
-language\cite{NickelRepo}. Nickel is a configuration programming language,
-meaning that a Nickel program eventually evaluates to a value that can be
+The explorative work presented in this paper is motived by the design of the
+Nickel language\cite{NickelRepo}. Nickel is a configuration programming
+language, meaning that a Nickel program evaluates to a value that can be
 directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
@@
 rethink of the language of the Nix package manager, used to describe one of the
-biggest sets of software packages ever built\cite{repology}.
-Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform or
-NixOps) as well.  While there's no hesitation in incorporating recent research
-when it serves a well-defined purpose, the design decisions are guided by the
-industrial use cases.
+biggest sets of software packages ever built\cite{repology}.  Nickel targets
+build systems and cloud deployment tools (Kubernetes, Terraform or NixOps) as
+well. While there's no hesitation in incorporating recent research when it
+serves a well-defined purpose, the design decisions are guided by the industrial
+use cases.
 
@@
 being both programmable and typed. In the configuration setting, there is a
-singular dilemma with respect to typing: if your program is only run once on
-fixed inputs and terminate, as are such programmable configurations, any type
-error will be reported at run-time anyway. Why bother with the complexity of a
-static type system? On the other hand, more and more software systems offload
-complexity to configurations, such as cloud infrastructure, build system, or
-operating systems, leading the corresponding configurations to become
-increasingly complex, to the point where static types are beneficial. For
-reusable code — that is, library functions — static types brings all the usual,
-widely acknowledged benefits.
+singular dilemma with respect to typing: since your program is run on fixed
+inputs and is expected to terminate, any relevant type error will be reported
+anyway.  Why bother with the complexity of a static type system? On the other
+hand, more and more software systems offload complexity to configurations, such
+as cloud infrastructure, build system, or operating systems, leading the
+corresponding configurations to become increasingly complex, to the point where
+static types are beneficial. For reusable code — that is, library functions —
+static types brings all the usual, widely acknowledged benefits.
 
-This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}, an
-approach capable of reuniting static typing and dynamic typing. Unlike
-gradually-typed language like TypeScript, we do not seek to build a complex type
-system that tries hard to type most of the idioms naturally arising in
-dynamically typed code.  Rather, we choose to provide a reasonably expressive
-type system with good inference properties, that should make writing typed
-generic function that operate on data relatively easy, but may require resorting
-to untyped code for more exotic expressions.  However, this is not a surrender,
-as Nickel provides a complementary mechanism for more complex data validation,
-contracts.
+This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
+that mixes both static typing and dynamic typing. Unlike gradually-typed
+language like TypeScript, we do not seek to build a complex type system that
+tries hard to make most of the idioms naturally arising in dynamically typed
+code typable. Rather, we choose to provide a reasonably expressive type system
+with good inference properties, that make usual generic function operating on
+data easily typable, but may require resorting to untyped code for more exotic
+expressions.  However, this is not a complete surrender, as Nickel provides a
+complementary mechanism for more advanced data validation: contracts.
 
@@
 In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
-Fellleisen introduces a principled approach to run-time predicate checking that
-nicely supports higher-order functions. It turns out these contracts are closely
-related to the type casts introduced by gradual typing. In
-\cite{WellTypedBlamed}, the authors precisely propose a system mixing gradual
-typing and contracts \textit{à la Findler \& Felleisen}. Nickel adopts a similar
-type system, with both statically typed terms, dynamically typed terms, and
-first-class contracts. In this setting, each construct of the type system -
-arrow types, record types, polymorphic types - must have a contract counterpart,
-that is able to check at run-time the membership of a value to the corresponding
-type.
+Fellleisen introduces a principled approach to run-time assertion checking that
+nicely support higher-order functions and introduce the notion of blame, which
+is fundamental for good error reporting. It turns out these contracts are
+closely related to the type casts introduced by gradual typing, modulo blame:
+both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of
+contracts as a safe interface between typed and untyped code. In
+\cite{WellTypedBlamed}, the authors precisely a system integrating gradual
+typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts a
+similar type system, with both statically typed terms, dynamically typed terms,
+and first-class contracts. In this setting, each type constructor of the static
+type system - arrows, records, foralls - must have a contract counterpart that
+is able to check at run-time that a value is indeed a member of the
+corresponding type.
 

### 2021-03-23 paper.tex
Fix color
Small fixes

@@
 Nickel language\cite{NickelRepo}. Nickel is a configuration programming
-language, meaning that a Nickel program evaluates to a value that can be
-directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
-easy to write and easy to embed in a larger system, to serve either directly as
-a native language to specify configuration, or as a meta-language to generate
-configuration. The focus is first and foremost practical: Nickel originated as a
-rethink of the language of the Nix package manager, used to describe one of the
-biggest sets of software packages ever built\cite{repology}.  Nickel targets
-build systems and cloud deployment tools (Kubernetes, Terraform or NixOps) as
-well. While there's no hesitation in incorporating recent research when it
-serves a well-defined purpose, the design decisions are guided by the industrial
-use cases.
+language, meaning that a Nickel program evaluates to a value that must be
+straightforwardly serializable to JSON, YAML or alike.  Nickel aims at being
+lightweight, easy to write and easy to embed in a larger system, to serve either
+directly as a native language to specify configuration, or as a meta-language to
+generate configuration. The focus is first and foremost practical: Nickel
+originated as a rethink of the language of the Nix package manager, used to
+describe one of the biggest sets of software packages ever built\cite{repology}.
+Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform
+or NixOps) as well. While there's no hesitation in incorporating recent research
+when it serves a well-defined purpose, the design decisions are guided by the
+industrial use cases.
+
+\subsection{Typing}
 
@@
 being both programmable and typed. In the configuration setting, there is a
-singular dilemma with respect to typing: since your program is run on fixed
-inputs and is expected to terminate, any relevant type error will be reported
-anyway.  Why bother with the complexity of a static type system? On the other
-hand, more and more software systems offload complexity to configurations, such
-as cloud infrastructure, build system, or operating systems, leading the
-corresponding configurations to become increasingly complex, to the point where
-static types are beneficial. For reusable code — that is, library functions —
-static types brings all the usual, widely acknowledged benefits.
+singular dilemma with respect to typing: since a program is run on fixed inputs
+and is expected to terminate, any relevant type error should show up anyway.
+Why bother with the complexity of a static type system? On the other hand, more
+and more software systems offload complexity to configurations, such as the
+infrastructure as code paradigm used in cloud deployment. When complexity grows,
+static types becomes attractive again. For reusable code — that is, library
+functions — static types brings all the usual widely acknowledged benefits.
 
@@
 that mixes both static typing and dynamic typing. Unlike gradually-typed
-language like TypeScript, we do not seek to build a complex type system that
-tries hard to make most of the idioms naturally arising in dynamically typed
-code typable. Rather, we choose to provide a reasonably expressive type system
-with good inference properties, that make usual generic function operating on
-data easily typable, but may require resorting to untyped code for more exotic
-expressions.  However, this is not a complete surrender, as Nickel provides a
-complementary mechanism for more advanced data validation: contracts.
+languages like TypeScript, we do not seek to build a complex type system that
+tries hard to accept most of the idioms naturally arising in dynamically typed
+code. Rather, we choose to provide a reasonably expressive type system with good
+inference properties, that make typing generic function operating on data easy ,
+but may require to resort to untyped code for more exotic expressions.  In any
+case, this is not a surrender, as Nickel provides a complementary mechanism for
+more advanced data validation: contracts.
 
@@
 
+\color{black}
+
 Enforcing pre- and post-conditions at runtime is a widely established practice.
@@
 nicely support higher-order functions and introduce the notion of blame, which
-is fundamental for good error reporting. It turns out these contracts are
-closely related to the type casts introduced by gradual typing, modulo blame:
-both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of
-contracts as a safe interface between typed and untyped code. In
-\cite{WellTypedBlamed}, the authors precisely a system integrating gradual
-typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts a
-similar type system, with both statically typed terms, dynamically typed terms,
-and first-class contracts. In this setting, each type constructor of the static
-type system - arrows, records, foralls - must have a contract counterpart that
-is able to check at run-time that a value is indeed a member of the
-corresponding type.
-
-\color{black}
+is crucial to good error reporting. It turns out these contracts are closely
+related to the type casts introduced by gradual typing, modulo blame: both
+\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
+as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
+the authors precisely a system integrating gradual typing with contracts
+\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
+both statically typed terms, dynamically typed terms, and first-class contracts.
+In this setting, each type constructor of the static type system - arrows,
+records, foralls - must have a contract counterpart that is able to check at
+run-time that a value is indeed a member of the corresponding type.
 

### 2021-04-06 paper.tex
WIP: reduce text

@@
 
-The explorative work presented in this paper is motived by the design of the
-Nickel language\cite{NickelRepo}. Nickel is a configuration programming
+The explorative work presented in this paper is motived by the development of
+the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
 language, meaning that a Nickel program evaluates to a value that must be
@@
 directly as a native language to specify configuration, or as a meta-language to
-generate configuration. The focus is first and foremost practical: Nickel
-originated as a rethink of the language of the Nix package manager, used to
-describe one of the biggest sets of software packages ever built\cite{repology}.
-Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform
-or NixOps) as well. While there's no hesitation in incorporating recent research
-when it serves a well-defined purpose, the design decisions are guided by the
-industrial use cases.
+generate them. The focus is first and foremost practical: Nickel originated as a
+rethink of the language of the Nix package manager, used to describe one of the
+biggest sets of software packages\cite{repology}.  Nickel targets build systems
+and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
+there's no hesitation in incorporating recent research when justified, the
+design decisions are guided by the industrial use cases.
 
@@
 
-Our aim is that writing simple Nickel should not be more complex than writing
-JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
-being both programmable and typed. In the configuration setting, there is a
-singular dilemma with respect to typing: since a program is run on fixed inputs
-and is expected to terminate, any relevant type error should show up anyway.
-Why bother with the complexity of a static type system? On the other hand, more
-and more software systems offload complexity to configurations, such as the
-infrastructure as code paradigm used in cloud deployment. When complexity grows,
-static types becomes attractive again. For reusable code — that is, library
-functions — static types brings all the usual widely acknowledged benefits.
+Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
+Unlike JSON, though, Nickel anticipates large configurations by being both
+programmable and typed. In the configuration setting, there is a singular
+dilemma with respect to typing: since a program is run on fixed inputs and is
+expected to terminate, any relevant type error will show up at evaluation.  Why
+bother with the complexity of a static type system? On the other hand, more and
+more software systems offload complexity to configurations, such as the
+infrastructure as code paradigm prevailing today in cloud deployment. When the
+complexity of a codebase grows, static types becomes attractive again.  For
+library code — that is, library functions —, static types are specifically
+adapated and bring all the usual benefits.
 
 This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
-that mixes both static typing and dynamic typing. Unlike gradually-typed
-languages like TypeScript, we do not seek to build a complex type system that
-tries hard to accept most of the idioms naturally arising in dynamically typed
-code. Rather, we choose to provide a reasonably expressive type system with good
-inference properties, that make typing generic function operating on data easy ,
-but may require to resort to untyped code for more exotic expressions.  In any
-case, this is not a surrender, as Nickel provides a complementary mechanism for
-more advanced data validation: contracts.
+which mixes both static and dynamic typing. Unlike gradually-typed languages
+like TypeScript, we do not seek to build a complex type system that tries hard
+to accept most of the idioms naturally arising in dynamically typed code.
+Rather, we choose to provide a reasonably expressive type system with good
+inference properties, that make typing generic functions operating on data easy,
+but may require using untyped code for more exotic expressions.  In any case,
+this is not a surrender, as Nickel provides a complementary mechanism for more
+advanced data validation: contracts.
 
@@
 as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
-the authors precisely a system integrating gradual typing with contracts
+the authors precisely introduce a system integrating gradual typing with contracts
 \textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
@@
 
-Unions are fundamental and ubiquitous building blocks of program data. Products
-represent conjunctions, that is several pieces of data whith potentially
-different shapes that are packed together in one value. Unions represent
-disjunctions, which are several exclusive alternatives stored in one value. When
-nested together, unions and intersections are able to represent aribrary
-tree-like structures, allowing to model a large class of data.
+Products and unions are fundamental and ubiquitous building blocks of program
+data. A product is a compound value that hold several pieces of potentially
+different types packed together in one value. Dually, a union only holds one
+data that may alternatives stored in one value. When nested together, unions and
+intersections are able to represent arbitrary tree-like structures, allowing to
+model a large class of data.
 
-Unions are also found in configuration, to represent nullable value in JSON,
-plain enumerations, or more general union types when several representations are
-supported.
+Unions are also useful for configurations, to represent nullable values in JSON,
+plain enumerations, or to support several representations for a field (a file
+path could be accepted either as one string of "/"-separated strings, or as a
+list of strings).
 
@@
 languages, unions are usually embodied by algebraic data types [references],
-also called \textbf{tagged unions}. Tagged means that there is an additional
-field, or constructor, that allows to discriminate between each alternative.
-When an union is constructed, the programmer specifies which alternative is
-chosen by using the corresponding tag. When a union is consumed, usually via
-pattern matching, code branches are tagged as well, and each branch is ensured
-to handle data of the right shape. Well-known programming languages with
+also called \textbf{tagged unions}. Well-known programming languages with
 first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
 Swift and surely many more.
-
-\begin{lstlisting}[caption={Example of a binary tree as a tagged union in OCaml}, captionpos=b]
-type binary_tree =
-  | Node of 'a binary_tree * 'a binary_tree
-  | Leaf of int
-
-let sum =
-  let rec aux total = function
-  | Leaf val -> val
-  | Node(left,right) =>
-    let left = aux total left in
-    aux total right
-\end{lstlisting}
 
@@
 an implementation detail: from the point of view of the programmer, a tagged
-union is a fist-class data type associated with constructors and destructors all
-with well-defined semantics. In contrast, in a gradually typed language like
-Nickel, the question of the representation of tagged unions as untyped values
-arises.
+union is a fist-class data type associated with constructors and destructors,
+together with well-defined semantics. In contrast, in a gradually typed language
+like Nickel, the question of the representation of tagged unions as untyped
+values arises.
 
-Unfortunately, it seems that tagged unions do not have a canonical counterpart
-as serializable untyped data. A natural representation is a record with a tag
-and a value which shape depends on the tag:
+Unfortunately, tagged unions do not have a serializable canonical counterpart as
+untyped data. One usual representation is a record with a tag and a value which
+shape depends on the tag (see Figure \ref{fig:union-encoding}).
 
+\begin{figure}
+  \begin{center}
     \begin{lstlisting}
-{ tag: <A, B, C>, value: Dyn }
+        { tag: <Integer, String>, value: Dyn }
     \end{lstlisting}
-
-\lstinline+<A, B, C>+ is the type of a C-like enumeration that can take value
-$A$, $B$ or $C$, while \lstinline+Dyn+ is the unknown dynamic type.
+  \end{center}
+\caption{Encoding of the union type: integer or a string}
+\label{fig:union-encoding}
+\medskip
+\small
+In Nickel, \lstinline+<A, B, C>+ is the type of a C-like enumeration that can
+take value \lstinline+A+, \lstinline+B+ or \lstinline+C+, while \lstinline+Dyn+
+is the dynamic unitype.
+\end{figure}
 
 A better way to describe it would be as a dependent pair which first component
-is a tag and second component is the data type corresponding to this tag, but
-the language does not feature dependent types:
+is a tag and second component is the data type corresponding to this tag:
+
 \[
@@
 
-Alas, choosing the first representation as a default is not fully satisfying.
-First, it does not map precisely to a JSON type: in consequence, a lot of
-illegal combinations are structurally valid as untyped JSON values. But this is
-to be expected of any representation. More importantly, this does not
-necessarily match with the configuration schemas one can find in the wild, that
-favor simple, untagged union.
+Alas, Nickel does not feature dependent types, which would open a whole
+different level of complexity. Choosing the first representation as a default is
+not fully satisfying.  First, it does not map precisely to a JSON value: in
+consequence, a lot of illegal combinations are structurally valid as untyped
+JSON values. But this is to be expected of any representation. More importantly,
+this does not necessarily match with the configuration schemas one can find in
+the wild, that favor simple, untagged union.
 

### 2021-04-06 paper.tex
Additonal pass, provide further refs for flow typing

@@
 nicely support higher-order functions and introduce the notion of blame, which
-is crucial to good error reporting. It turns out these contracts are closely
-related to the type casts introduced by gradual typing, modulo blame: both
-\cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the value of contracts
-as a safe interface between typed and untyped code. In \cite{WellTypedBlamed},
-the authors precisely introduce a system integrating gradual typing with contracts
-\textit{à la Findler \& Felleisen}. Nickel adopts a similar type system, with
-both statically typed terms, dynamically typed terms, and first-class contracts.
-In this setting, each type constructor of the static type system - arrows,
-records, foralls - must have a contract counterpart that is able to check at
-run-time that a value is indeed a member of the corresponding type.
+is crucial to good error reporting. It became apparent later that their
+contracts are closely related to the type casts introduced by gradual typing,
+modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
+value of contracts as a safe interface between typed and untyped code. In
+\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
+gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
+a similar type system, with both statically typed terms, dynamically typed
+terms, and first-class contracts.  In this setting, each type constructor of the
+static type system - arrows, records, foralls - must have a contract counterpart
+that check at run-time that a value is indeed a member of the corresponding
+type. Various type constructors have been investigated [All refs], with
+solutions of varying complexity. Amongst all the common extensions to the
+polymorphic lambda calculus, one of the most simple and ubiquitous turns out to
+be one of the most involved: \textit{union types}.
 
@@
 Products and unions are fundamental and ubiquitous building blocks of program
-data. A product is a compound value that hold several pieces of potentially
-different types packed together in one value. Dually, a union only holds one
-data that may alternatives stored in one value. When nested together, unions and
-intersections are able to represent arbitrary tree-like structures, allowing to
-model a large class of data.
+data. A product is a compound data that hold several values of potentially
+different shape packed together. Dually, a union only holds one value whose
+shape is only known to be in a fixed set of alternatives. When nested together,
+unions and intersections are able to represent arbitrary tree-like structures,
+allowing to model a large class of data.
 
 Unions are also useful for configurations, to represent nullable values in JSON,
-plain enumerations, or to support several representations for a field (a file
-path could be accepted either as one string of "/"-separated strings, or as a
-list of strings).
+plain enumerations, or to support several representations for a field (for
+example, a file path could be accepted either as one string of
+\lstinline+"/"+-separated strings, or as a list of strings).
 
 In statically typed languages, and more specifically in functional programming
-languages, unions are usually embodied by algebraic data types [references],
+languages, unions are usually implemented as algebraic data types [references],
 also called \textbf{tagged unions}. Well-known programming languages with
@@
 an implementation detail: from the point of view of the programmer, a tagged
-union is a fist-class data type associated with constructors and destructors,
-together with well-defined semantics. In contrast, in a gradually typed language
-like Nickel, the question of the representation of tagged unions as untyped
-values arises.
+union is a fist-class data type associated with constructors and destructors
+with well-defined semantics. In contrast, in a gradually typed language like
+Nickel, the question of the representation of tagged unions as untyped values
+arises.
 
 Unfortunately, tagged unions do not have a serializable canonical counterpart as
-untyped data. One usual representation is a record with a tag and a value which
+untyped data. One usual representation is a record with a tag and a value whose
 shape depends on the tag (see Figure \ref{fig:union-encoding}).
@@
 
-Alas, Nickel does not feature dependent types, which would open a whole
-different level of complexity. Choosing the first representation as a default is
-not fully satisfying.  First, it does not map precisely to a JSON value: in
+Nickel does not feature dependent types, which would open a whole new level of
+complexity. Alas, choosing the non dependent representation as a default is not
+fully satisfying.  First, it does not map precisely to a JSON value: in
 consequence, a lot of illegal combinations are structurally valid as untyped
@@
 tries as much as possible to be expressive enough to encode data schemas at the
-language level. If these schemas use untagged union, then this is a strong
-incentive to support them at the type level, or at least, at the contract level.
-
-\subsection{Union types}
+language level thanks to contracts. If these schemas use untagged union, then
+this is a strong incentive to support them at the type level, or at least, at
+the contract level.
 
@@
 Untagged union, together with the other existing features of the type system,
-are sufficient to implement the encoding mentioned in [cite previous section]:
+are sufficient to implement the encoding illustrated in Figure
+\ref{fig:union-encoding}. For example, let us represent an algebraic data type
+(written in ML syntax) to an imaginary Nickel union type:
 
 \begin{lstlisting}
-type sum = A of Num | B of Str
+type sum = Int of int | Str of string
 \end{lstlisting}
@@
 \begin{lstlisting}
-{ tag : <A>, val : Num} | { tag : <B>, val : Str}
+{ tag : <Int>, val : Num} | { tag : <Str>, val : Str}
 \end{lstlisting}
@@
 
-Flow typing[ ], as implemented for example by TypeScript, is capable of solving
-this issue by correctly propagating the type information in each branch. One can
-imagine doing the same in Nickel. Thus equipped with untagged unions and a
-typechecker smart enough, classic tagged unions can be simulated, but with the
+Flow typing, implemented in Racket\cite{FlowTypingRacket1, FlowTypingRacket2},
+Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
+TypeScript\cite{FlowTypingTypeScript}, solves the issue by being capable of
+refining appropriately the type information in each branch. One can imagine
+doing the same in Nickel. Thus equipped with untagged unions and a typechecker
+smart enough, classic tagged unions can be simulated, but with the additional
 benefit that the language is not imposing any representation. Untagged unions
 empower the programmer to represent and handle a wider variety of union idioms
-appearing in the wild. Thus, adding union types (and their natural and
-unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.
+appearing in the wild.  Thus, adding union types to Nickel sounds appealing.
 

### 2021-04-06 paper.tex
Add some bib on contracts, rewording

@@
 and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
-there's no hesitation in incorporating recent research when justified, the
-design decisions are guided by the industrial use cases.
+there is no hesitation in incorporating recent research or doing it ourselves,
+the design decisions are always guided by the practical industrial use cases.
 
@@
 complexity of a codebase grows, static types becomes attractive again.  For
-library code — that is, library functions —, static types are specifically
-adapated and bring all the usual benefits.
+reusable code — that is, library functions —, static types are specifically
+adapated, and bring all the usual benefits of early error detection, code
+robustness, better code structure, documentation, and so on.
 
@@
 Rather, we choose to provide a reasonably expressive type system with good
-inference properties, that make typing generic functions operating on data easy,
+inference properties, that make typing functions operating on generic data easy,
 but may require using untyped code for more exotic expressions.  In any case,
@@
 that check at run-time that a value is indeed a member of the corresponding
-type. Various type constructors have been investigated [All refs], with
-solutions of varying complexity. Amongst all the common extensions to the
-polymorphic lambda calculus, one of the most simple and ubiquitous turns out to
-be one of the most involved: \textit{union types}.
+type. Contracts for various extensions of the simply typed $\lambda$-caluclus
+have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
+RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
+varying complexity.  Amongst all the common extensions to the polymorphic lambda
+calculus, one of the most simple and ubiquitous turns out to be one of the most
+involved: \textit{union types}.
 

### 2021-04-13 paper.tex
Add proper intro, small fixes

@@
 
-\title{Union and intersection contracts: theory vs practice}
+\title{Contracts for unions and intersections are hard for Nickel}
 \author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}
@@
 \begin{abstract}
- In this paper we do some cool stuff.
+This paper reports an attempt to incorporate union and intersection types in
+Nickel, a configuration programming language with a gradual type system and
+first-class contracts. While the end result looks appealing, it turns out the
+concrete impact on both the design and the implementation of the language is
+unexpectedly high. We review the issues raised in particular by the
+implementation of contracts for unions and intersections, and why we think the
+cost ends up too high for the benefit. We conclude by suggesting different leads
+to represent and type unions.
 \end{abstract}
@@
 
-The explorative work presented in this paper is motived by the development of
-the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
-language, meaning that a Nickel program evaluates to a value that must be
+Build systems, package managers, operating systems, cloud infrastructures, and
+web services are examples of modern complex software systems that require an
+extensive and non-trivial configuration in order to make them adapted to each
+different use-case. To manage the growing complexity that is then offloaded to
+configurations, the declarative approach has become more and more popular,
+illustrated for example by the infrastructure-as-code paradigm. This leads
+configuration to play an important role in critical aspects of software
+engineering, including security, availability, and maintainability.
+
+However, static text-based configuration alone is falling short of expressivity,
+and is seldom sufficient. Data cannot be transformed, combined nor shared,
+resulting at best in boilerplate and duplication of information, or at worst, in
+data being invalid or inconsistent. Data validation isn’t supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
+done. Correctly configuring a modern system is hard and failures may have
+substantial negative consequences.
+
+This is why the authors have been developping the Nickel
+language\cite{NickelRepo}. Nickel is a configuration programming language,
+meaning that a Nickel program evaluates to a value that must be
 straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
-lightweight, easy to write and easy to embed in a larger system, to serve either
-directly as a native language to specify configuration, or as a meta-language to
-generate them. The focus is first and foremost practical: Nickel originated as a
-rethink of the language of the Nix package manager, used to describe one of the
-biggest sets of software packages\cite{repology}.  Nickel targets build systems
-and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
-there is no hesitation in incorporating recent research or doing it ourselves,
-the design decisions are always guided by the practical industrial use cases.
+lightweight and easy to write while still providing features for writing correct
+code and generating valid data. The focus is first and foremost practical:
+Nickel originated as a rethink of the language of the Nix package manager, used
+to describe one of the biggest sets of software packages\cite{repology}.  Nickel
+targets build systems and cloud deployment tools (Kubernetes, Terraform or
+NixOps) as well. While we are keen on incorporating existing or doing new
+research when it solves a well-defined problem, the general design goals are
+always guided by the practical industrial use cases.
 
@@
 bother with the complexity of a static type system? On the other hand, more and
-more software systems offload complexity to configurations, such as the
-infrastructure as code paradigm prevailing today in cloud deployment. When the
-complexity of a codebase grows, static types becomes attractive again.  For
+more software systems offload complexity to configurations, in particular with
+the infrastructure as code paradigm prevailing today in cloud deployment. When
+the complexity of a codebase grows, static types becomes attractive again.  For
 reusable code — that is, library functions —, static types are specifically
@@
 Rather, we choose to provide a reasonably expressive type system with good
-inference properties, that make typing functions operating on generic data easy,
+inference properties, which makes typing functions operating on generic data easy,
 but may require using untyped code for more exotic expressions.  In any case,
@@
 calculus, one of the most simple and ubiquitous turns out to be one of the most
-involved: \textit{union types}.
+involved: \emph{union types}.
 
@@
 languages, unions are usually implemented as algebraic data types [references],
-also called \textbf{tagged unions}. Well-known programming languages with
+also called \emph{tagged unions}. Well-known programming languages with
 first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
@@
 attach type tags to values anyway, which can usually be observed using functions
-like JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
-unions} has already been an idiom in dynamically typed language for a long time
+like JavaScript's \verb+typeof+. Using so-called \emph{untagged unions} has
+already been an idiom in dynamically typed language for a long time
 [ref/examples], and being able to type and handle these unions is an important
@@
 In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. But as a configuration language, Nickel
-does not evolve in the abstract. It aims at the best interoperability possible
-with JSON. It is designed to generate data for other softwares to consume, and
-tries as much as possible to be expressive enough to encode data schemas at the
-language level thanks to contracts. If these schemas use untagged union, then
-this is a strong incentive to support them at the type level, or at least, at
-the contract level.
+support, so we are freer of our choice. However, Nickel is a configuration
+language, whose contracts and types system is intended to be expressive enough
+to encode data schemas encoutered in practice. Since these schemas do use
+untagged unions, there is a strong incentive to support untagged uniohns at the
+type level, or at least, at the contract level.
 
@@
 \begin{lstlisting}
-type sum = Int of int | Str of string
+type either = Left of int | Right of string
 \end{lstlisting}
@@
 \begin{lstlisting}
-{ tag : <Int>, val : Num} | { tag : <Str>, val : Str}
+{ tag : <Left>, val : Num} | { tag : <Right>, val : Str}
 \end{lstlisting}
@@
 switch x.tag {
-  A => x.val + 1,
-  B => then strings.fromNum x.val
+  Left => x.val + 1,
+  Right => strings.fromNum x.val
 }
@@
 Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
-TypeScript\cite{FlowTypingTypeScript}, solves the issue by being capable of
-refining appropriately the type information in each branch. One can imagine
-doing the same in Nickel. Thus equipped with untagged unions and a typechecker
-smart enough, classic tagged unions can be simulated, but with the additional
-benefit that the language is not imposing any representation. Untagged unions
-empower the programmer to represent and handle a wider variety of union idioms
-appearing in the wild.  Thus, adding union types to Nickel sounds appealing.
+TypeScript\cite{FlowTypingTypeScript}, is capable of refining appropriately the
+type information in each branch. Thus equipped with untagged unions and a
+typechecker smart enough, classic tagged unions can be simulated, but with the
+additional benefit that the language is not imposing any representation.
+Untagged unions empower the programmer to represent and handle a wider variety
+of union idioms appearing in the wild.  Thus, adding union types to Nickel
+sounds appealing.
 

### 2021-04-13 paper.tex
Typos + TODOs

@@
 resulting at best in boilerplate and duplication of information, or at worst, in
-data being invalid or inconsistent. Data validation isn’t supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
+data being invalid or inconsistent. Data validation is not supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
 done. Correctly configuring a modern system is hard and failures may have
@@
 
-This is why the authors have been developping the Nickel
+This is why the authors have been developing the Nickel
 language\cite{NickelRepo}. Nickel is a configuration programming language,
@@
 reusable code — that is, library functions —, static types are specifically
-adapated, and bring all the usual benefits of early error detection, code
+adapted, and bring all the usual benefits of early error detection, code
 robustness, better code structure, documentation, and so on.
@@
 that check at run-time that a value is indeed a member of the corresponding
-type. Contracts for various extensions of the simply typed $\lambda$-caluclus
+type. Contracts for various extensions of the simply typed $\lambda$-calculus
 have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
@@
 language, whose contracts and types system is intended to be expressive enough
-to encode data schemas encoutered in practice. Since these schemas do use
-untagged unions, there is a strong incentive to support untagged uniohns at the
+to encode data schemas encountered in practice. Since these schemas do use
+untagged unions, there is a strong incentive to support untagged unions at the
 type level, or at least, at the contract level.
@@
 
+$(A -> B) /\ (C -> D) ~ (A \/ C) -> (B \/ D)$
+$(A \/ B) -> C ~ (A -> C) /\ (B -> C)$
+
+% TODO: lazyness + parallelisation
+
 \printbibliography

### 2021-04-27 paper.tex
Update paper.tex
Co-authored-by: Teodoro Freund <tfreund95@gmail.com>
Squashed commit of the following:
commit 29c6e1d668c9b945b186b4e9b92039dd799eb432
Merge: 7dd0b8b eb71bb3
Author: Yann Hamdaoui <yann.hamdaoui@gmail.com>
Date:   Tue Apr 20 16:12:24 2021 +0200

    Merge pull request #1 from tweag/motivations-background

    Background and motivations

commit eb71bb33aeee9be1b5b7441fbaea85295be737c7
Author: Yann Hamdaoui <yann.hamdaoui@gmail.com>
Date:   Tue Apr 20 16:03:11 2021 +0200

    Update paper.tex

    Co-authored-by: Teodoro Freund <tfreund95@gmail.com>

commit 1e89c5e092da83d33c8fa

@@
 \usepackage{url}
+
+%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%
+
+  % TOGGLE ME to turn off all the commentary:
+  \InputIfFileExists{no-editing-marks}{
+    \def\noeditingmarks{}
+  }
+
+  \usepackage{xargs}
+  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+  % ^^ Need for pgfsyspdfmark apparently?
+  \ifx\noeditingmarks\undefined
+      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
+      \newcommand{\Red}[1]{{\color{red}{#1}}}
+      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
+      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
+      \newenvironment{alt}{\color{red}}{}
+
+      \newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
+      \newcommandx{\info}[2][1=]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,#1]{#2}}
+      \newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
+      \newcommandx{\inconsistent}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
+      \newcommandx{\critical}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
+      \newcommand{\improvement}[1]{\todo[linecolor=pink,backgroundcolor=pink!25,bordercolor=pink]{#1}}
+      \newcommandx{\resolved}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}} % use this to mark a resolved question
+  \else
+  %    \newcommand{\Red}[1]{#1}
+      \newcommand{\Red}[1]{{\color{red}{#1}}}
+      \newcommand{\newaudit}[1]{#1}
+      \newcommand{\note}[1]{}
+      \newenvironment{alt}{}{}
+  %    \renewcommand\todo[2]{}
+      \newcommand{\unsure}[2][1=]{}
+      \newcommand{\info}[2][1=]{}
+      \newcommand{\change}[2]{}
+      \newcommand{\inconsistent}[2]{}
+      \newcommand{\critical}[2]{}
+      \newcommand{\improvement}[1]{}
+      \newcommand{\resolved}[2]{}
+  \fi
+
+%%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%
 
@@
 \section*{Introduction}
-Intro
+Intro\unsure{todo}
 
@@
 the infrastructure as code paradigm prevailing today in cloud deployment. When
-the complexity of a codebase grows, static types becomes attractive again.  For
+the complexity of a codebase grows, static types become attractive again.  For
 reusable code — that is, library functions —, static types are specifically
@@
 
-\subsection{Unions types}
+\subsection{Union types}
 \color{red}Goals of the section
@@
 
-$(A -> B) /\ (C -> D) ~ (A \/ C) -> (B \/ D)$
-$(A \/ B) -> C ~ (A -> C) /\ (B -> C)$
+$$(A -> B) \cap (C -> D) \sim (A \cup C) -> (B \cup D)$$
+$$(A \cup B) -> C \sim (A -> C) \cap (B -> C)$$
 
 % TODO: lazyness + parallelisation
+
+\section{Challenges}
+\input{challenges}
 

### 2021-04-27 paper.tex
Add hyperref
Stress union more than products + stress the importance of custom contracts

@@
 \usepackage{url}
+\usepackage{hyperref}
 
@@
 
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
@@
 typing and main design orientations, such as practice-oriented, lightweight,
-etc.\vspace{0.5cm}\color{black}
+etc.\vspace{0.5cm}}
 
@@
 \subsection{Contracts}
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
 A primer on contracts, which in the end are the problematic bit when confronted
-with unions.\vspace{0.5cm}
-
-\color{black}
+with unions.\vspace{0.5cm}}
 
@@
 a similar type system, with both statically typed terms, dynamically typed
-terms, and first-class contracts.  In this setting, each type constructor of the
-static type system - arrows, records, foralls - must have a contract counterpart
-that check at run-time that a value is indeed a member of the corresponding
-type. Contracts for various extensions of the simply typed $\lambda$-calculus
-have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
-RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
-varying complexity.  Amongst all the common extensions to the polymorphic lambda
-calculus, one of the most simple and ubiquitous turns out to be one of the most
-involved: \emph{union types}.
+terms, and first-class contracts. In "first-class contracts", we include the
+ability to write user-defined contracts. Because built-in types are too
+simplistic for the purpose of data validation, the ability of writing a small
+validation function for, say, an URL, and having it immediately available as a
+contract that can be composed with other type constructors, is in our
+opinion one of the main appeal of Nickel contracts system. In this hybrid
+system, each type constructor of the static type system - arrows, records,
+foralls - must have a contract counterpart that checks at run-time that a value
+is indeed a member of the corresponding type. Contracts for various extensions
+of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
+    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
+GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
+common extensions to the polymorphic lambda calculus, one of the most useful and
+widespread turns out to be one of the most challenging: \emph{union types}.
 
 \subsection{Union types}
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
-Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}\color{black}
+Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}}
 
-Products and unions are fundamental and ubiquitous building blocks of program
-data. A product is a compound data that hold several values of potentially
-different shape packed together. Dually, a union only holds one value whose
-shape is only known to be in a fixed set of alternatives. When nested together,
-unions and intersections are able to represent arbitrary tree-like structures,
-allowing to model a large class of data.
+Unions are fundamental and ubiquitous building blocks of program data. A union
+only holds one value whose shape is only known to be in a fixed set of
+alternatives: for eample, either a boolean or a string. Wuen used in combination
+with products, they can represent arbitrary tree-like structures, allowing to
+model a large class of data.
 
-Unions are also useful for configurations, to represent nullable values in JSON,
-plain enumerations, or to support several representations for a field (for
-example, a file path could be accepted either as one string of
-\lstinline+"/"+-separated strings, or as a list of strings).
+Unions are also useful for configurations:
+\begin{itemize}
+    \item To represent nullable values in JSON
+    \item For plain enumerations. For, the enabled protocols of a data server
+        would be either \lstinline+HTTP+, \lstinline+FTP+ or \lstinline+SFTP+.
+    \item To support different representations for a field. A file path could be
+        accepted either as one string of \lstinline+"/"+-separated strings, or
+        as a list of strings.
+\end{itemize}
 
@@
 \subsection{Tagged versus untagged}
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
 Interpreted languages already need to tag their values, hence bare union are more natural.
-It also corresponds better to prior usage in untyped code\vspace{0.5cm}\color{black}
+It also corresponds better to prior usage in untyped code\vspace{0.5cm}}
 
@@
 \subsection{Tagged as untagged}
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
@@
 tagged unions. They let the user free of the representation though, and encode
-more stuff, that's why they are appealing.\vspace{0.5cm}\color{black}
+more stuff, that's why they are appealing.\vspace{0.5cm}}
 
@@
 \subsection{Union and intersections}
-\color{red}Goals of the section
+{\color{red}Goals of the section
 
-Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}\color{black}
+Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}}
 

### 2021-04-27 paper.tex
Add note on intersections sneaking in

@@
 \begin{itemize}
-    \item To represent nullable values in JSON
-    \item For plain enumerations. For, the enabled protocols of a data server
+    \item To represent nullable values in JSON.
+    \item For plain enumerations. For example, the enabled protocols of a data server
         would be either \lstinline+HTTP+, \lstinline+FTP+ or \lstinline+SFTP+.
-    \item To support different representations for a field. A file path could be
+    \item To support different representations of the same data. A file path could be
         accepted either as one string of \lstinline+"/"+-separated strings, or
@@
 language, whose contracts and types system is intended to be expressive enough
-to encode data schemas encountered in practice. Since these schemas do use
+to encode data schemas encountered in the wild. Since these schemas do use
 untagged unions, there is a strong incentive to support untagged unions at the
@@
 \ref{fig:union-encoding}. For example, let us represent an algebraic data type
-(written in ML syntax) to an imaginary Nickel union type:
+(written in ML syntax) as an imaginary Nickel union type:
 
@@
 {\color{red}Goals of the section
-
 Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}}
 
-$$(A -> B) \cap (C -> D) \sim (A \cup C) -> (B \cup D)$$
-$$(A \cup B) -> C \sim (A -> C) \cap (B -> C)$$
+While our original motivation is in supporting unions, the rest of the paper
+explore complete systems featuring union and intersection types. One could
+wonder if the difficulties could then be sidestepped by removing intesection out
+of the equation. But
 
-% TODO: lazyness + parallelisation
+\begin{enumerate}
+    \item Most of the encountered issues already appear with bare unions.
+    \item Intersection being dual to unions, they are showing up in some way
+        when unions are put in negative positions. For example, a contract for
+        $(A \cup B) \to C$ in a system without intersections must be equivalent
+        to the contract $(A \to C) \cap (B \to C)$
+\end{enumerate}
 

### 2021-04-27 paper.tex
Add short section on performance/parallelization

@@
 
+\subsection{Performance considerations}
+\info{This a draft}
+\todo{transition, location}
+
+At first sight, performance should not be a critical issue given the use cases
+of Nickel. But generating large configurations can already hit some limits in
+current related languages [reference?]. A natural is parallelization:
+Nickel being an almost pure functional programming language, evaluating separate
+files or expressions in parallel is straightforward and effective. This
+parallelization potential is thus a design requirement, and must be when the
+addition of union types.
+
 \section{Challenges}

### 2021-04-27 paper.tex
Better size for the todo boxes
Some notes
Note: Typescript's type are static

@@
   \ifx\noeditingmarks\undefined
-      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
+      % \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
       \newcommand{\Red}[1]{{\color{red}{#1}}}
@@
 Intro\unsure{todo}
-
+\info{Goals of the paper: position paper: unions/intersections nice in
+theory problematic in practice. In particular non-orthogonal with the
+rest of the features of your language.}
 \section{Background on union types}
@@
 which mixes both static and dynamic typing. Unlike gradually-typed languages
-like TypeScript, we do not seek to build a complex type system that tries hard
+like TypeScript\unsure{Todo: explain that typescript uses unions as
+  static typing whereas we are speaking of dynamic contracts}, we do not seek to build a complex type system that tries hard
 to accept most of the idioms naturally arising in dynamically typed code.

### 2021-05-03 paper.tex
Starting section on comparison, outlines

--- 2021-04-27 paper.tex
@@
 
+\section{Current (Real life) implementations}
+
+\info{The goal of this section is to outline existing languages that implement
+union and intersection, in some form, and shortly compare these
+with the semantics tried to obtain on research (WMW, KT)}
+
+Unions and intersection are starting to become more and more common
+on non-academic programming languages, as stated on the introduction,
+unions are a natural abstraction for dynamic languages, where the type
+of an expression could have many different shapes, without necessarily
+having a mark to distinguish between them.
+With this in mind, a valid question is how do these languages solve the
+union/intersection conundrum? Or, even better, what is it that they
+solve?
+
+In the rest of this section we revisit some different approaches.
+
+\subsection{Racket}
+
+Racket is language based on Scheme mainly thought as a language to experiment
+with different language desings and ideas. It provides a very complete contract
+system.
+
+Among those contracts, they provide \texttt{and/c} and \texttt{or/c} contracts.
+These contracts are lightweight versions of unions and intersections, they simply
+check that every contract holds, in the case of and/c, and that at least one contract
+holds, in the case of or/c.
+
+So, for instance, the usual usecase of intersection to encode overloading
+does not work on Racket. Take a look at the next example, the function
+\texttt{overload} shoudl accept a call like \texttt{(overload 3)}, but it fails
+since it 3 is not a valid \texttt{string?}.
+
+\texttt{(define/contract overload
+(and/c (-> number? number?)
+       (-> string? string?))
+(λ (x) x))}
+
+Even worst, the or/c contract doesn't allow to have multiple higher order
+contracts that could (potentially) apply to a function, these have to be
+differentiated by first order markers (for instance, the number of arguments).
+
+\subsubsection*{case->}
+
+Racket does provide an alternative for overloading methods, called the
+\texttt{case->} contract, however, this overloading of methods only
+works over contracts of functions with different amount of parameters.
+
+\todo{Coming back to the problems of composing U/I with other typing
+features, Racket has made a clear choice and decided that higher order
+is more important (a wise decision)}
+
+\subsection{Python - MyPy}
+
 \printbibliography

### 2021-05-06 paper.tex
Move to the ACMart 10pt format
Use latest acmart
Some comments
+typos

@@
-\documentclass{article}
-\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
+\documentclass[sigplan,10pt]{acmart}
+%\usepackage{amssymb}
+\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
 \usepackage{listings}
@@
   \ifx\noeditingmarks\undefined
-      % \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
+      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
       \newcommand{\Red}[1]{{\color{red}{#1}}}
@@
 
-\usepackage{biblatex}
-\addbibresource{nickel.bib}
-
-\title{Contracts for unions and intersections are hard for Nickel}
-\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}
-
-\begin{document}
-
-\maketitle
+\bibliographystyle{ACM-Reference-Format}
 
@@
 \end{abstract}
+
+\title{Contracts for unions and intersections are hard for Nickel}
+\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}
+
+\begin{document}
+
+\maketitle
 
@@
 Build systems, package managers, operating systems, cloud infrastructures, and
-web services are examples of modern complex software systems that require an
+web services\unsure{Also continuous integration} are examples of modern complex software systems that require an
 extensive and non-trivial configuration in order to make them adapted to each
@@
 
-However, static text-based configuration alone is falling short of expressivity,
+However, static text-based configuration alone is falling short of expressiveness,
 and is seldom sufficient. Data cannot be transformed, combined nor shared,
@@
 In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
-Fellleisen introduces a principled approach to run-time assertion checking that
-nicely support higher-order functions and introduce the notion of blame, which
+Felleisen introduce a principled approach to run-time assertion checking that
+nicely supports higher-order functions and introduces the notion of blame, which
 is crucial to good error reporting. It became apparent later that their
@@
 \cite{WellTypedBlamed}, the authors precisely introduce a system integrating
-gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
+gradual typing with contracts \textit{à la Findler \&
+  Felleisen}.\unsure{The historical bit should probably be moved to
+  related works eventually} Nickel adopts
 a similar type system, with both statically typed terms, dynamically typed
@@
 simplistic for the purpose of data validation, the ability of writing a small
-validation function for, say, an URL, and having it immediately available as a
-contract that can be composed with other type constructors, is in our
-opinion one of the main appeal of Nickel contracts system. In this hybrid
+validation function for, say, a URL, and having it immediately available as a
+contract that can be composed with other type
+constructors\unsure{Needs code to illustrate.}, is in our
+opinion one of the main appeal of Nickel's contracts system. In this hybrid
 system, each type constructor of the static type system - arrows, records,
@@
 In statically typed languages, and more specifically in functional programming
-languages, unions are usually implemented as algebraic data types [references],
+languages, unions are usually implemented as algebraic data
+types\unsure{This is a very misleading statement at best. Sums are
+  "tagged union" sure, but that refers to an encoding a sums with
+  unions, certainly not an encoding of unions with sums!} [references],
 also called \emph{tagged unions}. Well-known programming languages with
@@
 an implementation detail: from the point of view of the programmer, a tagged
-union is a fist-class data type associated with constructors and destructors
+union is a first-class data type associated with constructors and destructors
 with well-defined semantics. In contrast, in a gradually typed language like
@@
 this does not necessarily match with the configuration schemas one can find in
-the wild, that favor simple, untagged union.
+the wild, which favor simple, untagged union.
 
-In dynamically typed languages, it is customary to directly store different kind
+In dynamically typed languages, it is customary to directly store different kinds
 of data in one variable without further ceremony. The runtime system needs to
 attach type tags to values anyway, which can usually be observed using functions
-like JavaScript's \verb+typeof+. Using so-called \emph{untagged unions} has
+like JavaScript's \verb+typeof+. Using so-called \emph{untagged
+  unions}\unsure{If you can observe with typeof, then it's
+  pretty much a tagged union, though the tag is implicit} has
 already been an idiom in dynamically typed language for a long time
@@
         when unions are put in negative positions. For example, a contract for
-        $(A \cup B) \to C$ in a system without intersections must be equivalent
+        $(A \cup B) \to C$ in a system without intersections must be
+        equivalent\unsure{``Must be'' is a rather strong statement, we
+        will find a way to tone it down}
         to the contract $(A \to C) \cap (B \to C)$
 \end{enumerate}
+
+\unsure{Some bits missing so far: we are interested in higher-order
+  contracts, Nickel is a higher-order \& pure language, contracts as
+  functions. We've said that user-defined contracts are a thing, it
+  probably matters so we may have to hammer it down.}
+
+\unsure{Something that we must mention somewhere: if we want to check
+  that something is a list of As lazily (to avoid making functions
+  like hd be O(n), for instance), then we have a lot of the problems of
+  higher-order contracts anyway.}
 
@@
 check that every contract holds, in the case of and/c, and that at least one contract
-holds, in the case of or/c.
+holds, in the case of or/c.\unsure{It doesn't really make sense for a
+  contract to “hold”, in general}
 
@@
 does not work on Racket. Take a look at the next example, the function
-\texttt{overload} shoudl accept a call like \texttt{(overload 3)}, but it fails
+\texttt{overload} should accept a call like \texttt{(overload 3)}, but it fails
 since it 3 is not a valid \texttt{string?}.
@@
 contracts that could (potentially) apply to a function, these have to be
-differentiated by first order markers (for instance, the number of arguments).
+differentiated by first order markers (for instance, the number of
+arguments).\unsure{This section is overall a bit hard to follow. I
+  think I've figured out what it means, but it needs more love. Also
+  give an example which doesn't work and one that does for or/c.}
 
@@
 
-\printbibliography
+\unsure{We ought to group all the static typing in one subsection}
+
+\bibliography{nickel}
 
 \end{document}
+
+% LocalWords:  Felleisen

### 2021-05-06 paper.tex
Move existing documentation to appendices
Tentative outline

@@
 rest of the features of your language.}
+
+\newpage
+
+\section{Features of languages}
+\info{Zoology of various features that we will eventually show
+  conflict with this or that property or implementation of union and
+  intersection. Including user-define contracts.}
+
+\newpage
+
+\section{Union \& intersection}
+\info{What are they, what are they used for.}
+
+\newpage
+
+\section{Issues coming from the semantics}
+\info{What goes wrong or difficult when trying to implement union and
+  intersection literally}
+
+\newpage
+
+\section{Issues in concrete restrictions from the literature}
+\info{The various papers and the tradeoffs they make}
+
+\newpage
+
+\section{Related work}
+\info{Including a mention of statically typed systems with union}
+
+\newpage
+
+\section{Conclusion}
+
+\appendix
+
 \section{Background on union types}

### 2021-05-07 paper.tex
Started survey on other papers

@@
 \info{The various papers and the tradeoffs they make}
+
+The papers that previously researched union and intersection
+as dynamic checks have, in our eyes, different shortcomings
+that makes them suboptimal solutions.
+
+Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
+they noticed that, since union contracts should share information between
+different application contexts, while intersection contracts should behave
+independently between each application context, the system becomes much
+simpler if every contract gets transformed into a union of intersections
+of other contracts, by using the usual De Morgan laws,
+and then eagerly opening up unions, but postponing intersection
+contracts as much as possible.
+\todo{This assumes the difficulty has been previously explained}
+
+This straightforward solutions makes composing the Keil and Thiemann
+system with other existing systems more difficult; in the words
+of Williams, Morrison, and Wadler:
+"(...) the monitoring semantics for contracts of intersection and union types given by Keil
+and Thiemann are not uniform. (...) If uniformity helps composition, then
+special cases can hinder composition."\cite{RootCauseOfBlame}
+
+Another difficult issue Keil and Thiemann were able to solve, is the
+fact that flat contracts and intersection do not usually play nice
+together.
+They provide two different set of rules for their calculus,
+the first one is non deterministic, which purpose is to provide
+semantics that are simple to understand.
+The second one is a deterministic set of rules, that are intended
+to be simply to implement. These semantics need to perform operations
+on the evaluation context, to decide whether or not to evaluate
+a given contract.
+
+We see two problems here, a minor one is the fact that two different
+semantics need to be defined, which translates either to a difficult
+to implement or a difficult to understand system. A second, major one,
+is the dependency added over the evaluation context, since, again,
+it makes composing this, with other features desired on programming languages,
+difficult, or even impossible (as the case with common subexpression elimination).
+
+Williams, Morrison, and Wadler, take a different approach on flat contracts,
+they are completely forbidden in their system, explicitly to not encounter
+the previously mentioned issue.
+This restriction goes against what is usually expected from dynamic checks:
+validation of data, and not just inclusion on types, is an expected
+feature of these systems.
+
+\todo{Monitoring sound problem}
 

### 2021-05-09 paper.tex
Some more on existing work

@@
 system with other existing systems more difficult; in the words
-of Williams, Morrison, and Wadler:
+of Williams, Morris, and Wadler:
 "(...) the monitoring semantics for contracts of intersection and union types given by Keil
@@
 
-Williams, Morrison, and Wadler, take a different approach on flat contracts,
+Williams, Morris, and Wadler, take a different approach on flat contracts,
 they are completely forbidden in their system, explicitly to not encounter
@@
 validation of data, and not just inclusion on types, is an expected
-feature of these systems.
+feature of contracts.
 
-\todo{Monitoring sound problem}
+On top of that, a fundamental problem arises on said paper,
+symptom of the difficulty of establishing good enough, and yet
+simple semantics for union and intersection dynamic checks.
+One of their main results, is the fact that their calculus, obeys
+all of their expected monitoring properties for higher ordered, union
+and, intersection contracts.
+However, one of these properties states the following:
+
+$$ K \in \llbracket A \cap B \rrbracket ~ if ~ K \in \llbracket A \rrbracket \lor K \in \llbracket B \rrbracket $$
+
+Which means that a continuation ($K$) complies negatively with an
+intersection of $A$ and $B$ if it complies with at least one of these.
+Altough this is true, it's a weaker statement of what we'd like to have,
+since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$, 
+where $K_1 \in \llbracket A \rrbracket$ but $K_1 \not \in \llbracket B \rrbracket$,
+and dually for $K_2$, which would mean the continuation presented above would not
+be in $\llbracket A \cap B \rrbracket$, even though our intuition (and the
+introduction to the paper) indicates that it should.
+
+While this mistake does not invalidate their work (the system they present
+handles these correctly), it shows that existing research is still not on
+good shape to handle the growing usecases being developed for
+union and intersection around dynamic languages.
+
+% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$
+
+% Keil and Thiemann did noticed this problem, and their contract satisfaction
+% rules only apply to elimination context, a particular kind of context
+% where the hole is either being applied to an argument
+% ($\square N$) or is used as an argument for a primitive
+% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).
+
+% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
+\unsure{Is this worth mentioning? Should we go so techincal?}
 
@@
        (-> string? string?))
-(λ (x) x))}
+($\lambda$ (x) x))}
 

### 2021-05-11 paper.tex
Started the zoology

@@
 
-\section*{Introduction}
+\section{Introduction}
 Intro\unsure{todo}
@@
   intersection. Including user-define contracts.}
+
+Programming language research no longer lives on the abstract world
+of ideal surfaces.
+Every addition needs to be compatible with years of research
+and development, and this is hard.
+Success was unavoidable.
+
+In this section we briefly introduce some aspects of real world
+programming languages that will cause trouble while interacting
+with union and intersection dynamic checks.
+
+\subsection*{User defined (flat) contracts}
+
+User defined contracts, also called flat contracts, are a particular
+kind of contract that behaves similar to an assertion, and whose
+main point is to apply some arbitrary predicate (as arbitrary as the
+host language allows it) to a value; if the predicate returns true,
+the contract passes, otherwise it fails.
+
+For instance, we could define a flat contract $even?$ as 
+$\lambda x. ~ x ~ \% ~ 2 ~ == ~ 0$, that checks
+whether a number is even.
+And then use it to check that a function always returns an
+even number, for example:
+
+$$let~f~=~(\lambda n. ~ n * 2)@Number ~ \rightarrow ~ even?$$
+
+\unsure{Do we need deeper examples?}
+
+This kind of contracts are present on many different languages,
+for instance, the Eiffel programming language, the precursor
+on the Design by Contract ideals, allows to assert
+exactly these kinds of expression as pre- and post-conditions on
+functions.
+
+The Racket programming language also has a system to work with
+contracts\footnote{And a gradual typing system
+built on top of that, but that goes beyond the scope of this
+work.} that provides a mechanism for defining them; and
+composing these with other kinds of constructs.
+
+Even if flat contracts may seem like too much, these are expected
+by programmers on dynamic languages, where testing and runtime
+assertions are a common pattern for safe programming, and 
+programmers are used to express complex predicates
+over values.
+
+\subsection*{Code manipulation (optimizations)}
+\info{Present inlining and CSE as two major code
+optimizations.
+Purity and immutability as two important factors.}
 

### 2021-05-11 paper.tex
Add lstlisting basic configuration

@@
 %%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%% LST listing confinguraiton %%%%%%%%%%%
+
+\lstdefinelanguage{nickel}{
+    keywords=[1]{
+      if,
+      then,
+      else,
+      switch,
+    },
+    keywords=[2]{
+      let,
+      rec,
+      fun,
+      in
+    },
+    keywords=[3]{
+      Num,
+      Bool,
+      Str,
+      List,
+      Dyn,
+      forall,
+    },
+    keywords=[4]{
+      true,
+      false,
+      null
+    },
+    keywords=[5]{
+      doc,
+      default,
+    },
+    sensitive=true, % keywords are case-sensitive
+    morecomment=[l]{//}, % l is for line comment
+    morecomment=**[is][\color{dark-grey}]{\$}{\$},
+    morestring=[b]", % defines that strings are enclosed in double quotes
+    morestring=[s]{m\#"}{"\#m},
+    moredelim=[is][\color{bright-grey}]{\$}{\$},
+    literate=
+        *{->}{$\rightarrow$}1
+        {=>}{$\Rightarrow$}1
+        {>=}{$\geq$}1
+        {<=}{$\leq$}1
+} %
+
+\lstset{
+  language={nickel},
+  basicstyle=\small\ttfamily, % Global Code Style
+  captionpos=b, % Position of the Caption (t for top, b for bottom)
+  extendedchars=true, % Allows 256 instead of 128 ASCII characters
+  tabsize=2, % number of spaces indented when discovering a tab
+  columns=fixed, % make all characters equal width
+  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
+  showstringspaces=false, % lets spaces in strings appear as real spaces
+  breaklines=true, % wrap lines if they don't fit
+  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
+  frameround=tttt, % make the frame round at all four corners
+  framesep=4pt, % quarter circle size of the round corners
+  numbers=left, % show line numbers at the left
+  numberstyle=\tiny\ttfamily, % style of the line numbers
+  % commentstyle=\color{dark-grey}, % style of comments
+  % keywordstyle=[1]\color{blue-portage}, % style of keywords
+  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
+  % keywordstyle=[3]\color{blue-portage}, % style of keywords
+  % keywordstyle=[4]\color{green-pea}, % style of keywords
+  % keywordstyle=[5]\color{pink-froly}, % style of keywords
+  % stringstyle=\color{blue-marguerite}, % style of strings
+}
 
@@
   \begin{center}
-    \begin{lstlisting}
-        { tag: <Integer, String>, value: Dyn }
+\begin{lstlisting}[language=nickel]
+{tag: <Num, Str>, value: Dyn}
 \end{lstlisting}
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=caml]
 type either = Left of int | Right of string
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[language=nickel]
 { tag : <Left>, val : Num} | { tag : <Right>, val : Str}

### 2021-05-11 paper.tex
Add custom contract example, small fixes

@@
 
+\definecolor{OliveGreen}{RGB}{128,128,0}
 %%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%
@@
     },
+    keywordsprefix=\#,
     keywords=[4]{
@@
 
-Build systems, package managers, operating systems, cloud infrastructures, and
-web services\unsure{Also continuous integration} are examples of modern complex software systems that require an
-extensive and non-trivial configuration in order to make them adapted to each
-different use-case. To manage the growing complexity that is then offloaded to
+Build systems, package managers, operating systems, cloud infrastructure,
+continuous integration and web services\resolved{Also continuous integration}
+are examples of modern complex software systems that require an extensive and
+non-trivial configuration in order to make them adapted to each different
+use-case. To manage the growing complexity that is then offloaded to
 configurations, the declarative approach has become more and more popular,
@@
 
-However, static text-based configuration alone is falling short of expressiveness,
-and is seldom sufficient. Data cannot be transformed, combined nor shared,
-resulting at best in boilerplate and duplication of information, or at worst, in
-data being invalid or inconsistent. Data validation is not supported either, and must be handed over to yet another tool down the configuration pipeline, if ever
-done. Correctly configuring a modern system is hard and failures may have
-substantial negative consequences.
+However, static text-based configuration alone is falling short of
+expressiveness, and is seldom sufficient. Data cannot be transformed, combined
+nor shared, resulting at best in boilerplate and duplication of information, or
+at worst, in data being invalid or inconsistent. Data validation is not
+supported either, and must be handed over to yet another tool down the
+configuration pipeline, if ever done. Correctly configuring a modern system is
+hard and failures may have substantial negative consequences.
 
@@
 \cite{WellTypedBlamed}, the authors precisely introduce a system integrating
-gradual typing with contracts \textit{à la Findler \&
-  Felleisen}.\unsure{The historical bit should probably be moved to
-  related works eventually} Nickel adopts
-a similar type system, with both statically typed terms, dynamically typed
-terms, and first-class contracts. In "first-class contracts", we include the
-ability to write user-defined contracts. Because built-in types are too
-simplistic for the purpose of data validation, the ability of writing a small
-validation function for, say, a URL, and having it immediately available as a
-contract that can be composed with other type
-constructors\unsure{Needs code to illustrate.}, is in our
-opinion one of the main appeal of Nickel's contracts system. In this hybrid
-system, each type constructor of the static type system - arrows, records,
-foralls - must have a contract counterpart that checks at run-time that a value
-is indeed a member of the corresponding type. Contracts for various extensions
-of the simply typed $\lambda$-calculus have been investigated \cite{BlameForAll,
-    KeilThiemannUnionIntersection, RootCauseOfBlame, DependentContracts,
-GradualTypingClasses}, with solutions of varying complexity.  Amongst all the
-common extensions to the polymorphic lambda calculus, one of the most useful and
-widespread turns out to be one of the most challenging: \emph{union types}.
+gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
+historical bit should probably be moved to related works eventually} Nickel
+adopts a similar type system, with both statically typed terms, dynamically
+typed terms, and first-class contracts. In "first-class contracts", we include
+the ability to write user-defined contracts. Because built-in types are too
+simplistic for the purpose of data validation, the ability to write a small
+validation function for, say, a port number, to have it immediately available as
+a contract that can be composed with other type constructors\unsure{Needs code
+to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
+system.
+
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel]
+let Port = contracts.fromPred (fun value =>
+  builtins.isNum value && value % 1 == 0
+  && value >= 0 && value <= 65353) in
+let Configuration = {
+  host | #Host,
+  port | #Port,
+} in
+{host = "nickel-lang.org", port = 80}
+| #Configuration
+\end{lstlisting}
+  \end{center}
+\caption{Contract composition}
+\label{fig:union-encoding}
+\medskip
+\small
+Defines a port contract from a predicate and use it in the definition of a
+\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
+role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
+contract to a value, similar to `:` for static types. We assume that a
+\lstinline+Host+ contract has also been defined.
+\end{figure}
+
+In this hybrid system, each type constructor of the static type system - arrows,
+records, foralls - must have a contract counterpart that checks at run-time that
+a value is indeed a member of the corresponding type. Contracts for various
+extensions of the simply typed $\lambda$-calculus have been investigated
+\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
+DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
+Amongst all the common extensions to the polymorphic lambda calculus, one of the
+most useful and widespread turns out to be one of the most challenging:
+\emph{union types}.
 
@@
 only holds one value whose shape is only known to be in a fixed set of
-alternatives: for eample, either a boolean or a string. Wuen used in combination
+alternatives: for example, either a boolean or a string. When used in combination
 with products, they can represent arbitrary tree-like structures, allowing to
@@
 first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
-Swift and surely many more.
+Swift and many more.
 
@@
 \begin{lstlisting}[language=caml]
-type either = Left of int | Right of string
+type either = Left of float | Right of string
 \end{lstlisting}
@@
 \begin{lstlisting}[language=nickel]
-{ tag : <Left>, val : Num} | { tag : <Right>, val : Str}
+{ tag : <Left>, val : Num}
+| { tag : <Right>, val : Str}
 \end{lstlisting}

### 2021-05-12 paper.tex
Flat contracts improvementes

@@
 
-\subsection*{User defined (flat) contracts}
+\subsection*{Flat contracts}
 
@@
 main point is to apply some arbitrary predicate (as arbitrary as the
-host language allows it) to a value; if the predicate returns true,
-the contract passes, otherwise it fails.
+host language allows it) to a value; if the predicate evaluates to
+true, the contract passes, otherwise it fails.
 
@@
 $\lambda x. ~ x ~ \% ~ 2 ~ == ~ 0$, that checks
-whether a number is even.
+whether a number is even by checking the remainder of dividing it by 2.
 And then use it to check that a function always returns an
-even number, for example:
+even number, when applied to a $Number$:
 
-$$let~f~=~(\lambda n. ~ n * 2)@Number ~ \rightarrow ~ even?$$
+$$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
 
-\unsure{Do we need deeper examples?}
+\unsure{Do we need deeper examples?
+Maybe something showing that flat contracts could evaluate
+a function.}
 
@@
 for instance, the Eiffel programming language, the precursor
-on the Design by Contract ideals, allows to assert
-exactly these kinds of expression as pre- and post-conditions on
-functions.
+of the Design by Contract ideals, allows to assert
+these kinds of expression as pre- and post-conditions on
+functions\cite{EiffelDesignByContract}.
+A somewhat newer language, the D programming
+language, allows, among other things, to have invariants on classes,
+written with flat contracts\cite{DLangContracts}.
 
-The Racket programming language also has a system to work with
+Finally, the Racket programming language also has a system to work with
 contracts\footnote{And a gradual typing system
 built on top of that, but that goes beyond the scope of this
-work.} that provides a mechanism for defining them; and
-composing these with other kinds of constructs.
+work.} that allows to define flat contracts, and
+to compose them with other kinds of dynamic checks,
+like higher order contracts or a lightweight take on union
+and intersection contracts\cite{RacketContracts}.
+\todo{Talk about this, probably later on the paper}
 
-Even if flat contracts may seem like too much, these are expected
+Even if flat contracts may seem like too much, the ability to make
+non trivial checks on values (to validate their output) is expected
 by programmers on dynamic languages, where testing and runtime

### 2021-05-16 paper.tex
Tiny review on flat contracts

@@
 
+\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
+
 \bibliographystyle{ACM-Reference-Format}
@@
 
-For instance, we could define a flat contract $even?$ as
-$\lambda x. ~ x ~ \% ~ 2 ~ == ~ 0$, that checks
+For instance, we could define a flat contract \nickel{even} as
+\nickel{fun x => x \% 2 == 0}, that checks
 whether a number is even by checking the remainder of dividing it by 2.
 And then use it to check that a function always returns an
-even number, when applied to a $Number$:
+even number, when applied to a \nickel{Number}:
 
-$$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=nickel]
+let f = (fun n => n * 2) @ (Number -> even)
+\end{lstlisting}
+  \end{center}
+\caption{A function wrapped with a contract stating that it always returns
+an even number}
+\label{fig:num-to-even}
+\end{figure}
+
+% $$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
 
@@
 language, allows, among other things, to have invariants on classes,
-written with flat contracts\cite{DLangContracts}.
+written with flat contracts, that are
+dinamically checked at the frontiers of every public method \cite{DLangContracts}.
 
 Finally, the Racket programming language also has a system to work with
-contracts\footnote{And a gradual typing system
+contracts that allows to define flat contracts
+\footnote{And a gradual typing system
 built on top of that, but that goes beyond the scope of this
-work.} that allows to define flat contracts, and
+work.}
+, and
 to compose them with other kinds of dynamic checks,
@@
 and intersection contracts\cite{RacketContracts}.
-\todo{Talk about this, probably later on the paper}
+\todo{Talk about this, probably later on the paper.
+See Related Works section}
 
-Even if flat contracts may seem like too much, the ability to make
+Even if flat contracts may seem like too permissive, the ability to make
 non trivial checks on values (to validate their output) is expected
@@
 \info{Including a mention of statically typed systems with union}
+\info{Include something about Racket's and and or contracts}
 

### 2021-05-16 paper.tex
Review on other papers and Racket

@@
 \newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
+\newcommand{\racket}[1]{\lstinline[language=lisp]{#1}}
 
@@
 
+For instance, if the value \nickel{V} is contracted with
+\nickel{A /\\ (B \\/ C)}, then, before the contract is checked,
+the contract gets opened up to something equivalent to
+\nickel{(A /\\ B) \\/ (A /\\ C)}.
+
 This straightforward solutions makes composing the Keil and Thiemann
@@
 together.
-They provide two different set of rules for their calculus,
-the first one is non deterministic, which purpose is to provide
+\todo{This assumes the difficulty was already explained}
+They provide two different sets of rules for their calculus,
+the first one is non deterministic, whose purpose is to provide
 semantics that are simple to understand.
 The second one is a deterministic set of rules, that are intended
-to be simply to implement. These semantics need to perform operations
+to be simple to implement.
+This second set needs to perform operations
 on the evaluation context, to decide whether or not to evaluate
@@
 semantics need to be defined, which translates either to a difficult
-to implement or a difficult to understand system. A second, major one,
+to implement or a difficult to understand system.
+A second, major one,
 is the dependency added over the evaluation context, since, again,
@@
 
-$$ K \in \llbracket A \cap B \rrbracket ~ if ~ K \in \llbracket A \rrbracket \lor K \in \llbracket B \rrbracket $$
+$$ K \in \llbracket A \cap B \rrbracket^- ~ if ~ K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
 
 Which means that a continuation ($K$) complies negatively with an
-intersection of $A$ and $B$ if it complies with at least one of these.
+intersection of $A$ and $B$ if it complies negatively with at least one of these.
 Altough this is true, it's a weaker statement of what we'd like to have,
 since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
-where $K_1 \in \llbracket A \rrbracket$ but $K_1 \not \in \llbracket B \rrbracket$,
+where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
 and dually for $K_2$, which would mean the continuation presented above would not
-be in $\llbracket A \cap B \rrbracket$, even though our intuition (and the
+be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
 introduction to the paper) indicates that it should.
@@
 While this mistake does not invalidate their work (the system they present
-handles these correctly), it shows that existing research is still not on
+handles these cases correctly), it shows that existing research is still not on
 good shape to handle the growing usecases being developed for
@@
 % K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
-\unsure{Is this worth mentioning? Should we go so techincal?}
+\unsure{Is it worth mentioning how continuations are formed?
+Should we go so techincal?}
 
@@
 of an expression could have many different shapes, without necessarily
-having a mark to distinguish between them.
+having a mark to distinguish between them at type level.
 With this in mind, a valid question is how do these languages solve the
@@
 
-Racket is language based on Scheme mainly thought as a language to experiment
+Racket is a language based on Scheme, mainly thought as a language to experiment
 with different language desings and ideas. It provides a very complete contract
-system.
+system\cite{RacketContracts}.
 
-Among those contracts, they provide \texttt{and/c} and \texttt{or/c} contracts.
-These contracts are lightweight versions of unions and intersections, they simply
-check that every contract holds, in the case of and/c, and that at least one contract
-holds, in the case of or/c.\unsure{It doesn't really make sense for a
-  contract to “hold”, in general}
+Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
+These contracts are lightweight versions of intersections and unions, they simply
+check that every contract is valid, in the case of \racket{and/c},
+and that at least one contract
+is valid, in the case of \racket{or/c}.
 
-So, for instance, the usual usecase of intersection to encode overloading
+So, for instance, the usual usecase of intersection to encode bounded overloading
 does not work on Racket. Take a look at the next example, the function
-\texttt{overload} should accept a call like \texttt{(overload 3)}, but it fails
-since it 3 is not a valid \texttt{string?}.
+\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
+since 3 is not a valid \racket{string?}.
 
-\texttt{(define/contract overload
+\begin{figure}[h]
+
+\begin{lstlisting}[language=lisp]
+(define/contract overload
 (and/c (-> number? number?)
        (-> string? string?))
-($\lambda$ (x) x))}
+(lambda (x) x))
+\end{lstlisting}
+\caption{Non working function overloading on Racket.}
 
-Even worst, the or/c contract doesn't allow to have multiple higher order
+\end{figure}
+
+Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
 contracts that could (potentially) apply to a function, these have to be
-differentiated by first order markers (for instance, the number of
-arguments).\unsure{This section is overall a bit hard to follow. I
+differentiated by first order markers.
+For instance, the code on Figure \ref{code:racket:or/c:working}
+would correctly allow only to call
+\racket{united} with a \racket{number?},
+and it can make this distinction since it knows that
+\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
+However, the code on Figure \ref{code:racket:or/c:non-working} would fail
+before running, since the \racket{or/c} contract is unable to know
+which of the two function contracts to use.
+Ideally, this last example should just behave as wrapping
+the function with the \racket{(-> even? even?)} contract.
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=lisp]
+(define/contract united
+(or/c (-> number? number?)
+      (-> string? string? string?))
+(lambda (x) x))
+\end{lstlisting}
+\caption{Working \racket{or/c} example on Racket.}
+\label{code:racket:or/c:working}
+
+\end{figure}
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=lisp]
+(define/contract united
+(or/c (-> number? number?)
+      (-> even? even?))
+(lambda (x) x))
+\end{lstlisting}
+\caption{Non working \racket{or/c} example on Racket.}
+\label{code:racket:or/c:non-working}
+
+\end{figure}
+
+\unsure{This section is overall a bit hard to follow. I
   think I've figured out what it means, but it needs more love. Also
@@
 Racket does provide an alternative for overloading methods, called the
-\texttt{case->} contract, however, this overloading of methods only
+\racket{case->} contract, however, this overloading of methods only
 works over contracts of functions with different amount of parameters.
+It's a good alternative to allow limited overloading of functions.
+As an example, consider the following non failing piece of code, and
+try to understand how Racket can decide, at each application,
+which function contract to use.
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=lisp]
+(define/contract overcase
+  (case-> (-> string? string?)
+          (-> number? number? number?)
+          )
+  (lambda (x [y 0]) (if (number? x)
+                   (+ x y)
+                   x)))
+
+(overcase 1 2)
+
+(overcase "hello")
+\end{lstlisting}
+\end{figure}
 

### 2021-05-17 paper.tex
[WIP] Start filling the optimizations section

@@
 
+The good performances of modern programs heavily rely on complex optimizations
+performed by the compiler or the interpreter. Even more so for functional
+languages, because the functional which model is very different from what actual
+computers are good at executing, naive code generation can exhibit unacceptable
+slowdowns. Functional programs tend to make heavy use of functions, and function
+is a costly operation from the point of view of low-level execution.
+
+One such crucial optimization is inlining. Inlining eliminates a function
+application by substituting the function for its definition at compile time (or
+before execution, for interpreted language). This is especially important for
+short functions that are called repeatedly.
+
+% EXAMPLE INLINING
+
+Another optimization is common subexpression elimination (CSE). CSE somehow
+goes into the other direction: it factors out equal subexpressions to avoid
+recomputing them several times.
+
+% EXAMPLE CSE
+
+Beyond CSE, there exists other similar transformations, such as Haskell's
+let-floating~\ref{} or loop-invariant code motion, which amount to hoist
+invariant out from repeated blocks.
+
+More generally, there is a whole family of opitmizations, such as let-floating,
+\begin{lstlisting}[language=caml]
+type Ordering = Smaller | Equals | Greater
+sort : 'a list
+       -> ('a -> 'a -> Ordering)
+       -> 'a list
+random_list : unit -> int list
+
+sort (random_list ()) (fun x y ->
+  if x < y then Smaller
+  else if x == y then Equals
+  else Greater)
+\end{lstlisting}
+
+Common subexpression elemination (CSE) etc.
+
+\begin{lstlisting}[language=caml]
+
+\end{lstlisting}
+
 \newpage
@@
 validation function for, say, a port number, to have it immediately available as
-a contract that can be composed with other type constructors\unsure{Needs code
+a contract that can be composed with other type constructors\resolved{Needs code
 to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
-system.
+system. An example is given in Figure~\ref{fig:contract-composition}. The first
+line defines a port contract from a predicate and use it in the definition of a
+larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
+contract checks that a value is a record with the following shape: it must have
+a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
+\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
+a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
+entails that a value respect a contract, akin to \lstinline+:+ for types. We
+assume that a \lstinline+Host+ contract has been defined in a similar way.
 
@@
 \caption{Contract composition}
-\label{fig:union-encoding}
+\label{fig:contract-composition}
 \medskip
@@
 records, foralls - must have a contract counterpart that checks at run-time that
-a value is indeed a member of the corresponding type. Contracts for various
+a value is a member of the corresponding type. Contracts for various
 extensions of the simply typed $\lambda$-calculus have been investigated

### 2021-05-17 paper.tex
More on optimizations

@@
     morecomment=[l]{//}, % l is for line comment
-    morecomment=**[is][\color{dark-grey}]{\$}{\$},
+    morecomment=**[is][\color{gray}]{\$}{\$},
     morestring=[b]", % defines that strings are enclosed in double quotes
     morestring=[s]{m\#"}{"\#m},
-    moredelim=[is][\color{bright-grey}]{\$}{\$},
+    moredelim=[is][\color{gray}]{\$}{\$},
     literate=
@@
 
-The good performances of modern programs heavily rely on complex optimizations
+The performance of modern programs heavily relies on non-trivial optimizations
 performed by the compiler or the interpreter. Even more so for functional
-languages, because the functional which model is very different from what actual
-computers are good at executing, naive code generation can exhibit unacceptable
-slowdowns. Functional programs tend to make heavy use of functions, and function
-is a costly operation from the point of view of low-level execution.
+languages, whose operational semantics is usually very different from the actual
+primitives of modern processors: naive code generation can exhibit unacceptable
+slowdowns. Functional programs tend to make heavy use of functions, and a function
+call is a costly operation from the point of view of low-level execution.
 
 One such crucial optimization is inlining. Inlining eliminates a function
-application by substituting the function for its definition at compile time (or
-before execution, for interpreted language). This is especially important for
-short functions that are called repeatedly.
+application by directly substituting the function for its definition at compile
+time (or before execution, for interpreted language). This is especially
+important for small functions that are called repeatedly.
 
-% EXAMPLE INLINING
+\begin{lstlisting}[language=nickel]
+let elem = fun elt =>
+  lists.any (fun x => x == elt) in
+
+let subList = fun l1 l2 =>
+  elem (lists.head l1) l2
+  && subList (list.tail l1) l2
+\end{lstlisting}
 
@@
 
-% EXAMPLE CSE
-
-Beyond CSE, there exists other similar transformations, such as Haskell's
-let-floating~\ref{} or loop-invariant code motion, which amount to hoist
-invariant out from repeated blocks.
-
-More generally, there is a whole family of opitmizations, such as let-floating,
-\begin{lstlisting}[language=caml]
-type Ordering = Smaller | Equals | Greater
-sort : 'a list
-       -> ('a -> 'a -> Ordering)
-       -> 'a list
-random_list : unit -> int list
-
-sort (random_list ()) (fun x y ->
-  if x < y then Smaller
-  else if x == y then Equals
-  else Greater)
+\begin{lstlisting}[language=nickel]
+let elemAtOrLast = fun index list =>
+  if index == lists.length list then
+    lists.elemAt (lists.length list - 1) list
+  else
+    lists.elemAt index list
 \end{lstlisting}
 
-Common subexpression elemination (CSE) etc.
+Similar transformations exist beyond CSE, such as loop-invariant code motion or
+some of Haskell's let-floating~\ref{}, which amount to hoist invariant
+expressions out from repeated blocks in order to avoid recomputing them. In
+particular, take the following example:
 
-\begin{lstlisting}[language=caml]
+\begin{lstlisting}[language=nickel]
+let f = fun x =>
+  g y (x + y)
 
+// We want
+let f =
+  let _g = g y in
+  fun x => \_g (x + y)
 \end{lstlisting}
 
+The soundness of these optimizations is tied the invariance of the semantics of
+programs with respect to substitution. Inlining requires that one can replace
+the application of a function by its body: this is $\beta$-reduction, and it is
+hard to think of a language when it is not a valid transformations. On the other
+hand, CSE-like transformations require an expansion property, of the form
+
+$$M[N/x] \simeq let~x~=~M~in~M[x]$$
+
+This does not necessarily holds, and is in particular trivially violated in
+presence of side effects:
+
+$$let~f~x~=~print~"hi";~(x+1)~in~(f~1, f~1) \\
+\not \simeq let~f~x~=~print~"hi";~(x+1)~in~let~y~=~f~1~in~(y,y)$$
+
+However, this property does hold for \emph{pure} terms, that are terms without
+side-effects. Pure languages, or languages with a form of effect tracking such
+as Haskell, are able to identify such pure terms. Other languages rely on
+sub-approximations (for instance, applying optimizations to arithmetic
+expressions).
 \newpage

### 2021-05-18 paper.tex
Rework the optimizations section and its examples

@@
   numberstyle=\tiny\ttfamily, % style of the line numbers
+  escapeinside={(*@}{@*)} % escape sequence to insert latex
   % commentstyle=\color{dark-grey}, % style of comments
@@
 performed by the compiler or the interpreter. Even more so for functional
-languages, whose operational semantics is usually very different from the actual
-primitives of modern processors: naive code generation can exhibit unacceptable
-slowdowns. Functional programs tend to make heavy use of functions, and a function
-call is a costly operation from the point of view of low-level execution.
+languages, whose operational semantics is usually far removed from the actual
+primitives of modern processors: naive execution can exhibit unacceptable
+slowdowns. Functional programs tend to make heavy use of functions, and a
+function call is a costly operation from the point of view of low-level
+execution.
 
-One such crucial optimization is inlining. Inlining eliminates a function
-application by directly substituting the function for its definition at compile
-time (or before execution, for interpreted language). This is especially
-important for small functions that are called repeatedly.
+One crucial optimization is inlining(Figure~\ref{fig:optimizations-inlining-ex}). Inlining eliminates a function application
+by directly substituting the function for its definition at compile time (or
+before execution, for interpreted language). This is especially important for
+small functions that are called repeatedly.
 
-\begin{lstlisting}[language=nickel]
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
 let elem = fun elt =>
@@
 \end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let subList = fun l1 l2 =>
+  lists.any (fun x => x == elt)
+  && subList (list.tail l1) l2
+\end{lstlisting}
+  \end{center}
+\caption{Inlining}
+\label{fig:optimizations-inlining-ex}
+\end{figure}
 
-Another optimization is common subexpression elimination (CSE). CSE somehow
-goes into the other direction: it factors out equal subexpressions to avoid
-recomputing them several times.
+Another optimization is common subexpression elimination (CSE,
+Figure~\ref{fig:optimizations-cse-ex}). CSE goes into the other direction: it
+factors out equal subexpressions to avoid recomputing them multiple times.
 
-\begin{lstlisting}[language=nickel]
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
 let elemAtOrLast = fun index list =>
-  if index == lists.length list then
+  if index > lists.length list - 1 then
     lists.elemAt (lists.length list - 1) list
@@
 \end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let elemAtOrLast = fun index list =>
+  let l = lists.length list - 1 in
+  if index > l then
+    lists.elemAt l list
+  else
+    lists.elemAt index list
+\end{lstlisting}
+  \end{center}
+\caption{Common subexpression elimination}
+\label{fig:optimizations-cse-ex}
+\end{figure}
 
@@
 expressions out from repeated blocks in order to avoid recomputing them. In
-particular, take the following example:
+particular, take the example of Figure~\ref{fig:optimizations-let-floating-ex}.
 
-\begin{lstlisting}[language=nickel]
-let f = fun x =>
-  g y (x + y)
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
+let f = fun x => g y (x + 1) 
+\end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let f' = let g' = g y in
+  fun x => g' (x + 1)
+\end{lstlisting}
+  \end{center}
+\caption{Let-floating}
+\label{fig:optimizations-let-floating-ex}
+\end{figure}
 
-// We want
-let f =
-  let _g = g y in
-  fun x => \_g (x + y)
+The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
+called. This may be costly, in particular in the presence of contracts: if
+\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
+on it, the penalty could be linear in the size of \lstinline+y+. A sensible
+thing to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
+Figure~\ref{fig:optimizations-let-floating-ex}, which is something Haskell's
+let-floating transformations could indeed do, given \lstinline+g+ is pure.
+
+The soundness of these optimizations is tied the invariance of the semantics of
+programs with respect to some substitutive transformations\todo{bad naming}.
+Inlining requires that one can replace the application of a function by its
+body: this is $\beta$-reduction, and it is hard to think of a language where it
+is not a valid transformations. On the other hand, CSE-like transformations
+require an expansion property of the form
+
+$$M[N/x] \simeq let~x~=~N~in~M[x]$$
+
+This equation clearly fails in a call-by-value language with side effects:
+
+\begin{lstlisting}[language=Nickel]
+let f x = print "hi";(x+1) in (f 1,f 1) 
+(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
+(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
 \end{lstlisting}
 
-The soundness of these optimizations is tied the invariance of the semantics of
-programs with respect to substitution. Inlining requires that one can replace
-the application of a function by its body: this is $\beta$-reduction, and it is
-hard to think of a language when it is not a valid transformations. On the other
-hand, CSE-like transformations require an expansion property, of the form
+However, this property does hold for \emph{pure} terms, that are terms without
+side-effects. In a pure language, or a language with effect tracking like
+Haskell, pure terms can be safely identified for applying optimizations. Other
+languages rely on under-approximations.
 
-$$M[N/x] \simeq let~x~=~M~in~M[x]$$
-
-This does not necessarily holds, and is in particular trivially violated in
-presence of side effects:
-
-$$let~f~x~=~print~"hi";~(x+1)~in~(f~1, f~1) \\
-\not \simeq let~f~x~=~print~"hi";~(x+1)~in~let~y~=~f~1~in~(y,y)$$
-
-However, this property does hold for \emph{pure} terms, that are terms without
-side-effects. Pure languages, or languages with a form of effect tracking such
-as Haskell, are able to identify such pure terms. Other languages rely on
-sub-approximations (for instance, applying optimizations to arithmetic
-expressions).
+\unsure{Should we tease already here that unions and intersections break this
+even for pure terms?}
 \newpage

### 2021-05-18 paper.tex
Add examples to Union/intersection section

@@
 
+% Content:
+%   - Unions are useful as a fundamental data block
+%     Examples: nullable, tree-like data structure, multiple representations
+
+\begin{lstlisting}
+type BinaryTree =
+Integer | (BinaryTree, BinaryTree)
+\end{lstlisting}
+
+\begin{lstlisting}
+optionalValue : String | Null
+\end{lstlisting}
+
+\begin{lstlisting}
+person : {name : String, age : Number}
+  | {name : String, dateOfBirth : String}
+\end{lstlisting}
+
+%   - Intersections are useful for subtyping or refinement (integers/floats
+%       e.g)
+\begin{lstlisting}
+1 : Int & Float
+\end{lstlisting}
+%   - Intersections are useful to encode overloading
+\begin{lstlisting}
+incr: (Int -> Int) & (Float -> Float)
+\end{lstlisting}
+
+%   - Intersections are useful to encode mixins (see Scala3)
+\begin{lstlisting}
+Comparator: {cmp: T -> T -> Bool}
+Shower: {show: T -> String}
+
+showSmaller : (dict: Comparator & Shower) -> T -> T -> String
+\end{lstlisting}
+%   - Intersections appears in some form as unions' dual
+\begin{lstlisting}
+\end{lstlisting}
+
+% Things to settle: how to motivate them, how to compare them to ADTs. Purely
+% dynamic approach, or typed approach too?
 \newpage

### 2021-05-18 paper.tex
Racket highlighting

@@
 
+% Copied from https://hal.inria.fr/hal-01140459/file/racket.inc.tex
+\lstdefinelanguage{racket} {
+  morekeywords=[1]{define, define-syntax, define-macro, lambda, define-stream, stream-lambda},
+  morekeywords=[2]{begin, call-with-current-continuation, call/cc,
+    call-with-input-file, call-with-output-file, case, cond,
+    do, else, for-each, if,
+    let*, let, let-syntax, letrec, letrec-syntax,
+    define-context, define-controller, Integer, Boolean, get, when-required, when-provided,
+    maybe_publish, require, submod, or/c, ->, \#\%module-begin,
+    always_publish, with-syntax, define-struct/contract, syntax-case,
+    define/contract,
+    let-values, let*-values,
+    module, provide,
+    and, or, not, delay, force,
+    \#`, \#',
+    \#lang, implement, begin-for-syntax, rename-out,
+    quasiquote, quote, unquote, unquote-splicing,
+    map, fold, syntax, syntax-rules, eval, environment, query },
+  morekeywords=[3]{import, export},
+  alsoletter={',`,-,/,>,<,\#,\%},
+  morecomment=[l]{;},
+%  literate={lambda}{{\lambdaup}}1, % lambda -- look at https://tex.stackexchange.com/questions/119879/math-symbols-in-tt-font
+  moredelim=**[is][\color{light-gray}]{<<@<<}{>>@>>},
+  moredelim=**[is][\itshape\color{OliveGreen}]{<<;<<}{>>;>>},
+  morecomment=[s]{\#|}{|\#},
+  sensitive=true,
+}
+
 \lstset{
@@
 \newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
-\newcommand{\racket}[1]{\lstinline[language=lisp]{#1}}
+\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
 
@@
 
+\unsure{Should we define contracts prior to this?}
 User defined contracts, also called flat contracts, are a particular
@@
 And then use it to check that a function always returns an
-even number, when applied to a \nickel{Number}:
+even number, when applied to a \nickel{Integer}:
 
@@
 \begin{lstlisting}[language=nickel]
-let f = (fun n => n * 2) @ (Number -> even)
+let f = (fun n => n * 2) @ (Integer -> even)
 \end{lstlisting}
@@
 
-\begin{lstlisting}[language=lisp]
+\begin{lstlisting}[language=racket]
 (define/contract overload
@@
 
-\begin{lstlisting}[language=lisp]
+\begin{lstlisting}[language=racket]
 (define/contract united
@@
 
-\begin{lstlisting}[language=lisp]
+\begin{lstlisting}[language=racket]
 (define/contract overcase

### 2021-05-18 paper.tex
some comments
In-call modifications

@@
 \begin{lstlisting}[language=nickel,title={Optimized program}]
-let f' = let g' = g y in
-  fun x => g' (x + 1)
+let g' = g y
+let f' = fun x => g' (x + 1)
 \end{lstlisting}
@@
 side-effects. In a pure language, or a language with effect tracking like
-Haskell, pure terms can be safely identified for applying optimizations. Other
-languages rely on under-approximations.
+Haskell\todo{Cite a different language, or remove citation}, pure terms can be
+safely identified for applying optimizations. Other languages rely on
+under-approximations.
+
+\subsubsection*{Lazyness}
+\unsure{Should we talk about lazyness?
+this would be a good place to introduce it}
 
@@
 even for pure terms?}
+
+\subsection*{Laziness}
+
 \newpage
@@
 
-% Content:
-%   - Unions are useful as a fundamental data block
-%     Examples: nullable, tree-like data structure, multiple representations
+Unions are useful as a fundamental data block.
+Examples: nullable, tree-like data structure, multiple representations
 
-\begin{lstlisting}
+\begin{lstlisting}[title={Tree-like data structure}]
 type BinaryTree =
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[title={Plain enumeration}]
+protocol : `Http | `Ftp | `Sftp
+\end{lstlisting}
+
+\begin{lstlisting}[title={Nullable values}]
 optionalValue : String | Null
@@
 
-\begin{lstlisting}
-person : {name : String, age : Number}
+\begin{lstlisting}[title={Alternative representations}]
+person : {name : String, age : Int}
   | {name : String, dateOfBirth : String}
@@
 %       e.g)
-\begin{lstlisting}
+\begin{lstlisting}[title={Value subtyping}]
 1 : Int & Float
@@
 %   - Intersections are useful to encode overloading
-\begin{lstlisting}
-incr: (Int -> Int) & (Float -> Float)
+\begin{lstlisting}[title={Overloading}]
+incr: (Int -> Int) & (Float -> Float) =
+  fun x => x + 1
 \end{lstlisting}
@@
 %   - Intersections are useful to encode mixins (see Scala3)
-\begin{lstlisting}
+\begin{lstlisting}[title={Structural mixins}]
 Comparator: {cmp: T -> T -> Bool}
@@
 %   - Intersections appears in some form as unions' dual
-\begin{lstlisting}
+\begin{lstlisting}[title={Intersection as the dual of unions}]
+f : (Num | String) -> Bool
+(*@ $\simeq$ @*) f : (Num -> Bool) (*@ $\cup$ @*) (String -> Bool)
 \end{lstlisting}
 
+\todo{Either here or the next section, show examples of unintuitive behaviour
+of U and I. For example, I is not jos composition of contracts.
+And union has to be treated as any one of the options}
 % Things to settle: how to motivate them, how to compare them to ADTs. Purely
 % dynamic approach, or typed approach too?
+
 \newpage
@@
 
-\begin{lstlisting}[language=racket]
+\begin{lstlisting}[language=lisp]
 (define/contract united

### 2021-05-19 paper.tex
[WIP]Union & intersection

@@
 
-However, this property does hold for \emph{pure} terms, that are terms without
-side-effects. In a pure language, or a language with effect tracking like
-Haskell\todo{Cite a different language, or remove citation}, pure terms can be
-safely identified for applying optimizations. Other languages rely on
-under-approximations.
+In the example above, the first term prints \nickel{"hi"} two times while the
+second term prints it only once. However, this property does hold for
+\emph{pure} terms, that are terms without side-effects. In a pure language, or a
+language with effect tracking like Haskell\todo{Cite a different language, or
+remove citation}, pure terms can be safely identified for applying
+optimizations. Other languages rely on under-approximations.
+
+\unsure{Should we tease already here that unions and intersections break this
+even for pure terms?}
 
@@
 
-\unsure{Should we tease already here that unions and intersections break this
-even for pure terms?}
-
-\subsection*{Laziness}
-
 \newpage
@@
 
-Unions are useful as a fundamental data block.
-Examples: nullable, tree-like data structure, multiple representations
-
+\subsection*{Unions}
+Unions are fundamental building blocks of programs. A union holds a value whose
+shape is known to be in a fixed set of alternatives: for example, a value that
+can be either a boolean or a string. They corresponds logically to conjunctions.
+Used in combination with products, they can represent arbitrary tree-like
+structures, allowing to model a large class of data
+(\ref{fig:union-adt-ex}).
+
+\label{fig:union-adt-ex}
 \begin{lstlisting}[title={Tree-like data structure}]
@@
 
+Other use-cases for unions include:
+\begin{itemize}
+    \item To represent nullable values in JSON (\ref{fig:union-nullable-ex}).
+    \item For plain enumerations (\ref{fig:union-enum-ex}).
+    \item To support different representations of the same data
+        (\ref{fig:union-alt-ex}).
+\end{itemize}
+All of which are useful in a configuration language.
+
+\label{fig:union-nullable-ex}
+\begin{lstlisting}[title={Nullable values}]
+optionalValue : String | Null
+\end{lstlisting}
+
+\label{fig:union-enum-ex}
 \begin{lstlisting}[title={Plain enumeration}]
@@
 
-\begin{lstlisting}[title={Nullable values}]
-optionalValue : String | Null
-\end{lstlisting}
-
+\label{fig:union-alt-ex}
 \begin{lstlisting}[title={Alternative representations}]
@@
 
-%   - Intersections are useful for subtyping or refinement (integers/floats
-%       e.g)
+In statically typed languages, and more specifically in functional programming
+languages, sums (also called tagged unions, or discriminated unions) are usually
+preferred, as constructors for algebraic datatypes. \unsure{This is a very
+    misleading statement at best. Sums are "tagged union" sure, but that refers
+to an encoding a sums with unions, certainly not an encoding of unions with
+sums!} [references]. Sums are unions' cousin where the alternatives are
+syntactically separated by an associated discriminating label. Well-known
+programming languages with first-class tagged unions include Haskell, OCaml,
+Rust, Scala, PureScript, Elm, Swift and many more.
+
+On the other hand, in dynamically typed languages, it is customary to directly
+store different kinds of data in one variable without further ceremony. The
+runtime system needs to attach type tags to values anyway, which can usually be
+observed using functions like JavaScript's \verb+typeof+. Using such bare
+unions\resolved{If you can observe with typeof, then it's pretty much a tagged
+union, though the tag is implicit} has already been an idiom in dynamically
+typed language for a long time. Being able to type and handle these unions is an
+important aspect of the migration to a gradual type system. Indeed, a lot of
+gradually typed languages either existed originally as dynamically typed
+language or aim at backward compatibility with a dynamically typed base
+language. This is why despite the presence of a static type system, gradually
+typed languages tend to favor bare unions types which is the typed counterpart
+of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
+
+In the case of Nickel, there is no pre-existing untyped language that we have to
+support, so we are freer of our choice. However, Nickel is a configuration
+language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
+untagged unions, there is a strong incentive to support untagged unions at the
+type level, or at least, at the contract level.
+
+\subsection*{Intersections}
+
+Intersections are less prevalent. An intersection is a value that belongs to to
+two types at the same time. A simple example is a number literal \nickel{1}
+which could be accepted as both an \nickel{Int} and a \nickel{Float} without
+having to perform conversions(Figure~\ref{fig:intersection-subytping-ex}). In
+this example, the intersection \nickel{Int & Float} plays the role of a subtype
+of both \nickel{Int} and \nickel{Float}.
+
+\label{fig:intersection-subytping-ex}
 \begin{lstlisting}[title={Value subtyping}]
@@
 \end{lstlisting}
-%   - Intersections are useful to encode overloading
+
+Intersections of higher-order types corresponding to overloading
+(Figure~\ref{fig:intersection-overloading-ex}).
+
+\label{fig:intersection-overloading-ex}
 \begin{lstlisting}[title={Overloading}]
@@
 
-%   - Intersections are useful to encode mixins (see Scala3)
+They can also be used to combine constraints (such as traits or interfaces), as
+done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
+
+\label{fig:intersection-mixins-ex}
 \begin{lstlisting}[title={Structural mixins}]
@@
 \end{lstlisting}
-%   - Intersections appears in some form as unions' dual
+
+Finally, intersection naturally show up as the dual of unions. Thus, in a system
+featuring unions, a union in a negative positions is a form of intersection
+(Figure~\ref{fig:intersection-dual}).
+
+\label{fig:intersection-dual}
 \begin{lstlisting}[title={Intersection as the dual of unions}]

### 2021-05-20 paper.tex
Smallfixes + review of section 5

@@
 having to perform conversions(Figure~\ref{fig:intersection-subytping-ex}). In
-this example, the intersection \nickel{Int & Float} plays the role of a subtype
-of both \nickel{Int} and \nickel{Float}.
+this example, the intersection \nickel{Int & Float} can be used as a subtype of
+both \nickel{Int} and \nickel{Float}.
 
@@
 
-For instance, if the value \nickel{V} is contracted with
-\nickel{A /\\ (B \\/ C)}, then, before the contract is checked,
+For instance, if the value \nickel{V} is contracted\todo{(Yann) Contracted means
+something else I think} with
+\nickel{A & (B | C)}, then, before the contract is checked,
 the contract gets opened up to something equivalent to
-\nickel{(A /\\ B) \\/ (A /\\ C)}.
+\nickel{(A & B) | (A & C)}.
 
@@
 of Williams, Morris, and Wadler:
-"(...) the monitoring semantics for contracts of intersection and union types given by Keil
-and Thiemann are not uniform. (...) If uniformity helps composition, then
-special cases can hinder composition."\cite{RootCauseOfBlame}
+"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
+and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
+special cases can hinder composition."\cite{RootCauseOfBlame}\info{(Yann) Does
+it also make it less efficient? If $(A \& B)$ fails because of $B$, their
+development causes to recheck the contract $A$ that has been duplicated?}
 
@@
 semantics need to be defined, which translates either to a difficult
-to implement or a difficult to understand system.
+to implement or a difficult to understand system\unsure{(Yann) Is it really a
+    problem per se? It is common thing to do (a nice declarative system
+    for proofs and an algorithmic one for implementation, proved equivalent).
+    Maybe the point is that the algorithmic system is way more complex than the
+declarative one}.
 A second, major one,
@@
 it makes composing this, with other features desired on programming languages,
-difficult, or even impossible (as the case with common subexpression elimination).
+difficult, or even impossible (as the case with common subexpression
+elimination)\unsure{(Yann) I think we need to substantiate this claim. If this
+context dependency explains why CSE is invalid, maybe have a little example?}
+\todo{(Yann) TODO: see if there's a low hanging complexity
+    argument we can make (like it's quadratic in the number of open contracts or
+whatever)}.
 
@@
 
-$$ K \in \llbracket A \cap B \rrbracket^- ~ if ~ K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
+$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
 
@@
 Should we go so techincal?}
+\todo{(Yann) Maybe adding a concrete example of this (it can be the same term but with
+    concrete types like Number and String and simple contexts like application
+to a dumb argument) is sufficient to make the point}
 
@@
 differentiated by first order markers.
-For instance, the code on Figure \ref{code:racket:or/c:working}
+For instance, the code on Figure~\ref{code:racket:or/c:working}
 would correctly allow only to call
@@
 \racket{(lambda (x) x)} can not be a function that takes 2 parameters.
-However, the code on Figure \ref{code:racket:or/c:non-working} would fail
+However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
 before running, since the \racket{or/c} contract is unable to know

### 2021-05-20 paper.tex
Pass on the optimization section

@@
 
-The performance of modern programs heavily relies on non-trivial optimizations
-performed by the compiler or the interpreter. Even more so for functional
-languages, whose operational semantics is usually far removed from the actual
-primitives of modern processors: naive execution can exhibit unacceptable
-slowdowns. Functional programs tend to make heavy use of functions, and a
-function call is a costly operation from the point of view of low-level
-execution.
-
-One crucial optimization is inlining(Figure~\ref{fig:optimizations-inlining-ex}). Inlining eliminates a function application
-by directly substituting the function for its definition at compile time (or
-before execution, for interpreted language). This is especially important for
-small functions that are called repeatedly.
+The performance of modern programs heavily relies on the optimizations performed
+by the compiler or the interpreter. Even more so for functional languages, whose
+execution model is often far removed from actual processors, causing naive
+execution to exhibit unacceptable slowdowns.
+
+One crucial optimization is
+inlining(Figure~\ref{fig:optimizations-inlining-ex}). Functional programs tend
+to make heavy use of functions, and a function call is a costly operation from
+the point of view of low-level execution. Inlining eliminates a function
+application by directly substituting the function for its definition at compile
+time (or before execution, for interpreted language). This is especially
+efficient for small functions that are called repeatedly.
 
@@
 
-Another optimization is common subexpression elimination (CSE,
-Figure~\ref{fig:optimizations-cse-ex}). CSE goes into the other direction: it
-factors out equal subexpressions to avoid recomputing them multiple times.
+While inlining expands an expression by substituting a definition for its value,
+an opposite transformation is sometimes beneficial, for example when a composite
+expression is repeated several times. This causes the same expression to be
+wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
+consist in introducing a new identifier for this expression and using the
+identifier in place of the original occurrences
+(Figure~\ref{fig:optimizations-cse-ex}), computing the result once and for all.
 
@@
 
-Similar transformations exist beyond CSE, such as loop-invariant code motion or
-some of Haskell's let-floating~\ref{}, which amount to hoist invariant
-expressions out from repeated blocks in order to avoid recomputing them. In
-particular, take the example of Figure~\ref{fig:optimizations-let-floating-ex}.
+Beyond CSE, optimizations such as loop-invariant code motion or
+let-floating~\ref{} apply the same principle of extracting an invariant
+expression to avoid recomputation respectively across loop iterations and
+function calls. For example, take the code of
+Figure~\ref{fig:optimizations-let-floating-ex}.
 
@@
 \lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
-on it, the penalty could be linear in the size of \lstinline+y+. A sensible
-thing to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
-Figure~\ref{fig:optimizations-let-floating-ex}, which is something Haskell's
-let-floating transformations could indeed do, given \lstinline+g+ is pure.
-
-The soundness of these optimizations is tied the invariance of the semantics of
-programs with respect to some substitutive transformations\todo{bad naming}.
-Inlining requires that one can replace the application of a function by its
-body: this is $\beta$-reduction, and it is hard to think of a language where it
-is not a valid transformations. On the other hand, CSE-like transformations
-require an expansion property of the form
+on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
+to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
+Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
+let-floating transformation could indeed do (given \lstinline+g+ is pure, as
+explained below).
+
+The soundness of all these optimizations is tied to the invariance of the
+semantics of programs with respect to specific substitutions. Inlining requires
+that one can replace the application of a function by its body, which is
+basically $\beta$-reduction: as long as the arguments are evaluated following
+the language's strategy, it is a valid transformation. On the other hand,
+CSE-like transformations require an expansion property of the form
 
@@
 remove citation}, pure terms can be safely identified for applying
-optimizations. Other languages rely on under-approximations.
+optimizations.
+
+Strikingly, we will see in section~\ref{} that the introduction of unions and
+intersections can make this equivalence \todo{name it} unsound \emph{even in a
+pure setting}, making the optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
 

### 2021-05-20 paper.tex
Small fixes unions section

@@
 shape is known to be in a fixed set of alternatives: for example, a value that
-can be either a boolean or a string. They corresponds logically to conjunctions.
+can be either a boolean or a string. They correspond logically to conjunctions.
 Used in combination with products, they can represent arbitrary tree-like
@@
 
-Other use-cases for unions include:
+In addition to internal program representations, unions are also useful to model
+external data:
 \begin{itemize}
-    \item To represent nullable values in JSON (\ref{fig:union-nullable-ex}).
+    \item To represent nullable values(\ref{fig:union-nullable-ex}).
     \item For plain enumerations (\ref{fig:union-enum-ex}).
@@
 \end{itemize}
-All of which are useful in a configuration language.
+All of which are common in configurations.
 

### 2021-05-20 paper.tex
small fixes 5 and 2.1

@@
 And then use it to check that a function always returns an
-even number, when applied to a \nickel{Integer}:
+even number, when applied to an \nickel{Integer}, as shown in
+Figure \ref{fig:num-to-even}.
+Or even, we could come up with a contract that checks that a function
+maps \nickel{0} to \nickel{0}, as can be seen on Figure
+\ref{fig:0-to-0}.
 
@@
 
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel]
+fun f => f 0 == 0
+\end{lstlisting}
+    \end{center}
+  \caption{A contract that checks that a function maps 0 to 0}
+  \label{fig:0-to-0}
+\end{figure}
 \unsure{Do we need deeper examples?
@@
 Even if flat contracts may seem like too permissive, the ability to make
-non trivial checks on values (to validate their output) is expected
+non trivial checks on values (that is, to validate their output) is expected
 by programmers on dynamic languages, where testing and runtime
-assertions are a common pattern for safe programming, and
+assertions are extremely useful tools for safe programming, and
 programmers are used to express complex predicates
@@
 let subList = fun l1 l2 =>
-  lists.any (fun x => x == elt)
+  lists.any (fun x => x == (lists.head l1)) l2
   && subList (list.tail l1) l2
@@
 
-$$M[N/x] \simeq let~x~=~N~in~M[x]$$
+$$M[N] \simeq let~x~=~N~in~M[x]$$
 
@@
 can be either a boolean or a string. They correspond logically to conjunctions.
+\unsure{disjunctions, right?}
 Used in combination with products, they can represent arbitrary tree-like
@@
 
-For instance, if the value \nickel{V} is contracted\todo{(Yann) Contracted means
-something else I think} with
+For instance, if the value \nickel{V} is wrapped with contract
+\todo{(Yann) Contracted means
+something else I think.
+(Teo) changed}
 \nickel{A & (B | C)}, then, before the contract is checked,
-the contract gets opened up to something equivalent to
+it will get opened up to something equivalent to
 \nickel{(A & B) | (A & C)}.
@@
 and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition."\cite{RootCauseOfBlame}\info{(Yann) Does
+special cases can hinder composition."\cite{RootCauseOfBlame}
+\info{(Yann) Does
 it also make it less efficient? If $(A \& B)$ fails because of $B$, their
 development causes to recheck the contract $A$ that has been duplicated?}
+And it also could harm performance, or at least complicate an implementation,
+since some sharing mechanism should be used to avoid checking the same contract,
+over the same value, multiple times.
 
@@
 semantics need to be defined, which translates either to a difficult
-to implement or a difficult to understand system\unsure{(Yann) Is it really a
+to implement or a difficult to understand system
+\unsure{(Yann) Is it really a
     problem per se? It is common thing to do (a nice declarative system
@@
     Maybe the point is that the algorithmic system is way more complex than the
-declarative one}.
+declarative one
+(Teo) I think you're right, I'll try to reword it}.
 A second, major one,
@@
 simple semantics for union and intersection dynamic checks.
-One of their main results, is the fact that their calculus, obeys
+One of their main results, is the fact that their calculus obeys
 all of their expected monitoring properties for higher ordered, union
@@
 intersection of $A$ and $B$ if it complies negatively with at least one of these.
+
 Altough this is true, it's a weaker statement of what we'd like to have,
-since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
-where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
-and dually for $K_2$, which would mean the continuation presented above would not
-be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
+since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
+where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
+but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
+and dually for $not~(f ~ true)$.
+Which would mean, according to their semantics, that
+the continuation presented above would not
+be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
+even though our intuition (and the
 introduction to the paper) indicates that it should.
+
+% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
+% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
+% and dually for $K_2$, which would mean the continuation presented above would not
+% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
+% introduction to the paper) indicates that it should.
 

### 2021-05-21 paper.tex
[wip] issues in semantics

@@
 \subsection*{Code manipulation (optimizations)}
+\label{sec:optimizations}
 \info{Present inlining and CSE as two major code
@@
   intersection literally}
+
+Type systems featuring intersection types are hard in a fundamental way,
+from their discovery, they've been studied as a way to characterize
+lambda-calculus terms normalizability; which means that typechecking a
+term on a system with intersection types is undecidable,
+similar to how the halting problem is undecidable\todo{reference}.
+
+Usually, when thinking about programming languages, there's an intuition
+that static checks (like typechecking) are more difficult and less powerful
+than dynamic checks (like testing).
+For instance, it is very easy to check if,
+in a given execution of a program, a variable holds an integer value; while it
+may become more challenging to check if a variable will always hold an integer value,
+no matter the particular execution.
+
+As we intend to show on this paper, unions and intersections are not only
+statically hard to check, but they are also dynamically hard;
+\emph{they are fundamentally hard}.
+In this section, we present intuitively the main challenges that arise
+when attempting to develop a runtime check system for unions and intersections.
+
+\subsubsection*{Union contracts are hard to brake}
+
+When programming with dynamic checks it is assumed that lack of
+failure does not mean that the program works perfectly.
+For instance, imagine having a function wrapped in a contract
+that checks that maps positive numbers to positive numbers, like
+function \nickel{f} on Figure \ref{fig:pos-to-pos}.
+
+Clearly, there is an error on this program, but if the first call
+to \nickel{f} that gets executed is \nickel{f 10}, the error will not
+be found at that point, and the first operand of the addition would be
+evaluated to \nickel{3}.
+If, later on, \nickel{f 5} gets executed, the contract would get violated,
+since \nickel{f} gets called with a positive value, but returns a negative
+one (\nickel{-2}).
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (fun x => x - 7) @ Positive -> Positive in
+(f 10) + (f 5) 
+\end{lstlisting}
+\caption{Wrong positive to positive function}
+\label{fig:pos-to-pos}
+\end{figure}
+
+Even if the program performed two different applications of \nickel{f}
+before finding the contract violation, the first one is not actually
+needed, and executing only \nickel{f 5} is enough to find the problem.
+
+Having a witness of size 1 for every contract failure is a nice to have property
+since it means, for example, that every time that a contract is violated,
+the stack trace is enough to understand how it was violated.
+However, when unions get introduced, it is lost.
+Consider now the program presented at Figure \ref{fig:wrong-union-function},
+this program (and in particular, \nickel{f}'s contract) still presents a problem,
+since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
+nor does it map positive numbers to non-positive numbers, so it won't be a valid
+value for the union of those contracts.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
+(f 10) + (f 5) 
+\end{lstlisting}
+\caption{Still wrong positive to positive or positive to non-positive function}
+\label{fig:wrong-union-function}
+\end{figure}
+
+Still, the program presents enough function calls to find that issue
+and violate the contract,
+since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
+and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
+But, what makes this case completely different from the one presented before, is that
+both function calls are needed, and, what is more, there is not a single way to
+use the wrapped value \nickel{f} that would find a violation to its contract;
+at least two are needed.
+
+This seemingly simple consequence generates major issues when considered in real
+life languages.
+For example, remember the different mechanisms for code manipulation presented
+in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
+both of these rely on the fact that substituting a pure function call by its body,
+or joining equal pure expressions and sharing their result, is a semantically valid
+transformation.
+However, consider the examples presented at Figure \ref{fig:optimized-programs},
+the first one is the inlined version of the second one, while the second
+one has something like a CSE applied to it.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Inlined]
+let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
+(f 1 true, f 1 false)
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
+let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
+let g x = f 1 x in
+(g true, g false)
+\end{lstlisting}
+\caption{Equivalent programs, with inlining or CSE applied}
+\label{fig:optimized-programs}
+\end{figure}
+\todo{Improve example}
+
+However, this two programs behave differently, the first one will return
+\nickel{(1, "False)}, without failing.
+While the second one will fail, since the single application of \nickel{f 1},
+applied to both \nickel{true} and \nickel{false} will be able to find the error.
+
+Summing it up, wrapping a value with a union of higer order contracts means that
+that value now has a state, that may get modified silently at every use site.
+Both reviewed papers suffer from this defect since it is the semantics
+of unions that interact badly with code manipulation operations,
+not a particular implementation.
 

### 2021-05-21 paper.tex
check on unions and inters

@@
 important aspect of the migration to a gradual type system. Indeed, a lot of
+\unsure{This is the first time discussing gradual type systems, maybe it should
+be introduced earlier, in particular how they relate to contracts 
+(I think is mentioned on the appendix)}
 gradually typed languages either existed originally as dynamically typed
@@
 
-Intersections are less prevalent. An intersection is a value that belongs to to
+Intersections are less prevalent. An intersection is a value that belongs to
 two types at the same time. A simple example is a number literal \nickel{1}
 which could be accepted as both an \nickel{Int} and a \nickel{Float} without
-having to perform conversions(Figure~\ref{fig:intersection-subytping-ex}). In
+having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
 this example, the intersection \nickel{Int & Float} can be used as a subtype of
@@
 f : (Num | String) -> Bool
-(*@ $\simeq$ @*) f : (Num -> Bool) (*@ $\cup$ @*) (String -> Bool)
+(*@ $\simeq$ @*) f : (Num -> Bool) & (String -> Bool)
 \end{lstlisting}

### 2021-05-24 paper.tex
Review mode, preparing for submission
Typo
Smallfixes
sec 2 small changes

@@
-\documentclass[sigplan,10pt]{acmart}
+\documentclass[sigplan,10pt,review]{acmart}
 %\usepackage{amssymb}
@@
 Or even, we could come up with a contract that checks that a function
-maps \nickel{0} to \nickel{0}, as can be seen on Figure
-\ref{fig:0-to-0}.
+maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
+and use it as a contract for the \nickel{not} function, as can be seen on Figure
+\ref{fig:checked-not}.
 
@@
 \begin{lstlisting}[language=nickel]
-fun f => f 0 == 0
+let notC = fun f => (f true == false) and (f false == true) in
+let checkedNot = not @ notC
 \end{lstlisting}
 \end{center}
-  \caption{A contract that checks that a function maps 0 to 0}
-  \label{fig:0-to-0}
-\end{figure}
-\unsure{Do we need deeper examples?
-Maybe something showing that flat contracts could evaluate
-a function.}
-
-This kind of contracts are present on many different languages,
+\caption{A contract that checks the not function behaves as expected}
+\label{fig:checked-not}
+\end{figure}
+
+This type of contracts are present on many different languages,
 for instance, the Eiffel programming language, the precursor
@@
 Finally, the Racket programming language also has a system to work with
-contracts that allows to define flat contracts
+contracts, powerful enough to define flat contracts
 \footnote{And a gradual typing system
@@
 assertions are extremely useful tools for safe programming, and
-programmers are used to express complex predicates
+programmers are accustomed to express complex predicates
 over values.
@@
 One crucial optimization is
-inlining(Figure~\ref{fig:optimizations-inlining-ex}). Functional programs tend
+inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
 to make heavy use of functions, and a function call is a costly operation from
@@
 
-\begin{figure}
+\info{I think it's better to use [h] on figures, we can discuss it}
+\begin{figure}[h]
   \begin{center}
@@
 wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
-consist in introducing a new identifier for this expression and using the
+consists in introducing a new identifier for this expression and using the
 identifier in place of the original occurrences
-(Figure~\ref{fig:optimizations-cse-ex}), computing the result once and for all.
-
-\begin{figure}
+(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.
+
+\begin{figure}[h]
   \begin{center}
@@
 shape is known to be in a fixed set of alternatives: for example, a value that
-can be either a boolean or a string. They correspond logically to conjunctions.
+can be either a boolean or a string. They correspond logically to disjunctions.
 \unsure{disjunctions, right?}
@@
 
-\subsubsection*{Union contracts are hard to brake}
+\subsubsection*{Union contracts are hard to break}
 
@@
 introduction to the paper) indicates that it should.
+\todo{Simplify example}
 
@@
 
+\bibliography{nickel}
+
 \appendix
+
+\todo{Remove parts that are already on the main part}
 
@@
 
-Unions are fundamental and ubiquitous building blocks of program data. A union
+\todo{Specify: union types? more general concept?}{Unions} are fundamental and ubiquitous building blocks of program data. A union
 only holds one value whose shape is only known to be in a fixed set of
@@
 
-\bibliography{nickel}
-
 \end{document}

### 2021-05-25 paper.tex
Wip on flat contracts and intersection

@@
 \subsection*{Flat contracts}
+\label{sec:flat-contracts}
 
@@
 can be either a boolean or a string. They correspond logically to disjunctions.
-\unsure{disjunctions, right?}
 Used in combination with products, they can represent arbitrary tree-like
@@
 
+\change{These next few labels need to be inside figures,
+they are not working}
 \label{fig:union-adt-ex}
@@
 For instance, imagine having a function wrapped in a contract
-that checks that maps positive numbers to positive numbers, like
+that checks that it maps positive numbers to positive numbers, like
 function \nickel{f} on Figure \ref{fig:pos-to-pos}.
@@
 since it means, for example, that every time that a contract is violated,
-the stack trace is enough to understand how it was violated.
+the stack trace is enough to understand how it was violated, and to potentially
+come up with a minimal example that breaks the contract.
 However, when unions get introduced, it is lost.
@@
 But, what makes this case completely different from the one presented before, is that
-both function calls are needed, and, what is more, there is not a single way to
-use the wrapped value \nickel{f} that would find a violation to its contract;
+both function calls are needed, and, what is more, there is not a way to
+use the wrapped value \nickel{f} only once that would find a violation to its contract;
 at least two are needed.
@@
 
-However, this two programs behave differently, the first one will return
-\nickel{(1, "False)}, without failing.
+Unexpectedly, this two programs behave differently, the first one will return
+\nickel{(1, "False")}, without failing.
 While the second one will fail, since the single application of \nickel{f 1},
@@
 not a particular implementation.
+This addition of global state complicates both the work of the language
+implementor, as well as the language user, since it adds non trivial
+side effects to your code.
+\unsure{Maybe mention CSE is actualy commonly done by hand as well}
+
+\subsubsection*{[WIP] Flat contracts are too permissive}
+
+Stop for a second and think how you could implement an intersection
+contract. The first thought you may have is "just apply
+both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
+Of course, if it were that simple we wouldn't be writing this paper,
+this idea starts to show problems when you first combine it with
+higher order contracts, since it would mean not complying with
+at least one of these contracts negatively would mean breaking the
+whole thing, as shown on the second example on the same Figure.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Implementation]
+x @ A & B
+(*@ $\simeq$ @*)
+(x @ A) @ B
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Problem]
+let g = (fun x => x) @ (Number -> Number) & (String -> String) in
+g 1
+\end{lstlisting}
+\caption{Implementations of intersection contracts: first iteration}
+\label{fig:inter-contracts}
+\end{figure}
+
+Still, not all hope is lost, we could pass some extra shared state
+to each one of the sub contracts and let them only fail if the other
+side of the contract has failed as well, as shown in Figure
+\ref{fig:inter-contracts-2}, where the sared state is
+represented by a label \nickel{l}.
+This idea is still quite reasonable, for instance, when executing
+\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
+that gets checked is the argument of \nickel{String -> String} over the function parameter, so
+something like \nickel{1 @ String}.
+Of course, this fails, but before raising a violation, this contract could
+check if the other side of the intersection (\nickel{Number -> Number}) has
+also tried to raise negative blame (on the parameter \nickel{1}), since
+it hasn't, the contract would return normally, and would mark itself as failed.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Implementation]
+x @ A & B
+(*@ $\simeq$ @*)
+(x @ A[l]) @ B[l]
+\end{lstlisting}
+\caption{Implementations of intersection contracts: second iteration}
+\label{fig:inter-contracts-2}
+\end{figure}
+
+This idea, which is an oversimplification of the ideas
+presented by Williams, Morris, and Wadler, seems to be spot on.
+However, it presents a major problem, it's not compatible with flat contracts,
+presented in Section \ref{sec:flat-contracts}.
+Consider the example presented on Figure \ref{fig:inter-flat-contracts},
+if the implementation presented previously were to be used, we may end up with
+something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
+\nickel{l} represents the shared state.
+At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
+\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
+This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
+fails negatively.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let C = fun f => (f 0) == 0 in
+let g = (fun x => x) @ (String -> String) & C
+\end{lstlisting}
+\caption{Intersection and flat contracts}
+\label{fig:inter-flat-contracts}
+\end{figure}
+
+But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
+\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
+Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
+failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
+should not have been used inside the evaluation of \nickel{C}, which
+adds an extra layer of complexity, contracts should be guarded, and execution
+depends on the shape of the surrounding evaluation context.
 

### 2021-05-25 paper.tex
small fixes sec 5

@@
 together.
-\todo{This assumes the difficulty was already explained}
 They provide two different sets of rules for their calculus,
@@
 a given contract.
-
-We see two problems here, a minor one is the fact that two different
-semantics need to be defined, which translates either to a difficult
-to implement or a difficult to understand system
+\unsure{Worth showing how the non deterministic version works? (See section B.1.1)}
+
+We see two problems here, a minor one is the fact that the two
+different semantics defined for the calculus are really far apart
+one from the other, both on the strategies used for contract resolution,
+as well as on the complexity levels.
 \unsure{(Yann) Is it really a
@@
 declarative one
-(Teo) I think you're right, I'll try to reword it}.
+(Teo) reworded}.
 A second, major one,
@@
 it makes composing this, with other features desired on programming languages,
-difficult, or even impossible (as the case with common subexpression
-elimination)\unsure{(Yann) I think we need to substantiate this claim. If this
-context dependency explains why CSE is invalid, maybe have a little example?}
+difficult, or even impossible
+\unsure{(Yann) I think we need to substantiate this claim. If this
+context dependency explains why CSE is invalid, maybe have a little example?
+(Teo) I removed it, I don't think there's an example, mainly since
+this context dependency is not part of the language. I do think it complicates
+implementation of said feature}
 \todo{(Yann) TODO: see if there's a low hanging complexity

### 2021-05-25 paper.tex
[WIP] Add intro, rework typing

@@
 
-\newpage
-
-\section{Features of languages}
-\info{Zoology of various features that we will eventually show
-  conflict with this or that property or implementation of union and
-  intersection. Including user-define contracts.}
-
-Programming language research no longer lives on the abstract world
-of ideal surfaces.
-Every addition needs to be compatible with years of research
-and development, and this is hard.
-Success was unavoidable.
-
-In this section we briefly introduce some aspects of real world
-programming languages that will cause trouble while interacting
-with union and intersection dynamic checks.
-
-\subsection*{Flat contracts}
-\label{sec:flat-contracts}
-
-\unsure{Should we define contracts prior to this?}
-User defined contracts, also called flat contracts, are a particular
-kind of contract that behaves similar to an assertion, and whose
-main point is to apply some arbitrary predicate (as arbitrary as the
-host language allows it) to a value; if the predicate evaluates to
-true, the contract passes, otherwise it fails.
-
-For instance, we could define a flat contract \nickel{even} as
-\nickel{fun x => x \% 2 == 0}, that checks
-whether a number is even by checking the remainder of dividing it by 2.
-And then use it to check that a function always returns an
-even number, when applied to an \nickel{Integer}, as shown in
-Figure \ref{fig:num-to-even}.
-Or even, we could come up with a contract that checks that a function
-maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
-and use it as a contract for the \nickel{not} function, as can be seen on Figure
-\ref{fig:checked-not}.
-
-\begin{figure}[h]
-\begin{center}
-\begin{lstlisting}[language=nickel]
-let f = (fun n => n * 2) @ (Integer -> even)
-\end{lstlisting}
-  \end{center}
-\caption{A function wrapped with a contract stating that it always returns
-an even number}
-\label{fig:num-to-even}
-\end{figure}
-
-% $$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let notC = fun f => (f true == false) and (f false == true) in
-let checkedNot = not @ notC
-\end{lstlisting}
-\end{center}
-\caption{A contract that checks the not function behaves as expected}
-\label{fig:checked-not}
-\end{figure}
-
-This type of contracts are present on many different languages,
-for instance, the Eiffel programming language, the precursor
-of the Design by Contract ideals, allows to assert
-these kinds of expression as pre- and post-conditions on
-functions\cite{EiffelDesignByContract}.
-A somewhat newer language, the D programming
-language, allows, among other things, to have invariants on classes,
-written with flat contracts, that are
-dinamically checked at the frontiers of every public method \cite{DLangContracts}.
-
-Finally, the Racket programming language also has a system to work with
-contracts, powerful enough to define flat contracts
-\footnote{And a gradual typing system
-built on top of that, but that goes beyond the scope of this
-work.}
-, and
-to compose them with other kinds of dynamic checks,
-like higher order contracts or a lightweight take on union
-and intersection contracts\cite{RacketContracts}.
-\todo{Talk about this, probably later on the paper.
-See Related Works section}
-
-Even if flat contracts may seem like too permissive, the ability to make
-non trivial checks on values (that is, to validate their output) is expected
-by programmers on dynamic languages, where testing and runtime
-assertions are extremely useful tools for safe programming, and
-programmers are accustomed to express complex predicates
-over values.
-
-\subsection*{Code manipulation (optimizations)}
-\label{sec:optimizations}
-\info{Present inlining and CSE as two major code
-optimizations.
-Purity and immutability as two important factors.}
-
-The performance of modern programs heavily relies on the optimizations performed
-by the compiler or the interpreter. Even more so for functional languages, whose
-execution model is often far removed from actual processors, causing naive
-execution to exhibit unacceptable slowdowns.
-
-One crucial optimization is
-inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
-to make heavy use of functions, and a function call is a costly operation from
-the point of view of low-level execution. Inlining eliminates a function
-application by directly substituting the function for its definition at compile
-time (or before execution, for interpreted language). This is especially
-efficient for small functions that are called repeatedly.
-
-\info{I think it's better to use [h] on figures, we can discuss it}
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let elem = fun elt =>
-  lists.any (fun x => x == elt) in
-
-let subList = fun l1 l2 =>
-  elem (lists.head l1) l2
-  && subList (list.tail l1) l2
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let subList = fun l1 l2 =>
-  lists.any (fun x => x == (lists.head l1)) l2
-  && subList (list.tail l1) l2
-\end{lstlisting}
-  \end{center}
-\caption{Inlining}
-\label{fig:optimizations-inlining-ex}
-\end{figure}
-
-While inlining expands an expression by substituting a definition for its value,
-an opposite transformation is sometimes beneficial, for example when a composite
-expression is repeated several times. This causes the same expression to be
-wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
-consists in introducing a new identifier for this expression and using the
-identifier in place of the original occurrences
-(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let elemAtOrLast = fun index list =>
-  if index > lists.length list - 1 then
-    lists.elemAt (lists.length list - 1) list
-  else
-    lists.elemAt index list
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let elemAtOrLast = fun index list =>
-  let l = lists.length list - 1 in
-  if index > l then
-    lists.elemAt l list
-  else
-    lists.elemAt index list
-\end{lstlisting}
-  \end{center}
-\caption{Common subexpression elimination}
-\label{fig:optimizations-cse-ex}
-\end{figure}
-
-Beyond CSE, optimizations such as loop-invariant code motion or
-let-floating~\ref{} apply the same principle of extracting an invariant
-expression to avoid recomputation respectively across loop iterations and
-function calls. For example, take the code of
-Figure~\ref{fig:optimizations-let-floating-ex}.
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let f = fun x => g y (x + 1)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let g' = g y
-let f' = fun x => g' (x + 1)
-\end{lstlisting}
-  \end{center}
-\caption{Let-floating}
-\label{fig:optimizations-let-floating-ex}
-\end{figure}
-
-The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
-called. This may be costly, in particular in the presence of contracts: if
-\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
-on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
-to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
-Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
-let-floating transformation could indeed do (given \lstinline+g+ is pure, as
-explained below).
-
-The soundness of all these optimizations is tied to the invariance of the
-semantics of programs with respect to specific substitutions. Inlining requires
-that one can replace the application of a function by its body, which is
-basically $\beta$-reduction: as long as the arguments are evaluated following
-the language's strategy, it is a valid transformation. On the other hand,
-CSE-like transformations require an expansion property of the form
-
-$$M[N] \simeq let~x~=~N~in~M[x]$$
-
-This equation clearly fails in a call-by-value language with side effects:
-
-\begin{lstlisting}[language=Nickel]
-let f x = print "hi";(x+1) in (f 1,f 1)
-(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
-(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
-\end{lstlisting}
-
-In the example above, the first term prints \nickel{"hi"} two times while the
-second term prints it only once. However, this property does hold for
-\emph{pure} terms, that are terms without side-effects. In a pure language, or a
-language with effect tracking like Haskell\todo{Cite a different language, or
-remove citation}, pure terms can be safely identified for applying
-optimizations.
-
-Strikingly, we will see in section~\ref{} that the introduction of unions and
-intersections can make this equivalence \todo{name it} unsound \emph{even in a
-pure setting}, making the optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
-
-\unsure{Should we tease already here that unions and intersections break this
-even for pure terms?}
-
-\subsubsection*{Lazyness}
-\unsure{Should we talk about lazyness?
-this would be a good place to introduce it}
-
-\newpage
-
-\section{Union \& intersection}
-\info{What are they, what are they used for.}
-
-\subsection*{Unions}
-Unions are fundamental building blocks of programs. A union holds a value whose
-shape is known to be in a fixed set of alternatives: for example, a value that
-can be either a boolean or a string. They correspond logically to disjunctions.
-Used in combination with products, they can represent arbitrary tree-like
-structures, allowing to model a large class of data
-(\ref{fig:union-adt-ex}).
-
-\change{These next few labels need to be inside figures,
-they are not working}
-\label{fig:union-adt-ex}
-\begin{lstlisting}[title={Tree-like data structure}]
-type BinaryTree =
-Integer | (BinaryTree, BinaryTree)
-\end{lstlisting}
-
-In addition to internal program representations, unions are also useful to model
-external data:
-\begin{itemize}
-    \item To represent nullable values(\ref{fig:union-nullable-ex}).
-    \item For plain enumerations (\ref{fig:union-enum-ex}).
-    \item To support different representations of the same data
-        (\ref{fig:union-alt-ex}).
-\end{itemize}
-All of which are common in configurations.
-
-\label{fig:union-nullable-ex}
-\begin{lstlisting}[title={Nullable values}]
-optionalValue : String | Null
-\end{lstlisting}
-
-\label{fig:union-enum-ex}
-\begin{lstlisting}[title={Plain enumeration}]
-protocol : `Http | `Ftp | `Sftp
-\end{lstlisting}
-
-\label{fig:union-alt-ex}
-\begin{lstlisting}[title={Alternative representations}]
-person : {name : String, age : Int}
-  | {name : String, dateOfBirth : String}
-\end{lstlisting}
-
-In statically typed languages, and more specifically in functional programming
-languages, sums (also called tagged unions, or discriminated unions) are usually
-preferred, as constructors for algebraic datatypes. \unsure{This is a very
-    misleading statement at best. Sums are "tagged union" sure, but that refers
-to an encoding a sums with unions, certainly not an encoding of unions with
-sums!} [references]. Sums are unions' cousin where the alternatives are
-syntactically separated by an associated discriminating label. Well-known
-programming languages with first-class tagged unions include Haskell, OCaml,
-Rust, Scala, PureScript, Elm, Swift and many more.
-
-On the other hand, in dynamically typed languages, it is customary to directly
-store different kinds of data in one variable without further ceremony. The
-runtime system needs to attach type tags to values anyway, which can usually be
-observed using functions like JavaScript's \verb+typeof+. Using such bare
-unions\resolved{If you can observe with typeof, then it's pretty much a tagged
-union, though the tag is implicit} has already been an idiom in dynamically
-typed language for a long time. Being able to type and handle these unions is an
-important aspect of the migration to a gradual type system. Indeed, a lot of
-\unsure{This is the first time discussing gradual type systems, maybe it should
-be introduced earlier, in particular how they relate to contracts 
-(I think is mentioned on the appendix)}
-gradually typed languages either existed originally as dynamically typed
-language or aim at backward compatibility with a dynamically typed base
-language. This is why despite the presence of a static type system, gradually
-typed languages tend to favor bare unions types which is the typed counterpart
-of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
-
-In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. However, Nickel is a configuration
-language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
-untagged unions, there is a strong incentive to support untagged unions at the
-type level, or at least, at the contract level.
-
-\subsection*{Intersections}
-
-Intersections are less prevalent. An intersection is a value that belongs to
-two types at the same time. A simple example is a number literal \nickel{1}
-which could be accepted as both an \nickel{Int} and a \nickel{Float} without
-having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
-this example, the intersection \nickel{Int & Float} can be used as a subtype of
-both \nickel{Int} and \nickel{Float}.
-
-\label{fig:intersection-subytping-ex}
-\begin{lstlisting}[title={Value subtyping}]
-1 : Int & Float
-\end{lstlisting}
-
-Intersections of higher-order types corresponding to overloading
-(Figure~\ref{fig:intersection-overloading-ex}).
-
-\label{fig:intersection-overloading-ex}
-\begin{lstlisting}[title={Overloading}]
-incr: (Int -> Int) & (Float -> Float) =
-  fun x => x + 1
-\end{lstlisting}
-
-They can also be used to combine constraints (such as traits or interfaces), as
-done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
-
-\label{fig:intersection-mixins-ex}
-\begin{lstlisting}[title={Structural mixins}]
-Comparator: {cmp: T -> T -> Bool}
-Shower: {show: T -> String}
-
-showSmaller : (dict: Comparator & Shower) -> T -> T -> String
-\end{lstlisting}
-
-Finally, intersection naturally show up as the dual of unions. Thus, in a system
-featuring unions, a union in a negative positions is a form of intersection
-(Figure~\ref{fig:intersection-dual}).
-
-\label{fig:intersection-dual}
-\begin{lstlisting}[title={Intersection as the dual of unions}]
-f : (Num | String) -> Bool
-(*@ $\simeq$ @*) f : (Num -> Bool) & (String -> Bool)
-\end{lstlisting}
-
-\todo{Either here or the next section, show examples of unintuitive behaviour
-of U and I. For example, I is not jos composition of contracts.
-And union has to be treated as any one of the options}
-% Things to settle: how to motivate them, how to compare them to ADTs. Purely
-% dynamic approach, or typed approach too?
-
-\newpage
-
-\section{Issues coming from the semantics}
-\info{What goes wrong or difficult when trying to implement union and
-  intersection literally}
-
-Type systems featuring intersection types are hard in a fundamental way,
-from their discovery, they've been studied as a way to characterize
-lambda-calculus terms normalizability; which means that typechecking a
-term on a system with intersection types is undecidable,
-similar to how the halting problem is undecidable\todo{reference}.
-
-Usually, when thinking about programming languages, there's an intuition
-that static checks (like typechecking) are more difficult and less powerful
-than dynamic checks (like testing).
-For instance, it is very easy to check if,
-in a given execution of a program, a variable holds an integer value; while it
-may become more challenging to check if a variable will always hold an integer value,
-no matter the particular execution.
-
-As we intend to show on this paper, unions and intersections are not only
-statically hard to check, but they are also dynamically hard;
-\emph{they are fundamentally hard}.
-In this section, we present intuitively the main challenges that arise
-when attempting to develop a runtime check system for unions and intersections.
-
-\subsubsection*{Union contracts are hard to break}
-
-When programming with dynamic checks it is assumed that lack of
-failure does not mean that the program works perfectly.
-For instance, imagine having a function wrapped in a contract
-that checks that it maps positive numbers to positive numbers, like
-function \nickel{f} on Figure \ref{fig:pos-to-pos}.
-
-Clearly, there is an error on this program, but if the first call
-to \nickel{f} that gets executed is \nickel{f 10}, the error will not
-be found at that point, and the first operand of the addition would be
-evaluated to \nickel{3}.
-If, later on, \nickel{f 5} gets executed, the contract would get violated,
-since \nickel{f} gets called with a positive value, but returns a negative
-one (\nickel{-2}).
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f = (fun x => x - 7) @ Positive -> Positive in
-(f 10) + (f 5) 
-\end{lstlisting}
-\caption{Wrong positive to positive function}
-\label{fig:pos-to-pos}
-\end{figure}
-
-Even if the program performed two different applications of \nickel{f}
-before finding the contract violation, the first one is not actually
-needed, and executing only \nickel{f 5} is enough to find the problem.
-
-Having a witness of size 1 for every contract failure is a nice to have property
-since it means, for example, that every time that a contract is violated,
-the stack trace is enough to understand how it was violated, and to potentially
-come up with a minimal example that breaks the contract.
-However, when unions get introduced, it is lost.
-Consider now the program presented at Figure \ref{fig:wrong-union-function},
-this program (and in particular, \nickel{f}'s contract) still presents a problem,
-since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
-nor does it map positive numbers to non-positive numbers, so it won't be a valid
-value for the union of those contracts.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
-(f 10) + (f 5) 
-\end{lstlisting}
-\caption{Still wrong positive to positive or positive to non-positive function}
-\label{fig:wrong-union-function}
-\end{figure}
-
-Still, the program presents enough function calls to find that issue
-and violate the contract,
-since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
-and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
-But, what makes this case completely different from the one presented before, is that
-both function calls are needed, and, what is more, there is not a way to
-use the wrapped value \nickel{f} only once that would find a violation to its contract;
-at least two are needed.
-
-This seemingly simple consequence generates major issues when considered in real
-life languages.
-For example, remember the different mechanisms for code manipulation presented
-in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
-both of these rely on the fact that substituting a pure function call by its body,
-or joining equal pure expressions and sharing their result, is a semantically valid
-transformation.
-However, consider the examples presented at Figure \ref{fig:optimized-programs},
-the first one is the inlined version of the second one, while the second
-one has something like a CSE applied to it.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Inlined]
-let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
-(f 1 true, f 1 false)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
-let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
-let g x = f 1 x in
-(g true, g false)
-\end{lstlisting}
-\caption{Equivalent programs, with inlining or CSE applied}
-\label{fig:optimized-programs}
-\end{figure}
-\todo{Improve example}
-
-Unexpectedly, this two programs behave differently, the first one will return
-\nickel{(1, "False")}, without failing.
-While the second one will fail, since the single application of \nickel{f 1},
-applied to both \nickel{true} and \nickel{false} will be able to find the error.
-
-Summing it up, wrapping a value with a union of higer order contracts means that
-that value now has a state, that may get modified silently at every use site.
-Both reviewed papers suffer from this defect since it is the semantics
-of unions that interact badly with code manipulation operations,
-not a particular implementation.
-This addition of global state complicates both the work of the language
-implementor, as well as the language user, since it adds non trivial
-side effects to your code.
-\unsure{Maybe mention CSE is actualy commonly done by hand as well}
-
-\subsubsection*{[WIP] Flat contracts are too permissive}
-
-Stop for a second and think how you could implement an intersection
-contract. The first thought you may have is "just apply
-both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
-Of course, if it were that simple we wouldn't be writing this paper,
-this idea starts to show problems when you first combine it with
-higher order contracts, since it would mean not complying with
-at least one of these contracts negatively would mean breaking the
-whole thing, as shown on the second example on the same Figure.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
-x @ A & B
-(*@ $\simeq$ @*)
-(x @ A) @ B
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Problem]
-let g = (fun x => x) @ (Number -> Number) & (String -> String) in
-g 1
-\end{lstlisting}
-\caption{Implementations of intersection contracts: first iteration}
-\label{fig:inter-contracts}
-\end{figure}
-
-Still, not all hope is lost, we could pass some extra shared state
-to each one of the sub contracts and let them only fail if the other
-side of the contract has failed as well, as shown in Figure
-\ref{fig:inter-contracts-2}, where the sared state is
-represented by a label \nickel{l}.
-This idea is still quite reasonable, for instance, when executing
-\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
-that gets checked is the argument of \nickel{String -> String} over the function parameter, so
-something like \nickel{1 @ String}.
-Of course, this fails, but before raising a violation, this contract could
-check if the other side of the intersection (\nickel{Number -> Number}) has
-also tried to raise negative blame (on the parameter \nickel{1}), since
-it hasn't, the contract would return normally, and would mark itself as failed.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
-x @ A & B
-(*@ $\simeq$ @*)
-(x @ A[l]) @ B[l]
-\end{lstlisting}
-\caption{Implementations of intersection contracts: second iteration}
-\label{fig:inter-contracts-2}
-\end{figure}
-
-This idea, which is an oversimplification of the ideas
-presented by Williams, Morris, and Wadler, seems to be spot on.
-However, it presents a major problem, it's not compatible with flat contracts,
-presented in Section \ref{sec:flat-contracts}.
-Consider the example presented on Figure \ref{fig:inter-flat-contracts},
-if the implementation presented previously were to be used, we may end up with
-something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
-\nickel{l} represents the shared state.
-At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
-\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
-This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
-fails negatively.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let C = fun f => (f 0) == 0 in
-let g = (fun x => x) @ (String -> String) & C
-\end{lstlisting}
-\caption{Intersection and flat contracts}
-\label{fig:inter-flat-contracts}
-\end{figure}
-
-But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
-\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
-Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
-failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
-should not have been used inside the evaluation of \nickel{C}, which
-adds an extra layer of complexity, contracts should be guarded, and execution
-depends on the shape of the surrounding evaluation context.
-
-\newpage
-
-\section{Issues in concrete restrictions from the literature}
-\info{The various papers and the tradeoffs they make}
-
-The papers that previously researched union and intersection
-as dynamic checks have, in our eyes, different shortcomings
-that makes them suboptimal solutions.
-
-Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
-they noticed that, since union contracts should share information between
-different application contexts, while intersection contracts should behave
-independently between each application context, the system becomes much
-simpler if every contract gets transformed into a union of intersections
-of other contracts, by using the usual De Morgan laws,
-and then eagerly opening up unions, but postponing intersection
-contracts as much as possible.
-\todo{This assumes the difficulty has been previously explained}
-
-For instance, if the value \nickel{V} is wrapped with contract
-\todo{(Yann) Contracted means
-something else I think.
-(Teo) changed}
-\nickel{A & (B | C)}, then, before the contract is checked,
-it will get opened up to something equivalent to
-\nickel{(A & B) | (A & C)}.
-
-This straightforward solutions makes composing the Keil and Thiemann
-system with other existing systems more difficult; in the words
-of Williams, Morris, and Wadler:
-"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
-and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition."\cite{RootCauseOfBlame}
-\info{(Yann) Does
-it also make it less efficient? If $(A \& B)$ fails because of $B$, their
-development causes to recheck the contract $A$ that has been duplicated?}
-And it also could harm performance, or at least complicate an implementation,
-since some sharing mechanism should be used to avoid checking the same contract,
-over the same value, multiple times.
-
-Another difficult issue Keil and Thiemann were able to solve, is the
-fact that flat contracts and intersection do not usually play nice
-together.
-They provide two different sets of rules for their calculus,
-the first one is non deterministic, whose purpose is to provide
-semantics that are simple to understand.
-The second one is a deterministic set of rules, that are intended
-to be simple to implement.
-This second set needs to perform operations
-on the evaluation context, to decide whether or not to evaluate
-a given contract.
-\unsure{Worth showing how the non deterministic version works? (See section B.1.1)}
-
-We see two problems here, a minor one is the fact that the two
-different semantics defined for the calculus are really far apart
-one from the other, both on the strategies used for contract resolution,
-as well as on the complexity levels.
-\unsure{(Yann) Is it really a
-    problem per se? It is common thing to do (a nice declarative system
-    for proofs and an algorithmic one for implementation, proved equivalent).
-    Maybe the point is that the algorithmic system is way more complex than the
-declarative one
-(Teo) reworded}.
-A second, major one,
-is the dependency added over the evaluation context, since, again,
-it makes composing this, with other features desired on programming languages,
-difficult, or even impossible
-\unsure{(Yann) I think we need to substantiate this claim. If this
-context dependency explains why CSE is invalid, maybe have a little example?
-(Teo) I removed it, I don't think there's an example, mainly since
-this context dependency is not part of the language. I do think it complicates
-implementation of said feature}
-\todo{(Yann) TODO: see if there's a low hanging complexity
-    argument we can make (like it's quadratic in the number of open contracts or
-whatever)}.
-
-Williams, Morris, and Wadler, take a different approach on flat contracts,
-they are completely forbidden in their system, explicitly to not encounter
-the previously mentioned issue.
-This restriction goes against what is usually expected from dynamic checks:
-validation of data, and not just inclusion on types, is an expected
-feature of contracts.
-
-On top of that, a fundamental problem arises on said paper,
-symptom of the difficulty of establishing good enough, and yet
-simple semantics for union and intersection dynamic checks.
-One of their main results, is the fact that their calculus obeys
-all of their expected monitoring properties for higher ordered, union
-and, intersection contracts.
-However, one of these properties states the following:
-
-$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
-
-Which means that a continuation ($K$) complies negatively with an
-intersection of $A$ and $B$ if it complies negatively with at least one of these.
-
-Altough this is true, it's a weaker statement of what we'd like to have,
-since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
-where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
-but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
-and dually for $not~(f ~ true)$.
-Which would mean, according to their semantics, that
-the continuation presented above would not
-be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
-even though our intuition (and the
-introduction to the paper) indicates that it should.
-\todo{Simplify example}
-
-% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
-% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
-% and dually for $K_2$, which would mean the continuation presented above would not
-% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
-% introduction to the paper) indicates that it should.
-
-While this mistake does not invalidate their work (the system they present
-handles these cases correctly), it shows that existing research is still not on
-good shape to handle the growing usecases being developed for
-union and intersection around dynamic languages.
-
-% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$
-
-% Keil and Thiemann did noticed this problem, and their contract satisfaction
-% rules only apply to elimination context, a particular kind of context
-% where the hole is either being applied to an argument
-% ($\square N$) or is used as an argument for a primitive
-% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).
-
-% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
-\unsure{Is it worth mentioning how continuations are formed?
-Should we go so techincal?}
-\todo{(Yann) Maybe adding a concrete example of this (it can be the same term but with
-    concrete types like Number and String and simple contexts like application
-to a dumb argument) is sufficient to make the point}
-
-\newpage
-
-\section{Related work}
-\info{Including a mention of statically typed systems with union}
-\info{Include something about Racket's and and or contracts}
-
-\newpage
-
-\section{Conclusion}
-
-\bibliography{nickel}
-
-\appendix
-
-\todo{Remove parts that are already on the main part}
-
-\section{Background on union types}
-
-\subsection{Nickel design space}
-
-{\color{red}Goals of the section
-
-Introduce Nickel, explain the choice of gradual
-typing and main design orientations, such as practice-oriented, lightweight,
-etc.\vspace{0.5cm}}
-
 Build systems, package managers, operating systems, cloud infrastructure,
 continuous integration and web services\resolved{Also continuous integration}
-are examples of modern complex software systems that require an extensive and
-non-trivial configuration in order to make them adapted to each different
-use-case. To manage the growing complexity that is then offloaded to
-configurations, the declarative approach has become more and more popular,
-illustrated for example by the infrastructure-as-code paradigm. This leads
-configuration to play an important role in critical aspects of software
-engineering, including security, availability, and maintainability.
-
-However, static text-based configuration alone is falling short of
-expressiveness, and is seldom sufficient. Data cannot be transformed, combined
-nor shared, resulting at best in boilerplate and duplication of information, or
-at worst, in data being invalid or inconsistent. Data validation is not
-supported either, and must be handed over to yet another tool down the
-configuration pipeline, if ever done. Correctly configuring a modern system is
-hard and failures may have substantial negative consequences.
+are examples of modern complex software systems which require an extensive and
+non-trivial configuration. To handle this complexity, the declarative approach
+has become more and more prominent: instead of applying a sequence of
+transformations to a system in order to reach a desired state, this state is
+described as static textual data. Some tool is then in charge of applying the
+necessary transformations. This approach makes the configuration persistent,
+reproducible, analyzable and simulable. Text-based configurations are led to
+play an important role in critical aspects of software engineering, including
+security, availability, and maintainability.
+
+However, static text-based configurations show their limits beyond a given size
+and complexity. Data cannot be transformed, combined, shared, nor validated.
+This lack of expressivity results at best in boilerplate and duplication of
+information, but it can also make data invalid or inconsistent. Correctly
+configuring a modern system is hard and failures may have substantial negative
+consequences.
 
@@
 straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
-lightweight and easy to write while still providing features for writing correct
-code and generating valid data. The focus is first and foremost practical:
-Nickel originated as a rethink of the language of the Nix package manager, used
-to describe one of the biggest sets of software packages\cite{repology}.  Nickel
-targets build systems and cloud deployment tools (Kubernetes, Terraform or
-NixOps) as well. While we are keen on incorporating existing or doing new
-research when it solves a well-defined problem, the general design goals are
-always guided by the practical industrial use cases.
-
-\subsection{Typing}
-
-Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
-Unlike JSON, though, Nickel anticipates large configurations by being both
-programmable and typed. In the configuration setting, there is a singular
-dilemma with respect to typing: since a program is run on fixed inputs and is
-expected to terminate, any relevant type error will show up at evaluation.  Why
-bother with the complexity of a static type system? On the other hand, more and
-more software systems offload complexity to configurations, in particular with
-the infrastructure as code paradigm prevailing today in cloud deployment. When
-the complexity of a codebase grows, static types become attractive again.  For
-reusable code — that is, library functions —, static types are specifically
-adapted, and bring all the usual benefits of early error detection, code
-robustness, better code structure, documentation, and so on.
+lightweight and easy to write while providing support for writing correct code
+and generating valid data. Nickel originated as a rethink of the language of the
+Nix package manager, used to describe one of the biggest sets of packaged
+software in existence\cite{repology}.  Nickel targets build systems and cloud deployment
+tools (Kubernetes, Terraform or NixOps) as well.
+
+\subsection*{To type or not to type}
+
+The end goal being data generation, Nickel is a dynamic, pure and functional
+language. Our goal is that writing a simple configuration in Nickel feels as
+easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
+configurations by being programmable, and by providing means of ensuring
+correctness.
+
+In this context, correctness means two things:
+\begin{enumerate}
+    \item The program runs without errors. For example, an error is raised when
+        evaluating the nonsensical division \nickel{1 / "ftp"}.
+    \item The program generates valid data, i.e. data that conforms to a schema
+        (imposed by the system consuming the configuration). For example, when
+        configuring a systemd service, the schema may require the presence of a
+        field \lstinline+port+ that holds a valid port number.
+\end{enumerate}
+
+A traditional answer is to use types, and in particular static types for (1).
+However, Nickel programs are of a special kind: they are usually simple,
+terminating programs that run on fixed inputs. Errors of type (1) will show up
+at evaluation anyway, or will not matter (dead code). For validation (2),
+traditional types are not adapated: statically checking common requirements
+like being a valid port number or a valid URL is, while not technically
+impossible, really hard to do.
+
+In a dynamic language, it is natural and easier to validate the data at runtime,
+\textit{a posteriori}. This is the principle of dynamic typing. But dynamically
+typed languages usually only check data when used in a primitive operation, like
+addition. In Nickel, the interpreter could not have a clue that a field
+\nickel{port}
+
+In the configuration setting, there is a
+singular dilemma with respect to typing: since a program is run on fixed inputs
+and is expected to terminate, any relevant type error will show up at
+evaluation.  Why bother with the complexity of a static type system? On the
+other hand, as the configurated systems become more and more complex, so do
+configuration codebases. Medium-sized to big codebases necessarily , static
+types become attractive again.  For reusable code — that is, library functions
+—, static types are specifically adapted, and bring all the usual benefits of
+early error detection, code robustness, better code structure, documentation,
+and so on.
 
@@
 which mixes both static and dynamic typing. Unlike gradually-typed languages
-like TypeScript\unsure{Todo: explain that typescript uses unions as
-  static typing whereas we are speaking of dynamic contracts}, we do not seek to build a complex type system that tries hard
-to accept most of the idioms naturally arising in dynamically typed code.
-Rather, we choose to provide a reasonably expressive type system with good
-inference properties, which makes typing functions operating on generic data easy,
-but may require using untyped code for more exotic expressions.  In any case,
-this is not a surrender, as Nickel provides a complementary mechanism for more
-advanced data validation: contracts.
+like TypeScript\unsure{Todo: explain that typescript uses unions as static
+typing whereas we are speaking of dynamic contracts}, we do not seek to build a
+complex type system that tries hard to accept most of the idioms naturally
+arising in dynamically typed code.  Rather, we choose to provide a reasonably
+expressive type system with good inference properties, which makes typing
+functions operating on generic data easy, but may require using untyped code for
+more exotic expressions.  In any case, this is not a surrender, as Nickel
+provides a complementary mechanism for more advanced data validation: contracts.
 
@@
 
+\newpage
+
+\section{Features of languages}
+\info{Zoology of various features that we will eventually show
+  conflict with this or that property or implementation of union and
+  intersection. Including user-define contracts.}
+
+Programming language research no longer lives on the abstract world
+of ideal surfaces.
+Every addition needs to be compatible with years of research
+and development, and this is hard.
+Success was unavoidable.
+
+In this section we briefly introduce some aspects of real world
+programming languages that will cause trouble while interacting
+with union and intersection dynamic checks.
+
+\subsection*{Flat contracts}
+\label{sec:flat-contracts}
+
+\unsure{Should we define contracts prior to this?}
+User defined contracts, also called flat contracts, are a particular
+kind of contract that behaves similar to an assertion, and whose
+main point is to apply some arbitrary predicate (as arbitrary as the
+host language allows it) to a value; if the predicate evaluates to
+true, the contract passes, otherwise it fails.
+
+For instance, we could define a flat contract \nickel{even} as
+\nickel{fun x => x \% 2 == 0}, that checks
+whether a number is even by checking the remainder of dividing it by 2.
+And then use it to check that a function always returns an
+even number, when applied to an \nickel{Integer}, as shown in
+Figure \ref{fig:num-to-even}.
+Or even, we could come up with a contract that checks that a function
+maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
+and use it as a contract for the \nickel{not} function, as can be seen on Figure
+\ref{fig:checked-not}.
+
+\begin{figure}[h]
+\begin{center}
+\begin{lstlisting}[language=nickel]
+let f = (fun n => n * 2) @ (Integer -> even)
+\end{lstlisting}
+  \end{center}
+\caption{A function wrapped with a contract stating that it always returns
+an even number}
+\label{fig:num-to-even}
+\end{figure}
+
+% $$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
+
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel]
+let notC = fun f => (f true == false) and (f false == true) in
+let checkedNot = not @ notC
+\end{lstlisting}
+\end{center}
+\caption{A contract that checks the not function behaves as expected}
+\label{fig:checked-not}
+\end{figure}
+
+This type of contracts are present on many different languages,
+for instance, the Eiffel programming language, the precursor
+of the Design by Contract ideals, allows to assert
+these kinds of expression as pre- and post-conditions on
+functions\cite{EiffelDesignByContract}.
+A somewhat newer language, the D programming
+language, allows, among other things, to have invariants on classes,
+written with flat contracts, that are
+dinamically checked at the frontiers of every public method \cite{DLangContracts}.
+
+Finally, the Racket programming language also has a system to work with
+contracts, powerful enough to define flat contracts
+\footnote{And a gradual typing system
+built on top of that, but that goes beyond the scope of this
+work.}
+, and
+to compose them with other kinds of dynamic checks,
+like higher order contracts or a lightweight take on union
+and intersection contracts\cite{RacketContracts}.
+\todo{Talk about this, probably later on the paper.
+See Related Works section}
+
+Even if flat contracts may seem like too permissive, the ability to make
+non trivial checks on values (that is, to validate their output) is expected
+by programmers on dynamic languages, where testing and runtime
+assertions are extremely useful tools for safe programming, and
+programmers are accustomed to express complex predicates
+over values.
+
+\subsection*{Code manipulation (optimizations)}
+\label{sec:optimizations}
+\info{Present inlining and CSE as two major code
+optimizations.
+Purity and immutability as two important factors.}
+
+The performance of modern programs heavily relies on the optimizations performed
+by the compiler or the interpreter. Even more so for functional languages, whose
+execution model is often far removed from actual processors, causing naive
+execution to exhibit unacceptable slowdowns.
+
+One crucial optimization is
+inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
+to make heavy use of functions, and a function call is a costly operation from
+the point of view of low-level execution. Inlining eliminates a function
+application by directly substituting the function for its definition at compile
+time (or before execution, for interpreted language). This is especially
+efficient for small functions that are called repeatedly.
+
+\info{I think it's better to use [h] on figures, we can discuss it}
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
+let elem = fun elt =>
+  lists.any (fun x => x == elt) in
+
+let subList = fun l1 l2 =>
+  elem (lists.head l1) l2
+  && subList (list.tail l1) l2
+\end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let subList = fun l1 l2 =>
+  lists.any (fun x => x == (lists.head l1)) l2
+  && subList (list.tail l1) l2
+\end{lstlisting}
+  \end{center}
+\caption{Inlining}
+\label{fig:optimizations-inlining-ex}
+\end{figure}
+
+While inlining expands an expression by substituting a definition for its value,
+an opposite transformation is sometimes beneficial, for example when a composite
+expression is repeated several times. This causes the same expression to be
+wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
+consists in introducing a new identifier for this expression and using the
+identifier in place of the original occurrences
+(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.
+
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
+let elemAtOrLast = fun index list =>
+  if index > lists.length list - 1 then
+    lists.elemAt (lists.length list - 1) list
+  else
+    lists.elemAt index list
+\end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let elemAtOrLast = fun index list =>
+  let l = lists.length list - 1 in
+  if index > l then
+    lists.elemAt l list
+  else
+    lists.elemAt index list
+\end{lstlisting}
+  \end{center}
+\caption{Common subexpression elimination}
+\label{fig:optimizations-cse-ex}
+\end{figure}
+
+Beyond CSE, optimizations such as loop-invariant code motion or
+let-floating~\ref{} apply the same principle of extracting an invariant
+expression to avoid recomputation respectively across loop iterations and
+function calls. For example, take the code of
+Figure~\ref{fig:optimizations-let-floating-ex}.
+
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
+let f = fun x => g y (x + 1)
+\end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Optimized program}]
+let g' = g y
+let f' = fun x => g' (x + 1)
+\end{lstlisting}
+  \end{center}
+\caption{Let-floating}
+\label{fig:optimizations-let-floating-ex}
+\end{figure}
+
+The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
+called. This may be costly, in particular in the presence of contracts: if
+\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
+on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
+to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
+Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
+let-floating transformation could indeed do (given \lstinline+g+ is pure, as
+explained below).
+
+The soundness of all these optimizations is tied to the invariance of the
+semantics of programs with respect to specific substitutions. Inlining requires
+that one can replace the application of a function by its body, which is
+basically $\beta$-reduction: as long as the arguments are evaluated following
+the language's strategy, it is a valid transformation. On the other hand,
+CSE-like transformations require an expansion property of the form
+
+$$M[N] \simeq let~x~=~N~in~M[x]$$
+
+This equation clearly fails in a call-by-value language with side effects:
+
+\begin{lstlisting}[language=Nickel]
+let f x = print "hi";(x+1) in (f 1,f 1)
+(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
+(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
+\end{lstlisting}
+
+In the example above, the first term prints \nickel{"hi"} two times while the
+second term prints it only once. However, this property does hold for
+\emph{pure} terms, that are terms without side-effects. In a pure language, or a
+language with effect tracking like Haskell\todo{Cite a different language, or
+remove citation}, pure terms can be safely identified for applying
+optimizations.
+
+Strikingly, we will see in section~\ref{} that the introduction of unions and
+intersections can make this equivalence \todo{name it} unsound \emph{even in a
+pure setting}, making the optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
+
+\unsure{Should we tease already here that unions and intersections break this
+even for pure terms?}
+
+\subsubsection*{Lazyness}
+\unsure{Should we talk about lazyness?
+this would be a good place to introduce it}
+
+\newpage
+
+\section{Union \& intersection}
+\info{What are they, what are they used for.}
+
+\subsection*{Unions}
+Unions are fundamental building blocks of programs. A union holds a value whose
+shape is known to be in a fixed set of alternatives: for example, a value that
+can be either a boolean or a string. They correspond logically to disjunctions.
+Used in combination with products, they can represent arbitrary tree-like
+structures, allowing to model a large class of data
+(\ref{fig:union-adt-ex}).
+
+\change{These next few labels need to be inside figures,
+they are not working}
+\label{fig:union-adt-ex}
+\begin{lstlisting}[title={Tree-like data structure}]
+type BinaryTree =
+Integer | (BinaryTree, BinaryTree)
+\end{lstlisting}
+
+In addition to internal program representations, unions are also useful to model
+external data:
+\begin{itemize}
+    \item To represent nullable values(\ref{fig:union-nullable-ex}).
+    \item For plain enumerations (\ref{fig:union-enum-ex}).
+    \item To support different representations of the same data
+        (\ref{fig:union-alt-ex}).
+\end{itemize}
+All of which are common in configurations.
+
+\label{fig:union-nullable-ex}
+\begin{lstlisting}[title={Nullable values}]
+optionalValue : String | Null
+\end{lstlisting}
+
+\label{fig:union-enum-ex}
+\begin{lstlisting}[title={Plain enumeration}]
+protocol : `Http | `Ftp | `Sftp
+\end{lstlisting}
+
+\label{fig:union-alt-ex}
+\begin{lstlisting}[title={Alternative representations}]
+person : {name : String, age : Int}
+  | {name : String, dateOfBirth : String}
+\end{lstlisting}
+
+In statically typed languages, and more specifically in functional programming
+languages, sums (also called tagged unions, or discriminated unions) are usually
+preferred, as constructors for algebraic datatypes. \unsure{This is a very
+    misleading statement at best. Sums are "tagged union" sure, but that refers
+to an encoding a sums with unions, certainly not an encoding of unions with
+sums!} [references]. Sums are unions' cousin where the alternatives are
+syntactically separated by an associated discriminating label. Well-known
+programming languages with first-class tagged unions include Haskell, OCaml,
+Rust, Scala, PureScript, Elm, Swift and many more.
+
+On the other hand, in dynamically typed languages, it is customary to directly
+store different kinds of data in one variable without further ceremony. The
+runtime system needs to attach type tags to values anyway, which can usually be
+observed using functions like JavaScript's \verb+typeof+. Using such bare
+unions\resolved{If you can observe with typeof, then it's pretty much a tagged
+union, though the tag is implicit} has already been an idiom in dynamically
+typed language for a long time. Being able to type and handle these unions is an
+important aspect of the migration to a gradual type system. Indeed, a lot of
+\unsure{This is the first time discussing gradual type systems, maybe it should
+be introduced earlier, in particular how they relate to contracts 
+(I think is mentioned on the appendix)}
+gradually typed languages either existed originally as dynamically typed
+language or aim at backward compatibility with a dynamically typed base
+language. This is why despite the presence of a static type system, gradually
+typed languages tend to favor bare unions types which is the typed counterpart
+of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
+
+In the case of Nickel, there is no pre-existing untyped language that we have to
+support, so we are freer of our choice. However, Nickel is a configuration
+language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
+untagged unions, there is a strong incentive to support untagged unions at the
+type level, or at least, at the contract level.
+
+\subsection*{Intersections}
+
+Intersections are less prevalent. An intersection is a value that belongs to
+two types at the same time. A simple example is a number literal \nickel{1}
+which could be accepted as both an \nickel{Int} and a \nickel{Float} without
+having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
+this example, the intersection \nickel{Int & Float} can be used as a subtype of
+both \nickel{Int} and \nickel{Float}.
+
+\label{fig:intersection-subytping-ex}
+\begin{lstlisting}[title={Value subtyping}]
+1 : Int & Float
+\end{lstlisting}
+
+Intersections of higher-order types corresponding to overloading
+(Figure~\ref{fig:intersection-overloading-ex}).
+
+\label{fig:intersection-overloading-ex}
+\begin{lstlisting}[title={Overloading}]
+incr: (Int -> Int) & (Float -> Float) =
+  fun x => x + 1
+\end{lstlisting}
+
+They can also be used to combine constraints (such as traits or interfaces), as
+done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
+
+\label{fig:intersection-mixins-ex}
+\begin{lstlisting}[title={Structural mixins}]
+Comparator: {cmp: T -> T -> Bool}
+Shower: {show: T -> String}
+
+showSmaller : (dict: Comparator & Shower) -> T -> T -> String
+\end{lstlisting}
+
+Finally, intersection naturally show up as the dual of unions. Thus, in a system
+featuring unions, a union in a negative positions is a form of intersection
+(Figure~\ref{fig:intersection-dual}).
+
+\label{fig:intersection-dual}
+\begin{lstlisting}[title={Intersection as the dual of unions}]
+f : (Num | String) -> Bool
+(*@ $\simeq$ @*) f : (Num -> Bool) & (String -> Bool)
+\end{lstlisting}
+
+\todo{Either here or the next section, show examples of unintuitive behaviour
+of U and I. For example, I is not jos composition of contracts.
+And union has to be treated as any one of the options}
+% Things to settle: how to motivate them, how to compare them to ADTs. Purely
+% dynamic approach, or typed approach too?
+
+\newpage
+
+\section{Issues coming from the semantics}
+\info{What goes wrong or difficult when trying to implement union and
+  intersection literally}
+
+Type systems featuring intersection types are hard in a fundamental way,
+from their discovery, they've been studied as a way to characterize
+lambda-calculus terms normalizability; which means that typechecking a
+term on a system with intersection types is undecidable,
+similar to how the halting problem is undecidable\todo{reference}.
+
+Usually, when thinking about programming languages, there's an intuition
+that static checks (like typechecking) are more difficult and less powerful
+than dynamic checks (like testing).
+For instance, it is very easy to check if,
+in a given execution of a program, a variable holds an integer value; while it
+may become more challenging to check if a variable will always hold an integer value,
+no matter the particular execution.
+
+As we intend to show on this paper, unions and intersections are not only
+statically hard to check, but they are also dynamically hard;
+\emph{they are fundamentally hard}.
+In this section, we present intuitively the main challenges that arise
+when attempting to develop a runtime check system for unions and intersections.
+
+\subsubsection*{Union contracts are hard to break}
+
+When programming with dynamic checks it is assumed that lack of
+failure does not mean that the program works perfectly.
+For instance, imagine having a function wrapped in a contract
+that checks that it maps positive numbers to positive numbers, like
+function \nickel{f} on Figure \ref{fig:pos-to-pos}.
+
+Clearly, there is an error on this program, but if the first call
+to \nickel{f} that gets executed is \nickel{f 10}, the error will not
+be found at that point, and the first operand of the addition would be
+evaluated to \nickel{3}.
+If, later on, \nickel{f 5} gets executed, the contract would get violated,
+since \nickel{f} gets called with a positive value, but returns a negative
+one (\nickel{-2}).
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (fun x => x - 7) @ Positive -> Positive in
+(f 10) + (f 5) 
+\end{lstlisting}
+\caption{Wrong positive to positive function}
+\label{fig:pos-to-pos}
+\end{figure}
+
+Even if the program performed two different applications of \nickel{f}
+before finding the contract violation, the first one is not actually
+needed, and executing only \nickel{f 5} is enough to find the problem.
+
+Having a witness of size 1 for every contract failure is a nice to have property
+since it means, for example, that every time that a contract is violated,
+the stack trace is enough to understand how it was violated, and to potentially
+come up with a minimal example that breaks the contract.
+However, when unions get introduced, it is lost.
+Consider now the program presented at Figure \ref{fig:wrong-union-function},
+this program (and in particular, \nickel{f}'s contract) still presents a problem,
+since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
+nor does it map positive numbers to non-positive numbers, so it won't be a valid
+value for the union of those contracts.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
+(f 10) + (f 5) 
+\end{lstlisting}
+\caption{Still wrong positive to positive or positive to non-positive function}
+\label{fig:wrong-union-function}
+\end{figure}
+
+Still, the program presents enough function calls to find that issue
+and violate the contract,
+since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
+and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
+But, what makes this case completely different from the one presented before, is that
+both function calls are needed, and, what is more, there is not a way to
+use the wrapped value \nickel{f} only once that would find a violation to its contract;
+at least two are needed.
+
+This seemingly simple consequence generates major issues when considered in real
+life languages.
+For example, remember the different mechanisms for code manipulation presented
+in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
+both of these rely on the fact that substituting a pure function call by its body,
+or joining equal pure expressions and sharing their result, is a semantically valid
+transformation.
+However, consider the examples presented at Figure \ref{fig:optimized-programs},
+the first one is the inlined version of the second one, while the second
+one has something like a CSE applied to it.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Inlined]
+let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
+(f 1 true, f 1 false)
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
+let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
+let g x = f 1 x in
+(g true, g false)
+\end{lstlisting}
+\caption{Equivalent programs, with inlining or CSE applied}
+\label{fig:optimized-programs}
+\end{figure}
+\todo{Improve example}
+
+Unexpectedly, this two programs behave differently, the first one will return
+\nickel{(1, "False")}, without failing.
+While the second one will fail, since the single application of \nickel{f 1},
+applied to both \nickel{true} and \nickel{false} will be able to find the error.
+
+Summing it up, wrapping a value with a union of higer order contracts means that
+that value now has a state, that may get modified silently at every use site.
+Both reviewed papers suffer from this defect since it is the semantics
+of unions that interact badly with code manipulation operations,
+not a particular implementation.
+This addition of global state complicates both the work of the language
+implementor, as well as the language user, since it adds non trivial
+side effects to your code.
+\unsure{Maybe mention CSE is actualy commonly done by hand as well}
+
+\subsubsection*{[WIP] Flat contracts are too permissive}
+
+Stop for a second and think how you could implement an intersection
+contract. The first thought you may have is "just apply
+both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
+Of course, if it were that simple we wouldn't be writing this paper,
+this idea starts to show problems when you first combine it with
+higher order contracts, since it would mean not complying with
+at least one of these contracts negatively would mean breaking the
+whole thing, as shown on the second example on the same Figure.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Implementation]
+x @ A & B
+(*@ $\simeq$ @*)
+(x @ A) @ B
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Problem]
+let g = (fun x => x) @ (Number -> Number) & (String -> String) in
+g 1
+\end{lstlisting}
+\caption{Implementations of intersection contracts: first iteration}
+\label{fig:inter-contracts}
+\end{figure}
+
+Still, not all hope is lost, we could pass some extra shared state
+to each one of the sub contracts and let them only fail if the other
+side of the contract has failed as well, as shown in Figure
+\ref{fig:inter-contracts-2}, where the sared state is
+represented by a label \nickel{l}.
+This idea is still quite reasonable, for instance, when executing
+\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
+that gets checked is the argument of \nickel{String -> String} over the function parameter, so
+something like \nickel{1 @ String}.
+Of course, this fails, but before raising a violation, this contract could
+check if the other side of the intersection (\nickel{Number -> Number}) has
+also tried to raise negative blame (on the parameter \nickel{1}), since
+it hasn't, the contract would return normally, and would mark itself as failed.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Implementation]
+x @ A & B
+(*@ $\simeq$ @*)
+(x @ A[l]) @ B[l]
+\end{lstlisting}
+\caption{Implementations of intersection contracts: second iteration}
+\label{fig:inter-contracts-2}
+\end{figure}
+
+This idea, which is an oversimplification of the ideas
+presented by Williams, Morris, and Wadler, seems to be spot on.
+However, it presents a major problem, it's not compatible with flat contracts,
+presented in Section \ref{sec:flat-contracts}.
+Consider the example presented on Figure \ref{fig:inter-flat-contracts},
+if the implementation presented previously were to be used, we may end up with
+something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
+\nickel{l} represents the shared state.
+At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
+\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
+This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
+fails negatively.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let C = fun f => (f 0) == 0 in
+let g = (fun x => x) @ (String -> String) & C
+\end{lstlisting}
+\caption{Intersection and flat contracts}
+\label{fig:inter-flat-contracts}
+\end{figure}
+
+But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
+\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
+Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
+failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
+should not have been used inside the evaluation of \nickel{C}, which
+adds an extra layer of complexity, contracts should be guarded, and execution
+depends on the shape of the surrounding evaluation context.
+
+\newpage
+
+\section{Issues in concrete restrictions from the literature}
+\info{The various papers and the tradeoffs they make}
+
+The papers that previously researched union and intersection
+as dynamic checks have, in our eyes, different shortcomings
+that makes them suboptimal solutions.
+
+Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
+they noticed that, since union contracts should share information between
+different application contexts, while intersection contracts should behave
+independently between each application context, the system becomes much
+simpler if every contract gets transformed into a union of intersections
+of other contracts, by using the usual De Morgan laws,
+and then eagerly opening up unions, but postponing intersection
+contracts as much as possible.
+\todo{This assumes the difficulty has been previously explained}
+
+For instance, if the value \nickel{V} is wrapped with contract
+\todo{(Yann) Contracted means
+something else I think.
+(Teo) changed}
+\nickel{A & (B | C)}, then, before the contract is checked,
+it will get opened up to something equivalent to
+\nickel{(A & B) | (A & C)}.
+
+This straightforward solutions makes composing the Keil and Thiemann
+system with other existing systems more difficult; in the words
+of Williams, Morris, and Wadler:
+"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
+and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
+special cases can hinder composition."\cite{RootCauseOfBlame}
+\info{(Yann) Does
+it also make it less efficient? If $(A \& B)$ fails because of $B$, their
+development causes to recheck the contract $A$ that has been duplicated?}
+And it also could harm performance, or at least complicate an implementation,
+since some sharing mechanism should be used to avoid checking the same contract,
+over the same value, multiple times.
+
+Another difficult issue Keil and Thiemann were able to solve, is the
+fact that flat contracts and intersection do not usually play nice
+together.
+They provide two different sets of rules for their calculus,
+the first one is non deterministic, whose purpose is to provide
+semantics that are simple to understand.
+The second one is a deterministic set of rules, that are intended
+to be simple to implement.
+This second set needs to perform operations
+on the evaluation context, to decide whether or not to evaluate
+a given contract.
+\unsure{Worth showing how the non deterministic version works? (See section B.1.1)}
+
+We see two problems here, a minor one is the fact that the two
+different semantics defined for the calculus are really far apart
+one from the other, both on the strategies used for contract resolution,
+as well as on the complexity levels.
+\unsure{(Yann) Is it really a
+    problem per se? It is common thing to do (a nice declarative system
+    for proofs and an algorithmic one for implementation, proved equivalent).
+    Maybe the point is that the algorithmic system is way more complex than the
+declarative one
+(Teo) reworded}.
+A second, major one,
+is the dependency added over the evaluation context, since, again,
+it makes composing this, with other features desired on programming languages,
+difficult, or even impossible
+\unsure{(Yann) I think we need to substantiate this claim. If this
+context dependency explains why CSE is invalid, maybe have a little example?
+(Teo) I removed it, I don't think there's an example, mainly since
+this context dependency is not part of the language. I do think it complicates
+implementation of said feature}
+\todo{(Yann) TODO: see if there's a low hanging complexity
+    argument we can make (like it's quadratic in the number of open contracts or
+whatever)}.
+
+Williams, Morris, and Wadler, take a different approach on flat contracts,
+they are completely forbidden in their system, explicitly to not encounter
+the previously mentioned issue.
+This restriction goes against what is usually expected from dynamic checks:
+validation of data, and not just inclusion on types, is an expected
+feature of contracts.
+
+On top of that, a fundamental problem arises on said paper,
+symptom of the difficulty of establishing good enough, and yet
+simple semantics for union and intersection dynamic checks.
+One of their main results, is the fact that their calculus obeys
+all of their expected monitoring properties for higher ordered, union
+and, intersection contracts.
+However, one of these properties states the following:
+
+$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
+
+Which means that a continuation ($K$) complies negatively with an
+intersection of $A$ and $B$ if it complies negatively with at least one of these.
+
+Altough this is true, it's a weaker statement of what we'd like to have,
+since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
+where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
+but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
+and dually for $not~(f ~ true)$.
+Which would mean, according to their semantics, that
+the continuation presented above would not
+be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
+even though our intuition (and the
+introduction to the paper) indicates that it should.
+\todo{Simplify example}
+
+% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
+% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
+% and dually for $K_2$, which would mean the continuation presented above would not
+% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
+% introduction to the paper) indicates that it should.
+
+While this mistake does not invalidate their work (the system they present
+handles these cases correctly), it shows that existing research is still not on
+good shape to handle the growing usecases being developed for
+union and intersection around dynamic languages.
+
+% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$
+
+% Keil and Thiemann did noticed this problem, and their contract satisfaction
+% rules only apply to elimination context, a particular kind of context
+% where the hole is either being applied to an argument
+% ($\square N$) or is used as an argument for a primitive
+% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).
+
+% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
+\unsure{Is it worth mentioning how continuations are formed?
+Should we go so techincal?}
+\todo{(Yann) Maybe adding a concrete example of this (it can be the same term but with
+    concrete types like Number and String and simple contexts like application
+to a dumb argument) is sufficient to make the point}
+
+\newpage
+
+\section{Related work}
+\info{Including a mention of statically typed systems with union}
+\info{Include something about Racket's and and or contracts}
+
+\newpage
+
+\section{Conclusion}
+
+\bibliography{nickel}
+
+\appendix
+
+\todo{Remove parts that are already on the main part}
+
+\section{Background on union types}
+
+\subsection{Nickel design space}
+
+{\color{red}Goals of the section
+
+Introduce Nickel, explain the choice of gradual
+typing and main design orientations, such as practice-oriented, lightweight,
+etc.\vspace{0.5cm}}
+
+Build systems, package managers, operating systems, cloud infrastructure,
+continuous integration and web services\resolved{Also continuous integration}
+are examples of modern complex software systems that require an extensive and
+non-trivial configuration in order to make them adapted to each different
+use-case. To manage the growing complexity that is then offloaded to
+configurations, the declarative approach has become more and more popular,
+illustrated for example by the infrastructure-as-code paradigm. This leads
+configuration to play an important role in critical aspects of software
+engineering, including security, availability, and maintainability.
+
+However, static text-based configuration alone is falling short of
+expressiveness, and is seldom sufficient. Data cannot be transformed, combined
+nor shared, resulting at best in boilerplate and duplication of information, or
+at worst, in data being invalid or inconsistent. Data validation is not
+supported either, and must be handed over to yet another tool down the
+configuration pipeline, if ever done. Correctly configuring a modern system is
+hard and failures may have substantial negative consequences.
+
+This is why the authors have been developing the Nickel
+language\cite{NickelRepo}. Nickel is a configuration programming language,
+meaning that a Nickel program evaluates to a value that must be
+straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
+lightweight and easy to write while still providing features for writing correct
+code and generating valid data. The focus is first and foremost practical:
+Nickel originated as a rethink of the language of the Nix package manager, used
+to describe one of the biggest sets of software packages\cite{repology}.  Nickel
+targets build systems and cloud deployment tools (Kubernetes, Terraform or
+NixOps) as well. While we are keen on incorporating existing or doing new
+research when it solves a well-defined problem, the general design goals are
+always guided by the practical industrial use cases.
+
+\subsection{Typing}
+
+Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
+Unlike JSON, though, Nickel anticipates large configurations by being both
+programmable and typed. In the configuration setting, there is a singular
+dilemma with respect to typing: since a program is run on fixed inputs and is
+expected to terminate, any relevant type error will show up at evaluation.  Why
+bother with the complexity of a static type system? On the other hand, more and
+more software systems offload complexity to configurations, in particular with
+the infrastructure as code paradigm prevailing today in cloud deployment. When
+the complexity of a codebase grows, static types become attractive again.  For
+reusable code — that is, library functions —, static types are specifically
+adapted, and bring all the usual benefits of early error detection, code
+robustness, better code structure, documentation, and so on.
+
+This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
+which mixes both static and dynamic typing. Unlike gradually-typed languages
+like TypeScript\unsure{Todo: explain that typescript uses unions as
+  static typing whereas we are speaking of dynamic contracts}, we do not seek to build a complex type system that tries hard
+to accept most of the idioms naturally arising in dynamically typed code.
+Rather, we choose to provide a reasonably expressive type system with good
+inference properties, which makes typing functions operating on generic data easy,
+but may require using untyped code for more exotic expressions.  In any case,
+this is not a surrender, as Nickel provides a complementary mechanism for more
+advanced data validation: contracts.
+
+\subsection{Contracts}
+{\color{red}Goals of the section
+
+A primer on contracts, which in the end are the problematic bit when confronted
+with unions.\vspace{0.5cm}}
+
+Enforcing pre- and post-conditions at runtime is a widely established practice.
+In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
+Felleisen introduce a principled approach to run-time assertion checking that
+nicely supports higher-order functions and introduces the notion of blame, which
+is crucial to good error reporting. It became apparent later that their
+contracts are closely related to the type casts introduced by gradual typing,
+modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
+value of contracts as a safe interface between typed and untyped code. In
+\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
+gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
+historical bit should probably be moved to related works eventually} Nickel
+adopts a similar type system, with both statically typed terms, dynamically
+typed terms, and first-class contracts. In "first-class contracts", we include
+the ability to write user-defined contracts. Because built-in types are too
+simplistic for the purpose of data validation, the ability to write a small
+validation function for, say, a port number, to have it immediately available as
+a contract that can be composed with other type constructors\resolved{Needs code
+to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
+system. An example is given in Figure~\ref{fig:contract-composition}. The first
+line defines a port contract from a predicate and use it in the definition of a
+larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
+contract checks that a value is a record with the following shape: it must have
+a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
+\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
+a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
+entails that a value respect a contract, akin to \lstinline+:+ for types. We
+assume that a \lstinline+Host+ contract has been defined in a similar way.
+
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel]
+let Port = contracts.fromPred (fun value =>
+  builtins.isNum value && value % 1 == 0
+  && value >= 0 && value <= 65353) in
+let Configuration = {
+  host | #Host,
+  port | #Port,
+} in
+{host = "nickel-lang.org", port = 80}
+| #Configuration
+\end{lstlisting}
+  \end{center}
+\caption{Contract composition}
+\label{fig:contract-composition}
+\medskip
+\small
+Defines a port contract from a predicate and use it in the definition of a
+\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
+role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
+contract to a value, similar to `:` for static types. We assume that a
+\lstinline+Host+ contract has also been defined.
+\end{figure}
+
+In this hybrid system, each type constructor of the static type system - arrows,
+records, foralls - must have a contract counterpart that checks at run-time that
+a value is a member of the corresponding type. Contracts for various
+extensions of the simply typed $\lambda$-calculus have been investigated
+\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
+DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
+Amongst all the common extensions to the polymorphic lambda calculus, one of the
+most useful and widespread turns out to be one of the most challenging:
+\emph{union types}.
+
 \subsection{Union types}

### 2021-05-26 paper.tex
More style conformance (authors, anonymous version)
Reading sections 1 & 2

@@
-\documentclass[sigplan,10pt,review]{acmart}
+\documentclass[sigplan,10pt,review,anonymous]{acmart}
 %\usepackage{amssymb}
@@
 \title{Contracts for unions and intersections are hard for Nickel}
-\author{Teodoro Freund, Yann Hamdaoui, Arnaud Spiwack}
+\author{Teodoro Freund}
+\affiliation{
+  \institution{???}
+  \city{Buenos Aires}
+  \country{Argentina}
+}
+\email{???}
+\author{Yann Hamdaoui}
+\affiliation{
+  \institution{Tweag}
+  \city{Paris}
+  \country{France}
+}
+\email{yann.hamdaoui@tweag.io}
+\author{Arnaud Spiwack}
+\affiliation{
+  \institution{Tweag}
+  \city{Paris}
+  \country{France}
+}
+\email{arnaud.spiwack@tweag.io}
 
@@
 
+\unsure{TODO: CCS classification; Keywords}
 \section{Introduction}
-Intro\unsure{todo}
 \info{Goals of the paper: position paper: unions/intersections nice in
@@
 contracts are closely related to the type casts introduced by gradual typing,
-modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
+modulo blame\unsure{What does ``modulo blame'' means here?}: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
 value of contracts as a safe interface between typed and untyped code. In
@@
 entails that a value respect a contract, akin to \lstinline+:+ for types. We
-assume that a \lstinline+Host+ contract has been defined in a similar way.
+assume that a \lstinline+Host+ contract has been defined in a similar
+way.\unsure{What's missing in this section is an example of how
+  contracts differ from standard dynamic type checking on some
+  function. Namely, the quality of error messages.}
 
@@
 This type of contracts are present on many different languages,
-for instance, the Eiffel programming language, the precursor
+for instance, the Eiffel programming language\unsure{Citation?}, the precursor
 of the Design by Contract ideals, allows to assert
@@
 A somewhat newer language, the D programming
-language, allows, among other things, to have invariants on classes,
+language, allows, among other things, to have invariants on
+classes\unsure{I (Arnaud) believe Eiffel has this feature as well,
+  doesn't it?},
 written with flat contracts, that are
@@
 to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
-system. An example is given in Figure~\ref{fig:contract-composition}. The first
+system. An example is given in Figure~\ref{fig:contract-composition-appendix}. The first
 line defines a port contract from a predicate and use it in the definition of a
@@
 \caption{Contract composition}
-\label{fig:contract-composition}
+\label{fig:contract-composition-appendix}
 \medskip

### 2021-05-26 paper.tex
Reading sections 3–5

@@
 languages, sums (also called tagged unions, or discriminated unions) are usually
-preferred, as constructors for algebraic datatypes. \unsure{This is a very
-    misleading statement at best. Sums are "tagged union" sure, but that refers
-to an encoding a sums with unions, certainly not an encoding of unions with
-sums!} [references]. Sums are unions' cousin where the alternatives are
+preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
 syntactically separated by an associated discriminating label. Well-known
@@
 However, when unions get introduced, it is lost.
-Consider now the program presented at Figure \ref{fig:wrong-union-function},
+Consider now the program presented in Figure \ref{fig:wrong-union-function},
 this program (and in particular, \nickel{f}'s contract) still presents a problem,
@@
 
+\subsection{Keil and Thiemann}
+\label{sec:keil-thiemann}
+\unsure{Subsection title?}
 Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
@@
 For instance, if the value \nickel{V} is wrapped with contract
-\todo{(Yann) Contracted means
-something else I think.
-(Teo) changed}
 \nickel{A & (B | C)}, then, before the contract is checked,
@@
 system with other existing systems more difficult; in the words
-of Williams, Morris, and Wadler:
-"(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
+of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
+``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
 and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition."\cite{RootCauseOfBlame}
+special cases can hinder composition.''\cite{RootCauseOfBlame}
 \info{(Yann) Does
@@
 since some sharing mechanism should be used to avoid checking the same contract,
-over the same value, multiple times.
+over the same value, multiple times.\unsure{Arnaud says: this whole
+  section is too abstract give concrete code and explain what goes
+  wrong. It's fine to give code in Nickel syntax}
 
@@
 
+\subsection{Williams, Morris, and Wadler}
+\label{sec:will-morr-wadl}
+\unsure{Subsection title?}
+
 Williams, Morris, and Wadler, take a different approach on flat contracts,
@@
     to a dumb argument) is sufficient to make the point}
+
+\subsection{Racket}
+\label{sec:racket}
+
+\unsure{Todo}
 

### 2021-05-26 paper.tex
Rework the intro and the contract part

@@
 necessary transformations. This approach makes the configuration persistent,
-reproducible, analyzable and simulable. Text-based configurations are led to
-play an important role in critical aspects of software engineering, including
-security, availability, and maintainability.
-
-However, static text-based configurations show their limits beyond a given size
-and complexity. Data cannot be transformed, combined, shared, nor validated.
-This lack of expressivity results at best in boilerplate and duplication of
-information, but it can also make data invalid or inconsistent. Correctly
-configuring a modern system is hard and failures may have substantial negative
-consequences.
+reproducible, versionable and subject to static analysis. Text-based
+configurations are thus led to play an important role in critical aspects of
+software engineering, including security, availability, and maintainability.
+
+However, static text-based configurations show their limits beyond a certain
+size and complexity. Data cannot be transformed, combined, shared, nor validated
+(without resorting to separate schemas and tools).  This lack of expressivity
+results at best in boilerplate and duplication of information, but it can also
+make data invalid or inconsistent.  Correctly configuring a modern system is ard
+and failures may have substantial negative consequences.
 
@@
 Nix package manager, used to describe one of the biggest sets of packaged
-software in existence\cite{repology}.  Nickel targets build systems and cloud deployment
-tools (Kubernetes, Terraform or NixOps) as well.
+software in existence\cite{repology}.  Nickel targets build systems, cloud
+deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
+non-trivial configuration as well.
 
@@
 configurations by being programmable, and by providing means of ensuring
-correctness.
-
-In this context, correctness means two things:
-\begin{enumerate}
-    \item The program runs without errors. For example, an error is raised when
-        evaluating the nonsensical division \nickel{1 / "ftp"}.
-    \item The program generates valid data, i.e. data that conforms to a schema
-        (imposed by the system consuming the configuration). For example, when
-        configuring a systemd service, the schema may require the presence of a
-        field \lstinline+port+ that holds a valid port number.
-\end{enumerate}
-
-A traditional answer is to use types, and in particular static types for (1).
-However, Nickel programs are of a special kind: they are usually simple,
-terminating programs that run on fixed inputs. Errors of type (1) will show up
-at evaluation anyway, or will not matter (dead code). For validation (2),
-traditional types are not adapated: statically checking common requirements
-like being a valid port number or a valid URL is, while not technically
-impossible, really hard to do.
-
-In a dynamic language, it is natural and easier to validate the data at runtime,
-\textit{a posteriori}. This is the principle of dynamic typing. But dynamically
-typed languages usually only check data when used in a primitive operation, like
-addition. In Nickel, the interpreter could not have a clue that a field
-\nickel{port}
-
-In the configuration setting, there is a
-singular dilemma with respect to typing: since a program is run on fixed inputs
-and is expected to terminate, any relevant type error will show up at
-evaluation.  Why bother with the complexity of a static type system? On the
-other hand, as the configurated systems become more and more complex, so do
-configuration codebases. Medium-sized to big codebases necessarily , static
-types become attractive again.  For reusable code — that is, library functions
-—, static types are specifically adapted, and bring all the usual benefits of
-early error detection, code robustness, better code structure, documentation,
-and so on.
-
-This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
-which mixes both static and dynamic typing. Unlike gradually-typed languages
-like TypeScript\unsure{Todo: explain that typescript uses unions as static
-typing whereas we are speaking of dynamic contracts}, we do not seek to build a
-complex type system that tries hard to accept most of the idioms naturally
-arising in dynamically typed code.  Rather, we choose to provide a reasonably
-expressive type system with good inference properties, which makes typing
-functions operating on generic data easy, but may require using untyped code for
-more exotic expressions.  In any case, this is not a surrender, as Nickel
-provides a complementary mechanism for more advanced data validation: contracts.
-
-\subsection{Contracts}
-{\color{red}Goals of the section
-
-A primer on contracts, which in the end are the problematic bit when confronted
-with unions.\vspace{0.5cm}}
-
-Enforcing pre- and post-conditions at runtime is a widely established practice.
-In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
-Felleisen introduce a principled approach to run-time assertion checking that
-nicely supports higher-order functions and introduces the notion of blame, which
-is crucial to good error reporting. It became apparent later that their
-contracts are closely related to the type casts introduced by gradual typing,
-modulo blame\unsure{What does ``modulo blame'' means here?}: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
-value of contracts as a safe interface between typed and untyped code. In
-\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
-gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
-historical bit should probably be moved to related works eventually} Nickel
-adopts a similar type system, with both statically typed terms, dynamically
-typed terms, and first-class contracts. In "first-class contracts", we include
-the ability to write user-defined contracts. Because built-in types are too
-simplistic for the purpose of data validation, the ability to write a small
-validation function for, say, a port number, to have it immediately available as
-a contract that can be composed with other type constructors\resolved{Needs code
-to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
-system. An example is given in Figure~\ref{fig:contract-composition}. The first
-line defines a port contract from a predicate and use it in the definition of a
-larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
+both correctness and validity:
+
+\begin{description}
+    \item[(correctness)] The program runs without errors. Errors may be type
+        errors, raised for example when evaluating the nonsensical division
+        \nickel{1 / "ftp"}. More generally, they can any the violation of a pre-
+        or post-condition. When such an error cannot be prevented, we want at
+        least early failures and relevant error messages.
+    \item[(validity)] The program generates valid data, i.e. data that
+        conforms to a schema (imposed by the system consuming the
+        configuration). For example, when configuring a systemd service, the
+        schema may require the presence of a field \lstinline+port+ that holds a
+        valid port number.
+\end{description}
+
+A traditional answer is to use static types, in particular static types for
+correctness. However, Nickel programs are of a special kind: they are usually
+simple, terminating programs that run on fixed inputs. Correctness errors will
+show up at evaluation anyway, or will not matter (dead code). With respect to
+validation, traditional types seem too limited. Validation commonly involves
+checking requirements such as being a valid port number or a valid URL which
+are, while not technically impossible, really hard to check statically.
+
+In a dynamic language, it is natural and easier to validate data at runtime.
+This underlies the approach of dynamic typing. But dynamically typed languages
+usually only check the same simple predicates as static types (being a number,
+being a string, etc.), and solely for data that are arguments to a primitive
+operation. For validation, we want to check arbitrary predicates on arbitrary
+values. This is provided in a basic form by assertions, available in most
+languages out there. An assertion is an expression that verifies that a
+predicate holds, or abort the program otherwise.  While assertions are fine for
+simple sanity checks, they can quickly becomes verbose, clumsy to compose and to
+reuse, and provide very limited context attached to an error. It turns out there
+is a better tool than plain assertions: contracts.
+
+\subsection*{Contracts}
+
+Assertions are usually mainly used for correctness by enforcing pre- and
+post-conditions of functions at runtime. In their foundational
+paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
+principled approach to run-time assertion checking called contracts. As opposed
+to plain assertions, contracts can be composed naturally into more complex ones.
+Nickel features such a contract system. For example, contracts nicely supports
+higher-order functions. If the user has written an \nickel{Even} contract for
+number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
+illustrated in Figure~\ref{fig:contract-higher-order}. This contracts will check
+at each call that the argument is an even number (pre-condition) and that the
+return value is a integer (post-condition). The \nickel{#} character
+distinguishes contracts in Nickel. Contracts also introduce the notion of blame
+which is crucial to good error reporting in presence of higher-order contracts
+and other composed contracts.
+
+\begin{figure}
+  \begin{center}
+\begin{lstlisting}[language=nickel]
+let Even = contracts.fromPred (fun val =>
+  builtins.isNum val
+  && val % 2 == 0) in
+let Int = contracts.fromPred (fun val =>
+  builtins.isNum val
+  && val % 1 == 0) in
+{divBy2 | #Even -> #Int = fun x => x/2}
+\end{lstlisting}
+  \end{center}
+\caption{Higher-order contract}
+\label{fig:contract-higher-order}
+\end{figure}
+
+In our opinion, one of the main appeal of Nickel is this ability to write a
+simple validation function for, say, a port number, and have it immediately
+available as a contract that can be used to construct a other contracts by
+writing natural type-like expressions. Another example is given in
+Figure~\ref{fig:contract-composition}. The first line defines a port contract
+from a predicate and use it in the definition of a larger
+\lstinline+Configuration+ record contract. The \lstinline+Configuration+
 contract checks that a value is a record with the following shape: it must have
@@
 \lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
-a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
-entails that a value respect a contract, akin to \lstinline+:+ for types. We
-assume that a \lstinline+Host+ contract has been defined in a similar
-way.\unsure{What's missing in this section is an example of how
-  contracts differ from standard dynamic type checking on some
+a \emph{schema} used to validate the final data, which is exactly what we were
+looking for in the previous section. Not only it plays the role of a schema, but
+it also \emph{looks} like one, as opposed to a less readable plain validation
+function. The \lstinline+|+ operator entails that a value respect a contract,
+akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
+been defined in a similar way. \unsure{What's missing in this section is an
+example of how contracts differ from standard dynamic type checking on some
 function. Namely, the quality of error messages.}
@@
 \label{fig:contract-composition}
-\medskip
-\small
-Defines a port contract from a predicate and use it in the definition of a
-\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
-role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
-contract to a value, similar to `:` for static types. We assume that a
-\lstinline+Host+ contract has also been defined.
-\end{figure}
-
-In this hybrid system, each type constructor of the static type system - arrows,
-records, foralls - must have a contract counterpart that checks at run-time that
-a value is a member of the corresponding type. Contracts for various
-extensions of the simply typed $\lambda$-calculus have been investigated
-\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
-DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
-Amongst all the common extensions to the polymorphic lambda calculus, one of the
-most useful and widespread turns out to be one of the most challenging:
-\emph{union types}.
+\end{figure}
+
+Contracts have another perk: they correspond to the type casts of gradual
+typing, a discipline for mixing static and dynamic typing in the same language.
+As such, they are a precious ingredient for handling the interaction of
+statically typed and dynamically typed code. While we focused on the writing of
+configuration, Nickel is a proper programming language that is also designed to
+write generic and reusable library code. There, the typing trade-off is
+different, and static typing is more adapted. Nickel has a gradual type system,
+which is yet another argument for having contracts. Let us end this section by
+stressing the following points:
+\begin{itemize}
+    \item \textbf{Even in a purely dynamically typed language, contracts are an
+        appealing feature for validation and correctness}. The gradual
+        typing part deserved to be mentioned, but is is out of the scope of this
+        paper. 
+    \item \textbf{User-defined contracts are essential for validation}. Some contract
+        systems (as the one of~\ref{RootCauseOfBlame} that will be extensively
+        discussed later in the paper) do not feature user-defined contracts.
+        Rather, a fixed set of built-in contracts such as \nickel{Number} or
+        \nickel{String} are provided, very much like base types. The absence of
+        user-defined contracts is understandable when using contracts as a
+        dynamic typing discipline for correctness. But a system without
+        user-defined contracts is limited in the same way as static typing with
+        respect to validation.
+\end{itemize}
 

### 2021-05-27 paper.tex
read on intro

@@
 results at best in boilerplate and duplication of information, but it can also
-make data invalid or inconsistent.  Correctly configuring a modern system is ard
+make data invalid or inconsistent.  Correctly configuring a modern system is hard
 and failures may have substantial negative consequences.
@@
 to plain assertions, contracts can be composed naturally into more complex ones.
-Nickel features such a contract system. For example, contracts nicely supports
+Nickel features such a contract system. For example, contracts nicely support
 higher-order functions. If the user has written an \nickel{Even} contract for
 number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
-illustrated in Figure~\ref{fig:contract-higher-order}. This contracts will check
+illustrated in Figure~\ref{fig:contract-higher-order}. This contract will check
 at each call that the argument is an even number (pre-condition) and that the
@@
 
-In our opinion, one of the main appeal of Nickel is this ability to write a
+In our opinion, one of the main appeals of Nickel is this ability to write a
 simple validation function for, say, a port number, and have it immediately
-available as a contract that can be used to construct a other contracts by
+available as a contract that can be used to construct other contracts by
 writing natural type-like expressions. Another example is given in
@@
 it also \emph{looks} like one, as opposed to a less readable plain validation
-function. The \lstinline+|+ operator entails that a value respect a contract,
+function. The \lstinline+|+ operator entails that a value respects a contract,
 akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
@@
 \end{figure}
+\unsure{I wouldn't use nickel-lang.org unless you are
+planning on putting something up there. Maybe nixos.org?}
 
@@
     \item \textbf{User-defined contracts are essential for validation}. Some contract
-        systems (as the one of~\ref{RootCauseOfBlame} that will be extensively
+        systems (as the one of~\cite{RootCauseOfBlame} that will be extensively
         discussed later in the paper) do not feature user-defined contracts.

### 2021-05-27 paper.tex
Some more sec 4

@@
 Type systems featuring intersection types are hard in a fundamental way,
-from their discovery, they've been studied as a way to characterize
+from their discovery, they've been studied as a form to characterize
 lambda-calculus terms normalizability; which means that typechecking a
@@
 As we intend to show on this paper, unions and intersections are not only
-statically hard to check, but they are also dynamically hard;
+statically hard to check, but they are also dynamically hard:
 \emph{they are fundamentally hard}.
@@
 
-\subsubsection*{Union contracts are hard to break}
+\subsection*{Union contracts are hard to break}
 
@@
 failure does not mean that the program works perfectly.
+
 For instance, imagine having a function wrapped in a contract
-that checks that it maps positive numbers to positive numbers, like
-function \nickel{f} on Figure \ref{fig:pos-to-pos}.
-
+that checks that it maps positive numbers to positive numbers,
+and using it in different ways,
+like the program on Figure \ref{fig:pos-to-pos}.
 Clearly, there is an error on this program, but if the first call
@@
 
-Having a witness of size 1 for every contract failure is a nice to have property
+Having a witness of size 1 for every contract failure is a really nice to have property
 since it means, for example, that every time that a contract is violated,
@@
 come up with a minimal example that breaks the contract.
-However, when unions get introduced, it is lost.
+However, when unions get introduced, this property is lost.
+
 Consider now the program presented in Figure \ref{fig:wrong-union-function},
@@
 
-Still, the program presents enough function calls to find that issue
+As before, the program presents enough function calls to find that issue
 and violate the contract,
@@
 However, consider the examples presented at Figure \ref{fig:optimized-programs},
-the first one is the inlined version of the second one, while the second
-one has something like a CSE applied to it.
+the first one is an inlined version of the second one (where function
+\nickel{g} has been inlined); while the second is similar to the first, but
+the common subexpression \nickel{f 1} has been eliminated.
 
@@
 \end{figure}
-\todo{Improve example}
-
-Unexpectedly, this two programs behave differently, the first one will return
+
+Unexpectedly, these two programs behave very differently, the first one will return
 \nickel{(1, "False")}, without failing.
 While the second one will fail, since the single application of \nickel{f 1},
-applied to both \nickel{true} and \nickel{false} will be able to find the error.
+applied to both \nickel{true} and \nickel{false} will be able to find the error
+on the union contract (\nickel{(Boolean -> Number) | (Boolean -> String)}).
+
+A valid argument against this point would be stating that further work should be done
+on optimizers, so to not modify these tricky cases, even if this is true,
+programmers also modify their code on a daily basis, and if moving from the first
+code snippet at Figure \ref{fig:optimized-programs} to the second one changes
+the behaviour of a program, we argue that union contracts not only go against the
+implementation of languages, but also against their usability.
 
@@
 side effects to your code.
-\unsure{Maybe mention CSE is actualy commonly done by hand as well}
-
-\subsubsection*{[WIP] Flat contracts are too permissive}
+
+\subsection*{Flat contracts do not intersect}
 
@@
 contract. The first thought you may have is "just apply
-both contracts", as shown on the Implementation of Figure \ref{fig:inter-contracts}.
-Of course, if it were that simple we wouldn't be writing this paper,
-this idea starts to show problems when you first combine it with
+both contracts", as shown on the Implementation snippet
+from Figure \ref{fig:inter-contracts}.
+Of course, if it were that simple we wouldn't be writing this paper.
+This idea starts to show problems when you first combine it with
 higher order contracts, since it would mean not complying with
-at least one of these contracts negatively would mean breaking the
-whole thing, as shown on the second example on the same Figure.
+at least one of these contracts negatively would entail breaking the
+whole thing, as shown on the second example on the same Figure,
+that would fail.
 
@@
 
-Still, not all hope is lost, we could pass some extra shared state
+Not all hope is lost, we could pass some extra shared state
 to each one of the sub contracts and let them only fail if the other
 side of the contract has failed as well, as shown in Figure
-\ref{fig:inter-contracts-2}, where the sared state is
+\ref{fig:inter-contracts-2}, where the shared state is
 represented by a label \nickel{l}.
@@
 check if the other side of the intersection (\nickel{Number -> Number}) has
-also tried to raise negative blame (on the parameter \nickel{1}), since
+also tried to raise negative blame on the same parameter ()\nickel{1}), since
 it hasn't, the contract would return normally, and would mark itself as failed.
@@
 
+On the other hand, if you were to do something like
+\nickel{g true}, that would fail also negatively the contract
+\nickel{String -> String} and, when checking if \nickel{true}
+is a correct parameter for \nickel{Number -> Number} it would check
+if the other side of the intersection has failed, and since it
+has, a violation would be raised and the whole program would
+correctly fail.
+
 This idea, which is an oversimplification of the ideas
@@
 However, it presents a major problem, it's not compatible with flat contracts,
-presented in Section \ref{sec:flat-contracts}.
+as presented in Section \ref{sec:flat-contracts}.
 Consider the example presented on Figure \ref{fig:inter-flat-contracts},
-if the implementation presented previously were to be used, we may end up with
+if the implementation discussed previously were to be used, we may end up with
 something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
@@
 At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
-\nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
+if \nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
 This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
@@
 But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
-\nickel{g} is never applied to \nickel{0}, in fact, it is never applied.
+\nickel{g} is never applied to \nickel{0}, in fact, it is never applied
+to anything.
 Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
@@
 should not have been used inside the evaluation of \nickel{C}, which
-adds an extra layer of complexity, contracts should be guarded, and execution
+adds an extra layer of complexity: contracts should be guarded, and execution
 depends on the shape of the surrounding evaluation context.
+
+\subsection*{[WIP] Union and intersection are not a silver bullet}
+
+Before wrapping up this section, we'd like to show that maybe, union and intersection
+contracts are not exactly what we would want them to be.
+As stated throughout this work, programmers coming from dynamic languages
+are accustomed to perform safety checks (data validation) with
+great amounts of flexibility, and the whole power of their
+language of choice at their disposition.
+
+For instance, a Python programmer may have no problem in checking
+at runtime, that a given variable is an integer if a given flag is
+not set, as shown in Figure \ref{fig:python-assertion},
+while a Haskell programmer might close her eyes and leave the room
+instantly.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=python]
+result = 42 if flag else "Disabled"
+...
+assert ((not flag) or isinstance(result, int))
+\end{lstlisting}
+\caption{Asserting at runtime the type of a variable}
+\label{fig:python-assertion}
+\end{figure}
+
+These expectations are not really met by how unions and intersection
+contracts are supposed to work.
+For example, consider the Nickel code shown on Figure
+\ref{fig:intersection-distrbution}, at first sight, this code should
+complete execution succesfully, returning \nickel{(1, 1)},
+however, under both the Keil and Thiemann, as well as the
+Williams, Morris, and Wadler works, this program will fail, with a violation
+of the contract imposed to \nickel{g}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = fun x y => x in
+let g = f @ (Number -> Number -> Number) & (Number -> Boolean -> Number) in
+let h = g 1 in 
+(h 1, h true)
+\end{lstlisting}
+\caption{Intersection contracts don't distribute}
+\label{fig:intersection-distrbution}
+\end{figure}
+
+The reason this happens is that
+there is a fundamental characterization of how intersection
+contracts behave: the choice of which branch to choose is made exactly
+when the intersection is opened up, and not delayed as much as possible.
+This means, that the contract presented on Figure \ref{fig:intersection-distrbution},
+\nickel{(Number -> Number -> Number) & (Number -> Boolean -> Number)},
+will not behave the same way that
+\nickel{(Number -> (Number -> Number) & (Boolean -> Number)},
+when these two expressions should be logically equivalent.
 
@@
 a given contract.
-\unsure{Worth showing how the non deterministic version works? (See section B.1.1)}
+\unsure{Worth showing how the non deterministic version works? (See section B.1.1)
+try with an example}
 

### 2021-05-27 paper.tex
Pass on intro

@@
 deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
-non-trivial configuration as well.
-
-\subsection*{To type or not to type}
+non-trivial configuration.
+
+\subsection*{Correct programs \& valid data}
 
@@
         errors, raised for example when evaluating the nonsensical division
-        \nickel{1 / "ftp"}. More generally, they can any the violation of a pre-
-        or post-condition. When such an error cannot be prevented, we want at
-        least early failures and relevant error messages.
-    \item[(validity)] The program generates valid data, i.e. data that
-        conforms to a schema (imposed by the system consuming the
-        configuration). For example, when configuring a systemd service, the
-        schema may require the presence of a field \lstinline+port+ that holds a
-        valid port number.
+        \nickel{1 / "ftp"}. More generally, they can be any the violation of a
+        pre- or post-condition. When such an error cannot be prevented, we want
+        at least early failures and relevant error messages.\todo{make this
+        definition more precise}
+    \item[(validity)] The program generates valid data, i.e. data that conforms
+        to a schema (imposed by the system consuming the configuration). For
+        example, when configuring a web service, the schema may require the
+        presence of a field \lstinline+port+ that holds a valid port number.
 \end{description}
 
-A traditional answer is to use static types, in particular static types for
+A traditional answer is to use static types, and in particular static types for
 correctness. However, Nickel programs are of a special kind: they are usually
@@
 show up at evaluation anyway, or will not matter (dead code). With respect to
-validation, traditional types seem too limited. Validation commonly involves
+validation, traditional types are too limited. Validation commonly involves
 checking requirements such as being a valid port number or a valid URL which
@@
 being a string, etc.), and solely for data that are arguments to a primitive
-operation. For validation, we want to check arbitrary predicates on arbitrary
+operation. For validation, one needs to check arbitrary predicates on arbitrary
 values. This is provided in a basic form by assertions, available in most
@@
 predicate holds, or abort the program otherwise.  While assertions are fine for
-simple sanity checks, they can quickly becomes verbose, clumsy to compose and to
-reuse, and provide very limited context attached to an error. It turns out there
-is a better tool than plain assertions: contracts.
+simple sanity checks, their use can incur verbosity, they are clumsy to compose
+and to reuse, and they provide limited context attached to an error. It turns
+out there is a better tool than plain assertions: contracts.
 
@@
 higher-order functions. If the user has written an \nickel{Even} contract for
-number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
-illustrated in Figure~\ref{fig:contract-higher-order}. This contract will check
-at each call that the argument is an even number (pre-condition) and that the
-return value is a integer (post-condition). The \nickel{#} character
-distinguishes contracts in Nickel. Contracts also introduce the notion of blame
-which is crucial to good error reporting in presence of higher-order contracts
-and other composed contracts.
+numbers, they can use it right away to form a function contract \nickel{#Even ->
+#Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
+contract will check at each call that the argument is an even number
+(pre-condition) and that the return value is a integer (post-condition). The
+\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
+the notion of blame which is crucial to good error reporting in presence of
+higher-order contracts and other composed contracts.
 
@@
 simple validation function for, say, a port number, and have it immediately
-available as a contract that can be used to construct other contracts by
-writing natural type-like expressions. Another example is given in
-Figure~\ref{fig:contract-composition}. The first line defines a port contract
-from a predicate and use it in the definition of a larger
-\lstinline+Configuration+ record contract. The \lstinline+Configuration+
+available to construct other contracts by writing simple type-like expressions.
+Another example is given in Figure~\ref{fig:contract-composition}. The first
+line defines a port contract from a predicate and use it in the definition of a
+larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
 contract checks that a value is a record with the following shape: it must have
@@
 looking for in the previous section. Not only it plays the role of a schema, but
-it also \emph{looks} like one, as opposed to a less readable plain validation
+it also \emph{looks} like one, as opposed to the corresponding plain validation
 function. The \lstinline+|+ operator entails that a value respects a contract,
@@
 } in
-{host = "nickel-lang.org", port = 80}
+{host = "nixos.org", port = 80}
 | #Configuration
@@
 \end{figure}
-\unsure{I wouldn't use nickel-lang.org unless you are
+\resolved{I wouldn't use nickel-lang.org unless you are
 planning on putting something up there. Maybe nixos.org?}
@@
         paper. 
-    \item \textbf{User-defined contracts are essential for validation}. Some contract
-        systems (as the one of~\cite{RootCauseOfBlame} that will be extensively
-        discussed later in the paper) do not feature user-defined contracts.
-        Rather, a fixed set of built-in contracts such as \nickel{Number} or
-        \nickel{String} are provided, very much like base types. The absence of
-        user-defined contracts is understandable when using contracts as a
-        dynamic typing discipline for correctness. But a system without
-        user-defined contracts is limited in the same way as static typing with
-        respect to validation.
+    \item \textbf{User-defined contracts are essential for validation}. Some
+        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
+        extensively discussed later in the paper) do not feature user-defined
+        contracts.  Rather, a fixed set of built-in contracts such as
+        \nickel{Number} or \nickel{String} are provided, very much like base
+        types. The absence of user-defined contracts is understandable when
+        using contracts as a dynamic typing discipline for correctness. But a
+        system without user-defined contracts is limited in the same way as
+        static typing with respect to validation.
 \end{itemize}

### 2021-05-27 paper.tex
Small fixes and trailing spaces

@@
 Beyond CSE, optimizations such as loop-invariant code motion or
-let-floating~\ref{} apply the same principle of extracting an invariant
-expression to avoid recomputation respectively across loop iterations and
-function calls. For example, take the code of
-Figure~\ref{fig:optimizations-let-floating-ex}.
+let-floating~\ref{} apply the same principle of extracting out an invariant
+expression to avoid its recomputation respectively across loop iterations and
+function calls.
 
@@
 
+For example, take the code of Figure~\ref{fig:optimizations-let-floating-ex}.
 The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is

### 2021-05-27 paper.tex
[WIP]Add a paragraph on lazyness

@@
 
-\subsubsection*{Lazyness}
-\unsure{Should we talk about lazyness?
-this would be a good place to introduce it}
+\subsection*{Lazyness}
+Evaluation strategies relates to the evaluation of the expression bound to
+identifiers. Expressions are be bound by assignments (e.g. the \nickel{blah} construct in Nickel), or arguments are bound to
+parameters in a function application. We will use Figure~\ref{} as a working
+example.
+
+The eager strategy consists in evaluating right away a bound value before
+storing it in an identifier. If an identifier is not used, this leads to
+unnecessary work: in Figure~\ref{}, the expression \nickel{fibonacci 100} would
+be computed although it is never used. The lazy strategy, on the other hand,
+only evaluates expression if they are needed. In the lazy model, bound
+expressions are stored as in their current unevaluated form, also called a
+\emph{thunk}. Thunks are only evaluated at their first usage, and their value is
+memoized for future usages.
+
+The eager strategy has been chosen in the vast majority of languages: while
+theoretically less efficient, it is simpler to implement and more predictable
+for the programmer. Also, lazy evaluation can occur performance overhead and
+memory leaks in unexpected ways. While eager evaluation is the default, some
+languages however offer a form of lazy evaluation, either as an opt-in option
+(e.g. Ocaml and ReScript), or by default for specific primitives types such as
+like list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
+(and its web avatar PureScript). Although rare in general purpose languages,
+lazy evaluation is more widespread in the configuration space: the configuration
+languages Dhall, CUE and Jsonnet also implements full lazy evaluation. So are
+the Nix language, which inspired Nickel, and Nickel itself.
 

### 2021-05-28 paper.tex
Pass on sec 4

@@
 
-\subsection*{[WIP] Union and intersection are not a silver bullet}
-
-Before wrapping up this section, we'd like to show that maybe, union and intersection
+\subsection*{Union and intersection are not a silver bullet}
+
+Before wrapping up this section, we would like to show that maybe, union and intersection
 contracts are not exactly what we would want them to be.
@@
 are accustomed to perform safety checks (data validation) with
-great amounts of flexibility, and the whole power of their
-language of choice at their disposition.
+great amounts of flexibility and the whole power of their
+language of choice at their disposal.
 
@@
 at runtime, that a given variable is an integer if a given flag is
-not set, as shown in Figure \ref{fig:python-assertion},
+not set, as shown in Figure \ref{fig:python-assertion};
 while a Haskell programmer might close her eyes and leave the room
@@
 ...
-assert ((not flag) or isinstance(result, int))
+assert (
+  (not flag) 
+  or isinstance(result, int)
+)
 \end{lstlisting}
@@
 
-These expectations are not really met by how unions and intersection
+These expectations are not really met by how unions and intersections
 contracts are supposed to work.
@@
 The reason this happens is that
-there is a fundamental characterization of how intersection
+there is a fundamental property of how intersection
 contracts behave: the choice of which branch to choose is made exactly
-when the intersection is opened up, and not delayed as much as possible.
+when the intersection is opened up, and not delayed as much as possible,
+as one would like.
 This means, that the contract presented on Figure \ref{fig:intersection-distrbution},
@@
 \nickel{(Number -> (Number -> Number) & (Boolean -> Number)},
-when these two expressions should be logically equivalent.
+when, with an intuitive view of how intersections should behave,
+these two expressions should be logically equivalent.
+
+It is worth noting, that intersection types do have this property,
+since subtyping would allow to correctly type check a term like
+\nickel{g 1} as both \nickel{Number -> Number} as well as
+\nickel{Boolean -> Number}, and then deducing that \nickel{g 1}
+can be given the type \nickel{(Number -> Number) & (Boolean -> Number)}.
 
@@
 
-\subsection{Keil and Thiemann}
+\subsection*{Keil and Thiemann}
 \label{sec:keil-thiemann}
@@
 
-\subsection{Williams, Morris, and Wadler}
+\subsection*{Williams, Morris, and Wadler}
 \label{sec:will-morr-wadl}
@@
 
-\subsection{Racket}
+\subsection*{Racket}
 \label{sec:racket}

### 2021-05-28 paper.tex
Pass on section 5

@@
 \end{figure}
-\resolved{I wouldn't use nickel-lang.org unless you are
-planning on putting something up there. Maybe nixos.org?}
 
@@
 \subsection*{Flat contracts do not intersect}
+\label{sec:flat-and-inter}
 
@@
 
-The papers that previously researched union and intersection
+The existing work that previously researched and implemented
+union and intersection
 as dynamic checks have, in our eyes, different shortcomings
 that makes them suboptimal solutions.
+In this section, we present informally present those issues,
+starting with the foundational paper by Keil and Thiemann
+\cite{KeilThiemannUnionIntersection}, then following to a recent
+work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
+finally going over some similar contracts available in the Racket
+programming language.
 
@@
 \unsure{Subsection title?}
-Starting with the Keil and Thiemann work \cite{KeilThiemannUnionIntersection},
-they noticed that, since union contracts should share information between
+Keil and Thiemann noticed that, since union contracts should share information between
 different application contexts, while intersection contracts should behave
@@
 simpler if every contract gets transformed into a union of intersections
-of other contracts, by using the usual De Morgan laws,
+of other contracts, by using the common De Morgan laws,
 and then eagerly opening up unions, but postponing intersection
@@
 
-For instance, if the value \nickel{V} is wrapped with contract
-\nickel{A & (B | C)}, then, before the contract is checked,
-it will get opened up to something equivalent to
-\nickel{(A & B) | (A & C)}.
+For instance, if someone were to wrap a value \nickel{x}
+in a contract as shown on the Original code snippet on Figure
+\ref{fig:kt-contract-transf}, then, before actually executing that
+contract, it would get transformed, and intersections would
+be pushed inside the unions, as can be seen on the Transformed
+code snippet on the same figure.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Original]
+x @ A & (B | C)
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Transformed]
+x @ (A & B) | (A & C)
+\end{lstlisting}
+\caption{Contract transformation performed by the Keil and
+Thiemann work}
+\label{fig:kt-contract-transf}
+\end{figure}
 
 This straightforward solutions makes composing the Keil and Thiemann
-system with other existing systems more difficult; in the words
+solution with other existing systems more difficult; in the words
 of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
@@
 fact that flat contracts and intersection do not usually play nice
-together.
+together, as shown on Section \ref{sec:flat-and-inter}.
 They provide two different sets of rules for their calculus,
-the first one is non deterministic, whose purpose is to provide
+the first one is non deterministic, and its purpose is to provide
 semantics that are simple to understand.
-The second one is a deterministic set of rules, that are intended
+With these rules, the interaction between flat contracts and
+intersections becomes much simpler, since they are able to check
+both sides of an intersection on separate threads.
+This allows them to be sure that contracts from one branch of the intersection
+won't get blamed while executing the other branch.
+
+As an example, consider the programs shown on Figure
+\ref{fig:kt-non-deterministic}, if the Original program
+gets executed, it will spawn two instances of said program,
+one with the first branch of the intersection (the First Program),
+and the other one with the second one (the Second Program),
+and these will get joined if they reach an equal execution point.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Original]
+x @ A & B
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=First Program]
+x @ A
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Second Program]
+x @ B
+\end{lstlisting}
+\caption{Non deterministic intersection contract opening}
+\label{fig:kt-non-deterministic}
+\end{figure}
+
+The second set of rules is deterministic, and it is intended
 to be simple to implement.
-This second set needs to perform operations
+However, this second set needs to perform operations
 on the evaluation context, to decide whether or not to evaluate
 a given contract.
-\unsure{Worth showing how the non deterministic version works? (See section B.1.1)
-try with an example}
+\footnote{You can check more on this topic on Section 5.1 of
+\cite{KeilThiemannUnionIntersection}}
 
@@
 as well as on the complexity levels.
-\unsure{(Yann) Is it really a
+\resolved{(Yann) Is it really a
     problem per se? It is common thing to do (a nice declarative system
@@
 difficult, or even impossible
-\unsure{(Yann) I think we need to substantiate this claim. If this
+\resolved{(Yann) I think we need to substantiate this claim. If this
 context dependency explains why CSE is invalid, maybe have a little example?
@@
 
-\subsection*{Racket}
+\subsection*{[WIP] Racket}
 \label{sec:racket}
+
+Racket is a language based on Scheme, mainly thought as a language to experiment
+with different language desings and ideas. It provides a very complete contract
+system\cite{RacketContracts}.
+
+Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
+These contracts are lightweight versions of intersections and unions, they simply
+check that every contract is valid, in the case of \racket{and/c},
+and that at least one contract
+is valid, in the case of \racket{or/c}.
+
+So, for instance, the usual usecase of intersection to encode bounded overloading
+does not work on Racket. Take a look at the next example, the function
+\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
+since 3 is not a valid \racket{string?}.
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=racket]
+(define/contract overload
+(and/c (-> number? number?)
+       (-> string? string?))
+(lambda (x) x))
+\end{lstlisting}
+\caption{Non working function overloading on Racket.}
+
+\end{figure}
+
+Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
+contracts that could (potentially) apply to a function, these have to be
+differentiated by first order markers.
+For instance, the code on Figure~\ref{code:racket:or/c:working}
+would correctly allow only to call
+\racket{united} with a \racket{number?},
+and it can make this distinction since it knows that
+\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
+However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
+before running, since the \racket{or/c} contract is unable to know
+which of the two function contracts to use.
+Ideally, this last example should just behave as wrapping
+the function with the \racket{(-> even? even?)} contract.
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=racket]
+(define/contract united
+(or/c (-> number? number?)
+      (-> string? string? string?))
+(lambda (x) x))
+\end{lstlisting}
+\caption{Working \racket{or/c} example on Racket.}
+\label{code:racket:or/c:working}
+
+\end{figure}
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=lisp]
+(define/contract united
+(or/c (-> number? number?)
+      (-> even? even?))
+(lambda (x) x))
+\end{lstlisting}
+\caption{Non working \racket{or/c} example on Racket.}
+\label{code:racket:or/c:non-working}
+
+\end{figure}
+
+\unsure{This section is overall a bit hard to follow. I
+  think I've figured out what it means, but it needs more love. Also
+  give an example which doesn't work and one that does for or/c.}
+
+\subsubsection*{case->}
+
+Racket does provide an alternative for overloading methods, called the
+\racket{case->} contract, however, this overloading of methods only
+works over contracts of functions with different amount of parameters.
+It's a good alternative to allow limited overloading of functions.
+As an example, consider the following non failing piece of code, and
+try to understand how Racket can decide, at each application,
+which function contract to use.
+
+\begin{figure}[h]
+
+\begin{lstlisting}[language=racket]
+(define/contract overcase
+  (case-> (-> string? string?)
+          (-> number? number? number?)
+          )
+  (lambda (x [y 0]) (if (number? x)
+                   (+ x y)
+                   x)))
+
+(overcase 1 2)
+
+(overcase "hello")
+\end{lstlisting}
+\end{figure}
+
+\todo{Coming back to the problems of composing U/I with other typing
+features, Racket has made a clear choice and decided that higher order
+is more important (a wise decision)}
 
@@
 
-Racket is a language based on Scheme, mainly thought as a language to experiment
-with different language desings and ideas. It provides a very complete contract
-system\cite{RacketContracts}.
-
-Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
-These contracts are lightweight versions of intersections and unions, they simply
-check that every contract is valid, in the case of \racket{and/c},
-and that at least one contract
-is valid, in the case of \racket{or/c}.
-
-So, for instance, the usual usecase of intersection to encode bounded overloading
-does not work on Racket. Take a look at the next example, the function
-\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
-since 3 is not a valid \racket{string?}.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract overload
-(and/c (-> number? number?)
-       (-> string? string?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Non working function overloading on Racket.}
-
-\end{figure}
-
-Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
-contracts that could (potentially) apply to a function, these have to be
-differentiated by first order markers.
-For instance, the code on Figure~\ref{code:racket:or/c:working}
-would correctly allow only to call
-\racket{united} with a \racket{number?},
-and it can make this distinction since it knows that
-\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
-However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
-before running, since the \racket{or/c} contract is unable to know
-which of the two function contracts to use.
-Ideally, this last example should just behave as wrapping
-the function with the \racket{(-> even? even?)} contract.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract united
-(or/c (-> number? number?)
-      (-> string? string? string?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Working \racket{or/c} example on Racket.}
-\label{code:racket:or/c:working}
-
-\end{figure}
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=lisp]
-(define/contract united
-(or/c (-> number? number?)
-      (-> even? even?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Non working \racket{or/c} example on Racket.}
-\label{code:racket:or/c:non-working}
-
-\end{figure}
-
-\unsure{This section is overall a bit hard to follow. I
-  think I've figured out what it means, but it needs more love. Also
-  give an example which doesn't work and one that does for or/c.}
-
-\subsubsection*{case->}
-
-Racket does provide an alternative for overloading methods, called the
-\racket{case->} contract, however, this overloading of methods only
-works over contracts of functions with different amount of parameters.
-It's a good alternative to allow limited overloading of functions.
-As an example, consider the following non failing piece of code, and
-try to understand how Racket can decide, at each application,
-which function contract to use.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract overcase
-  (case-> (-> string? string?)
-          (-> number? number? number?)
-          )
-  (lambda (x [y 0]) (if (number? x)
-                   (+ x y)
-                   x)))
-
-(overcase 1 2)
-
-(overcase "hello")
-\end{lstlisting}
-\end{figure}
-
-\todo{Coming back to the problems of composing U/I with other typing
-features, Racket has made a clear choice and decided that higher order
-is more important (a wise decision)}
-
 \subsection{Python - MyPy}

### 2021-05-28 paper.tex
Add a few lines to explain the syntax

@@
         *{->}{$\rightarrow$}1
+        {...}{\ldots}1
         {=>}{$\Rightarrow$}1
@@
 
+We will use the syntax of Nickel to write most examples throughout this paper.
+Syntactically, Nickel looks like a simple ML language with records. \nickel{let
+id = value in exp} is used to bind a value to an identifier in an expression.
+Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
+are written as \nickel{[x1, ..., xn]} and records as \nickel{\{ field1 = value1,
+..., fieldn = valuen \}}. Some examples uses constructions that are not
+currently available in Nickel for illustration purpose: in particular, we will
+use \nickel{A & B} (resp.  \nickel{A | B}) for the intersection (resp. union) of
+contracts (or types) \nickel{A} and \nickel{B}.
+
 \subsection*{Correct programs \& valid data}

### 2021-05-28 paper.tex
Small changes

@@
 typing, a discipline for mixing static and dynamic typing in the same language.
-As such, they are a precious ingredient for handling the interaction of
+As such, they are a precious ingredient for handling the interaction between 
 statically typed and dynamically typed code. While we focused on the writing of
@@
     \item \textbf{Even in a purely dynamically typed language, contracts are an
-        appealing feature for validation and correctness}. The gradual
-        typing part deserved to be mentioned, but is is out of the scope of this
-        paper.
+        appealing feature for validation and correctness}. The gradual typing
+        part enters the general picture but is out of the scope of this paper.
     \item \textbf{User-defined contracts are essential for validation}. Some
@@
 Evaluation strategies relates to the evaluation of the expression bound to
-identifiers. Expressions are be bound by assignments (e.g. the \nickel{blah} construct in Nickel), or arguments are bound to
-parameters in a function application. We will use Figure~\ref{} as a working
-example.
+identifiers. Expressions are be bound by assignments (e.g. let bindings in
+Nickel), or by function applications (arguments are bound to parameters). We
+will use Figure~\ref{} as a working example.
 

### 2021-05-29 paper.tex
Nice shape secs 5.1 and 5.2

@@
 that makes them suboptimal solutions.
-In this section, we present informally present those issues,
+In this section, we informally present those issues,
 starting with the foundational paper by Keil and Thiemann
@@
 
-This straightforward solutions makes composing the Keil and Thiemann
+This straightforward idea makes composing the Keil and Thiemann
 solution with other existing systems more difficult; in the words
@@
 and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition.''\cite{RootCauseOfBlame}
+special cases can hinder composition.''\cite{RootCauseOfBlame}.
 \info{(Yann) Does
@@
 since some sharing mechanism should be used to avoid checking the same contract,
-over the same value, multiple times.\unsure{Arnaud says: this whole
+over the same value, multiple times, exemplified by contract \nickel{A} on
+Figure \ref{fig:kt-contract-transf}, that gets duplicated.
+\unsure{Arnaud says: this whole
 section is too abstract give concrete code and explain what goes
@@
 This allows them to be sure that contracts from one branch of the intersection
-won't get blamed while executing the other branch.
+won't get blamed while executing the other branch, and vice versa.
 
@@
 gets executed, it will spawn two instances of said program,
-one with the first branch of the intersection (the First Program),
-and the other one with the second one (the Second Program),
-and these will get joined if they reach an equal execution point.
+one with the first branch of the intersection (\nickel{3 @ Number}),
+and the other one with the second branch (\nickel{3 @ Positive}),
+and these will get joined once they reach an equal point, in this case
+it will be \nickel{3}.
 
@@
 \begin{lstlisting}[language=nickel, title=Original]
-x @ A & B
+3 @ Number & Positive
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=First Program]
-x @ A
+3 @ Number
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=Second Program]
-x @ B
+3 @ Positive
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=Joined Program]
+3
 \end{lstlisting}
@@
 The second set of rules is deterministic, and it is intended
-to be simple to implement.
+to be simpler to implement.
 However, this second set needs to perform operations
@@
 \cite{KeilThiemannUnionIntersection}}
+For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
+when using the value \nickel{(fun x => x) @ (String -> String)}
+inside the execution of contract \nickel{C}, and applying it to \nickel{0},
+the contract runtime would check the context of execution and it would
+deduce that contract \nickel{String -> String} does not apply in this case,
+skipping it.
 
@@
 it makes composing this, with other features desired on programming languages,
-difficult, or even impossible
+difficult, or even impossible.
 \resolved{(Yann) I think we need to substantiate this claim. If this
@@
     argument we can make (like it's quadratic in the number of open contracts or
-whatever)}.
+whatever)}
 
@@
 This restriction goes against what is usually expected from dynamic checks:
-validation of data, and not just inclusion on types, is an expected
-feature of contracts.
+validation of data, and not just inclusion on types.
 
@@
 all of their expected monitoring properties for higher ordered, union
-and, intersection contracts.
+and intersection contracts.
 However, one of these properties states the following:
 
+\begin{figure}[h]
 $$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
-
-Which means that a continuation ($K$) complies negatively with an
+\caption{Sound monitoring property for intersection from Williams, Morris, and Wadler.}
+\label{fig:wmw-semantics}
+\end{figure}
+
+Which means that a continuation $K$ complies negatively with an
 intersection of $A$ and $B$ if it complies negatively with at least one of these.
+Or, less formally, that if you have a program that is missing a value at a given point,
+you can put a value wrapped with a contract stating $A \cap B$,
+and expect not to get a negative violation of that contract,
+if you can either put a value
+wrapped with $A$, or a value wrapped with $B$.
 
 Altough this is true, it's a weaker statement of what we'd like to have,
-since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
-where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
-but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
-and dually for $not~(f ~ true)$.
-Which would mean, according to their semantics, that
-the continuation presented above would not
-be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
-even though our intuition (and the
-introduction to the paper) indicates that it should.
-\todo{Simplify example}
-
-% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
-% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
-% and dually for $K_2$, which would mean the continuation presented above would not
-% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
-% introduction to the paper) indicates that it should.
-
-While this mistake does not invalidate their work (the system they present
+consider the two contexts presented on Figure \ref{fig:valid-contexts},
+the first one is a valid context for a \nickel{Number -> Number} function,
+that is, you could replace the $\square$ by some function wrapped in a
+\nickel{Number -> Number} contract, and the context will not violate it.
+Similarly, the second context from the same figure is a valid
+\nickel{Boolean -> Boolean} context.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=\nickel{Number -> Number} context]
+(*@$\square$@*) 3
+\end{lstlisting}
+\begin{lstlisting}[language=nickel, title=\nickel{Boolean -> Boolean} context]
+(*@$\square$@*) true
+\end{lstlisting}
+\caption{Two different contexts in Nickel}
+\label{fig:valid-contexts}
+\end{figure}
+
+Next, we can combine these two contexts into one, that essentially
+has one single missing value, that will be used independently in both contexts
+from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
+This context is not a valid \nickel{Number -> Number} context, since it will apply
+the value put on $\square$ (through \nickel{f}) to a non \nickel{Number} (\nickel{true}).
+In a similar way, it is not a valid \nickel{Boolean -> Boolean} context either.
+Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
+shown on Figure \ref{fig:wmw-semantics}, it is not a valid
+\nickel{(Number -> Number) & (Boolean -> Boolean)} context.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (*@$\square$@*) in
+(f 3, f true) 
+\end{lstlisting}
+\caption{Invalid \nickel{(Number -> Number) & (Boolean -> Boolean)} context,
+combining contexts from Figure \ref{fig:valid-contexts}}
+\label{fig:invalid-context}
+\end{figure}
+
+However, both our intuition, as the most common use case of intersections
+(that is, overloading), tells us that this should work.
+While this problem does not invalidate their work (in fact, the system they present
 handles these cases correctly), it shows that existing research is still not on
 good shape to handle the growing usecases being developed for
-union and intersection around dynamic languages.
+unions and intersections around dynamic languages.
 
@@
 % K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
-\unsure{Is it worth mentioning how continuations are formed?
-Should we go so techincal?}
-\todo{(Yann) Maybe adding a concrete example of this (it can be the same term but with
+\resolved{(Yann) Maybe adding a concrete example of this (it can be the same term but with
     concrete types like Number and String and simple contexts like application

### 2021-05-29 paper.tex
Took away a lot of parts from the appendix

@@
 
-\section{Current (Real life) implementations}
-
-\info{The goal of this section is to outline existing languages that implement
-union and intersection, in some form, and shortly compare these
-with the semantics tried to obtain on research (WMW, KT)}
-
-Unions and intersection are starting to become more and more common
-on non-academic programming languages, as stated on the introduction,
-unions are a natural abstraction for dynamic languages, where the type
-of an expression could have many different shapes, without necessarily
-having a mark to distinguish between them at type level.
-With this in mind, a valid question is how do these languages solve the
-union/intersection conundrum? Or, even better, what is it that they
-solve?
-
-In the rest of this section we revisit some different approaches.
-
-\subsection{Racket}
-
-\subsection{Python - MyPy}
-
-\unsure{We ought to group all the static typing in one subsection}
-
 \end{document}

### 2021-05-29 paper.tex
Pass on Racket

@@
     define-context, define-controller, Integer, Boolean, get, when-required, when-provided,
-    maybe_publish, require, submod, or/c, ->, \#\%module-begin,
+    maybe_publish, require, submod, or/c, and/c, ->, \#\%module-begin,
     always_publish, with-syntax, define-struct/contract, syntax-case,
@@
 
-\subsection*{[WIP] Racket}
+\subsection*{Racket}
 \label{sec:racket}
@@
 
-\todo{Coming back to the problems of composing U/I with other typing
-features, Racket has made a clear choice and decided that higher order
-is more important (a wise decision)}
-
-\unsure{Todo}
+Even if Racket's intent is not to implement union and intersection
+dynamic checks, they do provide contracts that are, in some sense,
+simplified versions of those.
+In our eyes, Racket has accepted the difficulty of correctly and practically
+combine union, intersection, and higher order contracts, and has choose
+to prioritise the highe order ones.
 

### 2021-05-30 paper.tex
Paper structure

@@
 \affiliation{
-  \institution{???}
+  \institution{Universidad de Buenos Aires}
   \city{Buenos Aires}
@@
 }
-\email{???}
+\email{tfreund95@gmail.com}
+
 \author{Yann Hamdaoui}
@@
 \section{Introduction}
+\label{sec:intro}
 \info{Goals of the paper: position paper: unions/intersections nice in
@@
 
+\subsection*{Structure of this paper}
+This paper is structured as follows, Section \ref{sec:intro} introduces
+this paper; Section \ref{sec:feat-lang} outlines various features
+of programming languages that may interact badly with union and intersection
+contracts.
+Section \ref{sec:union-inter} gives a primer on Unions and Intersections,
+enough to understand this work.
+
+Section \ref{sec:issues-sem} shows what difficulties or issues, proper of
+unions and intersections, one may
+encounter when trying to develop a union and intersection contracts.
+Section \ref{sec:issues-literature} presents issues and shortcomings found
+in existing attempts to solve this problem.
+Section \ref{sec:related-work} provides some related work, and Section
+\ref{sec:conclusion} concludes.
+
 \newpage
@@
 \section{Features of languages}
+\label{sec:feat-lang}
 \info{Zoology of various features that we will eventually show
@@
 \section{Union \& intersection}
+\label{sec:union-inter}
 \info{What are they, what are they used for.}
@@
 \section{Issues coming from the semantics}
+\label{sec:issues-sem}
 \info{What goes wrong or difficult when trying to implement union and
@@
 \section{Issues in concrete restrictions from the literature}
+\label{sec:issues-literature}
 \info{The various papers and the tradeoffs they make}
@@
 \section{Related work}
+\label{sec:related-work}
 \info{Including a mention of statically typed systems with union}
-\info{Include something about Racket's and and or contracts}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2021-05-30 paper.tex
Pass on 1, 2

@@
 
-For instance, we could define a flat contract \nickel{even} as
-\nickel{fun x => x \% 2 == 0}, that checks
-whether a number is even by checking the remainder of dividing it by 2.
-And then use it to check that a function always returns an
-even number, when applied to an \nickel{Integer}, as shown in
-Figure \ref{fig:num-to-even}.
-Or even, we could come up with a contract that checks that a function
-maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
-and use it as a contract for the \nickel{not} function, as can be seen on Figure
+For instance, the contract \nickel{Even} shown on Figure
+\ref{fig:contract-higher-order} is a flat contract, defined
+by taking the value that needs to be checked, and validating
+that is both a number and that the remainder of dividing it by 2
+is 0.
+And then, as seen on the same figure, we can use it to check that a function always
+returns an \nickel{Int} when applied to an \nickel{Even}.
+
+% \begin{figure}[h]
+% \begin{center}
+% \begin{lstlisting}[language=nickel]
+% let f = (fun n => n * 2) @ (Integer -> even)
+% \end{lstlisting}
+%   \end{center}
+% \caption{A function wrapped with a contract stating that it always returns
+% an even number}
+% \label{fig:num-to-even}
+% \end{figure}
+
+Or even more interesting, we could come up with a contract that checks that a function
+maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
+by applying the function and checking its results,
+and then use it as a contract for the \nickel{not} function, as can be seen on Figure
 \ref{fig:checked-not}.
@@
 \begin{lstlisting}[language=nickel]
-let f = (fun n => n * 2) @ (Integer -> even)
-\end{lstlisting}
-  \end{center}
-\caption{A function wrapped with a contract stating that it always returns
-an even number}
-\label{fig:num-to-even}
-\end{figure}
-
-% $$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let notC = fun f => (f true == false) and (f false == true) in
-let checkedNot = not @ notC
+let NotC = contracts.fromPred (
+  fun f => builtins.isFun f && (f true == false) && (f false == true)
+) in
+let checkedNot = not | #NotC
 \end{lstlisting}
@@
 This type of contracts are present on many different languages,
-for instance, the Eiffel programming language\unsure{Citation?}, the precursor
+for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
 of the Design by Contract ideals, allows to assert
 these kinds of expression as pre- and post-conditions on
-functions\cite{EiffelDesignByContract}.
-A somewhat newer language, the D programming
-language, allows, among other things, to have invariants on
-classes\unsure{I (Arnaud) believe Eiffel has this feature as well,
-  doesn't it?},
-written with flat contracts, that are
-dinamically checked at the frontiers of every public method \cite{DLangContracts}.
+functions and as invariants on classes\cite{EiffelDesignByContract}.
+\resolved{I (Arnaud) believe Eiffel has this feature as well,
+  doesn't it? (Teo) You're right},
+
+% A somewhat newer language, the D programming
+% language, allows, among other things, to have invariants on
+% classes
+% written with flat contracts, that are
+% dinamically checked at the frontiers of every public method \cite{DLangContracts}.
 
@@
 and intersection contracts\cite{RacketContracts}.
-\todo{Talk about this, probably later on the paper.
-See Related Works section}
 
@@
 Beyond CSE, optimizations such as loop-invariant code motion or
-let-floating~\ref{} apply the same principle of extracting out an invariant
+let-floating~\cite{letFloating} apply the same principle of extracting out an invariant
+\unsure{Was this the inteded citation?}
 expression to avoid its recomputation respectively across loop iterations and
@@
 
-Strikingly, we will see in section~\ref{} that the introduction of unions and
+Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of unions and
 intersections can make this equivalence \todo{name it} unsound \emph{even in a
@@
 \subsection*{Lazyness}
+\todo{Lazyness does not appear on sec 4, if we don't put somethign there, I'd remove it}
 Evaluation strategies relates to the evaluation of the expression bound to
-identifiers. Expressions are be bound by assignments (e.g. let bindings in
+identifiers. Expressions are bound by assignments (e.g. let bindings in
 Nickel), or by function applications (arguments are bound to parameters). We
@@
 (e.g. Ocaml and ReScript), or by default for specific primitives types such as
-like list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
+list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
 (and its web avatar PureScript). Although rare in general purpose languages,
@@
 \info{What are they, what are they used for.}
+\todo{Since nickel uses | for contract application, I'd move unions
+to some other symbol.
+I think the best option is to move to $\cap / \cup$}
 

### 2021-05-31 paper.tex
Rework the plan section

@@
 use \nickel{A & B} (resp.  \nickel{A | B}) for the intersection (resp. union) of
-contracts (or types) \nickel{A} and \nickel{B}.
+contracts (or types) \nickel{A} and \nickel{B}. Finally, we will write
+\nickel{exp @ C}\footnote{The current Nickel implementation actually uses the
+syntax \nickel{exp | C} for contract application, but we rather keep \nickel{|}
+for unions here}  for the application of a contract \nickel{C} to the expression
+\nickel{exp}.
 
@@
 \subsection*{Structure of this paper}
-This paper is structured as follows, Section \ref{sec:intro} introduces
-this paper; Section \ref{sec:feat-lang} outlines various features
-of programming languages that may interact badly with union and intersection
-contracts.
-Section \ref{sec:union-inter} gives a primer on Unions and Intersections,
-enough to understand this work.
-
-Section \ref{sec:issues-sem} shows what difficulties or issues, proper of
-unions and intersections, one may
-encounter when trying to develop a union and intersection contracts.
-Section \ref{sec:issues-literature} presents issues and shortcomings found
-in existing attempts to solve this problem.
-Section \ref{sec:related-work} provides some related work, and Section
-\ref{sec:conclusion} concludes.
+Section~\ref{sec:intro} introduced the perspective of designing a configuration
+language. We presented the concept of contracts and detailed why contracts are
+particularly fitted for a configuration language, and more generally for data
+validation or program correctness in a dynamically typed language. The rest of
+the paper is structured as follows:
+\begin{itemize}
+    \item Section~\ref{sec:feat-lang} outlines various common features of
+        programming languages that may interact in a bad way with union and
+        intersection contracts.
+    \item Section~\ref{sec:union-inter} introduces unions and intersections and
+        lists the possible motivations for adding them to a language.
+    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
+        encounter when trying to implement union and intersection contracts in
+        practice.
+    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
+        existing attempts to solve said difficulties.
+    \item Section~\ref{sec:related-work} provides some related work.
+\end{itemize}
+
+Section~\ref{sec:conclusion} concludes the paper.
 

### 2021-05-31 paper.tex
Make the syntax of examples uniform + fix typos

@@
     keywords=[3]{
-      Num,
-      Bool,
-      Str,
-      List,
-      Dyn,
       forall,
@@
         {...}{\ldots}1
+        {@&}{$\cap$}1
+        {@|}{$\cup$}1
         {=>}{$\Rightarrow$}1
@@
 Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
-are written as \nickel{[x1, ..., xn]} and records as \nickel{\{ field1 = value1,
-..., fieldn = valuen \}}. Some examples uses constructions that are not
-currently available in Nickel for illustration purpose: in particular, we will
-use \nickel{A & B} (resp.  \nickel{A | B}) for the intersection (resp. union) of
-contracts (or types) \nickel{A} and \nickel{B}. Finally, we will write
-\nickel{exp @ C}\footnote{The current Nickel implementation actually uses the
-syntax \nickel{exp | C} for contract application, but we rather keep \nickel{|}
-for unions here}  for the application of a contract \nickel{C} to the expression
-\nickel{exp}.
+are written as \nickel{[x1, ..., xn]} and records as \nickel{\{field1 = value1,
+..., fieldn = valuen\}}. The application of a contract \nickel{C} to the
+expression \nickel{exp} is written \nickel{exp | C}\footnote{The current
+    implementation requires a \nickel{#} character in front of custom contracts,
+    whose application would rather be \nickel{exp | #C}, but it is irrelevant
+and subject to change in the future.}. We will sometimes attach
+contracts directly to an identifier instead of the corresponding bound
+expression, that is: \nickel{let id | C = value in exp} is syntactic sugar for
+\nickel{let id = (value | C) in exp}.
+
+Some examples resort to constructions that are not currently available in Nickel
+for illustration purpose: we will use \nickel{A @& B} (resp. \nickel{A @| B})
+for the intersection (resp. union) of contracts (or types) \nickel{A} and
+\nickel{B}. \nickel{@|} and \nickel{@&} bind more loosely than arrows, meaning
+that \nickel{A -> B @& C -> D} is to be parsed as \nickel{(A -> B) @& (C -> D)}.
 
@@
 higher-order functions. If the user has written an \nickel{Even} contract for
-numbers, they can use it right away to form a function contract \nickel{#Even ->
-#Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
+numbers, they can use it right away to form a function contract \nickel{Even ->
+Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
 contract will check at each call that the argument is an even number
@@
   && val % 1 == 0) in
-{divBy2 | #Even -> #Int = fun x => x/2}
+{divBy2 | Even -> Int = fun x => x/2}
 \end{lstlisting}
@@
 let Configuration = {
-  host | #Host,
-  port | #Port,
+  host | Host,
+  port | Port,
 } in
 {host = "nixos.org", port = 80}
-| #Configuration
+| Configuration
 \end{lstlisting}
@@
         contracts.  Rather, a fixed set of built-in contracts such as
-        \nickel{Number} or \nickel{String} are provided, very much like base
+        \nickel{Num} or \nickel{Str} are provided, very much like base
         types. The absence of user-defined contracts is understandable when
@@
 % \begin{lstlisting}[language=nickel]
-% let f = (fun n => n * 2) @ (Integer -> even)
+% let f = fun n => n * 2 @ (Integer -> even)
 % \end{lstlisting}
@@
 the language's strategy, it is a valid transformation. On the other hand,
-CSE-like transformations require an expansion property of the form
-
-$$M[N] \simeq let~x~=~N~in~M[x]$$
-
-This equation clearly fails in a call-by-value language with side effects:
+a CSE-like transformation on a term $M$ requires an expansion property of the form
+
+\begin{equation}\label{eq:cbn-expansion}
+M[x/N] \simeq let~x~=~N~in~M
+\end{equation}
+
+Here, $M[x/N]$ is the substitution of $x$ for $N$ in the term $M$.
+\ref{eq:cbn-expansion} has an evident call-by-name flavour, and consequently
+fails in general for a call-by-value language with side effects.  Indeed, define
+the following function $f$:
 
 \begin{lstlisting}[language=Nickel]
-let f x = print "hi";(x+1) in (f 1,f 1)
-(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
-(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
+let f x = print "hi";(x+1) in 
+\end{lstlisting}
+
+Then:
+
+\begin{lstlisting}[language=Nickel]
+(f 1,f 1) (*@ $\not \simeq$ @*) let y = f 1 in (y,y)
 \end{lstlisting}
@@
 In the example above, the first term prints \nickel{"hi"} two times while the
-second term prints it only once. However, this property does hold for
+second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
 \emph{pure} terms, that are terms without side-effects. In a pure language, or a
@@
 
-Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of unions and
-intersections can make this equivalence \todo{name it} unsound \emph{even in a
-pure setting}, making the optimization of
+Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
+unions and intersections can make \ref{eq:cbn-expansion} \resolved{it} unsound
+\emph{even in a pure setting}, making the optimization of
 Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
@@
 type BinaryTree =
-Integer | (BinaryTree, BinaryTree)
+Integer @| (BinaryTree, BinaryTree)
 \end{lstlisting}
@@
 \begin{itemize}
-    \item To represent nullable values(\ref{fig:union-nullable-ex}).
+    \item To represent nullable values (\ref{fig:union-nullable-ex}).
     \item For plain enumerations (\ref{fig:union-enum-ex}).
@@
 \begin{lstlisting}[title={Nullable values}]
-optionalValue : String | Null
+optionalValue : Str @| Null
 \end{lstlisting}
@@
 \begin{lstlisting}[title={Plain enumeration}]
-protocol : `Http | `Ftp | `Sftp
+protocol : `Http @| `Ftp @| `Sftp
 \end{lstlisting}
@@
 \begin{lstlisting}[title={Alternative representations}]
-person : {name : String, age : Int}
-  | {name : String, dateOfBirth : String}
+person : {name : Str, age : Int}
+  @| {name : Str, dateOfBirth : Str}
 \end{lstlisting}
@@
 \begin{lstlisting}[title={Value subtyping}]
-1 : Int & Float
+1 : Int @& Float
 \end{lstlisting}
@@
 \begin{lstlisting}[title={Overloading}]
-incr: (Int -> Int) & (Float -> Float) =
+incr: (Int -> Int) @& (Float -> Float) =
   fun x => x + 1
@@
 \label{fig:intersection-mixins-ex}
-\begin{lstlisting}[title={Structural mixins}]
+\begin{lstlisting}[language=Scala,title={Structural mixins}]
 Comparator: {cmp: T -> T -> Bool}
@@
 \begin{lstlisting}[title={Intersection as the dual of unions}]
-f : (Num | String) -> Bool
-(*@ $\simeq$ @*) f : (Num -> Bool) & (String -> Bool)
+f : (Num @| Str) -> Bool
+(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
 \end{lstlisting}
@@
 \begin{lstlisting}[language=nickel]
-let f = (fun x => x - 7) @ Positive -> Positive in
+let f | Positive -> Positive
+      = fun x => x - 7 in
 (f 10) + (f 5)
@@
 \begin{lstlisting}[language=nickel]
-let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
+let f | (Positive -> Positive)
+        @| (Positive -> NonPositive)
+      = fun x => x - 7 in
 (f 10) + (f 5)
@@
 \begin{lstlisting}[language=nickel, title=Inlined]
-let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
-(f 1 true, f 1 false)
+let f | Num -> (Bool -> Num @| Bool -> Str)
+      = fun x y => if y then x else "False"
+in (f 1 true, f 1 false)
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
-let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
-let g x = f 1 x in
+let f = (fun x y => if y then x else "False") | Num -> (Bool -> Num @| Bool -> Str) in
+let g = f 1 in
 (g true, g false)
@@
 
-Unexpectedly, these two programs behave very differently, the first one will return
-\nickel{(1, "False")}, without failing.
-While the second one will fail, since the single application of \nickel{f 1},
-applied to both \nickel{true} and \nickel{false} will be able to find the error
-on the union contract (\nickel{(Boolean -> Number) | (Boolean -> String)}).
+Unexpectedly, these two programs behave very differently, the first one will
+return \nickel{(1, "False")}, without failing.  While the second one will fail,
+since the single application of \nickel{f 1}, applied to both \nickel{true} and
+\nickel{false} will be able to find the error on the union contract
+\nickel{(Bool -> Num) @| (Bool -> Str)}.
 
@@
 
-Summing it up, wrapping a value with a union of higer order contracts means that
+Summing it up, wrapping a value with a union of higher order contracts means that
 that value now has a state, that may get modified silently at every use site.
@@
 \begin{lstlisting}[language=nickel, title=Implementation]
-x @ A & B
-(*@ $\simeq$ @*)
-(x @ A) @ B
+x | A @& B (*@ $\simeq$ @*) (x | A) | B
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=Problem]
-let g = (fun x => x) @ (Number -> Number) & (String -> String) in
+let g | Num -> Num @& Str -> Str
+      = fun x => x in
 g 1
@@
 \nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
-that gets checked is the argument of \nickel{String -> String} over the function parameter, so
-something like \nickel{1 @ String}.
+that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
+something like \nickel{1 | Str}.
 Of course, this fails, but before raising a violation, this contract could
@@
 \begin{lstlisting}[language=nickel, title=Implementation]
-x @ A & B
-(*@ $\simeq$ @*)
-(x @ A[l]) @ B[l]
+x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
 \end{lstlisting}
@@
 \nickel{g true}, that would fail also negatively the contract
-\nickel{String -> String} and, when checking if \nickel{true}
+\nickel{Str -> Str} and, when checking if \nickel{true}
 is a correct parameter for \nickel{Number -> Number} it would check
@@
 as presented in Section \ref{sec:flat-contracts}.
-Consider the example presented on Figure \ref{fig:inter-flat-contracts},
-if the implementation discussed previously were to be used, we may end up with
-something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
+Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
+implementation discussed previously were to be used, we may end up with
+something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
 \nickel{l} represents the shared state.
-At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
-if \nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
-This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
+At this point, the intuitive thing to do is to check the contract \nickel{C}, by
+checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
+since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
 fails negatively.
@@
 let C = fun f => (f 0) == 0 in
-let g = (fun x => x) @ (String -> String) & C
+let g | (String -> String) @& C
+      = (fun x => x) 
 \end{lstlisting}
@@
 \ref{fig:intersection-distrbution}, at first sight, this code should
-complete execution succesfully, returning \nickel{(1, 1)},
+complete execution successfully, returning \nickel{(1, 1)},
 however, under both the Keil and Thiemann, as well as the
@@
 let f = fun x y => x in
-let g = f @ (Number -> Number -> Number) & (Number -> Boolean -> Number) in
+let g = f | (Num -> Num -> Num)
+            @& (Num -> Bool -> Num) in
 let h = g 1 in
@@
 
-The reason this happens is that
-there is a fundamental property of how intersection
-contracts behave: the choice of which branch to choose is made exactly
-when the intersection is opened up, and not delayed as much as possible,
-as one would like.
-This means, that the contract presented on Figure \ref{fig:intersection-distrbution},
-\nickel{(Number -> Number -> Number) & (Number -> Boolean -> Number)},
-will not behave the same way that
-\nickel{(Number -> (Number -> Number) & (Boolean -> Number)},
-when, with an intuitive view of how intersections should behave,
-these two expressions should be logically equivalent.
+The reason this happens is that there is a fundamental property of how
+intersection contracts behave: the choice of which branch to choose is made
+exactly when the intersection is opened up, and not delayed as much as possible,
+as one would like.  This means, that the contract presented on Figure
+\ref{fig:intersection-distrbution}, \nickel{Num -> Num -> Num @& Num
+-> Bool -> Num}, will not behave the same way that \nickel{Num ->
+(Num -> Num @& Bool -> Num)}, when, with an intuitive view of how
+intersections should behave, these two expressions should be logically
+equivalent.
 
@@
 \nickel{g 1} as both \nickel{Number -> Number} as well as
-\nickel{Boolean -> Number}, and then deducing that \nickel{g 1}
-can be given the type \nickel{(Number -> Number) & (Boolean -> Number)}.
+\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
+can be given the type \nickel{Num -> Num @& Bool -> Num}.
 
@@
 \begin{lstlisting}[language=nickel, title=Original]
-x @ A & (B | C)
+x | A @& (B @| C)
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=Transformed]
-x @ (A & B) | (A & C)
+x | (A @& B) | (A @& C)
 \end{lstlisting}
@@
 \info{(Yann) Does
-it also make it less efficient? If $(A \& B)$ fails because of $B$, their
+it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
 development causes to recheck the contract $A$ that has been duplicated?}
@@
 gets executed, it will spawn two instances of said program,
-one with the first branch of the intersection (\nickel{3 @ Number}),
-and the other one with the second branch (\nickel{3 @ Positive}),
+one with the first branch of the intersection (\nickel{3 | Num}),
+and the other one with the second branch (\nickel{3 | Positive}),
 and these will get joined once they reach an equal point, in this case
@@
 \begin{lstlisting}[language=nickel, title=Original]
-3 @ Number & Positive
+3 | Num @& Positive
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=First Program]
-3 @ Number
+3 | Num
 \end{lstlisting}
 \begin{lstlisting}[language=nickel, title=Second Program]
-3 @ Positive
+3 | Positive
 \end{lstlisting}
@@
 For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
-when using the value \nickel{(fun x => x) @ (String -> String)}
+when using the value \nickel{(fun x => x) | Str -> Str)}
 inside the execution of contract \nickel{C}, and applying it to \nickel{0},
 the contract runtime would check the context of execution and it would
-deduce that contract \nickel{String -> String} does not apply in this case,
+deduce that contract \nickel{Str -> Str does not apply in this case,
 skipping it.
@@
 consider the two contexts presented on Figure \ref{fig:valid-contexts},
-the first one is a valid context for a \nickel{Number -> Number} function,
+the first one is a valid context for a \nickel{Num -> Num} function,
 that is, you could replace the $\square$ by some function wrapped in a
-\nickel{Number -> Number} contract, and the context will not violate it.
+\nickel{Num -> Num} contract, and the context will not violate it.
 Similarly, the second context from the same figure is a valid
-\nickel{Boolean -> Boolean} context.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=\nickel{Number -> Number} context]
+\nickel{Bool -> Bool} context.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=\nickel{Num -> Num} context]
 (*@$\square$@*) 3
 \end{lstlisting}
-\begin{lstlisting}[language=nickel, title=\nickel{Boolean -> Boolean} context]
+\begin{lstlisting}[language=nickel, title=\nickel{Bool -> Bool} context]
 (*@$\square$@*) true
@@
 from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
-This context is not a valid \nickel{Number -> Number} context, since it will apply
-the value put on $\square$ (through \nickel{f}) to a non \nickel{Number} (\nickel{true}).
-In a similar way, it is not a valid \nickel{Boolean -> Boolean} context either.
+This context is not a valid \nickel{Num -> Num} context, since it will apply
+the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
+In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
 Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
 shown on Figure \ref{fig:wmw-semantics}, it is not a valid
-\nickel{(Number -> Number) & (Boolean -> Boolean)} context.
+\nickel{(Num -> Num) & (Bool -> Bool)} context.
 
@@
 \end{lstlisting}
-\caption{Invalid \nickel{(Number -> Number) & (Boolean -> Boolean)} context,
+\caption{Invalid \nickel{(Num -> Num) & (Bool -> Bool)} context,
 combining contexts from Figure \ref{fig:valid-contexts}}

### 2021-05-31 paper.tex
Fix compilation + small typos

@@
 having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
-this example, the intersection \nickel{Int & Float} can be used as a subtype of
+this example, the intersection \nickel{Int @& Float} can be used as a subtype of
 both \nickel{Int} and \nickel{Float}.
@@
 Of course, this fails, but before raising a violation, this contract could
-check if the other side of the intersection (\nickel{Number -> Number}) has
+check if the other side of the intersection (\nickel{Num -> Num}) has
 also tried to raise negative blame on the same parameter ()\nickel{1}), since
@@
 \nickel{Str -> Str} and, when checking if \nickel{true}
-is a correct parameter for \nickel{Number -> Number} it would check
+is a correct parameter for \nickel{Num -> Num} it would check
 if the other side of the intersection has failed, and since it
@@
 since subtyping would allow to correctly type check a term like
-\nickel{g 1} as both \nickel{Number -> Number} as well as
+\nickel{g 1} as both \nickel{Num -> Num} as well as
 \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
@@
 the contract runtime would check the context of execution and it would
-deduce that contract \nickel{Str -> Str does not apply in this case,
+deduce that contract \nickel{Str -> Str} does not apply in this case,
 skipping it.
@@
 shown on Figure \ref{fig:wmw-semantics}, it is not a valid
-\nickel{(Num -> Num) & (Bool -> Bool)} context.
+\nickel{Num -> Num @& Bool -> Bool} context.
 
@@
 \end{lstlisting}
-\caption{Invalid \nickel{(Num -> Num) & (Bool -> Bool)} context,
+\caption{Invalid \nickel{Num -> Num @& Bool -> Bool} context,
 combining contexts from Figure \ref{fig:valid-contexts}}

### 2021-05-31 paper.tex
Add conclusion

@@
 
+The design and implementation of programming languages is a challenging task. It
+is a delicate exercise to balance complexity, expressiveness, efficiency and
+user experience all together. Unfortunately for designers, a language cannot
+usually be reduced to a simple sum of small independent features. They are
+rather composed of a lot of moving parts that interact in unexpected ways. The
+more a language grows, the more any new addition has to be weighted carefully,
+as it has the potential to interact with any of the existing features.
+
+Part of recent developments revolve around mixing static typing and dynamic
+typing via gradual typing, in order to get the best of both worlds. Contracts
+(or casts in the gradual typing terminology) are crucial to this endeavour by
+allowing for a safe interaction between typed and untyped code. But we also made
+a case that contracts are interesting in themselves in a fully dynamically typed
+language (and even in a fully statically typed language).  Contracts are a
+precious idiom for data validation or to check pre- and post-condition that
+would be hard to ensure statically. In spirit, contracts are similar to
+assertions, but they are part of a principled framework, they are composable,
+they have good error reporting and they are integrated as a language feature. In
+particular, contracts natively handles higher-order functions and built-in data
+structures. This makes them an great candidate for integration in the Nickel
+configuration language.
+
+This is only the beginning of the story. A language designer then has to choose
+what contract combinators to provide among a number of possibilities: there is
+the standard package including higher-order function contracts, record contracts
+and list contracts.  Then, one could add polymorphic
+contacts~\cite{BlameForAll}, dependent function
+contracts~\ref{DependentContracts}, and so on. For validation purpose, union and
+intersection contracts are really appealing. Unions have a intuitive
+interpretation as modeling alternatives and thus appear implicitly or explicitly
+in a lot of data structures.
+
+Alas, the implementation of union and intersection contracts is not as simple as
+it first appears. In conjunction with higher-order contracts, or even just basic
+contracts in a lazy language, they behave in a way that is reminiscent of
+side-effects. These behaviors were, for most of them, already known: we do not
+claim their discovery. However, as language implementers, we have made an actual
+attempt at including them in a concrete language. We exposed what, in our
+opinion, are the most salient flaws and their consequences. Even in a pure
+language, union and intersection contracts break the equivalence between
+call-by-name and call-by-value, and consequently can prevent CSE-like
+optimizations. There are discrepancies between the expected and actual semantics
+that make some behaviors unintuitive. Their implementation is a non-trivial
+change that contaminates the evaluation of all other contracts, and the
+performance penalty is not obvious to assess.
+
+To conclude, let us make clear that we do not think union and intersection
+contracts are fundamentally broken, that they can not be implemented correctly,
+or that they do not bear any value (quite the contrary). They may still make
+sense to have in a language, and some apparent difficulties in the
+implementation could be lifted some day. But as often, there are gaps between
+the theoretical framework, a proof-of-concept, a prototype, and the integration
+in an actual language. We hope that our attempt may serve as a cautionary tale,
+that for union and intersection contracts, these gaps are sometimes large in
+places.
+
 \bibliography{nickel}

### 2021-05-31 paper.tex
Small rewording conclusion

@@
 attempt at including them in a concrete language. We exposed what, in our
-opinion, are the most salient flaws and their consequences. Even in a pure
-language, union and intersection contracts break the equivalence between
-call-by-name and call-by-value, and consequently can prevent CSE-like
+opinion, are the most important barriers and their impact on the rest of the
+language. Even in a pure setting, union and intersection contracts break the
+equivalence between call-by-name and call-by-value, and can consequently prevent
 optimizations. There are discrepancies between the expected and actual semantics
@@
 implementation could be lifted some day. But as often, there are gaps between
-the theoretical framework, a proof-of-concept, a prototype, and the integration
-in an actual language. We hope that our attempt may serve as a cautionary tale,
-that for union and intersection contracts, these gaps are sometimes large in
-places.
+the theoretical foundation, a proof-of-concept, a prototype, and the integration
+in an actual language. We hope that our attempt may serve as a cautionary tale:
+for union and intersection contracts, these gaps might be quite large in places.
 

### 2021-06-01 paper.tex
Add an example of good error reporting

@@
 
+\lstset{
+  escapeinside={(*@}{@*)} % escape sequence to insert latex
+}
 \newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
@@
 contract will check at each call that the argument is an even number
-(pre-condition) and that the return value is a integer (post-condition). The
-\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
-the notion of blame which is crucial to good error reporting in presence of
-higher-order contracts and other composed contracts.
-
-\begin{figure}
+(pre-condition) and that the return value is a integer (post-condition).
+Contracts also introduce the notion of blame which is crucial to good error
+reporting in presence of higher-order contracts and other composed contracts.
+
+\begin{figure}[h]
   \begin{center}
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let Even = contracts.fromPred (fun val =>
@@
 
-In our opinion, one of the main appeals of Nickel is this ability to write a
-simple validation function for, say, a port number, and have it immediately
-available to construct other contracts by writing simple type-like expressions.
-Another example is given in Figure~\ref{fig:contract-composition}. The first
-line defines a port contract from a predicate and use it in the definition of a
+In a dynamic language, even more so in a lazy one (see the paragraph on Lazyness
+in Section~\ref{sec:feat-lang}), error messages may point to a location far
+removed from the offending code. Contracts help by making assumptions on the
+values passed at component boundaries explicit (components can be functions,
+modules, classes, and so on, depending on language).
+
+Figure~\ref{fig:contract-reporting-wo} and Figure~\ref{fig:contract-reporting-w}
+give an example of a situation where adding a simple contract annotation
+\nickel{Str -> Str} to a function greatly improves the error message . In the
+original program of Figure~\ref{fig:contract-reporting-wo}, the
+\nickel{catHosts} function takes a string argument, append it to a list that is
+then folded. However, the programmer forgot one parameter when calling to
+\nickel{makeHost}, thus passing a partially applied function instead of a string
+to \nickel{catHosts}. Because this argument is put in a list and folded, the
+error is raised later on and points to inside the code of \nickel{catHosts} and
+\nickel{makeHost}, which may be totally unknown to the caller.
+
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Source program}]
+let catHosts = fun last =>
+  let hosts = ["foo.com", "bar.org"] in
+  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in
+
+let makeHost = fun server ext => server ++ "." ++ ext in
+
+catHosts (makeHost "google")
+\end{lstlisting}
+
+\begin{lstlisting}[title={Error message},frame=none,numbers=none]
+error: Type error
+3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
+  |            ^^^
+  | This expression has type Fun,
+  | but Str was expected
+4 |
+5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
+  |              -----------------------
+  |              evaluated to this
+  = ++, 2nd argument
+\end{lstlisting}
+  \end{center}
+\caption{Error reporting without contract}
+\label{fig:contract-reporting-wo}
+\end{figure}
+
+By adding a simple annotation to \nickel{catHosts} in
+Figure~\ref{fig:contract-reporting-w}, the interpreter fails before even
+entering the body of catHosts and is able to deliver a clearer message: the
+first parameter of \nickel{catHosts} should be a string but the user provided
+the body of the function \nickel{makeHost}. A stack trace is also shown,
+including the faulty call to \nickel{catHost}.
+
+\begin{figure}[h]
+  \begin{center}
+\begin{lstlisting}[language=nickel,title={Added contract}]
+let catHosts | Str -> Str = fun last =>
+\end{lstlisting}
+
+\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none]
+error: Blame error: contract broken by the caller.
+  |  Str -> Str
+  |  --- expected type of the argument [(*@{\ldots}@*)]
+[(*@{\ldots}@*)]
+1 | let catHosts | Str -> Str = fun last =>
+  |                ^^^^^^^^^^ bound here
+[(*@{\ldots}@*)]
+6 | catHosts (makeHost "google")
+  | ---------------------------- (2) calling <func>
+\end{lstlisting}
+  \end{center}
+\caption{Error reporting with contract}
+\label{fig:contract-reporting-w}
+\end{figure}
+
+The examples of Figure~\ref{fig:contract-reporting-wo} and
+Figure~\ref{fig:contract-reporting-w} are about correctness. For validation, one
+of the main appeals of Nickel in our opinion is the ability to write a simple
+validation function for, say, a port number, and have it immediately available
+to construct other contracts by writing simple type-like expressions.  Another
+example is given in Figure~\ref{fig:contract-composition}. The first line
+defines a port contract from a predicate and use it in the definition of a
 larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
@@
 akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
-been defined in a similar way. \unsure{What's missing in this section is an
+been defined in a similar way. \resolved{What's missing in this section is an
 example of how contracts differ from standard dynamic type checking on some
@@
 
-\begin{figure}
+\begin{figure}[h]
   \begin{center}

### 2021-06-01 appendix-old.tex
Move the old appendix in a separate file

--- 2021-05-31 paper.tex
@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
-%\usepackage{amssymb}
-\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
-\usepackage{listings}
-\usepackage{xcolor}
-\usepackage{array}
-\usepackage[toc,page]{appendix}
-\usepackage{graphicx}
-\usepackage{varioref}
-\usepackage{verbatim,comment}
-\usepackage{longtable}
-\usepackage{tikz}
-\usepackage{url}
-\usepackage{hyperref}
-
-\definecolor{OliveGreen}{RGB}{128,128,0}
-%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%
-
-  % TOGGLE ME to turn off all the commentary:
-  \InputIfFileExists{no-editing-marks}{
-    \def\noeditingmarks{}
-  }
-
-  \usepackage{xargs}
-  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
-  % ^^ Need for pgfsyspdfmark apparently?
-  \ifx\noeditingmarks\undefined
-      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
-      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
-      \newenvironment{alt}{\color{red}}{}
-
-      \newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\info}[2][1=]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,#1]{#2}}
-      \newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
-      \newcommandx{\inconsistent}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\critical}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommand{\improvement}[1]{\todo[linecolor=pink,backgroundcolor=pink!25,bordercolor=pink]{#1}}
-      \newcommandx{\resolved}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}} % use this to mark a resolved question
-  \else
-  %    \newcommand{\Red}[1]{#1}
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{#1}
-      \newcommand{\note}[1]{}
-      \newenvironment{alt}{}{}
-  %    \renewcommand\todo[2]{}
-      \newcommand{\unsure}[2][1=]{}
-      \newcommand{\info}[2][1=]{}
-      \newcommand{\change}[2]{}
-      \newcommand{\inconsistent}[2]{}
-      \newcommand{\critical}[2]{}
-      \newcommand{\improvement}[1]{}
-      \newcommand{\resolved}[2]{}
-  \fi
-
-%%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%% LST listing confinguraiton %%%%%%%%%%%
-
-\lstdefinelanguage{nickel}{
-    keywords=[1]{
-      if,
-      then,
-      else,
-      switch,
-    },
-    keywords=[2]{
-      let,
-      rec,
-      fun,
-      in
-    },
-    keywords=[3]{
-      forall,
-    },
-    keywordsprefix=\#,
-    keywords=[4]{
-      true,
-      false,
-      null
-    },
-    keywords=[5]{
-      doc,
-      default,
-    },
-    sensitive=true, % keywords are case-sensitive
-    morecomment=[l]{//}, % l is for line comment
-    morecomment=**[is][\color{gray}]{\$}{\$},
-    morestring=[b]", % defines that strings are enclosed in double quotes
-    morestring=[s]{m\#"}{"\#m},
-    moredelim=[is][\color{gray}]{\$}{\$},
-    literate=
-        *{->}{$\rightarrow$}1
-        {...}{\ldots}1
-        {@&}{$\cap$}1
-        {@|}{$\cup$}1
-        {=>}{$\Rightarrow$}1
-        {>=}{$\geq$}1
-        {<=}{$\leq$}1
-} %
-
-% Copied from https://hal.inria.fr/hal-01140459/file/racket.inc.tex
-\lstdefinelanguage{racket} {
-  morekeywords=[1]{define, define-syntax, define-macro, lambda, define-stream, stream-lambda},
-  morekeywords=[2]{begin, call-with-current-continuation, call/cc,
-    call-with-input-file, call-with-output-file, case, cond,
-    do, else, for-each, if,
-    let*, let, let-syntax, letrec, letrec-syntax,
-    define-context, define-controller, Integer, Boolean, get, when-required, when-provided,
-    maybe_publish, require, submod, or/c, and/c, ->, \#\%module-begin,
-    always_publish, with-syntax, define-struct/contract, syntax-case,
-    define/contract,
-    let-values, let*-values,
-    module, provide,
-    and, or, not, delay, force,
-    \#`, \#',
-    \#lang, implement, begin-for-syntax, rename-out,
-    quasiquote, quote, unquote, unquote-splicing,
-    map, fold, syntax, syntax-rules, eval, environment, query },
-  morekeywords=[3]{import, export},
-  alsoletter={',`,-,/,>,<,\#,\%},
-  morecomment=[l]{;},
-%  literate={lambda}{{\lambdaup}}1, % lambda -- look at https://tex.stackexchange.com/questions/119879/math-symbols-in-tt-font
-  moredelim=**[is][\color{light-gray}]{<<@<<}{>>@>>},
-  moredelim=**[is][\itshape\color{OliveGreen}]{<<;<<}{>>;>>},
-  morecomment=[s]{\#|}{|\#},
-  sensitive=true,
-}
-
-\lstset{
-  language={nickel},
-  basicstyle=\small\ttfamily, % Global Code Style
-  captionpos=b, % Position of the Caption (t for top, b for bottom)
-  extendedchars=true, % Allows 256 instead of 128 ASCII characters
-  tabsize=2, % number of spaces indented when discovering a tab
-  columns=fixed, % make all characters equal width
-  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
-  showstringspaces=false, % lets spaces in strings appear as real spaces
-  breaklines=true, % wrap lines if they don't fit
-  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
-  frameround=tttt, % make the frame round at all four corners
-  framesep=4pt, % quarter circle size of the round corners
-  numbers=left, % show line numbers at the left
-  numberstyle=\tiny\ttfamily, % style of the line numbers
-  escapeinside={(*@}{@*)} % escape sequence to insert latex
-  % commentstyle=\color{dark-grey}, % style of comments
-  % keywordstyle=[1]\color{blue-portage}, % style of keywords
-  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
-  % keywordstyle=[3]\color{blue-portage}, % style of keywords
-  % keywordstyle=[4]\color{green-pea}, % style of keywords
-  % keywordstyle=[5]\color{pink-froly}, % style of keywords
-  % stringstyle=\color{blue-marguerite}, % style of strings
-}
-
-\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
-\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\begin{abstract}
-This paper reports an attempt to incorporate union and intersection types in
-Nickel, a configuration programming language with a gradual type system and
-first-class contracts. While the end result looks appealing, it turns out the
-concrete impact on both the design and the implementation of the language is
-unexpectedly high. We review the issues raised in particular by the
-implementation of contracts for unions and intersections, and why we think the
-cost ends up too high for the benefit. We conclude by suggesting different leads
-to represent and type unions.
-\end{abstract}
-
-\title{Contracts for unions and intersections are hard for Nickel}
-\author{Teodoro Freund}
-\affiliation{
-  \institution{Universidad de Buenos Aires}
-  \city{Buenos Aires}
-  \country{Argentina}
-}
-\email{tfreund95@gmail.com}
-
-\author{Yann Hamdaoui}
-\affiliation{
-  \institution{Tweag}
-  \city{Paris}
-  \country{France}
-}
-\email{yann.hamdaoui@tweag.io}
-\author{Arnaud Spiwack}
-\affiliation{
-  \institution{Tweag}
-  \city{Paris}
-  \country{France}
-}
-\email{arnaud.spiwack@tweag.io}
-
-\begin{document}
-
-\maketitle
-
-\unsure{TODO: CCS classification; Keywords}
-\section{Introduction}
-\label{sec:intro}
-\info{Goals of the paper: position paper: unions/intersections nice in
-theory problematic in practice. In particular non-orthogonal with the
-rest of the features of your language.}
-
-Build systems, package managers, operating systems, cloud infrastructure,
-continuous integration and web services\resolved{Also continuous integration}
-are examples of modern complex software systems which require an extensive and
-non-trivial configuration. To handle this complexity, the declarative approach
-has become more and more prominent: instead of applying a sequence of
-transformations to a system in order to reach a desired state, this state is
-described as static textual data. Some tool is then in charge of applying the
-necessary transformations. This approach makes the configuration persistent,
-reproducible, versionable and subject to static analysis. Text-based
-configurations are thus led to play an important role in critical aspects of
-software engineering, including security, availability, and maintainability.
-
-However, static text-based configurations show their limits beyond a certain
-size and complexity. Data cannot be transformed, combined, shared, nor validated
-(without resorting to separate schemas and tools).  This lack of expressivity
-results at best in boilerplate and duplication of information, but it can also
-make data invalid or inconsistent.  Correctly configuring a modern system is hard
-and failures may have substantial negative consequences.
-
-This is why the authors have been developing the Nickel
-language\cite{NickelRepo}. Nickel is a configuration programming language,
-meaning that a Nickel program evaluates to a value that must be
-straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
-lightweight and easy to write while providing support for writing correct code
-and generating valid data. Nickel originated as a rethink of the language of the
-Nix package manager, used to describe one of the biggest sets of packaged
-software in existence\cite{repology}.  Nickel targets build systems, cloud
-deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
-non-trivial configuration.
-
-We will use the syntax of Nickel to write most examples throughout this paper.
-Syntactically, Nickel looks like a simple ML language with records. \nickel{let
-id = value in exp} is used to bind a value to an identifier in an expression.
-Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
-are written as \nickel{[x1, ..., xn]} and records as \nickel{\{field1 = value1,
-..., fieldn = valuen\}}. The application of a contract \nickel{C} to the
-expression \nickel{exp} is written \nickel{exp | C}\footnote{The current
-    implementation requires a \nickel{#} character in front of custom contracts,
-    whose application would rather be \nickel{exp | #C}, but it is irrelevant
-and subject to change in the future.}. We will sometimes attach
-contracts directly to an identifier instead of the corresponding bound
-expression, that is: \nickel{let id | C = value in exp} is syntactic sugar for
-\nickel{let id = (value | C) in exp}.
-
-Some examples resort to constructions that are not currently available in Nickel
-for illustration purpose: we will use \nickel{A @& B} (resp. \nickel{A @| B})
-for the intersection (resp. union) of contracts (or types) \nickel{A} and
-\nickel{B}. \nickel{@|} and \nickel{@&} bind more loosely than arrows, meaning
-that \nickel{A -> B @& C -> D} is to be parsed as \nickel{(A -> B) @& (C -> D)}.
-
-\subsection*{Correct programs \& valid data}
-
-The end goal being data generation, Nickel is a dynamic, pure and functional
-language. Our goal is that writing a simple configuration in Nickel feels as
-easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
-configurations by being programmable, and by providing means of ensuring
-both correctness and validity:
-
-\begin{description}
-    \item[(correctness)] The program runs without errors. Errors may be type
-        errors, raised for example when evaluating the nonsensical division
-        \nickel{1 / "ftp"}. More generally, they can be any the violation of a
-        pre- or post-condition. When such an error cannot be prevented, we want
-        at least early failures and relevant error messages.\todo{make this
-        definition more precise}
-    \item[(validity)] The program generates valid data, i.e. data that conforms
-        to a schema (imposed by the system consuming the configuration). For
-        example, when configuring a web service, the schema may require the
-        presence of a field \lstinline+port+ that holds a valid port number.
-\end{description}
-
-A traditional answer is to use static types, and in particular static types for
-correctness. However, Nickel programs are of a special kind: they are usually
-simple, terminating programs that run on fixed inputs. Correctness errors will
-show up at evaluation anyway, or will not matter (dead code). With respect to
-validation, traditional types are too limited. Validation commonly involves
-checking requirements such as being a valid port number or a valid URL which
-are, while not technically impossible, really hard to check statically.
-
-In a dynamic language, it is natural and easier to validate data at runtime.
-This underlies the approach of dynamic typing. But dynamically typed languages
-usually only check the same simple predicates as static types (being a number,
-being a string, etc.), and solely for data that are arguments to a primitive
-operation. For validation, one needs to check arbitrary predicates on arbitrary
-values. This is provided in a basic form by assertions, available in most
-languages out there. An assertion is an expression that verifies that a
-predicate holds, or abort the program otherwise.  While assertions are fine for
-simple sanity checks, their use can incur verbosity, they are clumsy to compose
-and to reuse, and they provide limited context attached to an error. It turns
-out there is a better tool than plain assertions: contracts.
-
-\subsection*{Contracts}
-
-Assertions are usually mainly used for correctness by enforcing pre- and
-post-conditions of functions at runtime. In their foundational
-paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
-principled approach to run-time assertion checking called contracts. As opposed
-to plain assertions, contracts can be composed naturally into more complex ones.
-Nickel features such a contract system. For example, contracts nicely support
-higher-order functions. If the user has written an \nickel{Even} contract for
-numbers, they can use it right away to form a function contract \nickel{Even ->
-Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
-contract will check at each call that the argument is an even number
-(pre-condition) and that the return value is a integer (post-condition). The
-\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
-the notion of blame which is crucial to good error reporting in presence of
-higher-order contracts and other composed contracts.
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let Even = contracts.fromPred (fun val =>
-  builtins.isNum val
-  && val % 2 == 0) in
-let Int = contracts.fromPred (fun val =>
-  builtins.isNum val
-  && val % 1 == 0) in
-{divBy2 | Even -> Int = fun x => x/2}
-\end{lstlisting}
-  \end{center}
-\caption{Higher-order contract}
-\label{fig:contract-higher-order}
-\end{figure}
-
-In our opinion, one of the main appeals of Nickel is this ability to write a
-simple validation function for, say, a port number, and have it immediately
-available to construct other contracts by writing simple type-like expressions.
-Another example is given in Figure~\ref{fig:contract-composition}. The first
-line defines a port contract from a predicate and use it in the definition of a
-larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
-contract checks that a value is a record with the following shape: it must have
-a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
-\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
-a \emph{schema} used to validate the final data, which is exactly what we were
-looking for in the previous section. Not only it plays the role of a schema, but
-it also \emph{looks} like one, as opposed to the corresponding plain validation
-function. The \lstinline+|+ operator entails that a value respects a contract,
-akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
-been defined in a similar way. \unsure{What's missing in this section is an
-    example of how contracts differ from standard dynamic type checking on some
-function. Namely, the quality of error messages.}
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let Port = contracts.fromPred (fun value =>
-  builtins.isNum value && value % 1 == 0
-  && value >= 0 && value <= 65353) in
-let Configuration = {
-  host | Host,
-  port | Port,
-} in
-{host = "nixos.org", port = 80}
-| Configuration
-\end{lstlisting}
-  \end{center}
-\caption{Contract composition}
-\label{fig:contract-composition}
-\end{figure}
-
-Contracts have another perk: they correspond to the type casts of gradual
-typing, a discipline for mixing static and dynamic typing in the same language.
-As such, they are a precious ingredient for handling the interaction between 
-statically typed and dynamically typed code. While we focused on the writing of
-configuration, Nickel is a proper programming language that is also designed to
-write generic and reusable library code. There, the typing trade-off is
-different, and static typing is more adapted. Nickel has a gradual type system,
-which is yet another argument for having contracts. Let us end this section by
-stressing the following points:
-\begin{itemize}
-    \item \textbf{Even in a purely dynamically typed language, contracts are an
-        appealing feature for validation and correctness}. The gradual typing
-        part enters the general picture but is out of the scope of this paper.
-    \item \textbf{User-defined contracts are essential for validation}. Some
-        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
-        extensively discussed later in the paper) do not feature user-defined
-        contracts.  Rather, a fixed set of built-in contracts such as
-        \nickel{Num} or \nickel{Str} are provided, very much like base
-        types. The absence of user-defined contracts is understandable when
-        using contracts as a dynamic typing discipline for correctness. But a
-        system without user-defined contracts is limited in the same way as
-        static typing with respect to validation.
-\end{itemize}
-
-\subsection*{Structure of this paper}
-Section~\ref{sec:intro} introduced the perspective of designing a configuration
-language. We presented the concept of contracts and detailed why contracts are
-particularly fitted for a configuration language, and more generally for data
-validation or program correctness in a dynamically typed language. The rest of
-the paper is structured as follows:
-\begin{itemize}
-    \item Section~\ref{sec:feat-lang} outlines various common features of
-        programming languages that may interact in a bad way with union and
-        intersection contracts.
-    \item Section~\ref{sec:union-inter} introduces unions and intersections and
-        lists the possible motivations for adding them to a language.
-    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
-        encounter when trying to implement union and intersection contracts in
-        practice.
-    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
-        existing attempts to solve said difficulties.
-    \item Section~\ref{sec:related-work} provides some related work.
-\end{itemize}
-
-Section~\ref{sec:conclusion} concludes the paper.
-
-\newpage
-
-\section{Features of languages}
-\label{sec:feat-lang}
-\info{Zoology of various features that we will eventually show
-  conflict with this or that property or implementation of union and
-  intersection. Including user-define contracts.}
-
-Programming language research no longer lives on the abstract world
-of ideal surfaces.
-Every addition needs to be compatible with years of research
-and development, and this is hard.
-Success was unavoidable.
-
-In this section we briefly introduce some aspects of real world
-programming languages that will cause trouble while interacting
-with union and intersection dynamic checks.
-
-\subsection*{Flat contracts}
-\label{sec:flat-contracts}
-
-\unsure{Should we define contracts prior to this?}
-User defined contracts, also called flat contracts, are a particular
-kind of contract that behaves similar to an assertion, and whose
-main point is to apply some arbitrary predicate (as arbitrary as the
-host language allows it) to a value; if the predicate evaluates to
-true, the contract passes, otherwise it fails.
-
-For instance, the contract \nickel{Even} shown on Figure
-\ref{fig:contract-higher-order} is a flat contract, defined
-by taking the value that needs to be checked, and validating
-that is both a number and that the remainder of dividing it by 2
-is 0.
-And then, as seen on the same figure, we can use it to check that a function always
-returns an \nickel{Int} when applied to an \nickel{Even}.
-
-% \begin{figure}[h]
-% \begin{center}
-% \begin{lstlisting}[language=nickel]
-% let f = fun n => n * 2 @ (Integer -> even)
-% \end{lstlisting}
-%   \end{center}
-% \caption{A function wrapped with a contract stating that it always returns
-% an even number}
-% \label{fig:num-to-even}
-% \end{figure}
-
-Or even more interesting, we could come up with a contract that checks that a function
-maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
-by applying the function and checking its results,
-and then use it as a contract for the \nickel{not} function, as can be seen on Figure
-\ref{fig:checked-not}.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let NotC = contracts.fromPred (
-  fun f => builtins.isFun f && (f true == false) && (f false == true)
-) in
-let checkedNot = not | #NotC
-\end{lstlisting}
-\end{center}
-\caption{A contract that checks the not function behaves as expected}
-\label{fig:checked-not}
-\end{figure}
-
-This type of contracts are present on many different languages,
-for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
-of the Design by Contract ideals, allows to assert
-these kinds of expression as pre- and post-conditions on
-functions and as invariants on classes\cite{EiffelDesignByContract}.
-\resolved{I (Arnaud) believe Eiffel has this feature as well,
-  doesn't it? (Teo) You're right},
-
-% A somewhat newer language, the D programming
-% language, allows, among other things, to have invariants on
-% classes
-% written with flat contracts, that are
-% dinamically checked at the frontiers of every public method \cite{DLangContracts}.
-
-Finally, the Racket programming language also has a system to work with
-contracts, powerful enough to define flat contracts
-\footnote{And a gradual typing system
-built on top of that, but that goes beyond the scope of this
-work.}
-, and
-to compose them with other kinds of dynamic checks,
-like higher order contracts or a lightweight take on union
-and intersection contracts\cite{RacketContracts}.
-
-Even if flat contracts may seem like too permissive, the ability to make
-non trivial checks on values (that is, to validate their output) is expected
-by programmers on dynamic languages, where testing and runtime
-assertions are extremely useful tools for safe programming, and
-programmers are accustomed to express complex predicates
-over values.
-
-\subsection*{Code manipulation (optimizations)}
-\label{sec:optimizations}
-\info{Present inlining and CSE as two major code
-optimizations.
-Purity and immutability as two important factors.}
-
-The performance of modern programs heavily relies on the optimizations performed
-by the compiler or the interpreter. Even more so for functional languages, whose
-execution model is often far removed from actual processors, causing naive
-execution to exhibit unacceptable slowdowns.
-
-One crucial optimization is
-inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
-to make heavy use of functions, and a function call is a costly operation from
-the point of view of low-level execution. Inlining eliminates a function
-application by directly substituting the function for its definition at compile
-time (or before execution, for interpreted language). This is especially
-efficient for small functions that are called repeatedly.
-
-\info{I think it's better to use [h] on figures, we can discuss it}
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let elem = fun elt =>
-  lists.any (fun x => x == elt) in
-
-let subList = fun l1 l2 =>
-  elem (lists.head l1) l2
-  && subList (list.tail l1) l2
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let subList = fun l1 l2 =>
-  lists.any (fun x => x == (lists.head l1)) l2
-  && subList (list.tail l1) l2
-\end{lstlisting}
-  \end{center}
-\caption{Inlining}
-\label{fig:optimizations-inlining-ex}
-\end{figure}
-
-While inlining expands an expression by substituting a definition for its value,
-an opposite transformation is sometimes beneficial, for example when a composite
-expression is repeated several times. This causes the same expression to be
-wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
-consists in introducing a new identifier for this expression and using the
-identifier in place of the original occurrences
-(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let elemAtOrLast = fun index list =>
-  if index > lists.length list - 1 then
-    lists.elemAt (lists.length list - 1) list
-  else
-    lists.elemAt index list
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let elemAtOrLast = fun index list =>
-  let l = lists.length list - 1 in
-  if index > l then
-    lists.elemAt l list
-  else
-    lists.elemAt index list
-\end{lstlisting}
-  \end{center}
-\caption{Common subexpression elimination}
-\label{fig:optimizations-cse-ex}
-\end{figure}
-
-Beyond CSE, optimizations such as loop-invariant code motion or
-let-floating~\cite{letFloating} apply the same principle of extracting out an invariant
-\unsure{Was this the inteded citation?}
-expression to avoid its recomputation respectively across loop iterations and
-function calls.
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
-let f = fun x => g y (x + 1)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
-let g' = g y
-let f' = fun x => g' (x + 1)
-\end{lstlisting}
-  \end{center}
-\caption{Let-floating}
-\label{fig:optimizations-let-floating-ex}
-\end{figure}
-
-For example, take the code of Figure~\ref{fig:optimizations-let-floating-ex}.
-The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
-called. This may be costly, in particular in the presence of contracts: if
-\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
-on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
-to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
-Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
-let-floating transformation could indeed do (given \lstinline+g+ is pure, as
-explained below).
-
-The soundness of all these optimizations is tied to the invariance of the
-semantics of programs with respect to specific substitutions. Inlining requires
-that one can replace the application of a function by its body, which is
-basically $\beta$-reduction: as long as the arguments are evaluated following
-the language's strategy, it is a valid transformation. On the other hand,
-a CSE-like transformation on a term $M$ requires an expansion property of the form
-
-\begin{equation}\label{eq:cbn-expansion}
-M[x/N] \simeq let~x~=~N~in~M
-\end{equation}
-
-Here, $M[x/N]$ is the substitution of $x$ for $N$ in the term $M$.
-\ref{eq:cbn-expansion} has an evident call-by-name flavour, and consequently
-fails in general for a call-by-value language with side effects.  Indeed, define
-the following function $f$:
-
-\begin{lstlisting}[language=Nickel]
-let f x = print "hi";(x+1) in 
-\end{lstlisting}
-
-Then:
-
-\begin{lstlisting}[language=Nickel]
-(f 1,f 1) (*@ $\not \simeq$ @*) let y = f 1 in (y,y)
-\end{lstlisting}
-
-In the example above, the first term prints \nickel{"hi"} two times while the
-second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
-\emph{pure} terms, that are terms without side-effects. In a pure language, or a
-language with effect tracking like Haskell\todo{Cite a different language, or
-remove citation}, pure terms can be safely identified for applying
-optimizations.
-
-Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
-unions and intersections can make \ref{eq:cbn-expansion} \resolved{it} unsound
-\emph{even in a pure setting}, making the optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
-
-\unsure{Should we tease already here that unions and intersections break this
-even for pure terms?}
-
-\subsection*{Lazyness}
-\todo{Lazyness does not appear on sec 4, if we don't put somethign there, I'd remove it}
-Evaluation strategies relates to the evaluation of the expression bound to
-identifiers. Expressions are bound by assignments (e.g. let bindings in
-Nickel), or by function applications (arguments are bound to parameters). We
-will use Figure~\ref{} as a working example.
-
-The eager strategy consists in evaluating right away a bound value before
-storing it in an identifier. If an identifier is not used, this leads to
-unnecessary work: in Figure~\ref{}, the expression \nickel{fibonacci 100} would
-be computed although it is never used. The lazy strategy, on the other hand,
-only evaluates expression if they are needed. In the lazy model, bound
-expressions are stored as in their current unevaluated form, also called a
-\emph{thunk}. Thunks are only evaluated at their first usage, and their value is
-memoized for future usages.
-
-The eager strategy has been chosen in the vast majority of languages: while
-theoretically less efficient, it is simpler to implement and more predictable
-for the programmer. Also, lazy evaluation can occur performance overhead and
-memory leaks in unexpected ways. While eager evaluation is the default, some
-languages however offer a form of lazy evaluation, either as an opt-in option
-(e.g. Ocaml and ReScript), or by default for specific primitives types such as
-list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
-(and its web avatar PureScript). Although rare in general purpose languages,
-lazy evaluation is more widespread in the configuration space: the configuration
-languages Dhall, CUE and Jsonnet also implements full lazy evaluation. So are
-the Nix language, which inspired Nickel, and Nickel itself.
-
-\newpage
-
-\section{Union \& intersection}
-\label{sec:union-inter}
-\info{What are they, what are they used for.}
-\todo{Since nickel uses | for contract application, I'd move unions
-to some other symbol.
-I think the best option is to move to $\cap / \cup$}
-
-\subsection*{Unions}
-Unions are fundamental building blocks of programs. A union holds a value whose
-shape is known to be in a fixed set of alternatives: for example, a value that
-can be either a boolean or a string. They correspond logically to disjunctions.
-Used in combination with products, they can represent arbitrary tree-like
-structures, allowing to model a large class of data
-(\ref{fig:union-adt-ex}).
-
-\change{These next few labels need to be inside figures,
-they are not working}
-\label{fig:union-adt-ex}
-\begin{lstlisting}[title={Tree-like data structure}]
-type BinaryTree =
-Integer @| (BinaryTree, BinaryTree)
-\end{lstlisting}
-
-In addition to internal program representations, unions are also useful to model
-external data:
-\begin{itemize}
-    \item To represent nullable values (\ref{fig:union-nullable-ex}).
-    \item For plain enumerations (\ref{fig:union-enum-ex}).
-    \item To support different representations of the same data
-        (\ref{fig:union-alt-ex}).
-\end{itemize}
-All of which are common in configurations.
-
-\label{fig:union-nullable-ex}
-\begin{lstlisting}[title={Nullable values}]
-optionalValue : Str @| Null
-\end{lstlisting}
-
-\label{fig:union-enum-ex}
-\begin{lstlisting}[title={Plain enumeration}]
-protocol : `Http @| `Ftp @| `Sftp
-\end{lstlisting}
-
-\label{fig:union-alt-ex}
-\begin{lstlisting}[title={Alternative representations}]
-person : {name : Str, age : Int}
-  @| {name : Str, dateOfBirth : Str}
-\end{lstlisting}
-
-In statically typed languages, and more specifically in functional programming
-languages, sums (also called tagged unions, or discriminated unions) are usually
-preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
-syntactically separated by an associated discriminating label. Well-known
-programming languages with first-class tagged unions include Haskell, OCaml,
-Rust, Scala, PureScript, Elm, Swift and many more.
-
-On the other hand, in dynamically typed languages, it is customary to directly
-store different kinds of data in one variable without further ceremony. The
-runtime system needs to attach type tags to values anyway, which can usually be
-observed using functions like JavaScript's \verb+typeof+. Using such bare
-unions\resolved{If you can observe with typeof, then it's pretty much a tagged
-union, though the tag is implicit} has already been an idiom in dynamically
-typed language for a long time. Being able to type and handle these unions is an
-important aspect of the migration to a gradual type system. Indeed, a lot of
-\unsure{This is the first time discussing gradual type systems, maybe it should
-be introduced earlier, in particular how they relate to contracts
-(I think is mentioned on the appendix)}
-gradually typed languages either existed originally as dynamically typed
-language or aim at backward compatibility with a dynamically typed base
-language. This is why despite the presence of a static type system, gradually
-typed languages tend to favor bare unions types which is the typed counterpart
-of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
-
-In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. However, Nickel is a configuration
-language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
-untagged unions, there is a strong incentive to support untagged unions at the
-type level, or at least, at the contract level.
-
-\subsection*{Intersections}
-
-Intersections are less prevalent. An intersection is a value that belongs to
-two types at the same time. A simple example is a number literal \nickel{1}
-which could be accepted as both an \nickel{Int} and a \nickel{Float} without
-having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
-this example, the intersection \nickel{Int @& Float} can be used as a subtype of
-both \nickel{Int} and \nickel{Float}.
-
-\label{fig:intersection-subytping-ex}
-\begin{lstlisting}[title={Value subtyping}]
-1 : Int @& Float
-\end{lstlisting}
-
-Intersections of higher-order types corresponding to overloading
-(Figure~\ref{fig:intersection-overloading-ex}).
-
-\label{fig:intersection-overloading-ex}
-\begin{lstlisting}[title={Overloading}]
-incr: (Int -> Int) @& (Float -> Float) =
-  fun x => x + 1
-\end{lstlisting}
-
-They can also be used to combine constraints (such as traits or interfaces), as
-done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
-
-\label{fig:intersection-mixins-ex}
-\begin{lstlisting}[language=Scala,title={Structural mixins}]
-Comparator: {cmp: T -> T -> Bool}
-Shower: {show: T -> String}
-
-showSmaller : (dict: Comparator & Shower) -> T -> T -> String
-\end{lstlisting}
-
-Finally, intersection naturally show up as the dual of unions. Thus, in a system
-featuring unions, a union in a negative positions is a form of intersection
-(Figure~\ref{fig:intersection-dual}).
-
-\label{fig:intersection-dual}
-\begin{lstlisting}[title={Intersection as the dual of unions}]
-f : (Num @| Str) -> Bool
-(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
-\end{lstlisting}
-
-\todo{Either here or the next section, show examples of unintuitive behaviour
-of U and I. For example, I is not jos composition of contracts.
-And union has to be treated as any one of the options}
-% Things to settle: how to motivate them, how to compare them to ADTs. Purely
-% dynamic approach, or typed approach too?
-
-\newpage
-
-\section{Issues coming from the semantics}
-\label{sec:issues-sem}
-\info{What goes wrong or difficult when trying to implement union and
-  intersection literally}
-
-Type systems featuring intersection types are hard in a fundamental way,
-from their discovery, they've been studied as a form to characterize
-lambda-calculus terms normalizability; which means that typechecking a
-term on a system with intersection types is undecidable,
-similar to how the halting problem is undecidable\todo{reference}.
-
-Usually, when thinking about programming languages, there's an intuition
-that static checks (like typechecking) are more difficult and less powerful
-than dynamic checks (like testing).
-For instance, it is very easy to check if,
-in a given execution of a program, a variable holds an integer value; while it
-may become more challenging to check if a variable will always hold an integer value,
-no matter the particular execution.
-
-As we intend to show on this paper, unions and intersections are not only
-statically hard to check, but they are also dynamically hard:
-\emph{they are fundamentally hard}.
-In this section, we present intuitively the main challenges that arise
-when attempting to develop a runtime check system for unions and intersections.
-
-\subsection*{Union contracts are hard to break}
-
-When programming with dynamic checks it is assumed that lack of
-failure does not mean that the program works perfectly.
-
-For instance, imagine having a function wrapped in a contract
-that checks that it maps positive numbers to positive numbers,
-and using it in different ways,
-like the program on Figure \ref{fig:pos-to-pos}.
-Clearly, there is an error on this program, but if the first call
-to \nickel{f} that gets executed is \nickel{f 10}, the error will not
-be found at that point, and the first operand of the addition would be
-evaluated to \nickel{3}.
-If, later on, \nickel{f 5} gets executed, the contract would get violated,
-since \nickel{f} gets called with a positive value, but returns a negative
-one (\nickel{-2}).
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f | Positive -> Positive
-      = fun x => x - 7 in
-(f 10) + (f 5)
-\end{lstlisting}
-\caption{Wrong positive to positive function}
-\label{fig:pos-to-pos}
-\end{figure}
-
-Even if the program performed two different applications of \nickel{f}
-before finding the contract violation, the first one is not actually
-needed, and executing only \nickel{f 5} is enough to find the problem.
-
-Having a witness of size 1 for every contract failure is a really nice to have property
-since it means, for example, that every time that a contract is violated,
-the stack trace is enough to understand how it was violated, and to potentially
-come up with a minimal example that breaks the contract.
-However, when unions get introduced, this property is lost.
-
-Consider now the program presented in Figure \ref{fig:wrong-union-function},
-this program (and in particular, \nickel{f}'s contract) still presents a problem,
-since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
-nor does it map positive numbers to non-positive numbers, so it won't be a valid
-value for the union of those contracts.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f | (Positive -> Positive)
-        @| (Positive -> NonPositive)
-      = fun x => x - 7 in
-(f 10) + (f 5)
-\end{lstlisting}
-\caption{Still wrong positive to positive or positive to non-positive function}
-\label{fig:wrong-union-function}
-\end{figure}
-
-As before, the program presents enough function calls to find that issue
-and violate the contract,
-since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
-and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
-But, what makes this case completely different from the one presented before, is that
-both function calls are needed, and, what is more, there is not a way to
-use the wrapped value \nickel{f} only once that would find a violation to its contract;
-at least two are needed.
-
-This seemingly simple consequence generates major issues when considered in real
-life languages.
-For example, remember the different mechanisms for code manipulation presented
-in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
-both of these rely on the fact that substituting a pure function call by its body,
-or joining equal pure expressions and sharing their result, is a semantically valid
-transformation.
-However, consider the examples presented at Figure \ref{fig:optimized-programs},
-the first one is an inlined version of the second one (where function
-\nickel{g} has been inlined); while the second is similar to the first, but
-the common subexpression \nickel{f 1} has been eliminated.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Inlined]
-let f | Num -> (Bool -> Num @| Bool -> Str)
-      = fun x y => if y then x else "False"
-in (f 1 true, f 1 false)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
-let f = (fun x y => if y then x else "False") | Num -> (Bool -> Num @| Bool -> Str) in
-let g = f 1 in
-(g true, g false)
-\end{lstlisting}
-\caption{Equivalent programs, with inlining or CSE applied}
-\label{fig:optimized-programs}
-\end{figure}
-
-Unexpectedly, these two programs behave very differently, the first one will
-return \nickel{(1, "False")}, without failing.  While the second one will fail,
-since the single application of \nickel{f 1}, applied to both \nickel{true} and
-\nickel{false} will be able to find the error on the union contract
-\nickel{(Bool -> Num) @| (Bool -> Str)}.
-
-A valid argument against this point would be stating that further work should be done
-on optimizers, so to not modify these tricky cases, even if this is true,
-programmers also modify their code on a daily basis, and if moving from the first
-code snippet at Figure \ref{fig:optimized-programs} to the second one changes
-the behaviour of a program, we argue that union contracts not only go against the
-implementation of languages, but also against their usability.
-
-Summing it up, wrapping a value with a union of higher order contracts means that
-that value now has a state, that may get modified silently at every use site.
-Both reviewed papers suffer from this defect since it is the semantics
-of unions that interact badly with code manipulation operations,
-not a particular implementation.
-This addition of global state complicates both the work of the language
-implementor, as well as the language user, since it adds non trivial
-side effects to your code.
-
-\subsection*{Flat contracts do not intersect}
-\label{sec:flat-and-inter}
-
-Stop for a second and think how you could implement an intersection
-contract. The first thought you may have is "just apply
-both contracts", as shown on the Implementation snippet
-from Figure \ref{fig:inter-contracts}.
-Of course, if it were that simple we wouldn't be writing this paper.
-This idea starts to show problems when you first combine it with
-higher order contracts, since it would mean not complying with
-at least one of these contracts negatively would entail breaking the
-whole thing, as shown on the second example on the same Figure,
-that would fail.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
-x | A @& B (*@ $\simeq$ @*) (x | A) | B
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Problem]
-let g | Num -> Num @& Str -> Str
-      = fun x => x in
-g 1
-\end{lstlisting}
-\caption{Implementations of intersection contracts: first iteration}
-\label{fig:inter-contracts}
-\end{figure}
-
-Not all hope is lost, we could pass some extra shared state
-to each one of the sub contracts and let them only fail if the other
-side of the contract has failed as well, as shown in Figure
-\ref{fig:inter-contracts-2}, where the shared state is
-represented by a label \nickel{l}.
-This idea is still quite reasonable, for instance, when executing
-\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
-that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
-something like \nickel{1 | Str}.
-Of course, this fails, but before raising a violation, this contract could
-check if the other side of the intersection (\nickel{Num -> Num}) has
-also tried to raise negative blame on the same parameter ()\nickel{1}), since
-it hasn't, the contract would return normally, and would mark itself as failed.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
-x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
-\end{lstlisting}
-\caption{Implementations of intersection contracts: second iteration}
-\label{fig:inter-contracts-2}
-\end{figure}
-
-On the other hand, if you were to do something like
-\nickel{g true}, that would fail also negatively the contract
-\nickel{Str -> Str} and, when checking if \nickel{true}
-is a correct parameter for \nickel{Num -> Num} it would check
-if the other side of the intersection has failed, and since it
-has, a violation would be raised and the whole program would
-correctly fail.
-
-This idea, which is an oversimplification of the ideas
-presented by Williams, Morris, and Wadler, seems to be spot on.
-However, it presents a major problem, it's not compatible with flat contracts,
-as presented in Section \ref{sec:flat-contracts}.
-Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
-implementation discussed previously were to be used, we may end up with
-something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
-\nickel{l} represents the shared state.
-At this point, the intuitive thing to do is to check the contract \nickel{C}, by
-checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
-since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
-fails negatively.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let C = fun f => (f 0) == 0 in
-let g | (String -> String) @& C
-      = (fun x => x) 
-\end{lstlisting}
-\caption{Intersection and flat contracts}
-\label{fig:inter-flat-contracts}
-\end{figure}
-
-But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
-\nickel{g} is never applied to \nickel{0}, in fact, it is never applied
-to anything.
-Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
-failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
-should not have been used inside the evaluation of \nickel{C}, which
-adds an extra layer of complexity: contracts should be guarded, and execution
-depends on the shape of the surrounding evaluation context.
-
-\subsection*{Union and intersection are not a silver bullet}
-
-Before wrapping up this section, we would like to show that maybe, union and intersection
-contracts are not exactly what we would want them to be.
-As stated throughout this work, programmers coming from dynamic languages
-are accustomed to perform safety checks (data validation) with
-great amounts of flexibility and the whole power of their
-language of choice at their disposal.
-
-For instance, a Python programmer may have no problem in checking
-at runtime, that a given variable is an integer if a given flag is
-not set, as shown in Figure \ref{fig:python-assertion};
-while a Haskell programmer might close her eyes and leave the room
-instantly.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=python]
-result = 42 if flag else "Disabled"
-...
-assert (
-  (not flag) 
-  or isinstance(result, int)
-)
-\end{lstlisting}
-\caption{Asserting at runtime the type of a variable}
-\label{fig:python-assertion}
-\end{figure}
-
-These expectations are not really met by how unions and intersections
-contracts are supposed to work.
-For example, consider the Nickel code shown on Figure
-\ref{fig:intersection-distrbution}, at first sight, this code should
-complete execution successfully, returning \nickel{(1, 1)},
-however, under both the Keil and Thiemann, as well as the
-Williams, Morris, and Wadler works, this program will fail, with a violation
-of the contract imposed to \nickel{g}.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f = fun x y => x in
-let g = f | (Num -> Num -> Num)
-            @& (Num -> Bool -> Num) in
-let h = g 1 in
-(h 1, h true)
-\end{lstlisting}
-\caption{Intersection contracts don't distribute}
-\label{fig:intersection-distrbution}
-\end{figure}
-
-The reason this happens is that there is a fundamental property of how
-intersection contracts behave: the choice of which branch to choose is made
-exactly when the intersection is opened up, and not delayed as much as possible,
-as one would like.  This means, that the contract presented on Figure
-\ref{fig:intersection-distrbution}, \nickel{Num -> Num -> Num @& Num
--> Bool -> Num}, will not behave the same way that \nickel{Num ->
-(Num -> Num @& Bool -> Num)}, when, with an intuitive view of how
-intersections should behave, these two expressions should be logically
-equivalent.
-
-It is worth noting, that intersection types do have this property,
-since subtyping would allow to correctly type check a term like
-\nickel{g 1} as both \nickel{Num -> Num} as well as
-\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
-can be given the type \nickel{Num -> Num @& Bool -> Num}.
-
-\newpage
-
-\section{Issues in concrete restrictions from the literature}
-\label{sec:issues-literature}
-\info{The various papers and the tradeoffs they make}
-
-The existing work that previously researched and implemented
-union and intersection
-as dynamic checks have, in our eyes, different shortcomings
-that makes them suboptimal solutions.
-In this section, we informally present those issues,
-starting with the foundational paper by Keil and Thiemann
-\cite{KeilThiemannUnionIntersection}, then following to a recent
-work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
-finally going over some similar contracts available in the Racket
-programming language.
-
-\subsection*{Keil and Thiemann}
-\label{sec:keil-thiemann}
-\unsure{Subsection title?}
-Keil and Thiemann noticed that, since union contracts should share information between
-different application contexts, while intersection contracts should behave
-independently between each application context, the system becomes much
-simpler if every contract gets transformed into a union of intersections
-of other contracts, by using the common De Morgan laws,
-and then eagerly opening up unions, but postponing intersection
-contracts as much as possible.
-\todo{This assumes the difficulty has been previously explained}
-
-For instance, if someone were to wrap a value \nickel{x}
-in a contract as shown on the Original code snippet on Figure
-\ref{fig:kt-contract-transf}, then, before actually executing that
-contract, it would get transformed, and intersections would
-be pushed inside the unions, as can be seen on the Transformed
-code snippet on the same figure.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Original]
-x | A @& (B @| C)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Transformed]
-x | (A @& B) | (A @& C)
-\end{lstlisting}
-\caption{Contract transformation performed by the Keil and
-Thiemann work}
-\label{fig:kt-contract-transf}
-\end{figure}
-
-This straightforward idea makes composing the Keil and Thiemann
-solution with other existing systems more difficult; in the words
-of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
-``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
-and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition.''\cite{RootCauseOfBlame}.
-\info{(Yann) Does
-it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
-development causes to recheck the contract $A$ that has been duplicated?}
-And it also could harm performance, or at least complicate an implementation,
-since some sharing mechanism should be used to avoid checking the same contract,
-over the same value, multiple times, exemplified by contract \nickel{A} on
-Figure \ref{fig:kt-contract-transf}, that gets duplicated.
-\unsure{Arnaud says: this whole
-section is too abstract give concrete code and explain what goes
-wrong. It's fine to give code in Nickel syntax}
-
-Another difficult issue Keil and Thiemann were able to solve, is the
-fact that flat contracts and intersection do not usually play nice
-together, as shown on Section \ref{sec:flat-and-inter}.
-They provide two different sets of rules for their calculus,
-the first one is non deterministic, and its purpose is to provide
-semantics that are simple to understand.
-With these rules, the interaction between flat contracts and
-intersections becomes much simpler, since they are able to check
-both sides of an intersection on separate threads.
-This allows them to be sure that contracts from one branch of the intersection
-won't get blamed while executing the other branch, and vice versa.
-
-As an example, consider the programs shown on Figure
-\ref{fig:kt-non-deterministic}, if the Original program
-gets executed, it will spawn two instances of said program,
-one with the first branch of the intersection (\nickel{3 | Num}),
-and the other one with the second branch (\nickel{3 | Positive}),
-and these will get joined once they reach an equal point, in this case
-it will be \nickel{3}.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Original]
-3 | Num @& Positive
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=First Program]
-3 | Num
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Second Program]
-3 | Positive
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Joined Program]
-3
-\end{lstlisting}
-\caption{Non deterministic intersection contract opening}
-\label{fig:kt-non-deterministic}
-\end{figure}
-
-The second set of rules is deterministic, and it is intended
-to be simpler to implement.
-However, this second set needs to perform operations
-on the evaluation context, to decide whether or not to evaluate
-a given contract.
-\footnote{You can check more on this topic on Section 5.1 of
-\cite{KeilThiemannUnionIntersection}}
-For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
-when using the value \nickel{(fun x => x) | Str -> Str)}
-inside the execution of contract \nickel{C}, and applying it to \nickel{0},
-the contract runtime would check the context of execution and it would
-deduce that contract \nickel{Str -> Str} does not apply in this case,
-skipping it.
-
-We see two problems here, a minor one is the fact that the two
-different semantics defined for the calculus are really far apart
-one from the other, both on the strategies used for contract resolution,
-as well as on the complexity levels.
-\resolved{(Yann) Is it really a
-    problem per se? It is common thing to do (a nice declarative system
-    for proofs and an algorithmic one for implementation, proved equivalent).
-    Maybe the point is that the algorithmic system is way more complex than the
-declarative one
-(Teo) reworded}.
-A second, major one,
-is the dependency added over the evaluation context, since, again,
-it makes composing this, with other features desired on programming languages,
-difficult, or even impossible.
-\resolved{(Yann) I think we need to substantiate this claim. If this
-context dependency explains why CSE is invalid, maybe have a little example?
-(Teo) I removed it, I don't think there's an example, mainly since
-this context dependency is not part of the language. I do think it complicates
-implementation of said feature}
-\todo{(Yann) TODO: see if there's a low hanging complexity
-    argument we can make (like it's quadratic in the number of open contracts or
-whatever)}
-
-\subsection*{Williams, Morris, and Wadler}
-\label{sec:will-morr-wadl}
-\unsure{Subsection title?}
-
-Williams, Morris, and Wadler, take a different approach on flat contracts,
-they are completely forbidden in their system, explicitly to not encounter
-the previously mentioned issue.
-This restriction goes against what is usually expected from dynamic checks:
-validation of data, and not just inclusion on types.
-
-On top of that, a fundamental problem arises on said paper,
-symptom of the difficulty of establishing good enough, and yet
-simple semantics for union and intersection dynamic checks.
-One of their main results, is the fact that their calculus obeys
-all of their expected monitoring properties for higher ordered, union
-and intersection contracts.
-However, one of these properties states the following:
-
-\begin{figure}[h]
-$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
-\caption{Sound monitoring property for intersection from Williams, Morris, and Wadler.}
-\label{fig:wmw-semantics}
-\end{figure}
-
-Which means that a continuation $K$ complies negatively with an
-intersection of $A$ and $B$ if it complies negatively with at least one of these.
-Or, less formally, that if you have a program that is missing a value at a given point,
-you can put a value wrapped with a contract stating $A \cap B$,
-and expect not to get a negative violation of that contract,
-if you can either put a value
-wrapped with $A$, or a value wrapped with $B$.
-
-Altough this is true, it's a weaker statement of what we'd like to have,
-consider the two contexts presented on Figure \ref{fig:valid-contexts},
-the first one is a valid context for a \nickel{Num -> Num} function,
-that is, you could replace the $\square$ by some function wrapped in a
-\nickel{Num -> Num} contract, and the context will not violate it.
-Similarly, the second context from the same figure is a valid
-\nickel{Bool -> Bool} context.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=\nickel{Num -> Num} context]
-(*@$\square$@*) 3
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=\nickel{Bool -> Bool} context]
-(*@$\square$@*) true
-\end{lstlisting}
-\caption{Two different contexts in Nickel}
-\label{fig:valid-contexts}
-\end{figure}
-
-Next, we can combine these two contexts into one, that essentially
-has one single missing value, that will be used independently in both contexts
-from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
-This context is not a valid \nickel{Num -> Num} context, since it will apply
-the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
-In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
-Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
-shown on Figure \ref{fig:wmw-semantics}, it is not a valid
-\nickel{Num -> Num @& Bool -> Bool} context.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f = (*@$\square$@*) in
-(f 3, f true) 
-\end{lstlisting}
-\caption{Invalid \nickel{Num -> Num @& Bool -> Bool} context,
-combining contexts from Figure \ref{fig:valid-contexts}}
-\label{fig:invalid-context}
-\end{figure}
-
-However, both our intuition, as the most common use case of intersections
-(that is, overloading), tells us that this should work.
-While this problem does not invalidate their work (in fact, the system they present
-handles these cases correctly), it shows that existing research is still not on
-good shape to handle the growing usecases being developed for
-unions and intersections around dynamic languages.
-
-% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$
-
-% Keil and Thiemann did noticed this problem, and their contract satisfaction
-% rules only apply to elimination context, a particular kind of context
-% where the hole is either being applied to an argument
-% ($\square N$) or is used as an argument for a primitive
-% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).
-
-% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
-\resolved{(Yann) Maybe adding a concrete example of this (it can be the same term but with
-    concrete types like Number and String and simple contexts like application
-    to a dumb argument) is sufficient to make the point}
-
-\subsection*{Racket}
-\label{sec:racket}
-
-Racket is a language based on Scheme, mainly thought as a language to experiment
-with different language desings and ideas. It provides a very complete contract
-system\cite{RacketContracts}.
-
-Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
-These contracts are lightweight versions of intersections and unions, they simply
-check that every contract is valid, in the case of \racket{and/c},
-and that at least one contract
-is valid, in the case of \racket{or/c}.
-
-So, for instance, the usual usecase of intersection to encode bounded overloading
-does not work on Racket. Take a look at the next example, the function
-\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
-since 3 is not a valid \racket{string?}.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract overload
-(and/c (-> number? number?)
-       (-> string? string?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Non working function overloading on Racket.}
-
-\end{figure}
-
-Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
-contracts that could (potentially) apply to a function, these have to be
-differentiated by first order markers.
-For instance, the code on Figure~\ref{code:racket:or/c:working}
-would correctly allow only to call
-\racket{united} with a \racket{number?},
-and it can make this distinction since it knows that
-\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
-However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
-before running, since the \racket{or/c} contract is unable to know
-which of the two function contracts to use.
-Ideally, this last example should just behave as wrapping
-the function with the \racket{(-> even? even?)} contract.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract united
-(or/c (-> number? number?)
-      (-> string? string? string?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Working \racket{or/c} example on Racket.}
-\label{code:racket:or/c:working}
-
-\end{figure}
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=lisp]
-(define/contract united
-(or/c (-> number? number?)
-      (-> even? even?))
-(lambda (x) x))
-\end{lstlisting}
-\caption{Non working \racket{or/c} example on Racket.}
-\label{code:racket:or/c:non-working}
-
-\end{figure}
-
-\unsure{This section is overall a bit hard to follow. I
-  think I've figured out what it means, but it needs more love. Also
-  give an example which doesn't work and one that does for or/c.}
-
-\subsubsection*{case->}
-
-Racket does provide an alternative for overloading methods, called the
-\racket{case->} contract, however, this overloading of methods only
-works over contracts of functions with different amount of parameters.
-It's a good alternative to allow limited overloading of functions.
-As an example, consider the following non failing piece of code, and
-try to understand how Racket can decide, at each application,
-which function contract to use.
-
-\begin{figure}[h]
-
-\begin{lstlisting}[language=racket]
-(define/contract overcase
-  (case-> (-> string? string?)
-          (-> number? number? number?)
-          )
-  (lambda (x [y 0]) (if (number? x)
-                   (+ x y)
-                   x)))
-
-(overcase 1 2)
-
-(overcase "hello")
-\end{lstlisting}
-\end{figure}
-
-Even if Racket's intent is not to implement union and intersection
-dynamic checks, they do provide contracts that are, in some sense,
-simplified versions of those.
-In our eyes, Racket has accepted the difficulty of correctly and practically
-combine union, intersection, and higher order contracts, and has choose
-to prioritise the highe order ones.
-
-\newpage
-
-\section{Related work}
-\label{sec:related-work}
-\info{Including a mention of statically typed systems with union}
-
-\newpage
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-The design and implementation of programming languages is a challenging task. It
-is a delicate exercise to balance complexity, expressiveness, efficiency and
-user experience all together. Unfortunately for designers, a language cannot
-usually be reduced to a simple sum of small independent features. They are
-rather composed of a lot of moving parts that interact in unexpected ways. The
-more a language grows, the more any new addition has to be weighted carefully,
-as it has the potential to interact with any of the existing features.
-
-Part of recent developments revolve around mixing static typing and dynamic
-typing via gradual typing, in order to get the best of both worlds. Contracts
-(or casts in the gradual typing terminology) are crucial to this endeavour by
-allowing for a safe interaction between typed and untyped code. But we also made
-a case that contracts are interesting in themselves in a fully dynamically typed
-language (and even in a fully statically typed language).  Contracts are a
-precious idiom for data validation or to check pre- and post-condition that
-would be hard to ensure statically. In spirit, contracts are similar to
-assertions, but they are part of a principled framework, they are composable,
-they have good error reporting and they are integrated as a language feature. In
-particular, contracts natively handles higher-order functions and built-in data
-structures. This makes them an great candidate for integration in the Nickel
-configuration language.
-
-This is only the beginning of the story. A language designer then has to choose
-what contract combinators to provide among a number of possibilities: there is
-the standard package including higher-order function contracts, record contracts
-and list contracts.  Then, one could add polymorphic
-contacts~\cite{BlameForAll}, dependent function
-contracts~\ref{DependentContracts}, and so on. For validation purpose, union and
-intersection contracts are really appealing. Unions have a intuitive
-interpretation as modeling alternatives and thus appear implicitly or explicitly
-in a lot of data structures.
-
-Alas, the implementation of union and intersection contracts is not as simple as
-it first appears. In conjunction with higher-order contracts, or even just basic
-contracts in a lazy language, they behave in a way that is reminiscent of
-side-effects. These behaviors were, for most of them, already known: we do not
-claim their discovery. However, as language implementers, we have made an actual
-attempt at including them in a concrete language. We exposed what, in our
-opinion, are the most salient flaws and their consequences. Even in a pure
-language, union and intersection contracts break the equivalence between
-call-by-name and call-by-value, and consequently can prevent CSE-like
-optimizations. There are discrepancies between the expected and actual semantics
-that make some behaviors unintuitive. Their implementation is a non-trivial
-change that contaminates the evaluation of all other contracts, and the
-performance penalty is not obvious to assess.
-
-To conclude, let us make clear that we do not think union and intersection
-contracts are fundamentally broken, that they can not be implemented correctly,
-or that they do not bear any value (quite the contrary). They may still make
-sense to have in a language, and some apparent difficulties in the
-implementation could be lifted some day. But as often, there are gaps between
-the theoretical framework, a proof-of-concept, a prototype, and the integration
-in an actual language. We hope that our attempt may serve as a cautionary tale,
-that for union and intersection contracts, these gaps are sometimes large in
-places.
-
-\bibliography{nickel}
-
 \appendix
@@
 \input{challenges}
-
-\end{document}
-
-% LocalWords:  Felleisen

### 2021-06-01 paper.tex
Move the old appendix in a separate file

@@
 \bibliography{nickel}
-
-\appendix
-
-\todo{Remove parts that are already on the main part}
-
-\section{Background on union types}
-
-\subsection{Nickel design space}
-
-{\color{red}Goals of the section
-
-Introduce Nickel, explain the choice of gradual
-typing and main design orientations, such as practice-oriented, lightweight,
-etc.\vspace{0.5cm}}
-
-Build systems, package managers, operating systems, cloud infrastructure,
-continuous integration and web services\resolved{Also continuous integration}
-are examples of modern complex software systems that require an extensive and
-non-trivial configuration in order to make them adapted to each different
-use-case. To manage the growing complexity that is then offloaded to
-configurations, the declarative approach has become more and more popular,
-illustrated for example by the infrastructure-as-code paradigm. This leads
-configuration to play an important role in critical aspects of software
-engineering, including security, availability, and maintainability.
-
-However, static text-based configuration alone is falling short of
-expressiveness, and is seldom sufficient. Data cannot be transformed, combined
-nor shared, resulting at best in boilerplate and duplication of information, or
-at worst, in data being invalid or inconsistent. Data validation is not
-supported either, and must be handed over to yet another tool down the
-configuration pipeline, if ever done. Correctly configuring a modern system is
-hard and failures may have substantial negative consequences.
-
-This is why the authors have been developing the Nickel
-language\cite{NickelRepo}. Nickel is a configuration programming language,
-meaning that a Nickel program evaluates to a value that must be
-straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
-lightweight and easy to write while still providing features for writing correct
-code and generating valid data. The focus is first and foremost practical:
-Nickel originated as a rethink of the language of the Nix package manager, used
-to describe one of the biggest sets of software packages\cite{repology}.  Nickel
-targets build systems and cloud deployment tools (Kubernetes, Terraform or
-NixOps) as well. While we are keen on incorporating existing or doing new
-research when it solves a well-defined problem, the general design goals are
-always guided by the practical industrial use cases.
-
-\subsection{Typing}
-
-Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
-Unlike JSON, though, Nickel anticipates large configurations by being both
-programmable and typed. In the configuration setting, there is a singular
-dilemma with respect to typing: since a program is run on fixed inputs and is
-expected to terminate, any relevant type error will show up at evaluation.  Why
-bother with the complexity of a static type system? On the other hand, more and
-more software systems offload complexity to configurations, in particular with
-the infrastructure as code paradigm prevailing today in cloud deployment. When
-the complexity of a codebase grows, static types become attractive again.  For
-reusable code — that is, library functions —, static types are specifically
-adapted, and bring all the usual benefits of early error detection, code
-robustness, better code structure, documentation, and so on.
-
-This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
-which mixes both static and dynamic typing. Unlike gradually-typed languages
-like TypeScript\unsure{Todo: explain that typescript uses unions as
-  static typing whereas we are speaking of dynamic contracts}, we do not seek to build a complex type system that tries hard
-to accept most of the idioms naturally arising in dynamically typed code.
-Rather, we choose to provide a reasonably expressive type system with good
-inference properties, which makes typing functions operating on generic data easy,
-but may require using untyped code for more exotic expressions.  In any case,
-this is not a surrender, as Nickel provides a complementary mechanism for more
-advanced data validation: contracts.
-
-\subsection{Contracts}
-{\color{red}Goals of the section
-
-A primer on contracts, which in the end are the problematic bit when confronted
-with unions.\vspace{0.5cm}}
-
-Enforcing pre- and post-conditions at runtime is a widely established practice.
-In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
-Felleisen introduce a principled approach to run-time assertion checking that
-nicely supports higher-order functions and introduces the notion of blame, which
-is crucial to good error reporting. It became apparent later that their
-contracts are closely related to the type casts introduced by gradual typing,
-modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
-value of contracts as a safe interface between typed and untyped code. In
-\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
-gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
-historical bit should probably be moved to related works eventually} Nickel
-adopts a similar type system, with both statically typed terms, dynamically
-typed terms, and first-class contracts. In "first-class contracts", we include
-the ability to write user-defined contracts. Because built-in types are too
-simplistic for the purpose of data validation, the ability to write a small
-validation function for, say, a port number, to have it immediately available as
-a contract that can be composed with other type constructors\resolved{Needs code
-to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
-system. An example is given in Figure~\ref{fig:contract-composition-appendix}. The first
-line defines a port contract from a predicate and use it in the definition of a
-larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
-contract checks that a value is a record with the following shape: it must have
-a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
-\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
-a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
-entails that a value respect a contract, akin to \lstinline+:+ for types. We
-assume that a \lstinline+Host+ contract has been defined in a similar way.
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let Port = contracts.fromPred (fun value =>
-  builtins.isNum value && value % 1 == 0
-  && value >= 0 && value <= 65353) in
-let Configuration = {
-  host | #Host,
-  port | #Port,
-} in
-{host = "nickel-lang.org", port = 80}
-| #Configuration
-\end{lstlisting}
-  \end{center}
-\caption{Contract composition}
-\label{fig:contract-composition-appendix}
-\medskip
-\small
-Defines a port contract from a predicate and use it in the definition of a
-\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
-role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
-contract to a value, similar to `:` for static types. We assume that a
-\lstinline+Host+ contract has also been defined.
-\end{figure}
-
-In this hybrid system, each type constructor of the static type system - arrows,
-records, foralls - must have a contract counterpart that checks at run-time that
-a value is a member of the corresponding type. Contracts for various
-extensions of the simply typed $\lambda$-calculus have been investigated
-\cite{BlameForAll, KeilThiemannUnionIntersection, RootCauseOfBlame,
-DependentContracts, GradualTypingClasses}, with solutions of varying complexity.
-Amongst all the common extensions to the polymorphic lambda calculus, one of the
-most useful and widespread turns out to be one of the most challenging:
-\emph{union types}.
-
-\subsection{Union types}
-{\color{red}Goals of the section
-
-Introduce union in general, and most usual form in statically typed languages, tagged unions\vspace{0.5cm}}
-
-\todo{Specify: union types? more general concept?}{Unions} are fundamental and ubiquitous building blocks of program data. A union
-only holds one value whose shape is only known to be in a fixed set of
-alternatives: for example, either a boolean or a string. When used in combination
-with products, they can represent arbitrary tree-like structures, allowing to
-model a large class of data.
-
-Unions are also useful for configurations:
-\begin{itemize}
-    \item To represent nullable values in JSON.
-    \item For plain enumerations. For example, the enabled protocols of a data server
-        would be either \lstinline+HTTP+, \lstinline+FTP+ or \lstinline+SFTP+.
-    \item To support different representations of the same data. A file path could be
-        accepted either as one string of \lstinline+"/"+-separated strings, or
-        as a list of strings.
-\end{itemize}
-
-In statically typed languages, and more specifically in functional programming
-languages, unions are usually implemented as algebraic data
-types\unsure{This is a very misleading statement at best. Sums are
-  "tagged union" sure, but that refers to an encoding a sums with
-  unions, certainly not an encoding of unions with sums!} [references],
-also called \emph{tagged unions}. Well-known programming languages with
-first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
-Swift and many more.
-
-\subsection{Tagged versus untagged}
-{\color{red}Goals of the section
-
-Interpreted languages already need to tag their values, hence bare union are more natural.
-It also corresponds better to prior usage in untyped code\vspace{0.5cm}}
-
-In the statically typed setting, the actual representation of a tagged union is
-an implementation detail: from the point of view of the programmer, a tagged
-union is a first-class data type associated with constructors and destructors
-with well-defined semantics. In contrast, in a gradually typed language like
-Nickel, the question of the representation of tagged unions as untyped values
-arises.
-
-Unfortunately, tagged unions do not have a serializable canonical counterpart as
-untyped data. One usual representation is a record with a tag and a value whose
-shape depends on the tag (see Figure \ref{fig:union-encoding}).
-
-\begin{figure}
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-{tag: <Num, Str>, value: Dyn}
-\end{lstlisting}
-  \end{center}
-\caption{Encoding of the union type: integer or a string}
-\label{fig:union-encoding}
-\medskip
-\small
-In Nickel, \lstinline+<A, B, C>+ is the type of a C-like enumeration that can
-take value \lstinline+A+, \lstinline+B+ or \lstinline+C+, while \lstinline+Dyn+
-is the dynamic unitype.
-\end{figure}
-
-A better way to describe it would be as a dependent pair which first component
-is a tag and second component is the data type corresponding to this tag:
-
-\[
-    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
-\]
-
-Nickel does not feature dependent types, which would open a whole new level of
-complexity. Alas, choosing the non dependent representation as a default is not
-fully satisfying.  First, it does not map precisely to a JSON value: in
-consequence, a lot of illegal combinations are structurally valid as untyped
-JSON values. But this is to be expected of any representation. More importantly,
-this does not necessarily match with the configuration schemas one can find in
-the wild, which favor simple, untagged union.
-
-In dynamically typed languages, it is customary to directly store different kinds
-of data in one variable without further ceremony. The runtime system needs to
-attach type tags to values anyway, which can usually be observed using functions
-like JavaScript's \verb+typeof+. Using so-called \emph{untagged
-  unions}\unsure{If you can observe with typeof, then it's
-  pretty much a tagged union, though the tag is implicit} has
-already been an idiom in dynamically typed language for a long time
-[ref/examples], and being able to type and handle these unions is an important
-aspect of the migration to a gradual type system. Indeed, a lot of gradually
-typed languages either existed originally as dynamically typed language or aim
-at backward compatibility with a dynamically typed base language. This is why
-they tend to favor supporting untagged union types, the typed counterpart of
-this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
-
-In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. However, Nickel is a configuration
-language, whose contracts and types system is intended to be expressive enough
-to encode data schemas encountered in the wild. Since these schemas do use
-untagged unions, there is a strong incentive to support untagged unions at the
-type level, or at least, at the contract level.
-
-\subsection{Tagged as untagged}
-{\color{red}Goals of the section
-
-Untagged unions, modulo a slightly smarter typechecker, can encode usual type-safe
-tagged unions. They let the user free of the representation though, and encode
-more stuff, that's why they are appealing.\vspace{0.5cm}}
-
-Untagged union, together with the other existing features of the type system,
-are sufficient to implement the encoding illustrated in Figure
-\ref{fig:union-encoding}. For example, let us represent an algebraic data type
-(written in ML syntax) as an imaginary Nickel union type:
-
-\begin{lstlisting}[language=caml]
-type either = Left of float | Right of string
-\end{lstlisting}
-
-\begin{lstlisting}[language=nickel]
-{ tag : <Left>, val : Num}
-| { tag : <Right>, val : Str}
-\end{lstlisting}
-
-There is a twist, though. As it is, the static type system is unable to refine
-the type of the value, making the following example rejected:
-
-\begin{lstlisting}
-switch x.tag {
-  Left => x.val + 1,
-  Right => strings.fromNum x.val
-}
-\end{lstlisting}
-
-Flow typing, implemented in Racket\cite{FlowTypingRacket1, FlowTypingRacket2},
-Groovy\cite{FlowTypingGroovy}, Whiley\cite{FlowTypingWhiley} or
-TypeScript\cite{FlowTypingTypeScript}, is capable of refining appropriately the
-type information in each branch. Thus equipped with untagged unions and a
-typechecker smart enough, classic tagged unions can be simulated, but with the
-additional benefit that the language is not imposing any representation.
-Untagged unions empower the programmer to represent and handle a wider variety
-of union idioms appearing in the wild.  Thus, adding union types to Nickel
-sounds appealing.
-
-\subsection{Union and intersections}
-{\color{red}Goals of the section
-Explain why having unions probably unavoidably leads one to have intersection.\vspace{0.5cm}}
-
-While our original motivation is in supporting unions, the rest of the paper
-explore complete systems featuring union and intersection types. One could
-wonder if the difficulties could then be sidestepped by removing intesection out
-of the equation. But
-
-\begin{enumerate}
-    \item Most of the encountered issues already appear with bare unions.
-    \item Intersection being dual to unions, they are showing up in some way
-        when unions are put in negative positions. For example, a contract for
-        $(A \cup B) \to C$ in a system without intersections must be
-        equivalent\unsure{``Must be'' is a rather strong statement, we
-        will find a way to tone it down}
-        to the contract $(A \to C) \cap (B \to C)$
-\end{enumerate}
-
-\unsure{Some bits missing so far: we are interested in higher-order
-  contracts, Nickel is a higher-order \& pure language, contracts as
-  functions. We've said that user-defined contracts are a thing, it
-  probably matters so we may have to hammer it down.}
-
-\unsure{Something that we must mention somewhere: if we want to check
-  that something is a list of As lazily (to avoid making functions
-  like hd be O(n), for instance), then we have a lot of the problems of
-  higher-order contracts anyway.}
-
-\subsection{Performance considerations}
-\info{This a draft}
-\todo{transition, location}
-
-At first sight, performance should not be a critical issue given the use cases
-of Nickel. But generating large configurations can already hit some limits in
-current related languages [reference?]. A natural is parallelization:
-Nickel being an almost pure functional programming language, evaluating separate
-files or expressions in parallel is straightforward and effective. This
-parallelization potential is thus a design requirement, and must be when the
-addition of union types.
-
-\section{Challenges}
-\input{challenges}
-
 \end{document}

### 2021-06-01 paper.tex
Typo
Fit error message example within the line size

@@
 give an example of a situation where adding a simple contract annotation
-\nickel{Str -> Str} to a function greatly improves the error message . In the
+\nickel{Str -> Str} to a function greatly improves the error message. In the
 original program of Figure~\ref{fig:contract-reporting-wo}, the
@@
 
-\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none]
+\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
 error: Blame error: contract broken by the caller.
@@
 6 | catHosts (makeHost "google")
-  | ---------------------------- (2) calling <func>
+  | --------------------------- (2) calling <func>
 \end{lstlisting}

### 2021-06-03 paper.tex
Fix comma at beginning of line
This is a bit brutal to be honest… and may be a bad idea.
Wip: Arnaud's pass through the introduction
Very not finished but I'll probably only be done tomorrow.

@@
 
-Build systems, package managers, operating systems, cloud infrastructure,
-continuous integration and web services\resolved{Also continuous integration}
-are examples of modern complex software systems which require an extensive and
-non-trivial configuration. To handle this complexity, the declarative approach
-has become more and more prominent: instead of applying a sequence of
-transformations to a system in order to reach a desired state, this state is
-described as static textual data. Some tool is then in charge of applying the
-necessary transformations. This approach makes the configuration persistent,
-reproducible, versionable and subject to static analysis. Text-based
-configurations are thus led to play an important role in critical aspects of
-software engineering, including security, availability, and maintainability.
-
-However, static text-based configurations show their limits beyond a certain
-size and complexity. Data cannot be transformed, combined, shared, nor validated
-(without resorting to separate schemas and tools).  This lack of expressivity
-results at best in boilerplate and duplication of information, but it can also
-make data invalid or inconsistent.  Correctly configuring a modern system is hard
-and failures may have substantial negative consequences.
+A configuration language is a language concerned with describing the
+configuration of an application. In traditional configuration
+languages, such as YAML, TOML, or JSON, the configuration is fully,
+and explicitly, spelt out.
+
+However, with the advent of DevOps, configurations have been extended
+to describe the entire state of a computer, or even a fleet of
+computers. For instance, with Kubernetes you need to configure a large
+fleet of (possibly replicated) docker containers. To describe this
+sort of configurations, you really want to be able to re-use and
+abstract parts of the configuration, like traditional programming
+languages do. To meet this need, languages such as Cue~\cite{?},
+Dhall~\cite{?}, Jsonnet~\cite{?}, or Nickel~\cite{NickelRepo}, where
+configurations are generated rather than spelt out, were created.
+
+Another example is continuous integration systems: it's fairly typical
+to need a matrix of job, wherein the same tests are run on different
+infrastructures, or with different versions of a compiler. Traditional
+configuration would have you write the same steps for each
+infrastructure. It's much better to write the steps once, and
+instantiate them for each infrastructure. Continuous integration
+systems typically do this using a templating system layered on top of
+YAML. Each of the configuration-generating languages above allow such
+job-matrix definition natively.
+
+\subsection*{Nickel}
+
+In this article, we will use the Nickel language~\cite{NickelRepo} as
+illustration and motivation. At its core, Nickel is the JSON data
+model, that is dictionaries (written \nickel{\{field1 = value1, ...,
+  fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
+with functions (\nickel{fun arg1 ... argn => body}) and
+let-definitions (\nickel{let id = value in exp}).\unsure{Arnaud: this
+  paragraph looks kind of bad, maybe it'll be better to put each
+  syntax entry in an enumeration, we will see if there is still room
+  for this}
+
+A Nickel configuration is then evaluated to an explicit configuration,
+\emph{e.g.}  in JSON, which can then be consumed by an
+application. Therefore a design constraint of Nickel is any Nickel
+data must have a straightforward interpretation in JSON.
+
+\subsection*{Contracts}
+
+A useful feature of a configuration language is to provide facilities
+for schema validation. That is, help answer questions like: does our
+configuration have all the required fields? does this field indeed
+contains a URL?
+
+These are an inherently dynamic question, as they are all questions
+about the evaluated configuration.
+
+\unsure{Arnaud: What's below this point in the introduction is being reworked}
 
@@
 then folded. However, the programmer forgot one parameter when calling to
-\nickel{makeHost}, thus passing a partially applied function instead of a string
+\hbox{\nickel{makeHost},} thus passing a partially applied function instead of a string
 to \nickel{catHosts}. Because this argument is put in a list and folded, the
@@
 
-% LocalWords:  Felleisen
+% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall
+% LocalWords:  Kubernetes DevOps natively

### 2021-06-04 paper.tex
Small fixes:
 * flat -> user defined
 * resolved some notes

@@
 
-\subsection*{Flat contracts}
+\subsection*{User defined contracts}
 \label{sec:flat-contracts}
 
-\unsure{Should we define contracts prior to this?}
-User defined contracts, also called flat contracts, are a particular
+User defined contracts, sometimes called flat contracts, are a particular
 kind of contract that behaves similar to an assertion, and whose
-main point is to apply some arbitrary predicate (as arbitrary as the
+main utility is to apply some arbitrary predicate (as arbitrary as the
 host language allows it) to a value; if the predicate evaluates to
@@
 For instance, the contract \nickel{Even} shown on Figure
-\ref{fig:contract-higher-order} is a flat contract, defined
+\ref{fig:contract-higher-order} is a user defined contract, defined
 by taking the value that needs to be checked, and validating
@@
 
-Or even more interesting, we could come up with a contract that checks that a function
+Even more interesting, we could come up with a contract that checks that a function
 maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
-by applying the function and checking its results,
+by applying the function exhaustively and checking its results,
 and then use it as a contract for the \nickel{not} function, as can be seen on Figure
@@
 
-Finally, the Racket programming language also has a system to work with
+The Racket programming language also has a system to work with
 contracts, powerful enough to define flat contracts
@@
 
-Even if flat contracts may seem like too permissive, the ability to make
+Even if user defined contracts may seem like too permissive, the ability to make
 non trivial checks on values (that is, to validate their output) is expected
@@
 Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
-unions and intersections can make \ref{eq:cbn-expansion} \resolved{it} unsound
+unions and intersections can make \ref{eq:cbn-expansion} unsound
 \emph{even in a pure setting}, making the optimization of
@@
 \info{What are they, what are they used for.}
-\todo{Since nickel uses | for contract application, I'd move unions
+\resolved{Since nickel uses | for contract application, I'd move unions
 to some other symbol.
@@
 important aspect of the migration to a gradual type system. Indeed, a lot of
-\unsure{This is the first time discussing gradual type systems, maybe it should
+\resolved{This is the first time discussing gradual type systems, maybe it should
 be introduced earlier, in particular how they relate to contracts
@@
 
-\subsection*{Flat contracts do not intersect}
+\subsection*{User defined contracts do not intersect}
 \label{sec:flat-and-inter}
@@
 presented by Williams, Morris, and Wadler, seems to be spot on.
-However, it presents a major problem, it's not compatible with flat contracts,
+However, it presents a major problem, it's not compatible with user defined contracts,
 as presented in Section \ref{sec:flat-contracts}.
@@
 \end{lstlisting}
-\caption{Intersection and flat contracts}
+\caption{Intersection and user defined contracts}
 \label{fig:inter-flat-contracts}
@@
 Another difficult issue Keil and Thiemann were able to solve, is the
-fact that flat contracts and intersection do not usually play nice
+fact that user defined contracts and intersection do not usually play nice
 together, as shown on Section \ref{sec:flat-and-inter}.
@@
 semantics that are simple to understand.
-With these rules, the interaction between flat contracts and
+With these rules, the interaction between user defined contracts and
 intersections becomes much simpler, since they are able to check
@@
 
-Williams, Morris, and Wadler, take a different approach on flat contracts,
+Williams, Morris, and Wadler, take a different approach on user defined contracts,
 they are completely forbidden in their system, explicitly to not encounter

### 2021-06-04 paper.tex
Acquire lock for section 4
Rework beginning of section 4

@@
 
-Type systems featuring intersection types are hard in a fundamental way,
-from their discovery, they've been studied as a form to characterize
-lambda-calculus terms normalizability; which means that typechecking a
-term on a system with intersection types is undecidable,
-similar to how the halting problem is undecidable\todo{reference}.
-
-Usually, when thinking about programming languages, there's an intuition
-that static checks (like typechecking) are more difficult and less powerful
-than dynamic checks (like testing).
-For instance, it is very easy to check if,
-in a given execution of a program, a variable holds an integer value; while it
-may become more challenging to check if a variable will always hold an integer value,
-no matter the particular execution.
-
-As we intend to show on this paper, unions and intersections are not only
-statically hard to check, but they are also dynamically hard:
-\emph{they are fundamentally hard}.
-In this section, we present intuitively the main challenges that arise
-when attempting to develop a runtime check system for unions and intersections.
+Static typing is about enforcing that some property always holds for a
+potentially unknown value, no matter the execution part. A dynamic check, on the
+other hand, just need to verify a property for a specific and known value, which
+is usually much easier. For example, dependent function types are complex and
+mostly implemented in proof assistants (although they are slowly making their
+way in some mainstream languages), whereas the corresponding dependent function
+contracts can be implemented in a few line of code (as is done in
+Racket~\cite{}).
+
+The gist of the present paper is to bring evidence that in the case of union and
+intersection, the dynamic checking part is complex as well: \emph{unions and
+intersections are fundamentally hard}.
+
+% Teo's version
+% Implementing a static type system is not an easy task.
+% Type systems featuring intersection types are hard in a fundamental way,
+% from their discovery, they've been studied as a form to characterize
+% lambda-calculus terms normalizability; which means that typechecking a
+% term on a system with intersection types is undecidable,
+% similar to how the halting problem is undecidable\todo{reference}.
+%
+% Usually, when thinking about programming languages, there's an intuition
+% that static checks (like typechecking) are more difficult and less powerful
+% than dynamic checks (like testing).
+% For instance, it is very easy to check if,
+% in a given execution of a program, a variable holds an integer value; while it
+% may become more challenging to check if a variable will always hold an integer value,
+% no matter the particular execution.
+%
+% As we intend to show on this paper, unions and intersections are not only
+% statically hard to check, but they are also dynamically hard:
+% \emph{they are fundamentally hard}.
+% In this section, we present intuitively the main challenges that arise
+% when attempting to develop a runtime check system for unions and intersections.
 
@@
 
-When programming with dynamic checks it is assumed that lack of
-failure does not mean that the program works perfectly.
-
-For instance, imagine having a function wrapped in a contract
-that checks that it maps positive numbers to positive numbers,
-and using it in different ways,
-like the program on Figure \ref{fig:pos-to-pos}.
-Clearly, there is an error on this program, but if the first call
-to \nickel{f} that gets executed is \nickel{f 10}, the error will not
-be found at that point, and the first operand of the addition would be
-evaluated to \nickel{3}.
-If, later on, \nickel{f 5} gets executed, the contract would get violated,
-since \nickel{f} gets called with a positive value, but returns a negative
-one (\nickel{-2}).
+In a pure language with standard contracts, the failure of a function contract
+can always be blamed on a single call. Figure~\ref{fig:pos-to-pos} shows a
+function with a simple contract attached that ensures that both the argument and
+the result of \nickel{f} are positive numbers. \nickel{f} is called twice, and
+the second call violates the requirement on the codomain, leading to blame
+\nickel{f}. The first call to \nickel{f} does not matter, and \nickel{5} is a
+single and independent witness of the violation of the contract. The user can be
+pointed to one location in practice.
+
+There is an intuitive justification for this fact. Apart from the error
+reporting part (which is crucial in practice!), the current contract system of
+Nickel can be implemented purely as a library, just given a \nickel{fail}
+primitive to abort the execution. In particular, applying a function contract to
+\nickel{f} replaces it with a guarding function \nickel{fWrapped} that performs
+the required checks. Thus, in a pure setting, the failure of \nickel{fWrapped 5}
+must be independant of its environment, and in particular of the call
+\nickel{fWrapped 10}.
+
+% When programming with dynamic checks it is assumed that lack of
+% failure does not mean that the program works perfectly.
+%
+% For instance, imagine having a function wrapped in a contract
+% that checks that it maps positive numbers to positive numbers,
+% and using it in different ways,
+% like the program on Figure \ref{fig:pos-to-pos}.
+% Clearly, there is an error on this program, but if the first call
+% to \nickel{f} that gets executed is \nickel{f 10}, the error will not
+% be found at that point, and the first operand of the addition would be
+% evaluated to \nickel{3}.
+% If, later on, \nickel{f 5} gets executed, the contract would get violated,
+% since \nickel{f} gets called with a positive value, but returns a negative
+% one (\nickel{-2}).
 
@@
 
-Even if the program performed two different applications of \nickel{f}
-before finding the contract violation, the first one is not actually
-needed, and executing only \nickel{f 5} is enough to find the problem.
-
-Having a witness of size 1 for every contract failure is a really nice to have property
-since it means, for example, that every time that a contract is violated,
-the stack trace is enough to understand how it was violated, and to potentially
-come up with a minimal example that breaks the contract.
-However, when unions get introduced, this property is lost.
-
-Consider now the program presented in Figure \ref{fig:wrong-union-function},
-this program (and in particular, \nickel{f}'s contract) still presents a problem,
-since \nickel{fun x => x - 7} does not map positive numbers to positive numbers,
-nor does it map positive numbers to non-positive numbers, so it won't be a valid
-value for the union of those contracts.
+% Even if the program performed two different applications of \nickel{f}
+% before finding the contract violation, the first one is not actually
+% needed, and executing only \nickel{f 5} is enough to find the problem.
+%
+% Having a witness of size 1 for every contract failure is a really nice to have property
+% since it means, for example, that every time that a contract is violated,
+% the stack trace is enough to understand how it was violated, and to potentially
+% come up with a minimal example that breaks the contract.
+% However, when unions get introduced, this property is lost.
+
+Union contracts are a different matter. Consider the program presented in
+Figure \ref{fig:wrong-union-function}. \nickel{f} is again violating the
+contract, as it neither maps positive numbers to positive numbers nor positive
+numbers to negative numbers.
 
@@
 
-As before, the program presents enough function calls to find that issue
-and violate the contract,
-since \nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
+In a system with union and intersection, this contract would fail, because
+\nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
 and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
-But, what makes this case completely different from the one presented before, is that
-both function calls are needed, and, what is more, there is not a way to
-use the wrapped value \nickel{f} only once that would find a violation to its contract;
-at least two are needed.
+But, as opposed to the previous example, removing one call or the other will make the program pass!
+Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
+the union,
+while the second one only fails the \nickel{Positive -> NonPositive} part. Both
+are needed to prove the violation of the full contract.
+
+Even in a pure setting, unions introduce a form of effect at a distance: the
+result of \nickel{f 5} is not local anymore as it depends on the previous
+execution of other calls to \nickel{f}. With unions,
+equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations} does not
+hold, and the candidate example of
+Figure~\ref{fig:optimizations-let-floating-ex} cannot be optimized in general.
+
+Figure~\ref{fig:optimized-programs} below shows another example of the same
+fact. A seemingly valid optimization ends up with a program that behave
+differently from the original one.
+
+\unsure{WARNING: what is below is being reworked by Yann}
 

### 2021-06-04 paper.tex
Hook directly with union at the beginning of the introduction

@@
 
+Union types, meaning a type \nickel{A @| B} containing values which
+belong either to a type \nickel{A} or \nickel{B}, are a popular tool
+when adding a static types to a dynamic language. In particular, both
+TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
+union typed to model the frequent practice to use the value
+\lstinline{null} (\lstinline{None} in Python) to represent an absent
+optional value. This is why the gradual typing literature, concerned
+with formalising the interplay between static and dynamic type
+systems, has been quite interested in union
+types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.
+
+On the other hand, unions are not a common feature of static type
+systems, mostly because they are quite difficult to verify
+statically. So unions are really only worth it in gradually typed
+language where they formalise existing dynamically typed patterns. On
+the other hand, surely, for dynamic tests, unions are really easy: it
+is simply the boolean disjunctions of two tests.
+
+Unfortunately, as we document in this article, as soon as you extend
+dynamic checks to \emph{contracts}~\cite{FindlerFelleisenHOContracts},
+unions become actually pretty difficult, and threaten desirable
+properties of your language.
+
+\subsection*{Configuration languages}
+
+To motivate contracts and the problem caused by unions, let's make a
+detour through configuration languages.
+%
 A configuration language is a language concerned with describing the
@@
 Another example is continuous integration systems: it's fairly typical
-to need a matrix of job, wherein the same tests are run on different
+to need a matrix of jobs, wherein the same tests are run on different
 infrastructures, or with different versions of a compiler. Traditional
-configuration would have you write the same steps for each
-infrastructure. It's much better to write the steps once, and
+configuration would have you copy the same steps for each
+infrastructure. This is tedious, hard to maintain, and error
+prone. It's much better, instead, to write the steps once, and
 instantiate them for each infrastructure. Continuous integration

### 2021-06-04 paper.tex
Arnaud working more of his way through the introduction

@@
 the other hand, surely, for dynamic tests, unions are really easy: it
-is simply the boolean disjunctions of two tests.
+is simply the Boolean disjunctions of two tests.
 
@@
 
-These are an inherently dynamic question, as they are all questions
-about the evaluated configuration.
+These are inherently dynamic questions, as they are all questions
+about the evaluated configuration. To this effect, Nickel lets us
+annotate any expression with a dynamic schema check: \nickel{exp |
+  C}. There is also syntactic sugar to annotate definitions:
+\nickel{let id | C = value in exp} stands for \nickel{let id = (value
+  | C) in exp}.
+
+Let us pause for a moment and consider the following: it is Nickel's
+ambition to be able to manipulate configurations like Nixpkgs. With
+over 50\,000 packages, is one of the largest repository of software
+packages in existence~\cite{repology}. Concretely, Nixpkgs is a
+dictionary mapping packages to build recipes. That is, a massive,
+over-50\,000-key-value-pair wide dictionary. It is absolutely out of the
+question to evaluate the entirety of this dictionary every time one
+needs to install 10 new packages: this would result in a painfully
+slow experience.
+
+To be able to support such large dictionaries, Nickel's dictionaries
+are \emph{lazy}, that is, the values are only evaluated when
+explicitly required. For instance, when writing
+\hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
+evaluated.
+
+But let's consider now writing something like \nickel{nixpkgs |
+  packages}, to guarantee that all the packages conform to the desired
+schema. If this were a simple Boolean test, it would have to evaluate
+all 50\,000 package to check their validity, hence breaking the
+laziness of dictionaries. Do we have to choose between laziness and
+schema validation? Fortunately, we don't! Enter
+\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
+which can be partially delayed, yet errors can be reported
+accurately. Contracts can respect laziness of dictionaries, and they
+can be use to add schema validation to functions as well (in fact
+functions were their original motivations).
+
+The is no Boolean function which can check that a value has type
+
+\rule{3cm}{3pt}
 
@@
 non-trivial configuration.
-
-We will use the syntax of Nickel to write most examples throughout this paper.
-Syntactically, Nickel looks like a simple ML language with records. \nickel{let
-id = value in exp} is used to bind a value to an identifier in an expression.
-Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
-are written as \nickel{[x1, ..., xn]} and records as \nickel{\{field1 = value1,
-..., fieldn = valuen\}}. The application of a contract \nickel{C} to the
-expression \nickel{exp} is written \nickel{exp | C}\footnote{The current
-    implementation requires a \nickel{#} character in front of custom contracts,
-    whose application would rather be \nickel{exp | #C}, but it is irrelevant
-and subject to change in the future.}. We will sometimes attach
-contracts directly to an identifier instead of the corresponding bound
-expression, that is: \nickel{let id | C = value in exp} is syntactic sugar for
-\nickel{let id = (value | C) in exp}.
-
-Some examples resort to constructions that are not currently available in Nickel
-for illustration purpose: we will use \nickel{A @& B} (resp. \nickel{A @| B})
-for the intersection (resp. union) of contracts (or types) \nickel{A} and
-\nickel{B}. \nickel{@|} and \nickel{@&} bind more loosely than arrows, meaning
-that \nickel{A -> B @& C -> D} is to be parsed as \nickel{(A -> B) @& (C -> D)}.
 
@@
 
-% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall
-% LocalWords:  Kubernetes DevOps natively
+% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
+% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
+% LocalWords:  TypeScript

### 2021-06-04 paper.tex
Continue reworking section 4 (WIP)

@@
 
-\subsection*{Union contracts are hard to break}
+\subsection*{Union contracts as an effect}
 
@@
 There is an intuitive justification for this fact. Apart from the error
-reporting part (which is crucial in practice!), the current contract system of
-Nickel can be implemented purely as a library, just given a \nickel{fail}
-primitive to abort the execution. In particular, applying a function contract to
-\nickel{f} replaces it with a guarding function \nickel{fWrapped} that performs
-the required checks. Thus, in a pure setting, the failure of \nickel{fWrapped 5}
-must be independant of its environment, and in particular of the call
-\nickel{fWrapped 10}.
+reporting part (but this is the crucial in practice!), the current contract
+system of Nickel can be implemented purely as a library, just given a
+\nickel{fail} primitive to abort the execution. In particular, applying a
+function contract to \nickel{f} replaces it with a guarding function
+\nickel{fWrapped} that performs the required checks. Thus, in a pure setting,
+the failure of \nickel{fWrapped 5} must be independant of its environment, and
+in particular of the call \nickel{fWrapped 10}.
 
@@
 \end{lstlisting}
-\caption{Wrong positive to positive function}
+\caption{Simple contract violation}
 \label{fig:pos-to-pos}
@@
 \end{lstlisting}
-\caption{Still wrong positive to positive or positive to non-positive function}
+\caption{Union contract violation}
 \label{fig:wrong-union-function}
@@
 
-In a system with union and intersection, this contract would fail, because
-\nickel{f 10} is a witness of \nickel{f} failing \nickel{Positive -> NonPositive},
-and \nickel{f 5} is a witness of \nickel{f} failing \nickel{Positive -> Positive}.
-But, as opposed to the previous example, removing one call or the other will make the program pass!
+In a system with unions, this contract would fail, because \nickel{f 10}
+(returns \nickel{3}) is a witness of \nickel{f} failing the contract
+\nickel{Positive -> NonPositive}, and \nickel{f 5} (returns \nickel{-2}) is a
+witness of \nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
+the previous example, removing one call or the other will make the program pass!
 Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
-the union,
-while the second one only fails the \nickel{Positive -> NonPositive} part. Both
-are needed to prove the violation of the full contract.
+the union, while the second one only fails the \nickel{Positive -> NonPositive}
+part. Both are needed to prove the violation of the full contract.
 
 Even in a pure setting, unions introduce a form of effect at a distance: the
-result of \nickel{f 5} is not local anymore as it depends on the previous
-execution of other calls to \nickel{f}. With unions,
-equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations} does not
-hold, and the candidate example of
-Figure~\ref{fig:optimizations-let-floating-ex} cannot be optimized in general.
-
-Figure~\ref{fig:optimized-programs} below shows another example of the same
-fact. A seemingly valid optimization ends up with a program that behave
-differently from the original one.
-
-\unsure{WARNING: what is below is being reworked by Yann}
-
-This seemingly simple consequence generates major issues when considered in real
-life languages.
-For example, remember the different mechanisms for code manipulation presented
-in Section \ref{sec:optimizations}, inlining and commmon subexpression elimination,
-both of these rely on the fact that substituting a pure function call by its body,
-or joining equal pure expressions and sharing their result, is a semantically valid
-transformation.
-However, consider the examples presented at Figure \ref{fig:optimized-programs},
-the first one is an inlined version of the second one (where function
-\nickel{g} has been inlined); while the second is similar to the first, but
-the common subexpression \nickel{f 1} has been eliminated.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Inlined]
+result of \nickel{f 5} is not local anymore, as it now depends on the previous
+execution of other calls to \nickel{f}. Unions contracts break the
+equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}, and the
+candidate example of Figure~\ref{fig:optimizations-let-floating-ex} cannot be
+optimized in general.
+
+Figure~\ref{fig:optimized-programs} illustrates this point further. It comprises
+an original program, and one where the common subexpression \nickel{f 1} have
+been eliminated. These two programs behave very differently:
+\begin{itemize}
+    \item The original version returns \nickel{(1, "False")} without failing.
+    \item The optimized version fails with a contract violation.
+\end{itemize}
+
+since the single application of \nickel{f 1}, applied to both \nickel{true} and
+\nickel{false} will be able to find the error on the union contract
+\nickel{(Bool -> Num) @| (Bool -> Str)}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel, title=Original]
 let f | Num -> (Bool -> Num @| Bool -> Str)
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
-let f = (fun x y => if y then x else "False") | Num -> (Bool -> Num @| Bool -> Str) in
+\begin{lstlisting}[language=nickel, title=Optimized]
+let f | Num -> (Bool -> Num @| Bool -> Str)
+      = fun x y => if y then x else "False"
 let g = f 1 in
@@
 
-Unexpectedly, these two programs behave very differently, the first one will
-return \nickel{(1, "False")}, without failing.  While the second one will fail,
-since the single application of \nickel{f 1}, applied to both \nickel{true} and
-\nickel{false} will be able to find the error on the union contract
-\nickel{(Bool -> Num) @| (Bool -> Str)}.
+\unsure{WARNING: what is below is being reworked by Yann}
+
+To sum up, union contracts introduce a form of effect at a distance. This
+impacts both the programmer and the compiler or interpreter:
+\begin{itemize}
+    \item As for any side-effect, this breaks local reasoning.
+    \item It prevents optimizations.
+\end{itemize}
+\todo{WIP}
 

### 2021-06-04 paper.tex
Most of the introduction

@@
 \usepackage{url}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{hyperref}
@@
 for schema validation. That is, help answer questions like: does our
-configuration have all the required fields? does this field indeed
+configuration have all the required fields? does the \nickel{url} field indeed
 contains a URL?
@@
 \nickel{let id | C = value in exp} stands for \nickel{let id = (value
-  | C) in exp}.
+  | C) in exp}.\unsure{This point used to be made, to we want to
+  reinsert it? However, Nickel programs are of a special kind: they are usually
+simple, terminating programs that run on fixed inputs. Correctness errors will
+show up at evaluation anyway, or will not matter (dead code).}
 
@@
 can be use to add schema validation to functions as well (in fact
-functions were their original motivations).
-
-The is no Boolean function which can check that a value has type
-
-\rule{3cm}{3pt}
-
-\unsure{Arnaud: What's below this point in the introduction is being reworked}
-
-This is why the authors have been developing the Nickel
-language\cite{NickelRepo}. Nickel is a configuration programming language,
-meaning that a Nickel program evaluates to a value that must be
-straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
-lightweight and easy to write while providing support for writing correct code
-and generating valid data. Nickel originated as a rethink of the language of the
-Nix package manager, used to describe one of the biggest sets of packaged
-software in existence\cite{repology}.  Nickel targets build systems, cloud
-deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
-non-trivial configuration.
-
-\subsection*{Correct programs \& valid data}
-
-The end goal being data generation, Nickel is a dynamic, pure and functional
-language. Our goal is that writing a simple configuration in Nickel feels as
-easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
-configurations by being programmable, and by providing means of ensuring
-both correctness and validity:
-
-\begin{description}
-    \item[(correctness)] The program runs without errors. Errors may be type
-        errors, raised for example when evaluating the nonsensical division
-        \nickel{1 / "ftp"}. More generally, they can be any the violation of a
-        pre- or post-condition. When such an error cannot be prevented, we want
-        at least early failures and relevant error messages.\todo{make this
-        definition more precise}
-    \item[(validity)] The program generates valid data, i.e. data that conforms
-        to a schema (imposed by the system consuming the configuration). For
-        example, when configuring a web service, the schema may require the
-        presence of a field \lstinline+port+ that holds a valid port number.
-\end{description}
-
-A traditional answer is to use static types, and in particular static types for
-correctness. However, Nickel programs are of a special kind: they are usually
-simple, terminating programs that run on fixed inputs. Correctness errors will
-show up at evaluation anyway, or will not matter (dead code). With respect to
-validation, traditional types are too limited. Validation commonly involves
-checking requirements such as being a valid port number or a valid URL which
-are, while not technically impossible, really hard to check statically.
-
-In a dynamic language, it is natural and easier to validate data at runtime.
-This underlies the approach of dynamic typing. But dynamically typed languages
-usually only check the same simple predicates as static types (being a number,
-being a string, etc.), and solely for data that are arguments to a primitive
-operation. For validation, one needs to check arbitrary predicates on arbitrary
-values. This is provided in a basic form by assertions, available in most
-languages out there. An assertion is an expression that verifies that a
-predicate holds, or abort the program otherwise.  While assertions are fine for
-simple sanity checks, their use can incur verbosity, they are clumsy to compose
-and to reuse, and they provide limited context attached to an error. It turns
-out there is a better tool than plain assertions: contracts.
-
-\subsection*{Contracts}
-
-Assertions are usually mainly used for correctness by enforcing pre- and
-post-conditions of functions at runtime. In their foundational
-paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
-principled approach to run-time assertion checking called contracts. As opposed
-to plain assertions, contracts can be composed naturally into more complex ones.
-Nickel features such a contract system. For example, contracts nicely support
-higher-order functions. If the user has written an \nickel{Even} contract for
-numbers, they can use it right away to form a function contract \nickel{Even ->
-Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
-contract will check at each call that the argument is an even number
-(pre-condition) and that the return value is a integer (post-condition).
-Contracts also introduce the notion of blame which is crucial to good error
-reporting in presence of higher-order contracts and other composed contracts.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=Nickel]
-let Even = contracts.fromPred (fun val =>
-  builtins.isNum val
-  && val % 2 == 0) in
-let Int = contracts.fromPred (fun val =>
-  builtins.isNum val
-  && val % 1 == 0) in
-{divBy2 | Even -> Int = fun x => x/2}
-\end{lstlisting}
-  \end{center}
-\caption{Higher-order contract}
-\label{fig:contract-higher-order}
-\end{figure}
-
-In a dynamic language, even more so in a lazy one (see the paragraph on Lazyness
-in Section~\ref{sec:feat-lang}), error messages may point to a location far
-removed from the offending code. Contracts help by making assumptions on the
-values passed at component boundaries explicit (components can be functions,
-modules, classes, and so on, depending on language).
-
-Figure~\ref{fig:contract-reporting-wo} and Figure~\ref{fig:contract-reporting-w}
-give an example of a situation where adding a simple contract annotation
-\nickel{Str -> Str} to a function greatly improves the error message. In the
-original program of Figure~\ref{fig:contract-reporting-wo}, the
-\nickel{catHosts} function takes a string argument, append it to a list that is
-then folded. However, the programmer forgot one parameter when calling to
-\hbox{\nickel{makeHost},} thus passing a partially applied function instead of a string
-to \nickel{catHosts}. Because this argument is put in a list and folded, the
-error is raised later on and points to inside the code of \nickel{catHosts} and
-\nickel{makeHost}, which may be totally unknown to the caller.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
+functions were the original motivation for contracts).
+
+There is no Boolean function which can check that a value has type
+\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str}
+checks for each call of the function whether
+\begin{enumerate}
+\item the argument has type \nickel{Str}, otherwise the caller of the
+  function is faulty
+\item if so, that the returned value has type \nickel{Str}, otherwise
+  the implementation of the function is faulty
+\end{enumerate}
+
+Like in the case of lazy dictionaries, the checks are delayed.
+Contracts keeps track of whether the caller or the implementation is
+at fault for a violation, hence it can report precise error
+messages. Compare Figure~\ref{fig:contract-reporting-wo} and
+Figure~\ref{fig:contract-reporting-w}: in
+Figure~\ref{fig:contract-reporting-wo} an error is reported inside the
+\nickel{catHosts} function, but \nickel{catHosts} is, in fact,
+correct, as is made clear by Figure~\ref{fig:contract-reporting-wo},
+where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
+contract, and correctly reports that the caller failed to call
+\nickel{catHosts} with a string argument.
+
+\begin{figure*}
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \begin{lstlisting}[language=nickel]
 let catHosts = fun last =>
@@
 
-let makeHost = fun server ext => server ++ "." ++ ext in
+let makeHost = fun server ext =>
+  server ++ "." ++ ext in
 
@@
 
-\begin{lstlisting}[title={Error message},frame=none,numbers=none]
+    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
 error: Type error
@@
 \end{lstlisting}
-  \end{center}
     \caption{Error reporting without contract}
     \label{fig:contract-reporting-wo}
-\end{figure}
-
-By adding a simple annotation to \nickel{catHosts} in
-Figure~\ref{fig:contract-reporting-w}, the interpreter fails before even
-entering the body of catHosts and is able to deliver a clearer message: the
-first parameter of \nickel{catHosts} should be a string but the user provided
-the body of the function \nickel{makeHost}. A stack trace is also shown,
-including the faulty call to \nickel{catHost}.
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel,title={Added contract}]
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.48\linewidth}
+    \begin{lstlisting}[language=nickel]
 let catHosts | Str -> Str = fun last =>
-\end{lstlisting}
-
-\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
+  [(*@{\ldots}@*)]
+\end{lstlisting}
+\vspace{10ex}
+    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
 error: Blame error: contract broken by the caller.
@@
     \end{lstlisting}
-  \end{center}
     \caption{Error reporting with contract}
     \label{fig:contract-reporting-w}
-\end{figure}
+  \end{subfigure}
+  \caption{Contracts improve error messages}
+\end{figure*}
+
+\rule{3cm}{3pt}
+
+\unsure{Arnaud: What's below this point in the introduction is being reworked}
+
+\subsection*{Correct programs \& valid data}
+
+With respect to
+validation, traditional types are too limited. Validation commonly involves
+checking requirements such as being a valid port number or a valid URL which
+are, while not technically impossible, really hard to check statically.
+
+\subsection*{Contracts}
 

### 2021-06-04 paper.tex
Use port as user-defined contract example

@@
 and then use it as a contract for the \nickel{not} function, as can be seen on Figure
-\ref{fig:checked-not}.
+\ref{fig:port-contract}.\unsure{These few paragraphs are out of date,
+  even more so since Arnaud changed the flat contract example}
 
@@
 \begin{lstlisting}[language=nickel]
-let NotC = contracts.fromPred (
-  fun f => builtins.isFun f && (f true == false) && (f false == true)
-) in
-let checkedNot = not | #NotC
+let Port = contracts.fromPred (fun value =>
+  builtins.isNum value && value % 1 == 0
+  && value >= 0 && value <= 65353) in
+80 | Port
 \end{lstlisting}
 \end{center}
-\caption{A contract that checks the not function behaves as expected}
-\label{fig:checked-not}
+\caption{A contract for valid ports}
+\label{fig:port-contract}
 \end{figure}

### 2021-06-04 paper.tex
Continue reworking of section 4 [WIP]

@@
 
-Static typing is about enforcing that some property always holds for a
-potentially unknown value, no matter the execution part. A dynamic check, on the
-other hand, just need to verify a property for a specific and known value, which
-is usually much easier. For example, dependent function types are complex and
-mostly implemented in proof assistants (although they are slowly making their
-way in some mainstream languages), whereas the corresponding dependent function
-contracts can be implemented in a few line of code (as is done in
-Racket~\cite{}).
+Static typing is about enforcing that some property about a (potentially
+unknown) value always holds, no matter the execution path. A dynamic check, on
+the other hand, just need to verify a property for a specific and known value,
+which is usually much easier. For example, dependent function types are complex
+and mostly implemented in proof assistants (although they are slowly making
+their way in some mainstream languages), whereas the corresponding dependent
+function contracts can be implemented in a few line of code (they are
+implemented in e.g. Racket~\cite{}).
 
@@
 
-\subsection*{Union contracts as an effect}
-
-In a pure language with standard contracts, the failure of a function contract
-can always be blamed on a single call. Figure~\ref{fig:pos-to-pos} shows a
-function with a simple contract attached that ensures that both the argument and
-the result of \nickel{f} are positive numbers. \nickel{f} is called twice, and
-the second call violates the requirement on the codomain, leading to blame
-\nickel{f}. The first call to \nickel{f} does not matter, and \nickel{5} is a
-single and independent witness of the violation of the contract. The user can be
-pointed to one location in practice.
-
-There is an intuitive justification for this fact. Apart from the error
-reporting part (but this is the crucial in practice!), the current contract
-system of Nickel can be implemented purely as a library, just given a
-\nickel{fail} primitive to abort the execution. In particular, applying a
-function contract to \nickel{f} replaces it with a guarding function
-\nickel{fWrapped} that performs the required checks. Thus, in a pure setting,
-the failure of \nickel{fWrapped 5} must be independant of its environment, and
-in particular of the call \nickel{fWrapped 10}.
+\subsection*{Union contracts as a side-effect}
+
+In Nickel, the failure of a function contract can always be traced back to a
+single call. For example, take the function \nickel{f} with a simple contract
+attached of Figure~\ref{fig:pos-to-pos}. The program fails with a contract error
+blaming \nickel{f}. Indeed, the second call \nickel{f 5} fails because the
+return value violates the \nickel{Positive} contract. The first call to
+\nickel{f} does not matter, and \nickel{f 5} is a single and independent witness
+of the violation of the contract. The user can be pointed to one location in
+practice.
+
+This single witness property can be justified as follows. Apart from the error
+reporting part (although this is the crucial bit in practice!), the current
+contract system of Nickel can be implemented purely as a library, requiring only
+a \nickel{fail} primitive to abort the execution. In practice, applying a
+function contract to \nickel{f} replaces it with a \nickel{f'} that additionally
+performs the required checks. Thus, since the core language is pure (the
+presence of \nickel{fail} isn't relevant here), the failure of \nickel{f' 5}
+must be independent of its environment and of any previous call to \nickel{f'}.
 
@@
 
-Union contracts are a different matter. Consider the program presented in
-Figure \ref{fig:wrong-union-function}. \nickel{f} is again violating the
-contract, as it neither maps positive numbers to positive numbers nor positive
-numbers to negative numbers.
+Union contracts are different. Consider the program presented in
+Figure~\ref{fig:wrong-union-function}. The same \nickel{f} is now given a union
+contract. \nickel{f} is violating this contract once again, as it neither maps
+positive numbers to positive numbers nor positive numbers to negative numbers.
 
@@
 
-In a system with unions, this contract would fail, because \nickel{f 10}
-(returns \nickel{3}) is a witness of \nickel{f} failing the contract
-\nickel{Positive -> NonPositive}, and \nickel{f 5} (returns \nickel{-2}) is a
-witness of \nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
-the previous example, removing one call or the other will make the program pass!
-Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
-the union, while the second one only fails the \nickel{Positive -> NonPositive}
-part. Both are needed to prove the violation of the full contract.
-
-Even in a pure setting, unions introduce a form of effect at a distance: the
-result of \nickel{f 5} is not local anymore, as it now depends on the previous
-execution of other calls to \nickel{f}. Unions contracts break the
-equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}, and the
-candidate example of Figure~\ref{fig:optimizations-let-floating-ex} cannot be
+This program must fail, because \nickel{f 10} is a witness of \nickel{f} failing
+the contract \nickel{Positive -> NonPositive}, and \nickel{f 5} is a witness of
+\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
+Figure~\ref{fig:pos-to-pos}, removing only one the call makes the program
+succeed! Indeed, the first call only fails the \nickel{Positive -> NonPositive}
+part of the union, while the second one only fails the \nickel{Positive ->
+NonPositive} part. In this particular example, there can't be a single call that
+is a witness of the violation of the contract: a minimum of two are needed.
+
+This behavior indicates that union contracts introduce side-effects. The result
+of \nickel{f 5} now depends on the previous execution and more generally on any
+prior call to \nickel{f}. This behavior of union contracts break the
+equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}. The
+candidate example of Figure~\ref{fig:optimizations-let-floating-ex} can't be
 optimized in general.
 
-Figure~\ref{fig:optimized-programs} illustrates this point further. It comprises
-an original program, and one where the common subexpression \nickel{f 1} have
-been eliminated. These two programs behave very differently:
+Figure~\ref{fig:optimized-programs} illustrates this point further. There is an
+original program, and an optimized version the common subexpression \nickel{f 1}
+has been eliminated. While equivalent in a pure language without contracts,
+these two programs behave differently here:
+
 \begin{itemize}
@@
 
-since the single application of \nickel{f 1}, applied to both \nickel{true} and
-\nickel{false} will be able to find the error on the union contract
-\nickel{(Bool -> Num) @| (Bool -> Str)}.
+In the original version, each partial application \nickel{f 1} gives rise to a
+fresh instance of the contract \nickel{Bool -> Num @| Bool -> Str}. These
+instances are independent, and can pick a different component of the union to
+satisfy. Although \nickel{f} doesn't actually respect the contract, these calls
+are not enough to prove so. In the optimized version, \nickel{g} is endowed with
+a single contract, that must pick one of the two components of the union. There,
+the two calls refers to the same union contracts, and shows that \nickel{f}
+violate its initial contract.
 
@@
 
+To sum up, the addition of union contracts introduce side-effects in a pure
+language. Side-effects have well-known pitfalls: 
+\begin{itemize}
+    \item For the programmer, they are hard to reason about. They prevent local
+        reasoning. In our previous examples, removing or adding a function call
+        somewhere can toggle a failure in a call at a totally different
+        location.
+    \item For the interpreter (or compiler), they inhibit many optimizations and
+        program transformations.
+\end{itemize}
+
+% A valid argument against this point would be stating that further work should be done
+% on optimizers, so to not modify these tricky cases, even if this is true,
+% programmers also modify their code on a daily basis, and if moving from the first
+% code snippet at Figure \ref{fig:optimized-programs} to the second one changes
+% the behaviour of a program, we argue that union contracts not only go against the
+% implementation of languages, but also against their usability.
+%
+% Summing it up, wrapping a value with a union of higher order contracts means that
+% that value now has a state, that may get modified silently at every use site.
+% Both reviewed papers suffer from this defect since it is the semantics
+% of unions that interact badly with code manipulation operations,
+% not a particular implementation.
+% This addition of global state complicates both the work of the language
+% implementor, as well as the language user, since it adds non trivial
+% side effects to your code.
+
+\subsection*{Intersection of user-defined contracts}
+\label{sec:flat-and-inter}
+
+A natural - but naive - implementation of intersection contracts could be the
+following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
+and \nickel{B} sequentially. This intuition works for simple contracts: surely
+checking that \nickel{x | Natural @& Odd} amounts to check that \nickel{x |
+Natural} and \nickel{x | Odd}. Unfortunately, this doesn't scale to higher-order
+contracts.
+
 \unsure{WARNING: what is below is being reworked by Yann}
 
-To sum up, union contracts introduce a form of effect at a distance. This
-impacts both the programmer and the compiler or interpreter:
-\begin{itemize}
-    \item As for any side-effect, this breaks local reasoning.
-    \item It prevents optimizations.
-\end{itemize}
-\todo{WIP}
-
-A valid argument against this point would be stating that further work should be done
-on optimizers, so to not modify these tricky cases, even if this is true,
-programmers also modify their code on a daily basis, and if moving from the first
-code snippet at Figure \ref{fig:optimized-programs} to the second one changes
-the behaviour of a program, we argue that union contracts not only go against the
-implementation of languages, but also against their usability.
-
-Summing it up, wrapping a value with a union of higher order contracts means that
-that value now has a state, that may get modified silently at every use site.
-Both reviewed papers suffer from this defect since it is the semantics
-of unions that interact badly with code manipulation operations,
-not a particular implementation.
-This addition of global state complicates both the work of the language
-implementor, as well as the language user, since it adds non trivial
-side effects to your code.
-
-\subsection*{User defined contracts do not intersect}
-\label{sec:flat-and-inter}
-
-Stop for a second and think how you could implement an intersection
-contract. The first thought you may have is "just apply
-both contracts", as shown on the Implementation snippet
-from Figure \ref{fig:inter-contracts}.
-Of course, if it were that simple we wouldn't be writing this paper.
-This idea starts to show problems when you first combine it with
-higher order contracts, since it would mean not complying with
-at least one of these contracts negatively would entail breaking the
-whole thing, as shown on the second example on the same Figure,
+works for a lot of simple contracts: Stop for
+a second and think how you could implement an intersection contract. The first
+thought you may have is "just apply both contracts", as shown on the
+Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if it
+were that simple we wouldn't be writing this paper.  This idea starts to show
+problems when you first combine it with higher order contracts, since it would
+mean not complying with at least one of these contracts negatively would entail
+breaking the whole thing, as shown on the second example on the same Figure,
 that would fail.

### 2021-06-04 paper.tex
Use num.isInt in the port example
Arnaud done with the introduction

@@
 
-\rule{3cm}{3pt}
-
-\unsure{Arnaud: What's below this point in the introduction is being reworked}
-
-\subsection*{Correct programs \& valid data}
-
-With respect to
-validation, traditional types are too limited. Validation commonly involves
-checking requirements such as being a valid port number or a valid URL which
-are, while not technically impossible, really hard to check statically.
-
-\subsection*{Contracts}
-
-The examples of Figure~\ref{fig:contract-reporting-wo} and
-Figure~\ref{fig:contract-reporting-w} are about correctness. For validation, one
-of the main appeals of Nickel in our opinion is the ability to write a simple
-validation function for, say, a port number, and have it immediately available
-to construct other contracts by writing simple type-like expressions.  Another
-example is given in Figure~\ref{fig:contract-composition}. The first line
-defines a port contract from a predicate and use it in the definition of a
-larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
-contract checks that a value is a record with the following shape: it must have
-a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
-\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
-a \emph{schema} used to validate the final data, which is exactly what we were
-looking for in the previous section. Not only it plays the role of a schema, but
-it also \emph{looks} like one, as opposed to the corresponding plain validation
-function. The \lstinline+|+ operator entails that a value respects a contract,
-akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
-been defined in a similar way. \resolved{What's missing in this section is an
-example of how contracts differ from standard dynamic type checking on some
-function. Namely, the quality of error messages.}
-
-\begin{figure}[h]
-  \begin{center}
-\begin{lstlisting}[language=nickel]
-let Port = contracts.fromPred (fun value =>
-  builtins.isNum value && value % 1 == 0
-  && value >= 0 && value <= 65353) in
-let Configuration = {
-  host | Host,
-  port | Port,
-} in
-{host = "nixos.org", port = 80}
-| Configuration
-\end{lstlisting}
-  \end{center}
-\caption{Contract composition}
-\label{fig:contract-composition}
-\end{figure}
-
-Contracts have another perk: they correspond to the type casts of gradual
-typing, a discipline for mixing static and dynamic typing in the same language.
-As such, they are a precious ingredient for handling the interaction between 
-statically typed and dynamically typed code. While we focused on the writing of
-configuration, Nickel is a proper programming language that is also designed to
-write generic and reusable library code. There, the typing trade-off is
-different, and static typing is more adapted. Nickel has a gradual type system,
-which is yet another argument for having contracts. Let us end this section by
-stressing the following points:
-\begin{itemize}
-    \item \textbf{Even in a purely dynamically typed language, contracts are an
-        appealing feature for validation and correctness}. The gradual typing
-        part enters the general picture but is out of the scope of this paper.
-    \item \textbf{User-defined contracts are essential for validation}. Some
-        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
-        extensively discussed later in the paper) do not feature user-defined
-        contracts.  Rather, a fixed set of built-in contracts such as
-        \nickel{Num} or \nickel{Str} are provided, very much like base
-        types. The absence of user-defined contracts is understandable when
-        using contracts as a dynamic typing discipline for correctness. But a
-        system without user-defined contracts is limited in the same way as
-        static typing with respect to validation.
-\end{itemize}
-
 \subsection*{Structure of this paper}
+\unsure{Make this section into a contributions section?}
 Section~\ref{sec:intro} introduced the perspective of designing a configuration
@@
 
-\subsection*{User defined contracts}
+\subsection*{User-defined contracts}
 \label{sec:flat-contracts}
+\unsure{Make sure that this point is made in the user-defined
+  contracts section: With respect to
+validation, traditional types are too limited. Validation commonly involves
+checking requirements such as being a valid port number or a valid URL which
+are, while not technically impossible, really hard to check statically.}
 
@@
 \begin{lstlisting}[language=nickel]
-let Port = contracts.fromPred (fun value =>
-  builtins.isNum value && value % 1 == 0
-  && value >= 0 && value <= 65353) in
+let Port = contracts.fromPred (fun p =>
+  num.isInt p && 0 <= p && p <= 65353) in
 80 | Port

### 2021-06-06 paper.tex
Remove the \newpage-s
I missed a newpage
Small fixes

@@
 TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
-union typed to model the frequent practice to use the value
+union types to model the frequent practice to use the value
 \lstinline{null} (\lstinline{None} in Python) to represent an absent
@@
 abstract parts of the configuration, like traditional programming
-languages do. To meet this need, languages such as Cue~\cite{?},
-Dhall~\cite{?}, Jsonnet~\cite{?}, or Nickel~\cite{NickelRepo}, where
+languages do. To meet this need, languages such as Cue~\cite{cueLang},
+Dhall~\cite{dhallLang}, Jsonnet~\cite{jsonnetLang}, or Nickel~\cite{NickelRepo}, where
 configurations are generated rather than spelt out, were created.
@@
 \nickel{let id | C = value in exp} stands for \nickel{let id = (value
-  | C) in exp}.\unsure{This point used to be made, to we want to
+  | C) in exp}.\unsure{This point used to be made, do we want to
   reinsert it? However, Nickel programs are of a special kind: they are usually
@@
 \nickel{catHosts} function, but \nickel{catHosts} is, in fact,
-correct, as is made clear by Figure~\ref{fig:contract-reporting-wo},
+correct, as is made clear by Figure~\ref{fig:contract-reporting-w},
 where \nickel{catHosts} is decorated with the \nickel{Str -> Str}
@@
 
-\newpage
-
 \section{Features of languages}
@@
 
-\newpage
-
 \section{Union \& intersection}
@@
 \info{What are they, what are they used for.}
-\resolved{Since nickel uses | for contract application, I'd move unions
-to some other symbol.
-I think the best option is to move to $\cap / \cup$}
 
@@
 
-\change{These next few labels need to be inside figures,
-they are not working}
-\label{fig:union-adt-ex}
+\begin{figure}[h]
 \begin{lstlisting}[title={Tree-like data structure}]
@@
 \end{lstlisting}
+\caption{A binary tree data type using unions}
+\label{fig:union-adt-ex}
+\end{figure}
 
 In addition to internal program representations, unions are also useful to model
-external data:
+external data (\ref{fig:unions-use-cases}):
 \begin{itemize}
-    \item To represent nullable values (\ref{fig:union-nullable-ex}).
-    \item For plain enumerations (\ref{fig:union-enum-ex}).
-    \item To support different representations of the same data
-        (\ref{fig:union-alt-ex}).
+    \item To represent nullable values.
+    \item For plain enumerations.
+    \item To support different representations of the same data.
 \end{itemize}
@@
 
-\label{fig:union-nullable-ex}
+\begin{figure}[h]
+% \label{fig:union-nullable-ex}
 \begin{lstlisting}[title={Nullable values}]
@@
 
-\label{fig:union-enum-ex}
+% \label{fig:union-enum-ex}
 \begin{lstlisting}[title={Plain enumeration}]
-protocol : `Http @| `Ftp @| `Sftp
-\end{lstlisting}
-
-\label{fig:union-alt-ex}
+protocol : `Http | `Ftp | `Sftp
+\end{lstlisting}
+
+% \label{fig:union-alt-ex}
 \begin{lstlisting}[title={Alternative representations}]
@@
 \end{lstlisting}
+\caption{Various use cases of unions}
+\label{fig:unions-use-cases}
+\end{figure}
 
@@
 % dynamic approach, or typed approach too?
-
-\newpage
 
@@
 can be given the type \nickel{Num -> Num @& Bool -> Num}.
-
-\newpage
 
@@
 
-\newpage
-
 \section{Related work}
@@
 \info{Including a mention of statically typed systems with union}
-
-\newpage
 
@@
 contacts~\cite{BlameForAll}, dependent function
-contracts~\ref{DependentContracts}, and so on. For validation purpose, union and
+contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
 intersection contracts are really appealing. Unions have a intuitive

### 2021-06-06 paper.tex
Rework on sec 3

@@
 
+% Copied from https://tex.stackexchange.com/questions/89574/language-option-supported-in-listings
+\lstdefinelanguage{JavaScript}{
+  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
+  % keywordstyle=\color{blue}\bfseries,
+  ndkeywords={class, export, boolean, throw, implements, import, this},
+  ndkeywordstyle=\color{darkgray}\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  commentstyle=\color{purple}\ttfamily,
+  % stringstyle=\color{red}\ttfamily,
+  morestring=[b]',
+  morestring=[b]"
+}
+
 \lstset{
@@
 \newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
+\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
+\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
 
@@
 
+% Goals:
+% Introduce unions and inter
+% Why are they useful?
+% How are they used
+% How de they relate with one another
+
 \subsection*{Unions}
-Unions are fundamental building blocks of programs. A union holds a value whose
-shape is known to be in a fixed set of alternatives: for example, a value that
-can be either a boolean or a string. They correspond logically to disjunctions.
-Used in combination with products, they can represent arbitrary tree-like
-structures, allowing to model a large class of data
-(\ref{fig:union-adt-ex}).
-
-\begin{figure}[h]
-\begin{lstlisting}[title={Tree-like data structure}]
-type BinaryTree =
-Integer @| (BinaryTree, BinaryTree)
-\end{lstlisting}
-\caption{A binary tree data type using unions}
-\label{fig:union-adt-ex}
-\end{figure}
-
-In addition to internal program representations, unions are also useful to model
-external data (\ref{fig:unions-use-cases}):
+Unions ($\cup$) are a fundamental building block of programs.
+A union holds a value whose
+shape is known to be in a fixed set of alterniteves, without exposing
+which of those alternatives the value belongs to.
+For example, a value that could be either a Number or a String can be
+modelled using unions as \nickel{Num @| String}.
+They correspond logically to disjunctions.
+
+Statically typed languages, and more specifically functional programming
+languages, provide a similar abstraction, sums
+(also called tagged unions, or discriminated unions), which
+also allow to model a value whose shape conforms to a fixed set of
+alternatives, but requiring an explicit mark (or tag) that differentiates
+to which of these alternatives it belongs to.
+For example, consider the Haskell code presented on Figure 
+\ref{fig:safediv-haskell}, it's a safe division that will
+check if the dividend is 0, and if it is, return an
+error message.
+Since the function can return two different types, a sum
+(\haskell{Either}) is used, and on each
+of the if branches an explicit tag has to be used (either
+\haskell{Left} or \haskell{Right}).
+
+\begin{figure}[h]
+\begin{lstlisting}[language=Haskell]
+safeDiv :: Int -> Int -> Either String Int
+safeDiv x y = 
+  if y == 0 then 
+    Left "Division by 0 is forbidden" 
+  else
+    Right (div x y)
+\end{lstlisting}
+\caption{Safe division implemented in Haskell}
+\label{fig:safediv-haskell}
+\end{figure}
+
+This incurred verbosity certainly facilitates the work of static checks,
+like typechecking, but dynamically typed languages permit much more flexibility,
+and programmers don't usually expect to tag their values.
+That is why, as mentioned on the Introduction, when designing
+type systems for dynamically typed languages, unions become of particular
+interest.
+See the implementation of a safe division function on TypeScript
+(Figure \ref{fig:safediv-ts}).
+Neither of the if branches has to tag their return value, and the return
+type of the function is modelled using unions, \typescript{string | number}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=JavaScript]
+function safeDiv(x: number, y: number): string | number {
+  if (y == 0) {
+      return "Division by 0 is forbidden"
+  } else {
+      return x/y
+  }
+}
+\end{lstlisting}
+\caption{Safe division implemented in TypeScript}
+\label{fig:safediv-ts}
+\end{figure}
+
+% \begin{figure}[h]
+% \begin{lstlisting}[title={Tree-like data structure}]
+% type BinaryTree =
+% Integer @| (BinaryTree, BinaryTree)
+% \end{lstlisting}
+% \caption{A binary tree data type using unions}
+% \label{fig:union-adt-ex}
+% \end{figure}
+
+Unions are not only useful to model the outcome of alternative
+branches of computations, but they can be used to
+formalize very common patterns of data types:
 \begin{itemize}
-    \item To represent nullable values.
-    \item For plain enumerations.
+    \item To represent nullable, or optional, values.
+\begin{lstlisting}[language=nickel]
+optionalValue : Str @| Null
+\end{lstlisting}
+
+    \item For plain enumerations of literals.
+\begin{lstlisting}[language=nickel]
+protocol : "Http" @| "Ftp" @| "Sftp"
+\end{lstlisting}
+
     \item To support different representations of the same data.
+\begin{lstlisting}[language=nickel]
+person : {name : Str, age : Int}
+  @| {name : Str, dateOfBirth : Date}
+\end{lstlisting}
+
 \end{itemize}
@@
 
-\begin{figure}[h]
-% \label{fig:union-nullable-ex}
-\begin{lstlisting}[title={Nullable values}]
-optionalValue : Str @| Null
-\end{lstlisting}
-
-% \label{fig:union-enum-ex}
-\begin{lstlisting}[title={Plain enumeration}]
-protocol : `Http | `Ftp | `Sftp
-\end{lstlisting}
-
-% \label{fig:union-alt-ex}
-\begin{lstlisting}[title={Alternative representations}]
-person : {name : Str, age : Int}
-  @| {name : Str, dateOfBirth : Str}
-\end{lstlisting}
-\caption{Various use cases of unions}
-\label{fig:unions-use-cases}
-\end{figure}
-
-In statically typed languages, and more specifically in functional programming
-languages, sums (also called tagged unions, or discriminated unions) are usually
-preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
-syntactically separated by an associated discriminating label. Well-known
-programming languages with first-class tagged unions include Haskell, OCaml,
-Rust, Scala, PureScript, Elm, Swift and many more.
-
-On the other hand, in dynamically typed languages, it is customary to directly
-store different kinds of data in one variable without further ceremony. The
-runtime system needs to attach type tags to values anyway, which can usually be
-observed using functions like JavaScript's \verb+typeof+. Using such bare
-unions\resolved{If you can observe with typeof, then it's pretty much a tagged
-union, though the tag is implicit} has already been an idiom in dynamically
-typed language for a long time. Being able to type and handle these unions is an
-important aspect of the migration to a gradual type system. Indeed, a lot of
-\resolved{This is the first time discussing gradual type systems, maybe it should
-be introduced earlier, in particular how they relate to contracts
-(I think is mentioned on the appendix)}
-gradually typed languages either existed originally as dynamically typed
-language or aim at backward compatibility with a dynamically typed base
-language. This is why despite the presence of a static type system, gradually
-typed languages tend to favor bare unions types which is the typed counterpart
-of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
-
-In the case of Nickel, there is no pre-existing untyped language that we have to
-support, so we are freer of our choice. However, Nickel is a configuration
-language, whose contract and type system is intended to encode data schemas encountered in the wild. Since these schemas do use
-untagged unions, there is a strong incentive to support untagged unions at the
+% \begin{figure}[h]
+% % \label{fig:union-nullable-ex}
+% \begin{lstlisting}[title={Nullable values}]
+% optionalValue : Str @| Null
+% \end{lstlisting}
+
+% % \label{fig:union-enum-ex}
+% \begin{lstlisting}[title={Plain enumeration}]
+% protocol : `Http | `Ftp | `Sftp
+% \end{lstlisting}
+
+% % \label{fig:union-alt-ex}
+% \begin{lstlisting}[title={Alternative representations}]
+% person : {name : Str, age : Int}
+%   @| {name : Str, dateOfBirth : Str}
+% \end{lstlisting}
+% \caption{Various use cases of unions}
+% \label{fig:unions-use-cases}
+% \end{figure}
+
+% In statically typed languages, and more specifically in functional programming
+% languages, sums (also called tagged unions, or discriminated unions) are usually
+% preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
+% syntactically separated by an associated discriminating label. Well-known
+% programming languages with first-class tagged unions include Haskell, OCaml,
+% Rust, Scala, PureScript, Elm, Swift and many more.
+
+% On the other hand, in dynamically typed languages, it is customary to directly
+% store different kinds of data in one variable without further ceremony. The
+% runtime system needs to attach type tags to values anyway, which can usually be
+% observed using functions like JavaScript's \verb+typeof+. Using such bare
+% unions\resolved{If you can observe with typeof, then it's pretty much a tagged
+% union, though the tag is implicit} has already been an idiom in dynamically
+% typed language for a long time. Being able to type and handle these unions is an
+% important aspect of the migration to a gradual type system. Indeed, a lot of
+% \resolved{This is the first time discussing gradual type systems, maybe it should
+% be introduced earlier, in particular how they relate to contracts
+% (I think is mentioned on the appendix)}
+% gradually typed languages either existed originally as dynamically typed
+% language or aim at backward compatibility with a dynamically typed base
+% language. This is why despite the presence of a static type system, gradually
+% typed languages tend to favor bare unions types which is the typed counterpart
+% of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
+
+Nickel has no pre-existing untyped language that we have to
+support, so we are freer of our choice.
+However, Nickel is a configuration language, whose contract and type system
+is intended to encode data schemas encountered in the wild. Since these schemas do not use
+tags to differentiate different kinds of values, there is a strong incentive to support untagged unions at the
 type level, or at least, at the contract level.
 
+We have shown how values of a union are built, is worth mentioning
+how they are used.
+For instance, consider the TypeScript function on Figure \ref{fig:addElem},
+we want it to insert an element to a list that is either an
+\typescript{Array<string>} or an \typescript{Array<number>},
+what value could we fill \typescript{...} with for this to typecheck?
+Since we don't know what's the shape of \typescript{list},
+we can't know whether to push a \typescript{string} or a
+\typescript{number}.
+That is, when using a union, we have to treat it as if it where any one
+of the alternatives, in the case of the example, we should provide
+it with a value that is both a \typescript{string} and a \typescript{number}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=JavaScript]
+function addElem(list: Array<string> | Array<number>) {
+  list.push(...)
+}
+\end{lstlisting}
+\caption{Adding an element to a union of two arrays}
+\label{fig:addElem}
+\end{figure}
+
 \subsection*{Intersections}
 
-Intersections are less prevalent. An intersection is a value that belongs to
-two types at the same time. A simple example is a number literal \nickel{1}
+Intersections ($\cap$) are less prevalent. An intersection is a value that belongs to
+two types at the same time. An example is a number literal \nickel{1}
 which could be accepted as both an \nickel{Int} and a \nickel{Float} without
-having to perform conversions (Figure~\ref{fig:intersection-subytping-ex}). In
-this example, the intersection \nickel{Int @& Float} can be used as a subtype of
-both \nickel{Int} and \nickel{Float}.
-
-\label{fig:intersection-subytping-ex}
-\begin{lstlisting}[title={Value subtyping}]
-1 : Int @& Float
-\end{lstlisting}
-
-Intersections of higher-order types corresponding to overloading
-(Figure~\ref{fig:intersection-overloading-ex}).
-
+having to perform conversions.
+In that case, a value of type \nickel{Int @& Float} can be used as a value of
+both type \nickel{Int} and \nickel{Float}.
+
+% \label{fig:intersection-subytping-ex}
+% \begin{lstlisting}[title={Value subtyping}]
+% 1 : Int @& Float
+% \end{lstlisting}
+
+The real usability of intersections shows up when intersecting
+functions, since it can be used to encode overloading.
+For instance, take a look at Figure
+\ref{fig:intersection-overloading-ex}, where the function
+\nickel{duplicate} works both as a function to duplicate
+arrays, as well as a function to duplicate strings.
+This is particular useful when using unions, since it's a
+good way to express that a function can deal with different
+shapes of data.
+For instance, on the same figure, you can use \nickel{duplicate}
+over some value with type \nickel{[String] @| String}, and know
+that it will work correctly.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let duplicate | ([String] -> [String]) @& (String -> String) =
+  fun x => x ++ x in
+let text | [String] @| String = ... in
+duplicate text
+\end{lstlisting}
+\caption{Duplicating an array of Strings or a String}
 \label{fig:intersection-overloading-ex}
-\begin{lstlisting}[title={Overloading}]
-incr: (Int -> Int) @& (Float -> Float) =
-  fun x => x + 1
-\end{lstlisting}
-
-They can also be used to combine constraints (such as traits or interfaces), as
-done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
-
-\label{fig:intersection-mixins-ex}
-\begin{lstlisting}[language=Scala,title={Structural mixins}]
-Comparator: {cmp: T -> T -> Bool}
-Shower: {show: T -> String}
-
-showSmaller : (dict: Comparator & Shower) -> T -> T -> String
-\end{lstlisting}
-
-Finally, intersection naturally show up as the dual of unions. Thus, in a system
+\end{figure}
+
+% \label{fig:intersection-overloading-ex}
+% \begin{lstlisting}[title={Overloading}]
+% incr: (Int -> Int) @& (Float -> Float) =
+%   fun x => x + 1
+% \end{lstlisting}
+
+They can also be used to combine records \unsure{Better
+naming}. For instance, in Figure \ref{fig:intersection-record}
+two types are defined \nickel{Animal} and \nickel{Pet}, and
+a variable that is compatible with both types is declared, with
+type \nickel{Animal @& Pet}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let Animal = { species: Str, race: Str } in
+let Pet = { owner: Str, name: Str } in
+let myDog | Animal & Pet = 
+  { species = "Canis Lupus",
+    race = "Australian Cattle Dog",
+    owner = "Anonymous Author",
+    name = "Juno" }
+\end{lstlisting}
+\caption{An animal that is also a pet}
+\label{fig:intersection-record}
+\end{figure}
+
+% They can also be used to combine constraints (such as traits or interfaces),as
+% done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
+
+% \label{fig:intersection-mixins-ex}
+% \begin{lstlisting}[language=Scala,title={Structural mixins}]
+% Comparator: {cmp: T -> T -> Bool}
+% Shower: {show: T -> String}
+
+% showSmaller : (dict: Comparator & Shower) -> T -> T -> String
+% \end{lstlisting}
+
+Finally, intersection naturally show up as the dual of unions.
+Thus, in a system
 featuring unions, a union in a negative positions is a form of intersection
@@
 
-\label{fig:intersection-dual}
-\begin{lstlisting}[title={Intersection as the dual of unions}]
+\begin{figure}[h]
+\begin{lstlisting}[language = nickel]
 f : (Num @| Str) -> Bool
@@
 \end{lstlisting}
-
-\todo{Either here or the next section, show examples of unintuitive behaviour
+\caption{Intersection as the dual of unions}
+\label{fig:intersection-dual}
+\end{figure}
+
+\resolved{Either here or the next section, show examples of unintuitive behaviour
 of U and I. For example, I is not jos composition of contracts.
-And union has to be treated as any one of the options}
+And union has to be treated as any one of the options.
+Kindo of resolved for unions}
 % Things to settle: how to motivate them, how to compare them to ADTs. Purely

### 2021-06-07 paper.tex
Remove trailing spaces
Rework of section 4 (subsection on intersections)

@@
 unknown) value always holds, no matter the execution path. A dynamic check, on
-the other hand, just need to verify a property for a specific and known value,
+the other hand, just needs to verify a property for a specific and known value,
 which is usually much easier. For example, dependent function types are complex
@@
 their way in some mainstream languages), whereas the corresponding dependent
-function contracts can be implemented in a few line of code (they are
-implemented in e.g. Racket~\cite{}).
-
-The gist of the present paper is to bring evidence that in the case of union and
-intersection, the dynamic checking part is complex as well: \emph{unions and
-intersections are fundamentally hard}.
+function contracts can be implemented in a straightforward way (they are
+supported in e.g. Racket~\cite{}).
+
+The gist of the present paper is to bring evidence that in the case of unions
+and intersections, the dynamic checking part is complex as well: \emph{unions
+and intersections are fundamentally hard}.
 
@@
 single call. For example, take the function \nickel{f} with a simple contract
-attached of Figure~\ref{fig:pos-to-pos}. The program fails with a contract error
-blaming \nickel{f}. Indeed, the second call \nickel{f 5} fails because the
-return value violates the \nickel{Positive} contract. The first call to
-\nickel{f} does not matter, and \nickel{f 5} is a single and independent witness
-of the violation of the contract. The user can be pointed to one location in
-practice.
+attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
+error blaming \nickel{f} because the return value of the second call \nickel{f 5}
+violates the \nickel{Positive} contract. The first call to \nickel{f} does
+not matter, and \nickel{f 5} is a single and independent witness of the contract
+violation. The user is pointed to this one location in practice.
 
@@
 a \nickel{fail} primitive to abort the execution. In practice, applying a
-function contract to \nickel{f} replaces it with a \nickel{f'} that additionally
-performs the required checks. Thus, since the core language is pure (the
-presence of \nickel{fail} isn't relevant here), the failure of \nickel{f' 5}
+function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
+additional checks. Thus, since the core language is pure (forgetting the
+presence of \nickel{fail}, which isn't relevant), the failure of \nickel{f' 5}
 must be independent of its environment and of any previous call to \nickel{f'}.
@@
 the contract \nickel{Positive -> NonPositive}, and \nickel{f 5} is a witness of
-\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
-Figure~\ref{fig:pos-to-pos}, removing only one the call makes the program
-succeed! Indeed, the first call only fails the \nickel{Positive -> NonPositive}
-part of the union, while the second one only fails the \nickel{Positive ->
-NonPositive} part. In this particular example, there can't be a single call that
-is a witness of the violation of the contract: a minimum of two are needed.
+\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to the
+example from Figure~\ref{fig:pos-to-pos}, removing only one of the calls makes the
+program succeed! Indeed, each call only unveils the violation of one component
+of the union. In this example, a single call to \nickel{f} that would be the
+witness of the violation of the whole contract doesn't even exist: a minimum of
+two are always needed.
 
 This behavior indicates that union contracts introduce side-effects. The result
-of \nickel{f 5} now depends on the previous execution and more generally on any
-prior call to \nickel{f}. This behavior of union contracts break the
-equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}. The
-candidate example of Figure~\ref{fig:optimizations-let-floating-ex} can't be
-optimized in general.
-
-Figure~\ref{fig:optimized-programs} illustrates this point further. There is an
-original program, and an optimized version the common subexpression \nickel{f 1}
-has been eliminated. While equivalent in a pure language without contracts,
-these two programs behave differently here:
+of \nickel{f 5} now depends on the previous execution and more specifically on
+any prior call to \nickel{f}. This behavior of union contracts break the
+property~\ref{eq:cbn-expansion} introduced in Section~\ref{sec:optimizations},
+that is required to perform CSE-like optimizations in all generality. The
+candidate example of Figure~\ref{fig:optimizations-let-floating-ex} in
+Section~\ref{sec:optimizations} can't be optimized in general.
+
+Figure~\ref{fig:optimized-programs} illustrates this point further. It contains
+an original program and an optimized version where the common subexpression
+\nickel{f 1} has been eliminated. While equivalent in a pure language without
+contracts, these two programs behave differently because of unions:
 
@@
 a single contract, that must pick one of the two components of the union. There,
-the two calls refers to the same union contracts, and shows that \nickel{f}
+the two calls refer to the same union contract, and shows that \nickel{f} does
 violate its initial contract.
@@
         location.
-    \item For the interpreter (or compiler), they inhibit many optimizations and
+    \item For the interpreter (or compiler), side-effects inhibit many optimizations and
         program transformations.
@@
 
-\subsection*{Intersection of user-defined contracts}
+\subsection*{Intersection with user-defined contracts}
 \label{sec:flat-and-inter}
@@
 following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
-and \nickel{B} sequentially. This intuition works for simple contracts: surely
-checking that \nickel{x | Natural @& Odd} amounts to check that \nickel{x |
-Natural} and \nickel{x | Odd}. Unfortunately, this doesn't scale to higher-order
-contracts.
-
-\unsure{WARNING: what is below is being reworked by Yann}
-
-works for a lot of simple contracts: Stop for
-a second and think how you could implement an intersection contract. The first
-thought you may have is "just apply both contracts", as shown on the
-Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if it
-were that simple we wouldn't be writing this paper.  This idea starts to show
-problems when you first combine it with higher order contracts, since it would
-mean not complying with at least one of these contracts negatively would entail
-breaking the whole thing, as shown on the second example on the same Figure,
-that would fail.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
-x | A @& B (*@ $\simeq$ @*) (x | A) | B
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Problem]
+and \nickel{B} sequentially:
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
+decomposition}]
+M | A @& B (*@ $\simeq$ @*) (M | A) | B
+\end{lstlisting}
+
+This intuition works for simple contracts: checking that \nickel{x | Natural @&
+Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
+Unfortunately, this doesn't scale to higher-order contracts.
+Figure~\ref{fig:inter-contracts} illustrates the use of an intersection to model
+a simple overloading of the identity function. If we were to apply the naive
+equivalence above, the argument \nickel{1} would fail the \nickel{Str -> Str}
+contract. The naive decomposition actually implies
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none]
+(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
+\end{lstlisting}
+
+This isn't the correct semantics for overloading. With this implementation, our
+contract for identity from Figure~\ref{fig:inter-contracts} would always
+fail because no argument satisfies \nickel{Num @& Str}.
+
+% Stop for a second and think how you could implement an intersection contract.
+% The first thought you may have is "just apply both contracts", as shown on the
+% Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if
+% it were that simple we wouldn't be writing this paper.  This idea starts to
+% show problems when you first combine it with higher order contracts, since it
+% would mean not complying with at least one of these contracts negatively would
+% entail breaking the whole thing, as shown on the second example on the same
+% Figure, that would fail.
+
+\begin{figure}[h]
+% \begin{lstlisting}[language=nickel, title=Implementation]
+% x | A @& B (*@ $\simeq$ @*) (x | A) | B
+% \end{lstlisting}
+\begin{lstlisting}[language=nickel]
 let g | Num -> Num @& Str -> Str
@@
 \end{lstlisting}
-\caption{Implementations of intersection contracts: first iteration}
+\caption{In}
 \label{fig:inter-contracts}
@@
 
-Not all hope is lost, we could pass some extra shared state
-to each one of the sub contracts and let them only fail if the other
-side of the contract has failed as well, as shown in Figure
-\ref{fig:inter-contracts-2}, where the shared state is
-represented by a label \nickel{l}.
-This idea is still quite reasonable, for instance, when executing
-\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
-that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
-something like \nickel{1 | Str}.
-Of course, this fails, but before raising a violation, this contract could
-check if the other side of the intersection (\nickel{Num -> Num}) has
-also tried to raise negative blame on the same parameter ()\nickel{1}), since
-it hasn't, the contract would return normally, and would mark itself as failed.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Implementation]
+In an higher-order intersection contract, blame is raised when:
+\begin{description}
+    \item[Negative blame] The argument fails \emph{both} components.
+    \item[Positive blame] The function fails \emph{one} component that the
+        argument previously satisfied.
+\end{description}
+
+% Not all hope is lost, we could pass some extra shared state
+% to each one of the sub contracts and let them only fail if the other
+% side of the contract has failed as well, as shown in Figure
+% \ref{fig:inter-contracts-2}, where the shared state is
+% represented by a label \nickel{l}.
+% This idea is still quite reasonable, for instance, when executing
+% \nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
+% that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
+% something like \nickel{1 | Str}.
+% Of course, this fails, but before raising a violation, this contract could
+% check if the other side of the intersection (\nickel{Num -> Num}) has
+% also tried to raise negative blame on the same parameter ()\nickel{1}), since
+% it hasn't, the contract would return normally, and would mark itself as failed.
+
+% On the other hand, if you were to do something like
+% \nickel{g true}, that would fail also negatively the contract
+% \nickel{Str -> Str} and, when checking if \nickel{true}
+% is a correct parameter for \nickel{Num -> Num} it would check
+% if the other side of the intersection has failed, and since it
+% has, a violation would be raised and the whole program would
+% correctly fail.
+%
+% This idea, which is an oversimplification of the ideas
+% presented by Williams, Morris, and Wadler, seems to be spot on.
+
+To fix the naive implementation, the interpreter can share state between the
+sub-contracts, in order to decide if blame must be raised or not when a
+sub-contract fails:
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none]
 x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
 \end{lstlisting}
-\caption{Implementations of intersection contracts: second iteration}
-\label{fig:inter-contracts-2}
-\end{figure}
-
-On the other hand, if you were to do something like
-\nickel{g true}, that would fail also negatively the contract
-\nickel{Str -> Str} and, when checking if \nickel{true}
-is a correct parameter for \nickel{Num -> Num} it would check
-if the other side of the intersection has failed, and since it
-has, a violation would be raised and the whole program would
-correctly fail.
-
-This idea, which is an oversimplification of the ideas
-presented by Williams, Morris, and Wadler, seems to be spot on.
-However, it presents a major problem, it's not compatible with user defined contracts,
-as presented in Section \ref{sec:flat-contracts}.
-Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
-implementation discussed previously were to be used, we may end up with
-something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
-\nickel{l} represents the shared state.
-At this point, the intuitive thing to do is to check the contract \nickel{C}, by
-checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
-since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
-fails negatively.
+
+Shared state is represented by the label \nickel{l}. This is in essence the
+approach proposed by Williams, Morris, and Wadler in~\cite{RootCauseOfBlame}.
+
+% \begin{figure}[h]
+% \begin{lstlisting}[language=nickel, title=Implementation]
+% x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
+% \end{lstlisting}
+% \caption{Implementations of intersection contracts: second iteration}
+% \label{fig:inter-contracts-2}
+% \end{figure}
+
+% However, it presents a major problem, it's not compatible with user defined contracts,
+% as presented in Section \ref{sec:flat-contracts}.
+% Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
+% implementation discussed previously were to be used, we may end up with
+% something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
+% \nickel{l} represents the shared state.
+% At this point, the intuitive thing to do is to check the contract \nickel{C}, by
+% checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
+% since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
+% fails negatively.
+
+However, their approach has a major drawback: it is not straightforwardly
+compatible with user-defined contracts (introduced in Section
+\ref{sec:flat-contracts}). The issue is similar to our initial issue with higher-order
+contracts and the naive decomposition: user-defined contracts may apply
+functions and thus make a sub-contract of the intersection fail, but this
+failure shouldn't always result in raising blame. An example is given on Figure
+\ref{fig:inter-flat-contracts}. Decomposing using the shared state approach, we
+end up with:
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Stateful
+decomposition}]
+((fun x => x) | (Str -> Str)[l]) | C[l]
+\end{lstlisting}
+
+where \nickel{l} represents the shared state.  At this point,
+applying the \nickel{C} contract results in evaluating:
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none]
+((fun x => x) | (Str -> Str)[l]) 0 == 0.
+\end{lstlisting}
+
+Applying a function wrapped in a \nickel{Str -> Str} contract to \nickel{0}
+fails negatively. This is not the expected behavior, since the identity function
+does respect semantically both contracts. As opposed to built-in higher-order
+contracts, user-defined contracts are black-box from the interpreter's point of
+view, and it is thus not obvious how to extend the shared state approach to
+handle user-defined contracts. The authors of~\cite{RootCauseOfBlame}
+acknowledge this fact: they decide to simply drop user-defined contracts
+altogether.
 
@@
 \begin{lstlisting}[language=nickel]
-let C = fun f => (f 0) == 0 in
+let C = fun f => f 0 == 0 in
 let g | (String -> String) @& C
-      = (fun x => x) 
+      = fun x => x
 \end{lstlisting}
@@
 
-But this should not have happened, since in the program of Figure \ref{fig:inter-flat-contracts}
-\nickel{g} is never applied to \nickel{0}, in fact, it is never applied
-to anything.
-Checking if \nickel{C} has already failed, and marking \nickel{String -> String} as
-failed negatively on \nickel{l} is not enough, the contract \nickel{String -> String}
-should not have been used inside the evaluation of \nickel{C}, which
-adds an extra layer of complexity: contracts should be guarded, and execution
-depends on the shape of the surrounding evaluation context.
+Once again, intersection contracts introduce side-effects in the picture. What's
+more, these side-effects interact with user-defined contracts in a non-trivial
+way, while they are an important feature for validation.
+\unsure{(Yann) Should we talk about thiemann here, who do support user-defined
+contracts?}
+% But this should not have happened, since in the program of Figure
+% \ref{fig:inter-flat-contracts} \nickel{g} is never applied to \nickel{0}, in
+% fact, it is never applied to anything.  Checking if \nickel{C} has already
+% failed, and marking \nickel{String -> String} as failed negatively on \nickel{l}
+% is not enough, the contract \nickel{String -> String} should not have been used
+% inside the evaluation of \nickel{C}, which adds an extra layer of complexity:
+% contracts should be guarded, and execution depends on the shape of the
+% surrounding evaluation context.
+
+\unsure{WARNING: what is below is being reworked by Yann}
 

### 2021-06-07 paper.tex
Reworking of section 4 (last subsection on semantics, WIP)

@@
 following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
-and \nickel{B} sequentially:
-
+and \nickel{B} sequentially, resulting in the naive decomposition rule of
+Figure~\ref{fig:naiv-impl}.
+
+\begin{figure}[h]
+\label{fig:naive-impl}
 \begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
@@
 \end{lstlisting}
+\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Exchange law}]
+(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
+\end{lstlisting}
+\begin{lstlisting}[language=nickel,title={Overloaded identity}]
+let g | Num -> Num @& Str -> Str
+      = fun x => x in
+g 1
+\end{lstlisting}
+\caption{Naive implementation of intersection}
+\end{figure}
 
@@
 Odd} amounts to check that \nickel{x | Natural} and \nickel{x | Odd}.
-Unfortunately, this doesn't scale to higher-order contracts.
-Figure~\ref{fig:inter-contracts} illustrates the use of an intersection to model
-a simple overloading of the identity function. If we were to apply the naive
-equivalence above, the argument \nickel{1} would fail the \nickel{Str -> Str}
-contract. The naive decomposition actually implies
-
-\begin{lstlisting}[language=nickel,frame=none,numbers=none]
-(A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
-\end{lstlisting}
-
-This isn't the correct semantics for overloading. With this implementation, our
-contract for identity from Figure~\ref{fig:inter-contracts} would always
-fail because no argument satisfies \nickel{Num @& Str}.
+Unfortunately, this doesn't scale to higher-order contracts.  The overloaded
+identity example of Figure~\ref{fig:naive-impl} illustrates the use of an
+intersection to model a simple overloading of the identity function. If we were
+to apply the naive decomposition, the argument \nickel{1} would fail the
+\nickel{Str -> Str} contract and abort the execution. Perhaps the exchange rule
+given in Figure~\ref{fig:naiv-impl}, which is a direct consequence of the naive
+decomposition, illustrates the issue better. It is clear that this exchange law
+isn't the right semantics for overloading. With this law, the contract for
+overloaded identity of Figure~\ref{fig:naiv-impl} would always fail because
+no argument can satisfy \nickel{Num @& Str}.
 
@@
 % Figure, that would fail.
-
-\begin{figure}[h]
-% \begin{lstlisting}[language=nickel, title=Implementation]
-% x | A @& B (*@ $\simeq$ @*) (x | A) | B
-% \end{lstlisting}
-\begin{lstlisting}[language=nickel]
-let g | Num -> Num @& Str -> Str
-      = fun x => x in
-g 1
-\end{lstlisting}
-\caption{In}
-\label{fig:inter-contracts}
-\end{figure}
 
@@
 
+\subsection*{Challenging the semantics}
+
 \unsure{WARNING: what is below is being reworked by Yann}
 
-\subsection*{Union and intersection are not a silver bullet}
-
-Before wrapping up this section, we would like to show that maybe, union and intersection
-contracts are not exactly what we would want them to be.
-As stated throughout this work, programmers coming from dynamic languages
-are accustomed to perform safety checks (data validation) with
-great amounts of flexibility and the whole power of their
-language of choice at their disposal.
-
-For instance, a Python programmer may have no problem in checking
-at runtime, that a given variable is an integer if a given flag is
-not set, as shown in Figure \ref{fig:python-assertion};
-while a Haskell programmer might close her eyes and leave the room
-instantly.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=python]
-result = 42 if flag else "Disabled"
-...
-assert (
-  (not flag)
-  or isinstance(result, int)
-)
-\end{lstlisting}
-\caption{Asserting at runtime the type of a variable}
-\label{fig:python-assertion}
-\end{figure}
-
-These expectations are not really met by how unions and intersections
-contracts are supposed to work.
-For example, consider the Nickel code shown on Figure
-\ref{fig:intersection-distrbution}, at first sight, this code should
-complete execution successfully, returning \nickel{(1, 1)},
-however, under both the Keil and Thiemann, as well as the
-Williams, Morris, and Wadler works, this program will fail, with a violation
-of the contract imposed to \nickel{g}.
+Previous points of this section were about how implementing the expected
+semantics for union and intersection has a profound impact on the language. Let
+us end the section by investigating a different aspect: does the
+semantics correspond to the programmer's intuition?
+
+But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
+as clearly wrong with respect to overloading. There is nothing wrong about it
+per se : we could chose a semantics for intersections that validate this rule.
+However, the resulting system wouldn't capture overloading, and is arguably not
+very interesting nor useful. In~\ref{KeilThiemannUnionIntersection}, the authors
+argue for basing the semantics of union and intersection contracts on union and
+intersection type systems, where they have a simple and natural definition from
+which the other properties follow. This approach makes sense, and has been
+continued in future works, and in particular in~\ref{RootCauseOfBlame}.
+
+But even this semantics is no panacea. Figure~\ref{fig:intersection-distrbution}
+shows an example that intuitively looks like it should succeeed, returning
+\nickel{(1, 1)}. However, under the semantics used both
+in~\ref{KeilThiemannUnionIntersection} and~\ref{RootCauseOfBlame}, this program
+fails with a contract violation.
+
+% Before wrapping up the section, we would like to show that maybe, union and
+% intersection contracts are not exactly what we would want them to be.  As stated
+% throughout this work, programmers coming from dynamic languages are accustomed
+% to perform safety checks (data validation) with great amounts of flexibility and
+% the whole power of their language of choice at their disposal.
+%
+% For instance, a Python programmer may have no problem in checking at runtime,
+% that a given variable is an integer if a given flag is not set, as shown in
+% Figure \ref{fig:python-assertion}; while a Haskell programmer might close her
+% eyes and leave the room instantly.
+%
+% \begin{figure}[h]
+% \begin{lstlisting}[language=python]
+% result = 42 if flag else "Disabled"
+% ...
+% assert (
+%   (not flag)
+%   or isinstance(result, int)
+% )
+% \end{lstlisting}
+% \caption{Asserting at runtime the type of a variable}
+% \label{fig:python-assertion}
+% \end{figure}
+%
+% These expectations are not really met by how unions and intersections
+% contracts are supposed to work.
+% For example, consider the Nickel code shown on Figure
+% \ref{fig:intersection-distrbution}, at first sight, this code should
+% complete execution successfully, returning \nickel{(1, 1)},
+% however, under both the Keil and Thiemann, as well as the
+% Williams, Morris, and Wadler works, this program will fail, with a violation
+% of the contract imposed to \nickel{g}.
 
@@
 
-The reason this happens is that there is a fundamental property of how
-intersection contracts behave: the choice of which branch to choose is made
-exactly when the intersection is opened up, and not delayed as much as possible,
-as one would like.  This means, that the contract presented on Figure
-\ref{fig:intersection-distrbution}, \nickel{Num -> Num -> Num @& Num
--> Bool -> Num}, will not behave the same way that \nickel{Num ->
-(Num -> Num @& Bool -> Num)}, when, with an intuitive view of how
-intersections should behave, these two expressions should be logically
-equivalent.
-
-It is worth noting, that intersection types do have this property,
-since subtyping would allow to correctly type check a term like
-\nickel{g 1} as both \nickel{Num -> Num} as well as
-\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
-can be given the type \nickel{Num -> Num @& Bool -> Num}.
+This discrepancy is inherent to the chosen semantics for intersections.  When
+applying a function with an intersection contract, only one fixed component of
+the intersection can be selected for the function.
+one components of the intersection. Not that it must necessarily known right at the
+application, but the result can't be used
+functions, the choice of the component is fixed once for all when the function
+is applied. In Figure~\ref{fig:intersection-distrbution}, the contract for
+\nickel{h} can be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but can't
+be both. This is a manifestation of the following more general fact: the
+distributivity law given in Figure~\ref{fig:distributivity-law} is not valid in
+the semantics, while it is both intuitive and sound.\unsure{What means sound, in
+this context?}
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+A -> B @& A -> C (*@ $\simeq$ @*) A -> (B @& C)
+\end{lstlisting}
+\caption{Distributivity law}
+\label{fig:distributivity-law}
+\end{figure}
+
+%
+% It is worth noting, that intersection types do have this property,
+% since subtyping would allow to correctly type check a term like
+% \nickel{g 1} as both \nickel{Num -> Num} as well as
+% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
+% can be given the type \nickel{Num -> Num @& Bool -> Num}.
+%
 

### 2021-06-07 paper.tex
Last bit of reworking section 4

@@
 and \nickel{B} sequentially, resulting in the naive decomposition rule of
-Figure~\ref{fig:naiv-impl}.
-
-\begin{figure}[h]
-\label{fig:naive-impl}
+Figure~\ref{fig:naive-impl}.
+
+\begin{figure}[h]
 \begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
@@
 \caption{Naive implementation of intersection}
+\label{fig:naive-impl}
 \end{figure}
@@
 \nickel{Str -> Str} contract and abort the execution. Perhaps the exchange rule
-given in Figure~\ref{fig:naiv-impl}, which is a direct consequence of the naive
+given in Figure~\ref{fig:naive-impl}, which is a direct consequence of the naive
 decomposition, illustrates the issue better. It is clear that this exchange law
 isn't the right semantics for overloading. With this law, the contract for
-overloaded identity of Figure~\ref{fig:naiv-impl} would always fail because
+overloaded identity of Figure~\ref{fig:naive-impl} would always fail because
 no argument can satisfy \nickel{Num @& Str}.
@@
 
-\unsure{WARNING: what is below is being reworked by Yann}
-
 Previous points of this section were about how implementing the expected
@@
 But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
-as clearly wrong with respect to overloading. There is nothing wrong about it
-per se : we could chose a semantics for intersections that validate this rule.
-However, the resulting system wouldn't capture overloading, and is arguably not
-very interesting nor useful. In~\ref{KeilThiemannUnionIntersection}, the authors
-argue for basing the semantics of union and intersection contracts on union and
-intersection type systems, where they have a simple and natural definition from
-which the other properties follow. This approach makes sense, and has been
-continued in future works, and in particular in~\ref{RootCauseOfBlame}.
-
-But even this semantics is no panacea. Figure~\ref{fig:intersection-distrbution}
+as "clearly wrong" with respect to overloading. There is nothing wrong per se :
+we could chose a semantics for intersections that validate this rule.  However,
+the resulting system wouldn't capture overloading, and is arguably not very
+interesting nor useful. In~\cite{KeilThiemannUnionIntersection}, the authors
+argue for basing the semantics of contracts on union and intersection type
+systems, which have a simple and natural definition from which the other
+properties follow. This approach makes sense, and has been continued in future
+works, and in particular in~\cite{RootCauseOfBlame}.
+
+But even this semantics is no panacea. Figure~\ref{fig:intersection-distribution}
 shows an example that intuitively looks like it should succeeed, returning
 \nickel{(1, 1)}. However, under the semantics used both
-in~\ref{KeilThiemannUnionIntersection} and~\ref{RootCauseOfBlame}, this program
+in~\cite{KeilThiemannUnionIntersection} and~\cite{RootCauseOfBlame}, this program
 fails with a contract violation.
@@
 \caption{Intersection contracts don't distribute}
-\label{fig:intersection-distrbution}
-\end{figure}
-
-This discrepancy is inherent to the chosen semantics for intersections.  When
-applying a function with an intersection contract, only one fixed component of
-the intersection can be selected for the function.
-one components of the intersection. Not that it must necessarily known right at the
-application, but the result can't be used
-functions, the choice of the component is fixed once for all when the function
-is applied. In Figure~\ref{fig:intersection-distrbution}, the contract for
-\nickel{h} can be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but can't
-be both. This is a manifestation of the following more general fact: the
-distributivity law given in Figure~\ref{fig:distributivity-law} is not valid in
-the semantics, while it is both intuitive and sound.\unsure{What means sound, in
-this context?}
+\label{fig:intersection-distribution}
+\end{figure}
+
+This discrepancy is inherent to the chosen semantics .  When applying a function
+with an intersection contract, only one component of the intersection must be
+selected for the function. The actual choice of component may be delayed to
+later, but the point is that it must be the same across all usages. In
+Figure~\ref{fig:intersection-distribution}, the contract for \nickel{h} can in
+consequence be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but not both.
+This is a manifestation of the following more general defect: the distributivity
+law given in Figure~\ref{fig:distributivity-law} is not valid in the semantics
+given in~\cite{KeilThiemannUnionIntersection}, while it is both sound to add and
+intuitive.
 
@@
 
-%
-% It is worth noting, that intersection types do have this property,
-% since subtyping would allow to correctly type check a term like
-% \nickel{g 1} as both \nickel{Num -> Num} as well as
-% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
-% can be given the type \nickel{Num -> Num @& Bool -> Num}.
-%
+Curiously, union and intersection type systems, on which
+\cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
+exhibit this property. Subtyping allows to correctly type check a term like
+\nickel{g 1} as both \nickel{Num -> Num} as well as
+\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
+can be given the type \nickel{Num -> Num @& Bool -> Num}.
+\unsure{(Yann) I just reworked the wording, but I'm not totally sold on the
+argument}
 
@@
 \info{The various papers and the tradeoffs they make}
+
+\unsure{WARNING: what is below is being reworked by Yann}
 

### 2021-06-07 paper.tex
Section 2: preamble and user-defined contracts

@@
 
+Unfortunately, the delayed check of contract, while essential to
+ensuring that schema validation doesn't affect performance (or, indeed,
+is possible at all on functions), make union contracts (and their less
+appreciated sibling, intersection contracts) quite problematic.
+
 \subsection*{Structure of this paper}
@@
 
-\section{Features of languages}
+\section{A typology of language features}
 \label{sec:feat-lang}
@@
 
-Programming language research no longer lives on the abstract world
-of ideal surfaces.
-Every addition needs to be compatible with years of research
-and development, and this is hard.
-Success was unavoidable.
-
-In this section we briefly introduce some aspects of real world
-programming languages that will cause trouble while interacting
-with union and intersection dynamic checks.
+Union contracts are not only difficult to implement, their
+unrestricted presence is incompatible with potentially desirable
+properties of the language. In this section we present some of these
+properties; we will show how these properties interact with union
+contracts in Sections~\ref{sec:issues-sem}
+and~\ref{sec:issues-literature}.
 
@@
 \label{sec:flat-contracts}
-\unsure{Make sure that this point is made in the user-defined
-  contracts section: With respect to
-validation, traditional types are too limited. Validation commonly involves
-checking requirements such as being a valid port number or a valid URL which
-are, while not technically impossible, really hard to check statically.}
-
-User defined contracts, sometimes called flat contracts, are a particular
-kind of contract that behaves similar to an assertion, and whose
-main utility is to apply some arbitrary predicate (as arbitrary as the
-host language allows it) to a value; if the predicate evaluates to
-true, the contract passes, otherwise it fails.
-
-For instance, the contract \nickel{Even} shown on Figure
-\ref{fig:contract-higher-order} is a user defined contract, defined
-by taking the value that needs to be checked, and validating
-that is both a number and that the remainder of dividing it by 2
-is 0.
-And then, as seen on the same figure, we can use it to check that a function always
-returns an \nickel{Int} when applied to an \nickel{Even}.
-
-% \begin{figure}[h]
-% \begin{center}
-% \begin{lstlisting}[language=nickel]
-% let f = fun n => n * 2 @ (Integer -> even)
-% \end{lstlisting}
-%   \end{center}
-% \caption{A function wrapped with a contract stating that it always returns
-% an even number}
-% \label{fig:num-to-even}
-% \end{figure}
-
-Even more interesting, we could come up with a contract that checks that a function
-maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
-by applying the function exhaustively and checking its results,
-and then use it as a contract for the \nickel{not} function, as can be seen on Figure
-\ref{fig:port-contract}.\unsure{These few paragraphs are out of date,
-  even more so since Arnaud changed the flat contract example}
+
+A strength of dynamic checking is that we can easily check properties which
+are impractical to check statically. For instance that a string
+represents a well-formed URL, or a number is a valid port.
+
+This same property is desirable of contracts as well, otherwise we
+lose an important benefits of dynamic checking. Preferably, we want to
+be able to extend the universe of contracts with user-defined
+predicates (sometimes called flat contracts).
+
+For instance, Figure~\ref{fig:port-contract} shows the definition of a
+contract for valid ports in Nickel syntax.
 
@@
 let Port = contracts.fromPred (fun p =>
-  num.isInt p && 0 <= p && p <= 65353) in
+  num.isInt p && 0 <= p && p <= 65535) in
 80 | Port
@@
 \end{figure}
+
+User-defined contracts can be combined with other contracts normally,
+for instance \nickel{Int -> Port} is a contract verified by functions
+which, given an integer returns a valid port.
 
@@
 functions and as invariants on classes\cite{EiffelDesignByContract}.
-\resolved{I (Arnaud) believe Eiffel has this feature as well,
-  doesn't it? (Teo) You're right},
-
-% A somewhat newer language, the D programming
-% language, allows, among other things, to have invariants on
-% classes
-% written with flat contracts, that are
-% dinamically checked at the frontiers of every public method \cite{DLangContracts}.
 
 The Racket programming language also has a system to work with
-contracts, powerful enough to define flat contracts
-\footnote{And a gradual typing system
-built on top of that, but that goes beyond the scope of this
-work.}
-, and
+contracts, powerful enough to define flat contracts, and
 to compose them with other kinds of dynamic checks,
@@
 and intersection contracts\cite{RacketContracts}.
-
-Even if user defined contracts may seem like too permissive, the ability to make
-non trivial checks on values (that is, to validate their output) is expected
-by programmers on dynamic languages, where testing and runtime
-assertions are extremely useful tools for safe programming, and
-programmers are accustomed to express complex predicates
-over values.
 

### 2021-06-07 paper.tex
Sections 2&3: superficial (spelling, etc…) pass

@@
 
-\subsection*{Code manipulation (optimizations)}
+\subsection*{Purity}
 \label{sec:optimizations}
@@
 Purity and immutability as two important factors.}
+
+\unsure{I (Arnaud) changed the title of this subsection. Maybe it's
+  not the best one, but if it is, then we ought to reorder the section
+  a little bit to fit the modified narrative.}
 
@@
 let-floating~\cite{letFloating} apply the same principle of extracting out an invariant
-\unsure{Was this the inteded citation?}
 expression to avoid its recomputation respectively across loop iterations and
@@
 
-\unsure{Should we tease already here that unions and intersections break this
-even for pure terms?}
-
-\subsection*{Lazyness}
-\todo{Lazyness does not appear on sec 4, if we don't put somethign there, I'd remove it}
-Evaluation strategies relates to the evaluation of the expression bound to
-identifiers. Expressions are bound by assignments (e.g. let bindings in
-Nickel), or by function applications (arguments are bound to parameters). We
-will use Figure~\ref{} as a working example.
-
-The eager strategy consists in evaluating right away a bound value before
-storing it in an identifier. If an identifier is not used, this leads to
-unnecessary work: in Figure~\ref{}, the expression \nickel{fibonacci 100} would
-be computed although it is never used. The lazy strategy, on the other hand,
-only evaluates expression if they are needed. In the lazy model, bound
-expressions are stored as in their current unevaluated form, also called a
-\emph{thunk}. Thunks are only evaluated at their first usage, and their value is
-memoized for future usages.
-
-The eager strategy has been chosen in the vast majority of languages: while
-theoretically less efficient, it is simpler to implement and more predictable
-for the programmer. Also, lazy evaluation can occur performance overhead and
-memory leaks in unexpected ways. While eager evaluation is the default, some
-languages however offer a form of lazy evaluation, either as an opt-in option
-(e.g. Ocaml and ReScript), or by default for specific primitives types such as
-list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
-(and its web avatar PureScript). Although rare in general purpose languages,
-lazy evaluation is more widespread in the configuration space: the configuration
-languages Dhall, CUE and Jsonnet also implements full lazy evaluation. So are
-the Nix language, which inspired Nickel, and Nickel itself.
-
 \section{Union \& intersection}
@@
 A union holds a value whose
-shape is known to be in a fixed set of alterniteves, without exposing
+shape is known to be in a fixed set of alternatives, without exposing
 which of those alternatives the value belongs to.
@@
 
-They can also be used to combine records \unsure{Better
-naming}. For instance, in Figure \ref{fig:intersection-record}
+They can also be used to combine dictionaries in an object-oriented fashion. For instance, in Figure \ref{fig:intersection-record}
 two types are defined \nickel{Animal} and \nickel{Pet}, and

### 2021-06-07 paper.tex
Add numbering to all the subsections

@@
 
-\subsection*{Configuration languages}
+\subsection{Configuration languages}
 
@@
 
-\subsection*{Nickel}
+\subsection{Nickel}
 
@@
 
-\subsection*{Contracts}
+\subsection{Contracts}
 
@@
 
-\subsection*{Structure of this paper}
+\subsection{Structure of this paper}
 \unsure{Make this section into a contributions section?}
@@
 
-\subsection*{User-defined contracts}
+\subsection{User-defined contracts}
 \label{sec:flat-contracts}
@@
 
-\subsection*{Purity}
+\subsection{Purity}
 \label{sec:optimizations}
@@
 
-\subsection*{Unions}
+\subsection{Unions}
 Unions ($\cup$) are a fundamental building block of programs.
@@
 
-\subsection*{Intersections}
+\subsection{Intersections}
 
@@
 
-\subsection*{Union contracts as a side-effect}
+\subsection{Union contracts as a side-effect}
 
@@
 function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
-additional checks. Thus, since the core language is pure (forgetting the
-presence of \nickel{fail}, which isn't relevant), the failure of \nickel{f' 5}
+additional checks. Thus, since the core language is pure (albeit
+partial, if only because \nickel{fail}), the failure of \nickel{f' 5}
 must be independent of its environment and of any previous call to \nickel{f'}.
@@
 \begin{lstlisting}[language=nickel, title=Optimized]
-let f | Num -> (Bool -> Num @| Bool -> Str)
+let f | Num -> (Bool -> Num @| Bool -> MStr)
       = fun x y => if y then x else "False"
@@
 
+\unsure{These arguments should go in the purity section}
 To sum up, the addition of union contracts introduce side-effects in a pure
@@
 
-\subsection*{Intersection with user-defined contracts}
+\subsection{Intersection with user-defined contracts}
 \label{sec:flat-and-inter}
@@
 
-\subsection*{Challenging the semantics}
+\subsection{Challenging the semantics}
 
@@
 
-\subsection*{Keil and Thiemann}
+\subsection{Keil and Thiemann}
 \label{sec:keil-thiemann}
@@
 
-\subsection*{Williams, Morris, and Wadler}
+\subsection{Williams, Morris, and Wadler}
 \label{sec:will-morr-wadl}
@@
 
-\subsection*{Racket}
+\subsection{Racket}
 \label{sec:racket}
@@
 
-\subsubsection*{case->}
+\subsubsection{case->}
 

### 2021-06-07 paper.tex
Superficial pass on sections 4 & 5

@@
 \begin{lstlisting}[language=nickel, title=Optimized]
-let f | Num -> (Bool -> Num @| Bool -> MStr)
+let f | Num -> (Bool -> Num @| Bool -> Str)
       = fun x y => if y then x else "False"
@@
 \begin{description}
-    \item[Negative blame] The argument fails \emph{both} components.
-    \item[Positive blame] The function fails \emph{one} component that the
+    \item[Faulty caller] The argument fails \emph{both} components.
+    \item[Faulty implementation] The function fails \emph{one} component that the
         argument previously satisfied.
@@
 \begin{lstlisting}[language=nickel]
-let C = fun f => f 0 == 0 in
+let C = contracts.fromPred (fun f =>
+  f 0 == 0) in
 let g | (String -> String) @& C

### 2021-06-08 paper.tex
CCS and keywords

@@
 
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10002944.10011122.10002945</concept_id>
+       <concept_desc>General and reference~Surveys and overviews</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011008.10011024</concept_id>
+       <concept_desc>Software and its engineering~Language features</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011074.10011099</concept_id>
+       <concept_desc>Software and its engineering~Software verification and validation</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[300]{General and reference~Surveys and overviews}
+\ccsdesc[500]{Software and its engineering~Language features}
+\ccsdesc[300]{Software and its engineering~Software verification and validation}
+
+\keywords{Dynamic languages, Contracts, Union, Intersection}
+
 \maketitle

### 2021-06-08 paper.tex
Simplifying conclusion

@@
 The design and implementation of programming languages is a challenging task. It
-is a delicate exercise to balance complexity, expressiveness, efficiency and
+is a delicate exercise to balance complexity, expressivity, efficiency, and
 user experience all together. Unfortunately for designers, a language cannot
@@
 
-Part of recent developments revolve around mixing static typing and dynamic
-typing via gradual typing, in order to get the best of both worlds. Contracts
-(or casts in the gradual typing terminology) are crucial to this endeavour by
-allowing for a safe interaction between typed and untyped code. But we also made
-a case that contracts are interesting in themselves in a fully dynamically typed
+Part of recent developments revolves around improving safety of
+languages, ideally with little to none effort required by the programmers.
+Contracts are crucial to this endeavour, they are fundamental to
+allow a safe interaction between typed and untyped code in gradually typed languages;
+but they are also interesting in themselves in a fully dynamically typed
 language (and even in a fully statically typed language). Contracts are a
-precious idiom for data validation or to check pre- and post-condition that
-would be hard to ensure statically. In spirit, contracts are similar to
-assertions, but they are part of a principled framework, they are composable,
-they have good error reporting and they are integrated as a language feature. In
-particular, contracts natively handles higher-order functions and built-in data
-structures. This makes them an great candidate for integration in the Nickel
-configuration language.
-
-This is only the beginning of the story. A language designer then has to choose
-what contract combinators to provide among a number of possibilities: there is
-the standard package including higher-order function contracts, record contracts
-and list contracts.  Then, one could add polymorphic
-contacts~\cite{BlameForAll}, dependent function
-contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
-intersection contracts are really appealing. Unions have a intuitive
-interpretation as modeling alternatives and thus appear implicitly or explicitly
-in a lot of data structures.
-
-Alas, the implementation of union and intersection contracts is not as simple as
+precious idiom for data validation or to enforce runtime pre- and post-condition that
+would be hard to ensure statically.
+
+% In spirit, contracts are similar to
+% assertions, but they are part of a principled framework, they are composable,
+% they have good error reporting and they are integrated as a language feature. In
+% particular, contracts natively handles higher-order functions and built-in data
+% structures. This makes them an great candidate for integration in the Nickel
+% configuration language.
+
+% This is only the beginning of the story. A language designer then has to choose
+% what contract combinators to provide among a number of possibilities: there is
+% the standard package including higher-order function contracts, record contracts
+% and list contracts.  Then, one could add polymorphic
+% contracts~\cite{BlameForAll}, dependent function
+% contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
+% intersection contracts are really appealing. Unions have a intuitive
+% interpretation as modeling alternatives and thus appear implicitly or explicitly
+% in a lot of data structures.
+
+When trying to enforce properties on dynamic languages, unions appear naturally,
+since they allow to express that a variable is in an unknown state, but within
+some bounded set of alternatives.
+Sadly, the implementation of union and intersection contracts is not as simple as
 it first appears. In conjunction with higher-order contracts, or even just basic

### 2021-06-08 paper.tex
Continue reworking section 5 (Keil&Thiemann)

@@
 \begin{lstlisting}[language=nickel]
-let C = contracts.fromPred (fun f =>
+let c = contracts.frompred (fun f =>
   f 0 == 0) in
-let g | (String -> String) @& C
+let g | (string -> string) @& c
       = fun x => x
 \end{lstlisting}
-\caption{Intersection and user defined contracts}
+\caption{intersection and user defined contracts}
 \label{fig:inter-flat-contracts}
@@
 
-\unsure{WARNING: what is below is being reworked by Yann}
-
-The existing work that previously researched and implemented
-union and intersection
-as dynamic checks have, in our eyes, different shortcomings
-that makes them suboptimal solutions.
-In this section, we informally present those issues,
-starting with the foundational paper by Keil and Thiemann
-\cite{KeilThiemannUnionIntersection}, then following to a recent
-work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
-finally going over some similar contracts available in the Racket
-programming language.
+In this section, we review some shortcomings of existing work on union and
+intersection contracts. We start with the work of Keil and
+Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
+convincing semantics for union and intersection contracts. We follow with the
+work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which builds
+upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
+uniform blame resolution.
+
+% The existing work that previously researched and implemented
+% union and intersection
+% as dynamic checks have, in our eyes, different shortcomings
+% that makes them suboptimal solutions.
+% In this section, we informally present those issues,
+% starting with the foundational paper by Keil and Thiemann
+% \cite{KeilThiemannUnionIntersection}, then following to a recent
+% work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
+% finally going over some similar contracts available in the Racket
+% programming language.
 
@@
 \unsure{Subsection title?}
-Keil and Thiemann noticed that, since union contracts should share information between
-different application contexts, while intersection contracts should behave
-independently between each application context, the system becomes much
-simpler if every contract gets transformed into a union of intersections
-of other contracts, by using the common De Morgan laws,
-and then eagerly opening up unions, but postponing intersection
-contracts as much as possible.
-\todo{This assumes the difficulty has been previously explained}
-
-For instance, if someone were to wrap a value \nickel{x}
-in a contract as shown on the Original code snippet on Figure
-\ref{fig:kt-contract-transf}, then, before actually executing that
-contract, it would get transformed, and intersections would
-be pushed inside the unions, as can be seen on the Transformed
-code snippet on the same figure.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Original]
-x | A @& (B @| C)
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Transformed]
-x | (A @& B) | (A @& C)
-\end{lstlisting}
-\caption{Contract transformation performed by the Keil and
-Thiemann work}
-\label{fig:kt-contract-transf}
-\end{figure}
-
-This straightforward idea makes composing the Keil and Thiemann
-solution with other existing systems more difficult; in the words
-of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
-``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
-and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
-special cases can hinder composition.''\cite{RootCauseOfBlame}.
-\info{(Yann) Does
+
+In Section~\ref{sec:issues-sem}, we've seen that different calls to a function
+with an union contract must share information: the behavior of one call is
+influenced by the previous ones, as the function must pick one component of the
+union to satisfy across all usages. Conversely, following the semantics of
+overloading, each application of a function with an intersection contract can
+select a different branch and is thus independent from the others. A general
+contract composed of nested unions, intersections and higher-order contracts
+appears to require complex book-keeping in order to correctly raise blame.
+
+An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
+these nested contracts using the De Morgan law of
+Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
+intersection as much as possible.  \todo{This assumes the difficulty has been
+previously explained} This transformation flattens arbitrarily nested unions and
+intersections to a union of intersections, making blame resolution more
+manageable.\unsure{(Yann) find better (more precise) wording}
+
+\begin{figure}[h]
+\begin{lstlisting}[frame=none, numbers=none, language=nickel]
+A @& (B @| C) (*@ $\simeq$ @*) (A @& B) @| (A @& C)
+\end{lstlisting}
+\caption{De Morgan law for contracts}
+\label{fig:de-morgan-law}
+\end{figure}
+
+To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
+specific reduction rules for unions and intersections which perform this
+rewriting on the fly. The result is that the reduction rules sometimes feel
+ad-hoc, and make us question the composability and extensibility of the system.
+In the words of Williams, Morris, and Wadler (see
+Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
+contracts of intersection and union types given by Keil and Thiemann are not
+uniform. (\ldots) If uniformity helps composition, then special cases can hinder
+composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
+that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
+potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
+re-evaluated in each branch of the right hand side of
+Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
+mechanism.
+\resolved{(Yann) Does
 it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
 development causes to recheck the contract $A$ that has been duplicated?}
-And it also could harm performance, or at least complicate an implementation,
-since some sharing mechanism should be used to avoid checking the same contract,
-over the same value, multiple times, exemplified by contract \nickel{A} on
-Figure \ref{fig:kt-contract-transf}, that gets duplicated.
 \unsure{Arnaud says: this whole
@@
 
-Another difficult issue Keil and Thiemann were able to solve, is the
-fact that user defined contracts and intersection do not usually play nice
-together, as shown on Section \ref{sec:flat-and-inter}.
-They provide two different sets of rules for their calculus,
-the first one is non deterministic, and its purpose is to provide
-semantics that are simple to understand.
-With these rules, the interaction between user defined contracts and
-intersections becomes much simpler, since they are able to check
-both sides of an intersection on separate threads.
-This allows them to be sure that contracts from one branch of the intersection
-won't get blamed while executing the other branch, and vice versa.
-
-As an example, consider the programs shown on Figure
-\ref{fig:kt-non-deterministic}, if the Original program
-gets executed, it will spawn two instances of said program,
-one with the first branch of the intersection (\nickel{3 | Num}),
-and the other one with the second branch (\nickel{3 | Positive}),
-and these will get joined once they reach an equal point, in this case
-it will be \nickel{3}.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Original]
-3 | Num @& Positive
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=First Program]
-3 | Num
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Second Program]
-3 | Positive
-\end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Joined Program]
-3
-\end{lstlisting}
-\caption{Non deterministic intersection contract opening}
-\label{fig:kt-non-deterministic}
-\end{figure}
-
-The second set of rules is deterministic, and it is intended
-to be simpler to implement.
-However, this second set needs to perform operations
-on the evaluation context, to decide whether or not to evaluate
-a given contract.
-\footnote{You can check more on this topic on Section 5.1 of
-\cite{KeilThiemannUnionIntersection}}
-For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
-when using the value \nickel{(fun x => x) | Str -> Str)}
-inside the execution of contract \nickel{C}, and applying it to \nickel{0},
-the contract runtime would check the context of execution and it would
-deduce that contract \nickel{Str -> Str} does not apply in this case,
-skipping it.
-
-We see two problems here, a minor one is the fact that the two
-different semantics defined for the calculus are really far apart
-one from the other, both on the strategies used for contract resolution,
-as well as on the complexity levels.
+One important advantage of~\cite{KeilThiemannUnionIntersection} is to handle
+user-defined contracts. We explained in Section~\ref{sec:issues-sem} why a
+direct approach for higher-order intersection contracts (as done
+in~\cite{RootCauseOfBlame}) is not trivial to adapt for user-defined contracts.
+\cite{KeilThiemannUnionIntersection} provides two operational semantics of the
+contract system: a non-deterministic one, more intuitive and proof-friendly, and
+a deterministic one for actual implementation. The non-deterministic semantics
+cleanly handle intersections with user-defined contracts: as each component is
+checked in a separate thread, there is no direct interaction.
+
+Let us recast the problematic example of Figure~\ref{fig:inter-flat-contracts}
+in Figure~\ref{fig:inter-flat-contracts-2}). The execution now splits the
+intersection into two sub-contracts that are checked separately. If one
+component fails, a flag in a global state. In particular, the two threads don't
+interact directly, and don't raise blame by themselves, leading this example run
+as expected.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel,title={Original program}]
+let C = contracts.fromPred (fun f =>
+  f 0 == 0) in
+let g | (String -> String) @& C
+      = fun x => x
+in g 0
+\end{lstlisting}
+\begin{lstlisting}[numbers=none,title={First component}]
+g 0 | Str -> Str
+\end{lstlisting}
+\begin{lstlisting}[numbers=none,title={Second component}]
+g 0 | C
+\end{lstlisting}
+\caption{Intersection and user defined contracts}
+\label{fig:inter-flat-contracts-2}
+\end{figure}
+
+In return, the complexity is offloaded to a separate procedure that must decide
+when to raise blame. This complexity is salient in the deterministic, sequential
+version\footnote{cf Section 5.1 of \cite{KeilThiemannUnionIntersection}}. There,
+the evaluation procedure has to deal again with the problem of eliding contracts
+that are not \emph{compatible}\footnote{You can check more on this topic on
+Section 5.1 of \cite{KeilThiemannUnionIntersection}} (the ones coming from a
+different branch that shouldn't be applied in the current context). In practice,
+the evaluation procedure maintains a global environment of contracts that is
+extended each time a non-trivial contract (union, intersection or higher-order)
+is opened. Then, any contract application must check the presence of the
+corresponding label in the environment. But the most surprising\todo{(Yann) find
+another word} part is that for contracts attached to a function that is applied,
+the whole \emph{context} (the continuation of the application) must be checked
+for compatibility with respect to each label in the contract
+environment!\todo{(Yann) TBC: find better wording}
+
+\unsure{WARNING: what is below is being reworked by Yann}
+
+% Another difficult issue Keil and Thiemann were able to solve, is the
+% fact that user defined contracts and intersection do not usually play nice
+% together, as shown on Section \ref{sec:flat-and-inter}.
+% They provide two different sets of rules for their calculus,
+% the first one is non deterministic, and its purpose is to provide
+% semantics that are simple to understand.
+% With these rules, the interaction between user defined contracts and
+% intersections becomes much simpler, since they are able to check
+% both sides of an intersection on separate threads.
+% This allows them to be sure that contracts from one branch of the intersection
+% won't get blamed while executing the other branch, and vice versa.
+%
+% As an example, consider the programs shown on Figure
+% \ref{fig:kt-non-deterministic}, if the Original program
+% gets executed, it will spawn two instances of said program,
+% one with the first branch of the intersection (\nickel{3 | Num}),
+% and the other one with the second branch (\nickel{3 | Positive}),
+% and these will get joined once they reach an equal point, in this case
+% it will be \nickel{3}.
+%
+% \begin{figure}[h]
+% \begin{lstlisting}[language=nickel, title=Original]
+% 3 | Num @& Positive
+% \end{lstlisting}
+% \begin{lstlisting}[language=nickel, title=First Program]
+% 3 | Num
+% \end{lstlisting}
+% \begin{lstlisting}[language=nickel, title=Second Program]
+% 3 | Positive
+% \end{lstlisting}
+% \begin{lstlisting}[language=nickel, title=Joined Program]
+% 3
+% \end{lstlisting}
+% \caption{Non deterministic intersection contract opening}
+% \label{fig:kt-non-deterministic}
+% \end{figure}
+%
+% The second set of rules is deterministic, and it is intended
+% to be simpler to implement.
+% However, this second set needs to perform operations
+% on the evaluation context, to decide whether or not to evaluate
+% a given contract.
+% \footnote{You can check more on this topic on Section 5.1 of
+% \cite{KeilThiemannUnionIntersection}}
+% For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
+% when using the value \nickel{(fun x => x) | Str -> Str)}
+% inside the execution of contract \nickel{C}, and applying it to \nickel{0},
+% the contract runtime would check the context of execution and it would
+% deduce that contract \nickel{Str -> Str} does not apply in this case,
+% skipping it.
+%
+% We see two problems here, a minor one is the fact that the two
+% different semantics defined for the calculus are really far apart
+% one from the other, both on the strategies used for contract resolution,
+% as well as on the complexity levels.
 \resolved{(Yann) Is it really a

### 2021-06-09 paper.tex
Change the title
Some words on Castagna.

@@
 
-\title{Contracts for unions and intersections are hard for Nickel}
+\title{Union and intersection contracts are hard, actually}
 \author{Teodoro Freund}
@@
 
-\section{Related work}
+\section{(More) Related work}
 \label{sec:related-work}
 \info{Including a mention of statically typed systems with union}
+
+\subsection{Gradual typing: a new perspective}
+
+Castagna et. al \cite{gradualCastagna} present a gradual type system that
+incorporates multiple non-trivial typing constructs, in particular, subtyping,
+unions, recursive types, negation types, and a bottom type.
+Which can easily encode intersection, difference, and top types.
+
+This system, even if related to the ones we studied, adheres to a different
+school of gradual typing, Greenberg would call it the static first
+school\cite{practiceTheoryGrad}.
+In this line of work, the goal is usually to develop a type system that can
+handle a dynamic type to represent the lack of precission on some type.
+The main reasons why we consider this work not comparable with what we attempted
+to solve are:
+\begin{itemize}
+  \item Like Williams et al\cite{RootCauseOfBlame}, it doesn't allow
+        to use user defined contracts.
+  \item Their base calculus needs to have a non-trivial type inference
+        algorithm, that makes the development of a programming language
+        much more cumbersome. For instance, it defines a set-theoretic
+        interpretation of types, and a subtyping relation based on it.
+  \item Contracts are completely handled by the execution engine (in the paper,
+        the rewriting system), and not added as a library with some basic
+        syntactic support.
+\end{itemize}
 

### 2021-06-09 paper.tex
Rework section 4 (WIP)

@@
 Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
-convincing semantics for union and intersection contracts. We follow with the
-work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which builds
-upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
+lean\todo{Not the right word} semantics for union and intersection contracts. We
+follow with the work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which
+builds upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
 uniform blame resolution.
@@
 that are not \emph{compatible}\footnote{You can check more on this topic on
-Section 5.1 of \cite{KeilThiemannUnionIntersection}} (the ones coming from a
-different branch that shouldn't be applied in the current context). In practice,
+Section 5.1 of \cite{KeilThiemannUnionIntersection}} (contracts are incompatible
+if they are coming from a different branch and should be ignored). In practice,
 the evaluation procedure maintains a global environment of contracts that is
@@
 is opened. Then, any contract application must check the presence of the
-corresponding label in the environment. But the most surprising\todo{(Yann) find
-another word} part is that for contracts attached to a function that is applied,
-the whole \emph{context} (the continuation of the application) must be checked
-for compatibility with respect to each label in the contract
-environment!\todo{(Yann) TBC: find better wording}
+corresponding label in the environment. What's more, each time a function with a contract attached is applied,
+the whole \emph{context} (that is, the continuation of the application) must be
+must be checked for compatibility with respect to each label of the contract
+environment. Checking for compatibility involves searching the context for
+contract applications.
+
+Keil and Thiemann realizes the tour de force of combining union and intersection
+contracts with user-defined contracts. The price to pay, though, is that
+implementing their solution in a reasonable way is far from obvious. Not only
+when starting from the more abstract, non-deterministic formulation, but even
+when starting from the algorithmic version of their system.
 
@@
 (Teo) reworded}.
-A second, major one,
-is the dependency added over the evaluation context, since, again,
-it makes composing this, with other features desired on programming languages,
-difficult, or even impossible.
+% A second, major one,
+% is the dependency added over the evaluation context, since, again,
+% it makes composing this, with other features desired on programming languages,
+% difficult, or even impossible.
 \resolved{(Yann) I think we need to substantiate this claim. If this
@@
 
-Williams, Morris, and Wadler, take a different approach on user defined contracts,
-they are completely forbidden in their system, explicitly to not encounter
-the previously mentioned issue.
-This restriction goes against what is usually expected from dynamic checks:
-validation of data, and not just inclusion on types.
-
+As mentioned in Section~\ref{sec:issues-sem}, Williams, Morris, and Wadler,
+downright exclude user-defined contracts from the picture. This choice gives
+them more freedom in the quest of a more uniform operational semantics. While
+this can be a sensible choice (in a gradual typing context e.g.), this is a show
+stopper for our initial use-case of the Nickel configuration language, the main
+motivation being to use contracts as customizable validation functions.
+
+\ref{sec:issues-sem}
+On the semantic side, one of the so-called the sound monitoring properties
+linking the operational semantics to the denotational semantics doesn't
+correspond to its attached description.
 On top of that, a fundamental problem arises on said paper,

### 2021-06-09 paper.tex
Section 4 intro + random stuff

@@
 contract for valid ports in Nickel syntax.
+%
+User-defined contracts can be combined with other contracts normally:
+\nickel{Int -> Port} is a contract verified by functions
+which, given an integer returns a valid port.
 
@@
 
-User-defined contracts can be combined with other contracts normally,
-for instance \nickel{Int -> Port} is a contract verified by functions
-which, given an integer returns a valid port.
-
-This type of contracts are present on many different languages,
+This type of contracts are present in many different languages,
 for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
-of the Design by Contract ideals, allows to assert
+of the Design by Contract philosophy, makes it possible to assert
 these kinds of expression as pre- and post-conditions on
@@
 
-\section{Issues coming from the semantics}
+\section{Incompatibilities}
 \label{sec:issues-sem}
-\info{What goes wrong or difficult when trying to implement union and
-  intersection literally}
-
-Static typing is about enforcing that some property about a (potentially
-unknown) value always holds, no matter the execution path. A dynamic check, on
-the other hand, just needs to verify a property for a specific and known value,
-which is usually much easier. For example, dependent function types are complex
-and mostly implemented in proof assistants (although they are slowly making
-their way in some mainstream languages), whereas the corresponding dependent
-function contracts can be implemented in a straightforward way (they are
-supported in e.g. Racket~\cite{}).
-
-The gist of the present paper is to bring evidence that in the case of unions
-and intersections, the dynamic checking part is complex as well: \emph{unions
-and intersections are fundamentally hard}.
-
-% Teo's version
-% Implementing a static type system is not an easy task.
-% Type systems featuring intersection types are hard in a fundamental way,
-% from their discovery, they've been studied as a form to characterize
-% lambda-calculus terms normalizability; which means that typechecking a
-% term on a system with intersection types is undecidable,
-% similar to how the halting problem is undecidable\todo{reference}.
-%
-% Usually, when thinking about programming languages, there's an intuition
-% that static checks (like typechecking) are more difficult and less powerful
-% than dynamic checks (like testing).
-% For instance, it is very easy to check if,
-% in a given execution of a program, a variable holds an integer value; while it
-% may become more challenging to check if a variable will always hold an integer value,
-% no matter the particular execution.
-%
-% As we intend to show on this paper, unions and intersections are not only
-% statically hard to check, but they are also dynamically hard:
-% \emph{they are fundamentally hard}.
-% In this section, we present intuitively the main challenges that arise
-% when attempting to develop a runtime check system for unions and intersections.
+
+However appealing union and intersection contracts may be, they happen
+to be fundamentally incompatible with the desirable language
+features from Section~\ref{sec:feat-lang}. At least in there
+full-blown form: in Section~\ref{sec:issues-literature} we will
+discuss pragmatic restrictions of union and intersection contracts to
+recover some or all of the features.
 

### 2021-06-09 paper.tex
Section 5 header

@@
 
-\section{Issues in concrete restrictions from the literature}
+\section{Pragmatic trade-offs}
 \label{sec:issues-literature}
-\info{The various papers and the tradeoffs they make}
-
-In this section, we review some shortcomings of existing work on union and
-intersection contracts. We start with the work of Keil and
-Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
-lean\todo{Not the right word} semantics for union and intersection contracts. We
-follow with the work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which
-builds upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
-uniform blame resolution.
-
-% The existing work that previously researched and implemented
-% union and intersection
-% as dynamic checks have, in our eyes, different shortcomings
-% that makes them suboptimal solutions.
-% In this section, we informally present those issues,
-% starting with the foundational paper by Keil and Thiemann
-% \cite{KeilThiemannUnionIntersection}, then following to a recent
-% work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
-% finally going over some similar contracts available in the Racket
-% programming language.
+
+Despite the difficulties of Section~\ref{sec:issues-sem}, union and
+intersection contracts are still sought after. In this section we turn
+to existing systems with union and intersection contracts in the
+literature and in implementations.
+
+These systems all make trade-offs, sacrificing some features of union
+and intersection contracts to preserve language features. We survey
+and discuss those trade-offs and their implications.
 
@@
 In Section~\ref{sec:issues-sem}, we've seen that different calls to a function
-with an union contract must share information: the behavior of one call is
+with a union contract must share information: the behavior of one call is
 influenced by the previous ones, as the function must pick one component of the

### 2021-06-09 paper.tex
Wibbles

@@
 when to raise blame. This complexity is salient in the deterministic, sequential
-version\footnote{cf Section 5.1 of \cite{KeilThiemannUnionIntersection}}. There,
+version~\cite[Section 5.1]{KeilThiemannUnionIntersection}. There,
 the evaluation procedure has to deal again with the problem of eliding contracts
-that are not \emph{compatible}\footnote{You can check more on this topic on
-Section 5.1 of \cite{KeilThiemannUnionIntersection}} (contracts are incompatible
+that are not \emph{compatible}~\cite[Section 5.1]{KeilThiemannUnionIntersection} (contracts are incompatible
 if they are coming from a different branch and should be ignored). In practice,

### 2021-06-09 paper.tex
Remove a bunch of marginal notes

@@
 \label{sec:intro}
-\info{Goals of the paper: position paper: unions/intersections nice in
-theory problematic in practice. In particular non-orthogonal with the
-rest of the features of your language.}
 
@@
 \label{sec:feat-lang}
-\info{Zoology of various features that we will eventually show
-  conflict with this or that property or implementation of union and
-  intersection. Including user-define contracts.}
 
@@
 \label{sec:optimizations}
-\info{Present inlining and CSE as two major code
-optimizations.
-Purity and immutability as two important factors.}
 
@@
 \label{sec:union-inter}
-\info{What are they, what are they used for.}
-
-% Goals:
-% Introduce unions and inter
-% Why are they useful?
-% How are they used
-% How de they relate with one another
 

### 2021-06-09 paper.tex
Section 3 header
Add a keyword + don't capitalise keywords
They seem uncapitalised in the acm-provided templates, so I just did
the same.
Finish reworking of 5.2

@@
 
-\keywords{Dynamic languages, Contracts, Union, Intersection}
+\keywords{contracts, higher-order contracts, union, intersection}
 
@@
 
-\unsure{TODO: CCS classification; Keywords}
 \section{Introduction}
@@
 \label{sec:union-inter}
+
+Let us now consider union and intersection contracts, before we
+explain, in Section~\ref{sec:issues-sem} how they can compromise the
+properties that we described in Section~\ref{sec:feat-lang}.
 
@@
 when starting from the algorithmic version of their system.
-
-\unsure{WARNING: what is below is being reworked by Yann}
 
@@
 
-\ref{sec:issues-sem}
 On the semantic side, one of the so-called the sound monitoring properties
-linking the operational semantics to the denotational semantics doesn't
-correspond to its attached description.
-On top of that, a fundamental problem arises on said paper,
-symptom of the difficulty of establishing good enough, and yet
-simple semantics for union and intersection dynamic checks.
-One of their main results, is the fact that their calculus obeys
-all of their expected monitoring properties for higher ordered, union
-and intersection contracts.
-However, one of these properties states the following:
+linking the operational semantics to the denotational semantics (supposed to
+capture the intuitive interpretation) is at odds with the given interpretation.
+% On top of that, a fundamental problem arises on said paper,
+% symptom of the difficulty of establishing good enough, and yet
+% simple semantics for union and intersection dynamic checks.
+% One of their main results, is the fact that their calculus obeys
+% all of their expected monitoring properties for higher ordered, union
+% and intersection contracts.
+
+For a contract $A$, $\llbracket A \rrbracket^+$ defines the set of terms that
+satisfy $A$. For example,
+
+\begin{lstlisting}[language=nickel,frame=none,numbers=none]
+    if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
+\end{lstlisting}
+
+For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
+contract can be violated in two ways: by the function returning a value that is
+not a string, or by the caller providing an argument that is not a number.  The
+former is embodied by the same notion as before: $f \in \llbracket Num \to Str
+\rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
+important contribution of~\cite{KeilThiemannUnionIntersection}, upon
+which~\cite{RootCauseOfBlame} builds upon, is to extend the notion of contract
+satisfaction to \emph{contexts} (i.e. a continuations), embodying the caller's
+obligations. A context $K$ is a term with a hole to be filled with a term (see
+Figure~\ref{fig:valid-contexts} for examples). Then, $K \in \llbracket Num \to
+Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
+if it only applies the hole to arguments satisfying the \nickel{Num} contract.
+
+These interpretations $\llbracket . \rrbracket^+$ and $\llbracket .
+\rrbracket^-$ then need to be extended to the setting of unions and
+intersections. In particular, one of the definitions of~\cite{RootCauseOfBlame}
+is the following:
 
@@
 $$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
-\caption{Sound monitoring property for intersection from Williams, Morris, and Wadler.}
+\caption{Sound monitoring property for intersection from~\cite{RootCauseOfBlame}}
 \label{fig:wmw-semantics}
@@
 
-Which means that a continuation $K$ complies negatively with an
-intersection of $A$ and $B$ if it complies negatively with at least one of these.
-Or, less formally, that if you have a program that is missing a value at a given point,
-you can put a value wrapped with a contract stating $A \cap B$,
-and expect not to get a negative violation of that contract,
-if you can either put a value
-wrapped with $A$, or a value wrapped with $B$.
-
-Altough this is true, it's a weaker statement of what we'd like to have,
-consider the two contexts presented on Figure \ref{fig:valid-contexts},
-the first one is a valid context for a \nickel{Num -> Num} function,
-that is, you could replace the $\square$ by some function wrapped in a
-\nickel{Num -> Num} contract, and the context will not violate it.
-Similarly, the second context from the same figure is a valid
-\nickel{Bool -> Bool} context.
+This reads as: a context $K$ satisfies the intersection of $A$ and $B$ if it
+satisfies at least one of the two. Morally, the $K$s in $\llbracket A \cap B
+\rrbracket^-$  should be the ones that can have their hole filled with a term
+satisfying $A \cap B$ without violating the $A \cap B$ contract.
+
+Although valid, this interpretation is weaker than what~\cite{RootCauseOfBlame}
+actually describes it to be and than the semantics given
+in~\cite{KeilThiemannUnionIntersection}. Consider the two contexts presented on
+Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
+\nickel{Num -> Num}, applying the hole to a number. Similarly, the second context
+from the same figure satisfies \nickel{Bool -> Bool}.
 
@@
 
-Next, we can combine these two contexts into one, that essentially
-has one single missing value, that will be used independently in both contexts
-from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
-This context is not a valid \nickel{Num -> Num} context, since it will apply
-the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
-In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
-Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
-shown on Figure \ref{fig:wmw-semantics}, it is not a valid
-\nickel{Num -> Num @& Bool -> Bool} context.
+Now, combining these two context as in~\ref{fig:invalid-context} gives a context
+that doesn't satisfy \nickel{Num -> Num} nor \nickel{Bool -> Bool}.  According
+to the semantics of~\cite{RootCauseOfBlame}, it thus doesn't satisfy \nickel{Num
+-> Num @& Bool -> Bool}. Yet, our intuition tells us that this example is a
+perfectly valid context for this contract: an overloaded function satisfying
+both \nickel{Num -> Num} and \nickel{Bool -> Bool} can safely be applied to a
+number and to a boolean in a row. And indeed, the denotational semantics given
+in~\cite{KeilThiemannUnionIntersection} does accept this context as valid.
+
+The issue is not as bad as it sounds: the system presented
+in~\cite{RootCauseOfBlame} actually corresponds to the wished, stronger
+semantics in practice. The specification is just not exactly the right one, but
+this can probably be fixed along the lines
+of~\ref{KeilThiemannUnionIntersection}, where the notion of context used to
+define satisfaction is carefully restricted (\emph{elimination contexts}). While
+this kind of oversight doesn't lower the quality of the work
+of~\ref{RootCauseOfBlame} by any measure, it illustrates that contract
+satisfaction in presence of unions, intersections and functions is a tricky
+notion. This difficulty contrasts with the standard framework of higher-order
+contracts where satisfaction is much more straightforward.
+% Next, we can combine these two contexts into one, that essentially
+% has one single missing value, that will be used independently in both contexts
+% from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
+% This context is not a valid \nickel{Num -> Num} context, since it will apply
+% the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
+% In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
+% Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
+% shown on Figure \ref{fig:wmw-semantics}, it is not a valid
+% \nickel{Num -> Num @& Bool -> Bool} context.
 
@@
 \end{lstlisting}
-\caption{Invalid \nickel{Num -> Num @& Bool -> Bool} context,
-combining contexts from Figure \ref{fig:valid-contexts}}
+\caption{Combined context}
 \label{fig:invalid-context}
@@
 
-However, both our intuition, as the most common use case of intersections
-(that is, overloading), tells us that this should work.
-While this problem does not invalidate their work (in fact, the system they present
-handles these cases correctly), it shows that existing research is still not on
-good shape to handle the growing usecases being developed for
-unions and intersections around dynamic languages.
+% However, both our intuition, as the most common use case of intersections
+% (that is, overloading), tells us that this should work.
+% While this problem does not invalidate their work (in fact, the system they present
+% handles these cases correctly), it shows that existing research is still not on
+% good shape to handle the growing usecases being developed for
+% unions and intersections around dynamic languages.
 
@@
 \label{sec:racket}
+
+\unsure{WARNING: what is below is being reworked by Yann}
 

### 2021-06-09 paper.tex
Pass on the union contract subsection

@@
 
+\lstdefinelanguage{Python}{
+  % keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
+  % keywordstyle=\color{blue}\bfseries,
+  % ndkeywords={class, export, boolean, throw, implements, import, this},
+  % ndkeywordstyle=\color{darkgray}\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  % comment=[l]{//},
+  % morecomment=[s]{/*}{*/},
+  % commentstyle=\color{purple}\ttfamily,
+  % stringstyle=\color{red}\ttfamily,
+  morestring=[b]',
+  morestring=[b]"
+}
+
 \lstset{
@@
 \newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
+\newcommand{\python}[1]{\lstinline[language=Python]{#1}}
 
@@
 union types to model the frequent practice to use the value
-\lstinline{null} (\lstinline{None} in Python) to represent an absent
+\typescript{null} (\python{None} in Python) to represent an absent
 optional value. This is why the gradual typing literature, concerned
@@
 \subsection{Unions}
-Unions ($\cup$) are a fundamental building block of programs.
-A union holds a value whose
-shape is known to be in a fixed set of alternatives, without exposing
-which of those alternatives the value belongs to.
-For example, a value that could be either a Number or a String can be
-modelled using unions as \nickel{Num @| String}.
-They correspond logically to disjunctions.
-
-Statically typed languages, and more specifically functional programming
-languages, provide a similar abstraction, sums
-(also called tagged unions, or discriminated unions), which
-also allow to model a value whose shape conforms to a fixed set of
-alternatives, but requiring an explicit mark (or tag) that differentiates
-to which of these alternatives it belongs to.
-For example, consider the Haskell code presented on Figure
-\ref{fig:safediv-haskell}, it's a safe division that will
-check if the dividend is 0, and if it is, return an
-error message.
-Since the function can return two different types, a sum
-(\haskell{Either}) is used, and on each
-of the if branches an explicit tag has to be used (either
-\haskell{Left} or \haskell{Right}).
-
-\begin{figure}[h]
-\begin{lstlisting}[language=Haskell]
-safeDiv :: Int -> Int -> Either String Int
-safeDiv x y =
-  if y == 0 then
-    Left "Division by 0 is forbidden"
-  else
-    Right (div x y)
-\end{lstlisting}
-\caption{Safe division implemented in Haskell}
-\label{fig:safediv-haskell}
-\end{figure}
-
-This incurred verbosity certainly facilitates the work of static checks,
-like typechecking, but dynamically typed languages permit much more flexibility,
-and programmers don't usually expect to tag their values.
-That is why, as mentioned on the Introduction, when designing
-type systems for dynamically typed languages, unions become of particular
-interest.
-See the implementation of a safe division function on TypeScript
-(Figure \ref{fig:safediv-ts}).
-Neither of the if branches has to tag their return value, and the return
-type of the function is modelled using unions, \typescript{string | number}.
-
-\begin{figure}[h]
+
+A union type \nickel{A @| B} is a type of values which are either of
+type \nickel{A} or of type \nickel{B}: literally the union of
+\nickel{A} and \nickel{B}. Union types are popular in gradual typed
+systems such as TypeScript~\cite{TypeScriptUnions} and
+MyPy~\cite{MyPyOptional}.
+
+\paragraph{In gradually typed systems}
+
+In these practical gradual type systems the problem is to be able to
+capture, in static types, as many programming patterns as possible
+from the underlying dynamically typed language. One such pattern is
+heterogeneous collections, in JavaScript (for TypeScript) and Python
+(for MyPy), heterogeneous collections, say arrays with values of
+different types, are frequent. For instance, in TypeScript, an array
+which can contain both strings and numbers would have type
+\typescript{Array<string|number>}.
+
+A probably even more common pattern is a variable which can contain
+either a value of type a (say, a number) or the \typescript{null}
+value. So much so, in fact, that MyPy defines a type alias
+\python{Optional[A]} for \python{Union[A,None]} (\python{None} is how
+Python renders the null value).
+
+Yet another application of union types is, rather than capturing a
+pattern from JavaScript or Python, to capture a pattern from
+traditional statically typed language: sum types. In statically typed
+languages, values of sum types are usually thought of as being built
+out of constructors. But neither JavaScript nor Python have such
+constructors. So instead, sums are construed as ``tagged unions'' (or
+discriminated unions), that is, quite literally, the union of two
+types which contain a discriminating tag. See
+Figure~\ref{fig:tagged-union} for an example from the TypeScript
+documentation: there the \typescript{kind} field is the tag, and its
+type in both alternatives is a singleton type which contains only the
+specified string.
+
+\begin{figure}
+  \centering
   \begin{lstlisting}[language=JavaScript]
-function safeDiv(x: number, y: number): string | number {
-  if (y == 0) {
-      return "Division by 0 is forbidden"
-  } else {
-      return x/y
-  }
-}
-\end{lstlisting}
-\caption{Safe division implemented in TypeScript}
-\label{fig:safediv-ts}
-\end{figure}
-
-% \begin{figure}[h]
-% \begin{lstlisting}[title={Tree-like data structure}]
-% type BinaryTree =
-% Integer @| (BinaryTree, BinaryTree)
+    interface Circle { kind: "circle"; radius: number; }
+
+    interface Square { kind: "square"; sideLength: number; }
+
+    type Shape = Circle | Square;
+  \end{lstlisting}
+
+  \caption{A sum type as a tagged union}
+  \label{fig:tagged-union}
+\end{figure}
+
+\paragraph{Union contracts}
+
+In the academic gradual type literature, it is common to use contracts
+as a glue between static and dynamic types. Therefore, the question of
+bringing union to contracts is natural, and have indeed been studied
+(\emph{e.g.}~\cite{KeilThiemannUnionIntersection,RootCauseOfBlame}).
+
+Like for static types, a valid value for the contract \nickel{A @| B}
+is a value which is valid either for contract \nickel{A} or for
+contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
+will see that it may be desirable to weaken this definition).
+
+Nickel is a language built from scratch with contracts, so it may be
+less clear why unions are useful. However, Nickel's ambition is to
+have its data model canonically interpretable in common serialization
+formats, in particular JSON. It means that it is very convenient to
+represent optional value by the \nickel{null} value like in
+JavaScript. It also means that Nickel doesn't have built-in
+constructors: constructors don't have a canonical representation in
+JSON. So it would be quite natural to represent optional contracts and
+sum contracts as unions.
+
+% Arnaud: I'm not deleting this quite yet because we may want to use
+% these examples, I just didn't manage to make them fit in the flow of
+% the narrative.
+%
+% Unions are not only useful to model the outcome of alternative
+% branches of computations, but they can be used to
+% formalize very common patterns of data types:
+% \begin{itemize}
+%     \item To represent nullable, or optional, values.
+% \begin{lstlisting}[language=nickel]
+% optionalValue | Str @| Null
 % \end{lstlisting}
-% \caption{A binary tree data type using unions}
-% \label{fig:union-adt-ex}
-% \end{figure}
-
-Unions are not only useful to model the outcome of alternative
-branches of computations, but they can be used to
-formalize very common patterns of data types:
-\begin{itemize}
-    \item To represent nullable, or optional, values.
-\begin{lstlisting}[language=nickel]
-optionalValue : Str @| Null
-\end{lstlisting}
-
-    \item For plain enumerations of literals.
-\begin{lstlisting}[language=nickel]
-protocol : "Http" @| "Ftp" @| "Sftp"
-\end{lstlisting}
-
-    \item To support different representations of the same data.
-\begin{lstlisting}[language=nickel]
-person : {name : Str, age : Int}
-  @| {name : Str, dateOfBirth : Date}
-\end{lstlisting}
-
-\end{itemize}
-All of which are common in configurations.
-
-% \begin{figure}[h]
-% % \label{fig:union-nullable-ex}
-% \begin{lstlisting}[title={Nullable values}]
-% optionalValue : Str @| Null
+
+%     \item For plain enumerations of literals.
+% \begin{lstlisting}[language=nickel]
+% protocol | "Http" @| "Ftp" @| "Sftp"
 % \end{lstlisting}
 
-% % \label{fig:union-enum-ex}
-% \begin{lstlisting}[title={Plain enumeration}]
-% protocol : `Http | `Ftp | `Sftp
+%     \item To support different representations of the same data.
+% \begin{lstlisting}[language=nickel]
+% person | {name : Str, age : Int}
+%   @| {name : Str, dateOfBirth : Date}
 % \end{lstlisting}
 
-% % \label{fig:union-alt-ex}
-% \begin{lstlisting}[title={Alternative representations}]
-% person : {name : Str, age : Int}
-%   @| {name : Str, dateOfBirth : Str}
-% \end{lstlisting}
-% \caption{Various use cases of unions}
-% \label{fig:unions-use-cases}
-% \end{figure}
-
-% In statically typed languages, and more specifically in functional programming
-% languages, sums (also called tagged unions, or discriminated unions) are usually
-% preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
-% syntactically separated by an associated discriminating label. Well-known
-% programming languages with first-class tagged unions include Haskell, OCaml,
-% Rust, Scala, PureScript, Elm, Swift and many more.
-
-% On the other hand, in dynamically typed languages, it is customary to directly
-% store different kinds of data in one variable without further ceremony. The
-% runtime system needs to attach type tags to values anyway, which can usually be
-% observed using functions like JavaScript's \verb+typeof+. Using such bare
-% unions\resolved{If you can observe with typeof, then it's pretty much a tagged
-% union, though the tag is implicit} has already been an idiom in dynamically
-% typed language for a long time. Being able to type and handle these unions is an
-% important aspect of the migration to a gradual type system. Indeed, a lot of
-% \resolved{This is the first time discussing gradual type systems, maybe it should
-% be introduced earlier, in particular how they relate to contracts
-% (I think is mentioned on the appendix)}
-% gradually typed languages either existed originally as dynamically typed
-% language or aim at backward compatibility with a dynamically typed base
-% language. This is why despite the presence of a static type system, gradually
-% typed languages tend to favor bare unions types which is the typed counterpart
-% of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).
-
-Nickel has no pre-existing untyped language that we have to
-support, so we are freer of our choice.
-However, Nickel is a configuration language, whose contract and type system
-is intended to encode data schemas encountered in the wild. Since these schemas do not use
-tags to differentiate different kinds of values, there is a strong incentive to support untagged unions at the
-type level, or at least, at the contract level.
-
+% \end{itemize}
+% All of which are common in configurations.
+
+\unsure{Arnaud: this paragraph is pretty good, on the other hand, there
+is no value which lets me complete the example, so this example falls
+a bit flat.}
 We have shown how values of a union are built, is worth mentioning
@@
 let Pet = { owner: Str, name: Str } in
-let myDog | Animal & Pet =
+let myDog | Animal @& Pet =
   { species = "Canis Lupus",
@@
 % LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
-% LocalWords:  TypeScript
+% LocalWords:  TypeScript serializable serialization

### 2021-06-09 paper.tex
Small wording issue

@@
 
-In these practical gradual type systems the problem is to be able to
+The problem that these practical gradual type systems are trying to solve is to
 capture, in static types, as many programming patterns as possible

### 2021-06-09 paper.tex
A bit more rewording

--- 2021-06-09 paper.tex
@@
 
-In these practical gradual type systems the problem is to be able to
+The problem that these practical gradual type systems are trying to solve is to
 capture, in static types, as many programming patterns as possible
-from the underlying dynamically typed language. One such pattern is
-heterogeneous collections, in JavaScript (for TypeScript) and Python
-(for MyPy), heterogeneous collections, say arrays with values of
-different types, are frequent. For instance, in TypeScript, an array
+from the underlying dynamically typed language (JavaScript for
+TypeScript and Python for MyPy). One such pattern is
+heterogeneous collections. For instance, in TypeScript, an array
 which can contain both strings and numbers would have type

### 2021-06-09 paper.tex
Rework section 5.3 [sec 5 finished]

@@
 
-Racket is a language based on Scheme, mainly thought as a language to experiment
-with different language desings and ideas. It provides a very complete contract
-system\cite{RacketContracts}.
-
-Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
-These contracts are lightweight versions of intersections and unions, they simply
-check that every contract is valid, in the case of \racket{and/c},
-and that at least one contract
-is valid, in the case of \racket{or/c}.
-
-So, for instance, the usual usecase of intersection to encode bounded overloading
-does not work on Racket. Take a look at the next example, the function
-\racket{overload} should accept a call like \racket{(overload 3)}, but it fails
-since 3 is not a valid \racket{string?}.
-
-\begin{figure}[h]
-
+Racket is a language based on the Scheme dialect of Lisp. Among established
+languages, Racket is probably the one with the most comprehensive contract
+system\cite{RacketContracts}. Regarding union and intersections, Racket provides
+the \racket{and/c} and \racket{or/c} combinators for contract.  These
+combinators correspond to the naive interpretation of unions and intersections
+(as in Figure~\ref{fig:naive-impl}): \racket{and/c} (resp.  \racket{or/c})
+simply checks sequentially that all (resp. one) contracts pass.  In consequence,
+unions or intersections of higher-order contracts don't match intuition, and in
+particular intersections don't model overloading. As in
+Figure~\ref{fig:naive-impl}, the example given in
+Figure~\ref{fig:racket-overloading} always fail because no argument satisfies
+both \racket{number?} and \racket{string?}.
+
+% Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
+% These contracts are lightweight versions of intersections and unions, they simply
+% check that every contract is valid, in the case of \racket{and/c},
+% and that at least one contract
+% is valid, in the case of \racket{or/c}.
+
+% So, for instance, the usual usecase of intersection to encode bounded overloading
+% does not work on Racket. Take a look at the next example, the function
+% \racket{overload} should accept a call like \racket{(overload 3)}, but it fails
+% since 3 is not a valid \racket{string?}.
+
+\begin{figure}[h]
 \begin{lstlisting}[language=racket]
@@
 \end{lstlisting}
-\caption{Non working function overloading on Racket.}
-
-\end{figure}
-
-Even worst, the \racket{or/c} contract doesn't allow to have multiple higher order
-contracts that could (potentially) apply to a function, these have to be
-differentiated by first order markers.
-For instance, the code on Figure~\ref{code:racket:or/c:working}
-would correctly allow only to call
-\racket{united} with a \racket{number?},
-and it can make this distinction since it knows that
-\racket{(lambda (x) x)} can not be a function that takes 2 parameters.
-However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
-before running, since the \racket{or/c} contract is unable to know
-which of the two function contracts to use.
-Ideally, this last example should just behave as wrapping
-the function with the \racket{(-> even? even?)} contract.
+\caption{\racket{and/c} and overloading}
+\label{fig:racket-overloading}
+\end{figure}
+
+When higher-order contracts are combined using \racket{or/c}, Racket imposes
+that contracts must distinguishable by their arity. Doing so, there is at most
+one candidate that can be selected directly. This is illustrated in
+Figure~\ref{code:racket:or/c:working} whose program is accepted. On the other
+hand, the program of Figure~\ref{code:racket:or/c:non-working} is rejected.
+
+% that could (potentially) apply to a function, these have to be
+% differentiated by first order markers.
+% For instance, the code on Figure~\ref{code:racket:or/c:working}
+% would correctly allow only to call
+% \racket{united} with a \racket{number?},
+% and it can make this distinction since it knows that
+% \racket{(lambda (x) x)} can not be a function that takes 2 parameters.
+% However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
+% before running, since the \racket{or/c} contract is unable to know
+% which of the two function contracts to use.
+% Ideally, this last example should just behave as wrapping
+% the function with the \racket{(-> even? even?)} contract.
 
@@
 \end{lstlisting}
-\caption{Working \racket{or/c} example on Racket.}
+\caption{Accepted use of \racket{or/c} with higher-order contracts}
 \label{code:racket:or/c:working}
@@
 \end{lstlisting}
-\caption{Non working \racket{or/c} example on Racket.}
+\caption{Rejected use of \racket{or/c} with higher-order contracts}
 \label{code:racket:or/c:non-working}
@@
 
-Racket does provide an alternative for overloading methods, called the
-\racket{case->} contract, however, this overloading of methods only
-works over contracts of functions with different amount of parameters.
-It's a good alternative to allow limited overloading of functions.
-As an example, consider the following non failing piece of code, and
-try to understand how Racket can decide, at each application,
-which function contract to use.
-
-\begin{figure}[h]
-
+Racket does provide the \racket{case->} combinator which(partially) supports
+overloading. As for the \racket{or/c}, the candidate contracts must have
+distinct arities to avoid ambiguity. An example is provided in
+Figure~\ref{fig:rafig:racket-case-fun}. The resulting possibilities are similar
+to static overloading, where one function can take additional parameters for
+example (e.g. as supported for Java methods). Unfortunately, it excludes the
+overloading of generic operations with fixed arity such as equality, comparison,
+arithmetic operators, and so on.
+
+% Racket does provide an alternative for overloading methods, called the
+% \racket{case->} contract, however, this overloading of methods only
+% works over contracts of functions with different amount of parameters.
+% It's a good alternative to allow limited overloading of functions.
+% As an example, consider the following non failing piece of code, and
+% try to understand how Racket can decide, at each application,
+% which function contract to use.
+
+\begin{figure}[h]
 \begin{lstlisting}[language=racket]
@@
 \end{lstlisting}
-\end{figure}
-
-Even if Racket's intent is not to implement union and intersection
-dynamic checks, they do provide contracts that are, in some sense,
-simplified versions of those.
-In our eyes, Racket has accepted the difficulty of correctly and practically
-combine union, intersection, and higher order contracts, and has choose
-to prioritise the highe order ones.
+\caption{Overloading with the \racket{case->} combinator}
+\label{fig:racket-case-fun}
+\end{figure}
+
+In conclusion, Racket also seems to avoid the difficulties of general unions and
+intersections. They make the pragmatic choice of a simple semantics that work
+for non higher-order contracts and a limited support for overloading.
+% Even if Racket's intent is not to implement union and intersection
+% dynamic checks, they do provide contracts that are, in some sense,
+% simplified versions of those.
+% In our eyes, Racket has accepted the difficulty of correctly and practically
+% combine union, intersection, and higher order contracts, and has choose
+% to prioritise the highe order ones.
 

### 2021-06-09 paper.tex
Acquire lock on sec 6
Rework of section 6 (related works)

@@
 Like in the case of lazy dictionaries, the checks are delayed.
-Contracts keeps track of whether the caller or the implementation is
+Contracts keep track of whether the caller or the implementation is
 at fault for a violation, hence it can report precise error
@@
 
-Keil and Thiemann realizes the tour de force of combining union and intersection
+Keil and Thiemann realize the tour de force of combining union and intersection
 contracts with user-defined contracts. The price to pay, though, is that
@@
 
-\unsure{WARNING: what is below is being reworked by Yann}
-
 Racket is a language based on the Scheme dialect of Lisp. Among established
@@
 
-\subsection{Gradual typing: a new perspective}
-
-Castagna et. al \cite{gradualCastagna} present a gradual type system that
-incorporates multiple non-trivial typing constructs, in particular, subtyping,
-unions, recursive types, negation types, and a bottom type.
-Which can easily encode intersection, difference, and top types.
-
-This system, even if related to the ones we studied, adheres to a different
-school of gradual typing, Greenberg would call it the static first
-school\cite{practiceTheoryGrad}.
-In this line of work, the goal is usually to develop a type system that can
-handle a dynamic type to represent the lack of precission on some type.
-The main reasons why we consider this work not comparable with what we attempted
-to solve are:
-\begin{itemize}
-  \item Like Williams et al\cite{RootCauseOfBlame}, it doesn't allow
-        to use user defined contracts.
-  \item Their base calculus needs to have a non-trivial type inference
-        algorithm, that makes the development of a programming language
-        much more cumbersome. For instance, it defines a set-theoretic
-        interpretation of types, and a subtyping relation based on it.
-  \item Contracts are completely handled by the execution engine (in the paper,
-        the rewriting system), and not added as a library with some basic
-        syntactic support.
-\end{itemize}
+\unsure{WARNING: what is below is being reworked by Yann}
+
+\subsection{Higher-order contracts}
+
+Enforcing pre- and post-conditions at runtime is a widely established practice.
+In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
+Felleisen introduce higher-order contracts, a principled approach to run-time
+assertion checking that nicely supports functions. They introduce the notion of
+blame, which is crucial to good error reporting. It became apparent later that
+their contracts are closely related to the type casts introduced by gradual
+typing, excluding blame: both \cite{FindlerMultiLang} and
+\cite{FelleisenInterLang} see the value of contracts as a safe interface between
+typed and untyped code. In \cite{WellTypedBlamed}, the authors precisely
+introduce a system integrating gradual typing with contracts \textit{à la
+Findler \& Felleisen}. Nickel adopts a similar type system, with both statically
+typed terms, dynamically typed terms, and first-class contracts. Higher-order
+contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
+and~\ref{RootCauseOfBlame} that this paper studied in detail.
+
+\subsection{Unions and intersections in gradual typing}
+
+Castagna et. al \cite{gradualCastagna} introduce a gradual type system based on
+so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types
+feature unions, intersections, negation types together with a notion of
+subtyping. This work adheres to the \emph{static first school} (see
+~\cite{practiceTheoryGrad}) of gradual typing: the reason for gradual types is
+to allow for a less precise type information. It follows that their goals and
+constraints are somehow different, resulting in the absence of first-class
+contracts. As for any gradual type system, they do have to implement casts -
+that are very close to contracts - for unions and intersections, but these casts
+are neither visible nor available to the programmer. Obviously, no contracts
+mean no user-defined contracts as well.
+
+The implementation of casts relies on abstract interpretation, leading to a
+complex specification, whose implementation is straightforward. It is not clear
+either how much these casts rely on static type information to function
+correctly, possibly hampering their transformation to more general standalone
+contracts.
 
@@
 The design and implementation of programming languages is a challenging task. It
-is a delicate exercise to balance complexity, expressivity, efficiency, and
-user experience all together. Unfortunately for designers, a language cannot
-usually be reduced to a simple sum of small independent features. They are
-rather composed of a lot of moving parts that interact in unexpected ways. The
-more a language grows, the more any new addition has to be weighted carefully,
-as it has the potential to interact with any of the existing features.
-
-Part of recent developments revolves around improving safety of
-languages, ideally with little to none effort required by the programmers.
-Contracts are crucial to this endeavour, they are fundamental to
-allow a safe interaction between typed and untyped code in gradually typed languages;
-but they are also interesting in themselves in a fully dynamically typed
-language (and even in a fully statically typed language). Contracts are a
-precious idiom for data validation or to enforce runtime pre- and post-condition that
-would be hard to ensure statically.
+is a delicate exercise to balance complexity, expressivity, efficiency, and user
+experience all together. Unfortunately for designers, a language cannot usually
+be reduced to a simple sum of small independent features. They are rather
+composed of a lot of moving parts that interact in unexpected ways. The more a
+language grows, the more any new addition has to be weighted carefully, as it
+has the potential to interact with any of the existing features.
+
+Part of recent developments revolves around improving safety of languages,
+ideally with little to none effort required by the programmers.  Contracts are
+crucial to this endeavour, they are fundamental to allow a safe interaction
+between typed and untyped code in gradually typed languages; but they are also
+interesting in themselves in a fully dynamically typed language (and even in a
+fully statically typed language). Contracts are a precious idiom for data
+validation or to enforce runtime pre- and post-condition that would be hard to
+ensure statically.
 
@@
 since they allow to express that a variable is in an unknown state, but within
-some bounded set of alternatives.
-Sadly, the implementation of union and intersection contracts is not as simple as
-it first appears. In conjunction with higher-order contracts, or even just basic
-contracts in a lazy language, they behave in a way that is reminiscent of
-side-effects. These behaviors were, for most of them, already known: we do not
-claim their discovery. However, as language implementers, we have made an actual
-attempt at including them in a concrete language. We exposed what, in our
-opinion, are the most important barriers and their impact on the rest of the
-language. Even in a pure setting, union and intersection contracts break the
-equivalence between call-by-name and call-by-value, and can consequently prevent
-optimizations. There are discrepancies between the expected and actual semantics
-that make some behaviors unintuitive. Their implementation is a non-trivial
-change that contaminates the evaluation of all other contracts, and the
-performance penalty is not obvious to assess.
+some bounded set of alternatives.  Sadly, the implementation of union and
+intersection contracts is not as simple as it first appears. In conjunction with
+higher-order contracts, they introduce side-effects in the language. These
+behaviors were, for most of them, already known: we do not claim their
+discovery. However, as language implementers, we have made an actual attempt at
+including them in a concrete language. We exposed what, in our opinion, are the
+most important barriers and their impact on the rest of the language. Even in a
+pure setting, union and intersection contracts break the equivalence between
+call-by-name and call-by-value, and can consequently prevent optimizations.
+There are discrepancies between the expected and actual semantics that make some
+behaviors unintuitive. Their implementation is a non-trivial change that
+contaminates the evaluation of all other contracts, and the performance penalty
+is not obvious to assess.
 

### 2021-06-09 paper.tex
Fix citation
Pass on intersection contracts

@@
 
-Intersections ($\cap$) are less prevalent. An intersection is a value that belongs to
-two types at the same time. An example is a number literal \nickel{1}
-which could be accepted as both an \nickel{Int} and a \nickel{Float} without
-having to perform conversions.
-In that case, a value of type \nickel{Int @& Float} can be used as a value of
-both type \nickel{Int} and \nickel{Float}.
-
-% \label{fig:intersection-subytping-ex}
-% \begin{lstlisting}[title={Value subtyping}]
-% 1 : Int @& Float
-% \end{lstlisting}
-
-The real usability of intersections shows up when intersecting
-functions, since it can be used to encode overloading.
-For instance, take a look at Figure
-\ref{fig:intersection-overloading-ex}, where the function
-\nickel{duplicate} works both as a function to duplicate
-arrays, as well as a function to duplicate strings.
-This is particular useful when using unions, since it's a
-good way to express that a function can deal with different
-shapes of data.
-For instance, on the same figure, you can use \nickel{duplicate}
-over some value with type \nickel{[String] @| String}, and know
-that it will work correctly.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let duplicate | ([String] -> [String]) @& (String -> String) =
-  fun x => x ++ x in
-let text | [String] @| String = ... in
-duplicate text
-\end{lstlisting}
-\caption{Duplicating an array of Strings or a String}
-\label{fig:intersection-overloading-ex}
-\end{figure}
-
-% \label{fig:intersection-overloading-ex}
-% \begin{lstlisting}[title={Overloading}]
-% incr: (Int -> Int) @& (Float -> Float) =
-%   fun x => x + 1
-% \end{lstlisting}
-
-They can also be used to combine dictionaries in an object-oriented fashion. For instance, in Figure \ref{fig:intersection-record}
-two types are defined \nickel{Animal} and \nickel{Pet}, and
-a variable that is compatible with both types is declared, with
-type \nickel{Animal @& Pet}.
+An intersection type \nickel{A @& B} validates values which are both
+valid for contract \nickel{A} and contract \nickel{B}.
+
+Intersection contracts (and types) are probably less prevalent than
+union in practical type systems. However, a function from a union is
+equivalent to an intersection. That is \nickel{(A @| B) ->
+  C}$\simeq$\nickel{(A -> C) @& (B -> C)}. So in a system with
+functions and unions, intersections are already morally present (and,
+for that matter, in a system with functions and intersections, unions
+are morally present). Some
+of our examples in Sections~\ref{sec:issues-sem}
+and~\ref{sec:issues-literature} are better expressed in terms of
+intersections, so it's best to include them.
+
+Yet intersections are useful in their own right. They can be used to
+combine dictionaries in the style of object-oriented inheritance. For instance, in
+Figure \ref{fig:intersection-record} two types are defined
+\nickel{Animal} and \nickel{Pet}, and a variable that is compatible
+with both types is declared, with type \nickel{Animal @& Pet}. This
+particular application is supported, for instance, by TypeScript.
 
@@
 
-% They can also be used to combine constraints (such as traits or interfaces),as
-% done in Scala 3 (Figure~\ref{fig:intersection-mixins-ex}).
-
-% \label{fig:intersection-mixins-ex}
-% \begin{lstlisting}[language=Scala,title={Structural mixins}]
-% Comparator: {cmp: T -> T -> Bool}
-% Shower: {show: T -> String}
-
-% showSmaller : (dict: Comparator & Shower) -> T -> T -> String
-% \end{lstlisting}
-
-Finally, intersection naturally show up as the dual of unions.
-Thus, in a system
-featuring unions, a union in a negative positions is a form of intersection
-(Figure~\ref{fig:intersection-dual}).
-
-\begin{figure}[h]
-\begin{lstlisting}[language = nickel]
-f : (Num @| Str) -> Bool
-(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
-\end{lstlisting}
-\caption{Intersection as the dual of unions}
-\label{fig:intersection-dual}
-\end{figure}
-
-\resolved{Either here or the next section, show examples of unintuitive behaviour
-of U and I. For example, I is not jos composition of contracts.
-And union has to be treated as any one of the options.
-Kindo of resolved for unions}
-% Things to settle: how to motivate them, how to compare them to ADTs. Purely
-% dynamic approach, or typed approach too?
+Another application of intersections shows up when intersecting
+functions: it can be used to encode overloading.
+For instance, take a look at Figure~\ref{fig:intersection-overloading-ex}, where the function
+\nickel{duplicate} works both as a function to duplicate
+arrays, as well as a function to duplicate strings.
+This is particularly useful when using unions, since it's a
+good way to express that a function can deal with different
+shapes of data.
+For instance, in the same figure, \nickel{duplicate} is (correctly) called
+on a value of type \nickel{[String] @| String}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let duplicate
+    | (List String -> List String)
+    @& (String -> String) =
+  fun x => x ++ x in
+let text | [String] @| String = ... in
+duplicate text
+\end{lstlisting}
+\caption{Duplicating an array of Strings or a String}
+\label{fig:intersection-overloading-ex}
+\end{figure}
 
@@
 contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
-and~\ref{RootCauseOfBlame} that this paper studied in detail.
+and~\cite{RootCauseOfBlame} that this paper explorted extensively.
 

### 2021-06-09 paper.tex
Typo

@@
 But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
-as "clearly wrong" with respect to overloading. There is nothing wrong per se :
+as ``clearly wrong'' with respect to overloading. There is nothing wrong per se :
 we could chose a semantics for intersections that validate this rule.  However,
@@
 component fails, a flag in a global state. In particular, the two threads don't
-interact directly, and don't raise blame by themselves, leading this example run
+interact directly, and don't raise blame by themselves, leading this example to run
 as expected.

### 2021-06-09 paper.tex
Add a sentence to introduce the word “blame”
We don't really use it a lot throughout the paper, so we may need to
remind the terminology in Section 5 when we discuss the literature.
Typo

@@
 at fault for a violation, hence it can report precise error
-messages. Compare Figure~\ref{fig:contract-reporting-wo} and
+messages. Contracts are said to \emph{blame} either the caller or the
+implementation. Compare Figure~\ref{fig:contract-reporting-wo} and
 Figure~\ref{fig:contract-reporting-w}: in
@@
 corresponding label in the environment. What's more, each time a function with a contract attached is applied,
-the whole \emph{context} (that is, the continuation of the application) must be
+the whole \emph{context} (that is, the continuation of the application)
 must be checked for compatibility with respect to each label of the contract

### 2021-06-09 paper.tex
Clean up

@@
 when starting from the algorithmic version of their system.
-
-% Another difficult issue Keil and Thiemann were able to solve, is the
-% fact that user defined contracts and intersection do not usually play nice
-% together, as shown on Section \ref{sec:flat-and-inter}.
-% They provide two different sets of rules for their calculus,
-% the first one is non deterministic, and its purpose is to provide
-% semantics that are simple to understand.
-% With these rules, the interaction between user defined contracts and
-% intersections becomes much simpler, since they are able to check
-% both sides of an intersection on separate threads.
-% This allows them to be sure that contracts from one branch of the intersection
-% won't get blamed while executing the other branch, and vice versa.
-%
-% As an example, consider the programs shown on Figure
-% \ref{fig:kt-non-deterministic}, if the Original program
-% gets executed, it will spawn two instances of said program,
-% one with the first branch of the intersection (\nickel{3 | Num}),
-% and the other one with the second branch (\nickel{3 | Positive}),
-% and these will get joined once they reach an equal point, in this case
-% it will be \nickel{3}.
-%
-% \begin{figure}[h]
-% \begin{lstlisting}[language=nickel, title=Original]
-% 3 | Num @& Positive
-% \end{lstlisting}
-% \begin{lstlisting}[language=nickel, title=First Program]
-% 3 | Num
-% \end{lstlisting}
-% \begin{lstlisting}[language=nickel, title=Second Program]
-% 3 | Positive
-% \end{lstlisting}
-% \begin{lstlisting}[language=nickel, title=Joined Program]
-% 3
-% \end{lstlisting}
-% \caption{Non deterministic intersection contract opening}
-% \label{fig:kt-non-deterministic}
-% \end{figure}
-%
-% The second set of rules is deterministic, and it is intended
-% to be simpler to implement.
-% However, this second set needs to perform operations
-% on the evaluation context, to decide whether or not to evaluate
-% a given contract.
-% \footnote{You can check more on this topic on Section 5.1 of
-% \cite{KeilThiemannUnionIntersection}}
-% For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
-% when using the value \nickel{(fun x => x) | Str -> Str)}
-% inside the execution of contract \nickel{C}, and applying it to \nickel{0},
-% the contract runtime would check the context of execution and it would
-% deduce that contract \nickel{Str -> Str} does not apply in this case,
-% skipping it.
-%
-% We see two problems here, a minor one is the fact that the two
-% different semantics defined for the calculus are really far apart
-% one from the other, both on the strategies used for contract resolution,
-% as well as on the complexity levels.
-\resolved{(Yann) Is it really a
-    problem per se? It is common thing to do (a nice declarative system
-    for proofs and an algorithmic one for implementation, proved equivalent).
-    Maybe the point is that the algorithmic system is way more complex than the
-declarative one
-(Teo) reworded}.
-% A second, major one,
-% is the dependency added over the evaluation context, since, again,
-% it makes composing this, with other features desired on programming languages,
-% difficult, or even impossible.
-\resolved{(Yann) I think we need to substantiate this claim. If this
-context dependency explains why CSE is invalid, maybe have a little example?
-(Teo) I removed it, I don't think there's an example, mainly since
-this context dependency is not part of the language. I do think it complicates
-implementation of said feature}
 \todo{(Yann) TODO: see if there's a low hanging complexity

### 2021-06-09 paper.tex
I don't think this adds anything, does it?

@@
 which~\cite{RootCauseOfBlame} builds upon, is to extend the notion of contract
-satisfaction to \emph{contexts} (i.e. a continuations), embodying the caller's
+satisfaction to \emph{contexts}, embodying the caller's
 obligations. A context $K$ is a term with a hole to be filled with a term (see

### 2021-06-09 paper.tex
Modify my bit for consistent terminology: values satisfy a contract

@@
 
-Like for static types, a valid value for the contract \nickel{A @| B}
-is a value which is valid either for contract \nickel{A} or for
+Like for static types, value which satisfies the contract \nickel{A @| B}
+is a value which satisfy either contract \nickel{A} or
 contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
@@
 
-An intersection type \nickel{A @& B} validates values which are both
-valid for contract \nickel{A} and contract \nickel{B}.
+An intersection type \nickel{A @& B} is satisfied by values which
+satisfy both contract \nickel{A} and contract \nickel{B}.
 
@@
 satisfaction to \emph{contexts}, embodying the caller's
-obligations. A context $K$ is a term with a hole to be filled with a term (see
-Figure~\ref{fig:valid-contexts} for examples). Then, $K \in \llbracket Num \to
+obligations. A context $K$ is a term with a hole to be filled with a term (Figure~\ref{fig:valid-contexts} has examples). Then, $K \in \llbracket Num \to
 Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})

### 2021-06-09 paper.tex
Complete pending sentence
Typo

@@
 intersection into two sub-contracts that are checked separately. If one
-component fails, a flag in a global state. In particular, the two threads don't
-interact directly, and don't raise blame by themselves, leading this example to run
-as expected.
+component fails, a corresponding flag is set in a global state. In particular,
+the two threads don't interact directly, and don't raise blame by themselves,
+leading this example to run as expected.
 
@@
 important contribution of~\cite{KeilThiemannUnionIntersection}, upon
-which~\cite{RootCauseOfBlame} builds upon, is to extend the notion of contract
+which~\cite{RootCauseOfBlame} builds, is to extend the notion of contract
 satisfaction to \emph{contexts}, embodying the caller's

### 2021-06-09 paper.tex
Kind of funny typo

@@
 this can probably be fixed along the lines
-of~\ref{KeilThiemannUnionIntersection}, where the notion of context used to
+of~\cite{KeilThiemannUnionIntersection}, where the notion of context used to
 define satisfaction is carefully restricted (\emph{elimination contexts}). While
 this kind of oversight doesn't lower the quality of the work
-of~\ref{RootCauseOfBlame} by any measure, it illustrates that contract
+of~\cite{RootCauseOfBlame} by any measure, it illustrates that contract
 satisfaction in presence of unions, intersections and functions is a tricky

### 2021-06-09 paper.tex
Typo

@@
 as ``clearly wrong'' with respect to overloading. There is nothing wrong per se :
-we could chose a semantics for intersections that validate this rule.  However,
+we could choose a semantics for intersections that validate this rule.  However,
 the resulting system wouldn't capture overloading, and is arguably not very
@@
 
-Now, combining these two context as in~\ref{fig:invalid-context} gives a context
+Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
 that doesn't satisfy \nickel{Num -> Num} nor \nickel{Bool -> Bool}.  According

### 2021-06-09 paper.tex
A modernised abstract

@@
 \begin{abstract}
-This paper reports an attempt to incorporate union and intersection types in
-Nickel, a configuration programming language with a gradual type system and
-first-class contracts. While the end result looks appealing, it turns out the
-concrete impact on both the design and the implementation of the language is
-unexpectedly high. We review the issues raised in particular by the
-implementation of contracts for unions and intersections, and why we think the
-cost ends up too high for the benefit. We conclude by suggesting different leads
-to represent and type unions.
+Union and intersection types are a staple of gradually typed language such
+as TypeScript. While it's long been recognized that union and
+intersections types are difficult to verify statically, it may appear
+at first that the dynamic part of gradual typing is actually pretty
+simple.
+
+It turns out however, that in presence of higher-order contracts union
+and intersection is deceptively difficult. The literature on
+higher-order contracts with union and intersection, while keenly aware
+of the fact, doesn't really explain why. We point and illustrate the
+problems and trade-offs inherent to union and intersection contracts,
+via example and a survey of the literature.
 \end{abstract}

### 2021-06-10 paper.tex
String -> Str, [] -> List

@@
 This same property is desirable of contracts as well, otherwise we
-lose an important benefits of dynamic checking. Preferably, we want to
+lose an important benefit of dynamic checking. Preferably, we want to
 be able to extend the universe of contracts with user-defined
@@
 For instance, in the same figure, \nickel{duplicate} is (correctly) called
-on a value of type \nickel{[String] @| String}.
+on a value of type \nickel{List Str @| Str}.
 
@@
 let duplicate
-    | (List String -> List String)
-    @& (String -> String) =
+    | (List Str -> List Str)
+    @& (Str -> Str) =
   fun x => x ++ x in
-let text | [String] @| String = ... in
+let text | List Str @| Str = ... in
 duplicate text
@@
   f 0 == 0) in
-let g | (string -> string) @& c
+let g | (Str -> Str) @& c
       = fun x => x
@@
 % fact, it is never applied to anything.  Checking if \nickel{C} has already
-% failed, and marking \nickel{String -> String} as failed negatively on \nickel{l}
-% is not enough, the contract \nickel{String -> String} should not have been used
+% failed, and marking \nickel{Str -> Str} as failed negatively on \nickel{l}
+% is not enough, the contract \nickel{Str -> Str} should not have been used
 % inside the evaluation of \nickel{C}, which adds an extra layer of complexity:
@@
   f 0 == 0) in
-let g | (String -> String) @& C
+let g | (Str -> Str) @& C
       = fun x => x

### 2021-06-10 paper.tex
Pass on whole text, fixed some comments

@@
 illustration and motivation. At its core, Nickel is the JSON data
-model, that is dictionaries (written \nickel{\{field1 = value1, ...,
-  fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
-with functions (\nickel{fun arg1 ... argn => body}) and
-let-definitions (\nickel{let id = value in exp}).\unsure{Arnaud: this
+model, and its conformed of:
+
+\begin{itemize}
+  \item dictionaries, written as:
+\begin{lstlisting}[language=nickel]
+{field1 = value1, ..., fieldn = valuen}
+\end{lstlisting}
+  \item arrays:
+\begin{lstlisting}[language=nickel]
+[x1, x2, ..., xn]
+\end{lstlisting}
+  \item functions:
+\begin{lstlisting}[language=nickel]
+fun arg1 ... argn => body
+\end{lstlisting}
+  \item and let-definitions:
+\begin{lstlisting}[language=nickel]
+let id = value in exp
+\end{lstlisting}  
+\end{itemize}
+
+% In this article, we will use the Nickel language~\cite{NickelRepo} as
+% illustration and motivation. At its core, Nickel is the JSON data
+% model, that is dictionaries (written \nickel{\{field1 = value1, ...,
+%   fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
+% with functions (\nickel{fun arg1 ... argn => body}) and
+% let-definitions (\nickel{let id = value in exp}).
+\resolved{Arnaud: this
   paragraph looks kind of bad, maybe it'll be better to put each
   syntax entry in an enumeration, we will see if there is still room
-  for this}
+  for this
+  Teo: Tried to enumerate the constructs, left the original commented}
 
@@
 A probably even more common pattern is a variable which can contain
-either a value of type a (say, a number) or the \typescript{null}
+either a value of type \typescript{A} (say, a number) or the \typescript{null}
 value. So much so, in fact, that MyPy defines a type alias
@@
 
-\unsure{Arnaud: this paragraph is pretty good, on the other hand, there
+\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
 is no value which lets me complete the example, so this example falls
-a bit flat.}
+a bit flat.
+Teo: Improved the example, also added another one commented that is, for me,
+more realistic, but I think the one uncommented is simpler.}
 We have shown how values of a union are built, is worth mentioning
@@
 we want it to insert an element to a list that is either an
-\typescript{Array<string>} or an \typescript{Array<number>},
-what value could we fill \typescript{...} with for this to typecheck?
+\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
+where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
+\footnote{Note that in TypeScript, union and intersection are written as
+\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
+as we do in Nickel.}.
+What value could we fill the square ($\square$) with for this to typecheck?
 Since we don't know what's the shape of \typescript{list},
-we can't know whether to push a \typescript{string} or a
-\typescript{number}.
-That is, when using a union, we have to treat it as if it where any one
-of the alternatives, in the case of the example, we should provide
-it with a value that is both a \typescript{string} and a \typescript{number}.
+we can't know whether to push a \typescript{42 | "Nickel"} or a
+\typescript{"Nickel" | 3.5}. In general,
+when using a union, we have to treat it as if it where any one
+of the alternatives. In the case of the example, we should provide
+it with a value that could be pushed to both arrays, the only possible one
+is \typescript{"Nickel"},
+since it's the only value that satisfies the intersection of
+\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.
+
+\info{Not planned, but nice place to smooth things out towards
+intersections}
 
@@
 \begin{lstlisting}[language=JavaScript]
-function addElem(list: Array<string> | Array<number>) {
-  list.push(...)
+function addElem(list: Array< 42 | "Nickel" > | Array< "Nickel" | 3.5 >) {
+    list.push((*@ $\square$ @*))
 }
@@
 \end{figure}
+
+% \begin{figure}[h]
+% \begin{lstlisting}[language=JavaScript]
+% function addElem(values: Array<{ Nix: string }> | Array<{ Nickel: number }>) {
+%     values.push({Nix: "Packages", Nickel: 32})
+% }
+% \end{lstlisting}
+% \caption{Adding an element to a union of two arrays}
+% \label{fig:addElem}
+% \end{figure}
 
@@
 equivalent to an intersection. That is \nickel{(A @| B) ->
-  C}$\simeq$\nickel{(A -> C) @& (B -> C)}. So in a system with
+  C} $\simeq$ \nickel{(A -> C) @& (B -> C)}. So in a system with
 functions and unions, intersections are already morally present (and,
@@
 \begin{lstlisting}[language=nickel]
-let Animal = { species: Str, race: Str } in
+let Animal =
+  { species: Str, breed: Str, name: Str } in
 let Pet = { owner: Str, name: Str } in
@@
   { species = "Canis Lupus",
-    race = "Australian Cattle Dog",
+    breed   = "Australian Cattle Dog",
     owner   = "Anonymous Author",
@@
 to be fundamentally incompatible with the desirable language
-features from Section~\ref{sec:feat-lang}. At least in there
+features from Section~\ref{sec:feat-lang}. At least in their
 full-blown form: in Section~\ref{sec:issues-literature} we will
@@
 
-In Nickel, the failure of a function contract can always be traced back to a
+In \unsure{Maybe specify its Nickel without Union or inter}Nickel,
+the failure of a function contract can always be traced back to a
 single call. For example, take the function \nickel{f} with a simple contract
@@
 
-In an higher-order intersection contract, blame is raised when:
+In a higher-order intersection contract, blame is raised when:
 \begin{description}
     \item[Faulty caller] The argument fails \emph{both} components.
-    \item[Faulty implementation] The function fails \emph{one} component that the
-        argument previously satisfied.
+    \item[Faulty implementation] The function fails at least \emph{one}
+        component that the argument previously satisfied.
 \end{description}
@@
 
-Curiously, union and intersection type systems, on which
-\cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
-exhibit this property. Subtyping allows to correctly type check a term like
-\nickel{g 1} as both \nickel{Num -> Num} as well as
-\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
-can be given the type \nickel{Num -> Num @& Bool -> Num}.
-\unsure{(Yann) I just reworked the wording, but I'm not totally sold on the
-argument}
+% Curiously, union and intersection type systems, on which
+% \cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
+% exhibit this property. Subtyping allows to correctly type check a term like
+% \nickel{g 1} as both \nickel{Num -> Num} as well as
+% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
+% can be given the type \nickel{Num -> Num @& Bool -> Num}.
+\resolved{(Yann) I just reworked the wording, but I'm not totally sold on the
+argument
+(Teo) I agree, I think it would need to be expanded for it to actually
+say something, I'll remove it}
 
@@
 Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
-intersection as much as possible.  \todo{This assumes the difficulty has been
-previously explained} This transformation flattens arbitrarily nested unions and
+intersection as much as possible.
+This transformation flattens arbitrarily nested unions and
 intersections to a union of intersections, making blame resolution more
@@
 development causes to recheck the contract $A$ that has been duplicated?}
-\unsure{Arnaud says: this whole
+\resolved{Arnaud says: this whole
 section is too abstract give concrete code and explain what goes
@@
 \begin{lstlisting}[numbers=none,title={First component}]
-g 0 | Str -> Str
+(g | Str -> Str) 0
 \end{lstlisting}
 \begin{lstlisting}[numbers=none,title={Second component}]
-g 0 | C
+(g | C) 0
 \end{lstlisting}
@@
 
-On the semantic side, one of the so-called the sound monitoring properties
+On the semantic side, one of the so-called sound monitoring properties
 linking the operational semantics to the denotational semantics (supposed to
@@
 in~\cite{KeilThiemannUnionIntersection} does accept this context as valid.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let f = (*@$\square$@*) in
+(f 3, f true)
+\end{lstlisting}
+\caption{Combined context}
+\label{fig:invalid-context}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let f = (*@$\square$@*) in
-(f 3, f true)
-\end{lstlisting}
-\caption{Combined context}
-\label{fig:invalid-context}
-\end{figure}
-
 % However, both our intuition, as the most common use case of intersections
@@
 combinators correspond to the naive interpretation of unions and intersections
-(as in Figure~\ref{fig:naive-impl}): \racket{and/c} (resp.  \racket{or/c})
+(as in Figure~\ref{fig:naive-impl} for intersections): \racket{and/c} (resp.  \racket{or/c})
 simply checks sequentially that all (resp. one) contracts pass.  In consequence,
@@
 When higher-order contracts are combined using \racket{or/c}, Racket imposes
-that contracts must distinguishable by their arity. Doing so, there is at most
+that contracts must be distinguishable by their arity. Doing so, there is at most
 one candidate that can be selected directly. This is illustrated in
@@
 distinct arities to avoid ambiguity. An example is provided in
-Figure~\ref{fig:rafig:racket-case-fun}. The resulting possibilities are similar
+Figure~\ref{fig:racket-case-fun}. The resulting possibilities are similar
 to static overloading, where one function can take additional parameters for
@@
 The implementation of casts relies on abstract interpretation, leading to a
-complex specification, whose implementation is straightforward. It is not clear
+complex specification, whose implementation is not straightforward. It is not clear
 either how much these casts rely on static type information to function

### 2021-06-10 paper.tex
Another pass, small fixes

@@
 as TypeScript. While it's long been recognized that union and
-intersections types are difficult to verify statically, it may appear
+intersection types are difficult to verify statically, it may appear
 at first that the dynamic part of gradual typing is actually pretty
@@
 belong either to a type \nickel{A} or \nickel{B}, are a popular tool
-when adding a static types to a dynamic language. In particular, both
+when adding static types to a dynamic language. In particular, both
 TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
@@
 illustration and motivation. At its core, Nickel is the JSON data
-model, and its conformed of:
+model, augmented with abstraction mechanisms, and its conformed of:
 
@@
   \item dictionaries, written as:
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=nickel]
 {field1 = value1, ..., fieldn = valuen}
@@
   \item arrays:
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=nickel]
 [x1, x2, ..., xn]
@@
   \item functions:
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=nickel]
 fun arg1 ... argn => body
@@
   \item and let-definitions:
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=nickel]
 let id = value in exp
@@
 schema validation? Fortunately, we don't! Enter
-\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic check
+\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic checks
 which can be partially delayed, yet errors can be reported
@@
 \subsection{Structure of this paper}
-\unsure{Make this section into a contributions section?}
+\unsure{Make this section into a contributions section?
+Or maybe just kill it?}
 Section~\ref{sec:intro} introduced the perspective of designing a configuration
@@
 
-\info{I think it's better to use [h] on figures, we can discuss it}
 \begin{figure}[h]
@@
   \begin{lstlisting}[language=JavaScript]
-    interface Circle { kind: "circle"; radius: number; }
-
-    interface Square { kind: "square"; sideLength: number; }
+    interface Circle {
+      kind: "circle";
+      radius: number; }
+
+    interface Square {
+      kind: "square";
+      sideLength: number; }
 
@@
 \typescript{"Nickel" | 3.5}. In general,
-when using a union, we have to treat it as if it where any one
+when using a union, we have to treat it as if it were any one
 of the alternatives. In the case of the example, we should provide
@@
 
+\unsure{Maybe it's better to give it in Typescript syntax?
+left it commented just in case}
+% \begin{figure}[h]
+% \begin{lstlisting}[language=JavaScript]
+% type Animal = {
+%   species: string,
+%   breed:   string,
+%   name:    string}
+
+% type Pet = {owner: string, name: string}
+
+% let myDog : Animal & Pet = {
+%   species : "Canis Lupus",
+%   breed   : "Australian Cattle Dog",
+%   owner   : "Anonymous Author",
+%   name    : "Juno" }
+% \end{lstlisting}
+% \caption{An animal that is also a pet}
+% \label{fig:intersection-record}
+% \end{figure}
+
 Another application of intersections shows up when intersecting
@@
 For instance, in the same figure, \nickel{duplicate} is (correctly) called
-on a value of type \nickel{List Str @| Str}.
+on a value of type \nickel{(List Str) @| Str}.
 
@@
   fun x => x ++ x in
-let text | List Str @| Str = ... in
+let text | (List Str) @| Str = ... in
 duplicate text
@@
 
-\subsubsection{case->}
+\paragraph{case->}
 

### 2021-06-10 paper.tex
Fix case in example
Typos

@@
 
-It turns out however, that in presence of higher-order contracts union
-and intersection is deceptively difficult. The literature on
-higher-order contracts with union and intersection, while keenly aware
-of the fact, doesn't really explain why. We point and illustrate the
-problems and trade-offs inherent to union and intersection contracts,
-via example and a survey of the literature.
+It turns out however, that in presence of higher-order contracts union and
+intersection are deceptively difficult. The literature on higher-order contracts
+with union and intersection, while keenly aware of the fact, doesn't really
+explain why. We point and illustrate the problems and trade-offs inherent to
+union and intersection contracts, via example and a survey of the literature.
 \end{abstract}
@@
 systems, has been quite interested in union
-types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.
+types~\cite{RootCauseOfBlame,gradualCastagna,ORTIN2011278,ToroTanterGradualUnion,KeilThiemannUnionIntersection}.
 
@@
 ambition to be able to manipulate configurations like Nixpkgs. With
-over 50\,000 packages, is one of the largest repository of software
+over 50\,000 packages, it is one of the largest repository of software
 packages in existence~\cite{repology}. Concretely, Nixpkgs is a
@@
 4 |
-5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
-  |              -----------------------
+5 | let makeHost = fun server ext => [(*@{\ldots}@*)] in
+  |                --------------------
   |                evaluated to this
@@
 be able to extend the universe of contracts with user-defined
-predicates (sometimes called flat contracts).
+predicates.
 
@@
 \typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
-where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
-\footnote{Note that in TypeScript, union and intersection are written as
+where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal
+types\footnote{Note that in TypeScript, union and intersection are written as
 \typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
@@
 
-Yet intersections are useful in their own right. They can be used to
+Yet, intersections are useful in their own right. They can be used to
 combine dictionaries in the style of object-oriented inheritance. For instance, in
@@
 \begin{lstlisting}[language=nickel]
-let c = contracts.frompred (fun f =>
+let c = contracts.fromPred (fun f =>
   f 0 == 0) in
@@
       = fun x => x
+in g 0
 \end{lstlisting}

### 2021-06-10 paper.tex
Fix code example

--- 2021-06-10 paper.tex
@@
 be able to extend the universe of contracts with user-defined
-predicates.
+predicates (sometimes called flat contracts).
 
@@
 \typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
-where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal
-types\footnote{Note that in TypeScript, union and intersection are written as
+where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
+\footnote{Note that in TypeScript, union and intersection are written as
 \typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
@@
 
-Yet, intersections are useful in their own right. They can be used to
+Yet intersections are useful in their own right. They can be used to
 combine dictionaries in the style of object-oriented inheritance. For instance, in

### 2021-06-10 paper.tex
Set US English dictionary in Emacs
Remove reference to flat contracts
User-defined contracts are a subset of flat contracts but these are not
the same thing. Flat contracts is basically non-composed contracts (base
types + user-defined).
Section 5: moving Section 4.3 in + subsection titles
*Multiple* inheritance

--- 2021-06-10 paper.tex
@@
 Yet, intersections are useful in their own right. They can be used to
-combine dictionaries in the style of object-oriented inheritance. For instance, in
+combine dictionaries in the style of object-oriented multiple inheritance. For instance, in
 Figure \ref{fig:intersection-record} two types are defined
@@
 contracts?}
-% But this should not have happened, since in the program of Figure
-% \ref{fig:inter-flat-contracts} \nickel{g} is never applied to \nickel{0}, in
-% fact, it is never applied to anything.  Checking if \nickel{C} has already
-% failed, and marking \nickel{Str -> Str} as failed negatively on \nickel{l}
-% is not enough, the contract \nickel{Str -> Str} should not have been used
-% inside the evaluation of \nickel{C}, which adds an extra layer of complexity:
-% contracts should be guarded, and execution depends on the shape of the
-% surrounding evaluation context.
-
-\subsection{Challenging the semantics}
+
+\section{Pragmatic trade-offs}
+\label{sec:issues-literature}
+
+Despite the difficulties of Section~\ref{sec:issues-sem}, union and
+intersection contracts are still sought after. In this section we turn
+to existing systems with union and intersection contracts in the
+literature and in implementations.
+
+These systems all make trade-offs, sacrificing some features of union
+and intersection contracts to preserve language features. We survey
+and discuss those trade-offs and their implications.
+
+\subsection{A semantics based on types}
 
@@
 
-% Curiously, union and intersection type systems, on which
-% \cite{KeilThiemannUnionIntersection} are basing their semantics on, don't
-% exhibit this property. Subtyping allows to correctly type check a term like
-% \nickel{g 1} as both \nickel{Num -> Num} as well as
-% \nickel{Bool -> Num}, and then deducing that \nickel{g 1}
-% can be given the type \nickel{Num -> Num @& Bool -> Num}.
-\resolved{(Yann) I just reworked the wording, but I'm not totally sold on the
-argument
-(Teo) I agree, I think it would need to be expanded for it to actually
-say something, I'll remove it}
-
-\section{Pragmatic trade-offs}
-\label{sec:issues-literature}
-
-Despite the difficulties of Section~\ref{sec:issues-sem}, union and
-intersection contracts are still sought after. In this section we turn
-to existing systems with union and intersection contracts in the
-literature and in implementations.
-
-These systems all make trade-offs, sacrificing some features of union
-and intersection contracts to preserve language features. We survey
-and discuss those trade-offs and their implications.
-
-\subsection{Keil and Thiemann}
+\subsection{A first realization}
 \label{sec:keil-thiemann}
@@
 
-\subsection{Williams, Morris, and Wadler}
+\subsection{Monitoring properties}
 \label{sec:will-morr-wadl}
@@
 
+% Local Variables:
+% ispell-dictionary: "american"
+% End:
+
 % LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy

### 2021-06-10 paper.tex
Section 5.1

@@
 
-\subsection{A semantics based on types}
-
-Previous points of this section were about how implementing the expected
-semantics for union and intersection has a profound impact on the language. Let
-us end the section by investigating a different aspect: does the
-semantics correspond to the programmer's intuition?
-
-But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
-as ``clearly wrong'' with respect to overloading. There is nothing wrong per se :
-we could choose a semantics for intersections that validate this rule.  However,
-the resulting system wouldn't capture overloading, and is arguably not very
-interesting nor useful. In~\cite{KeilThiemannUnionIntersection}, the authors
-argue for basing the semantics of contracts on union and intersection type
-systems, which have a simple and natural definition from which the other
-properties follow. This approach makes sense, and has been continued in future
-works, and in particular in~\cite{RootCauseOfBlame}.
-
-But even this semantics is no panacea. Figure~\ref{fig:intersection-distribution}
-shows an example that intuitively looks like it should succeeed, returning
-\nickel{(1, 1)}. However, under the semantics used both
-in~\cite{KeilThiemannUnionIntersection} and~\cite{RootCauseOfBlame}, this program
-fails with a contract violation.
-
-% Before wrapping up the section, we would like to show that maybe, union and
-% intersection contracts are not exactly what we would want them to be.  As stated
-% throughout this work, programmers coming from dynamic languages are accustomed
-% to perform safety checks (data validation) with great amounts of flexibility and
-% the whole power of their language of choice at their disposal.
-%
-% For instance, a Python programmer may have no problem in checking at runtime,
-% that a given variable is an integer if a given flag is not set, as shown in
-% Figure \ref{fig:python-assertion}; while a Haskell programmer might close her
-% eyes and leave the room instantly.
-%
-% \begin{figure}[h]
-% \begin{lstlisting}[language=python]
-% result = 42 if flag else "Disabled"
-% ...
-% assert (
-%   (not flag)
-%   or isinstance(result, int)
-% )
-% \end{lstlisting}
-% \caption{Asserting at runtime the type of a variable}
-% \label{fig:python-assertion}
-% \end{figure}
-%
-% These expectations are not really met by how unions and intersections
-% contracts are supposed to work.
-% For example, consider the Nickel code shown on Figure
-% \ref{fig:intersection-distrbution}, at first sight, this code should
-% complete execution successfully, returning \nickel{(1, 1)},
-% however, under both the Keil and Thiemann, as well as the
-% Williams, Morris, and Wadler works, this program will fail, with a violation
-% of the contract imposed to \nickel{g}.
+\subsection{A conductive semantics}
+
+In order to give a precise definition to what values ought to satisfy
+union and intersection contracts, \cite{KeilThiemannUnionIntersection}
+give a coinductively defined semantics inspired by union and
+intersection type systems. The key innovation of
+\cite{KeilThiemannUnionIntersection} is recognizing that giving a
+semantics to higher-order contracts requires defining not only what
+values satisfy a contract, but also what \emph{contexts} satisfy the
+contract. This models the situation where context may violate a
+contract by calling a function with an inappropriate argument.
+
+Concretely, given a contract $C$, \cite{KeilThiemannUnionIntersection}
+introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
+\rrbracket^-$ of values and contexts, respectively, satisfying the
+contract. They are defined by mutual induction and coinduction.
+
+This semantics has limited support for overloading. Consider the
+example in Figure~\ref{fig:intersection-distribution}: it could
+evaluate to the pair \nickel{(1, 1)}. But the coinductive semantics
+of~\cite{KeilThiemannUnionIntersection} rejects it as a contract
+violation. This can be phrased pithily as the fact that the
+coinductive semantics doesn't validate the property \nickel{A -> B @&
+  A -> C}$\simeq$\nickel{A -> (B @& C)}.
 
@@
 
-This discrepancy is inherent to the chosen semantics .  When applying a function
-with an intersection contract, only one component of the intersection must be
-selected for the function. The actual choice of component may be delayed to
-later, but the point is that it must be the same across all usages. In
-Figure~\ref{fig:intersection-distribution}, the contract for \nickel{h} can in
-consequence be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but not both.
-This is a manifestation of the following more general defect: the distributivity
-law given in Figure~\ref{fig:distributivity-law} is not valid in the semantics
-given in~\cite{KeilThiemannUnionIntersection}, while it is both sound to add and
-intuitive.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-A -> B @& A -> C (*@ $\simeq$ @*) A -> (B @& C)
-\end{lstlisting}
-\caption{Distributivity law}
-\label{fig:distributivity-law}
-\end{figure}
+Note that if the last two lines of
+Figure~\ref{fig:intersection-distribution} had read \nickel{(g 1 1, g
+  1 true)} instead, then the coinductive semantics would accept the
+example. It implies that under the coinductive semantics,
+common-subexpression elimination (see Section~\ref{sec:optimizations})
+is quite perilous.
 
@@
 % LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
-% LocalWords:  TypeScript serializable serialization
+% LocalWords:  TypeScript serializable serialization coinductively
+% LocalWords:  coinduction coinductive

### 2021-06-10 paper.tex
Move and rework example of union

@@
 \emph{pure} terms, that are terms without side-effects. In a pure language, or a
-language with effect tracking like Haskell\todo{Cite a different language, or
-remove citation}, pure terms can be safely identified for applying
-optimizations.
+language with effect tracking like PureScript\resolved{Cite a different
+    language, or remove citation. (ADDED) tbh PureScript is still
+statically-typed, compiled lang, but who knows: the "script" in the name and the
+web orientation may still be more appealing to the audience of DLS}, pure terms
+can be safely identified for applying optimizations.
 
 Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
-unions and intersections can make \ref{eq:cbn-expansion} unsound
-\emph{even in a pure setting}, making the optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
+unions and intersections can make \ref{eq:cbn-expansion} unsound \emph{even in a
+pure setting}, preventing the optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.
 
@@
 
-Like for static types, value which satisfies the contract \nickel{A @| B}
-is a value which satisfy either contract \nickel{A} or
+Like for static types, a value which satisfies the contract \nickel{A @| B}
+is a value which satisfies either contract \nickel{A} or
 contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
@@
 more realistic, but I think the one uncommented is simpler.}
-We have shown how values of a union are built, is worth mentioning
-how they are used.
-For instance, consider the TypeScript function on Figure \ref{fig:addElem},
-we want it to insert an element to a list that is either an
-\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
-where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal
-types\footnote{Note that in TypeScript, union and intersection are written as
-\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
-as we do in Nickel.}.
-What value could we fill the square ($\square$) with for this to typecheck?
-Since we don't know what's the shape of \typescript{list},
-we can't know whether to push a \typescript{42 | "Nickel"} or a
-\typescript{"Nickel" | 3.5}. In general,
-when using a union, we have to treat it as if it were any one
-of the alternatives. In the case of the example, we should provide
-it with a value that could be pushed to both arrays, the only possible one
-is \typescript{"Nickel"},
-since it's the only value that satisfies the intersection of
-\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.
-
-\info{Not planned, but nice place to smooth things out towards
-intersections}
-
-\begin{figure}[h]
-\begin{lstlisting}[language=JavaScript]
-function addElem(list: Array< 42 | "Nickel" > | Array< "Nickel" | 3.5 >) {
-    list.push((*@ $\square$ @*))
-}
-\end{lstlisting}
-\caption{Adding an element to a union of two arrays}
-\label{fig:addElem}
-\end{figure}
 
@@
 
-Intersection contracts (and types) are probably less prevalent than
-union in practical type systems. However, a function from a union is
-equivalent to an intersection. That is \nickel{(A @| B) ->
-  C} $\simeq$ \nickel{(A -> C) @& (B -> C)}. So in a system with
-functions and unions, intersections are already morally present (and,
-for that matter, in a system with functions and intersections, unions
-are morally present). Some
-of our examples in Sections~\ref{sec:issues-sem}
-and~\ref{sec:issues-literature} are better expressed in terms of
-intersections, so it's best to include them.
+Intersection contracts (and types) are probably less prevalent than union in
+practical type systems. However, a function from a union is equivalent to an
+intersection. That is \nickel{(A @| B) -> C} $\simeq$ \nickel{(A -> C) @& (B ->
+C)}. So in a system with functions and unions, intersections are already morally
+present (and, for that matter, in a system with functions and intersections,
+unions are morally present). Some of our examples in
+Sections~\ref{sec:issues-sem} and~\ref{sec:issues-literature} are better
+expressed in terms of intersections, so it's best to include them.
+
+Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
+function \nickel{appendDate} append an element to \nickel{list} whose type is
+only known to be in union of two lists, each using a different representation.
+Both alternatives support falling back to a simple string for unparsed dates.
+Faced with these two possibilities, \nickel{appendDate} can only append a
+value which fits both types: this is precisely the intersection
+\nickel{(Date @| Str) @& (DateWeek @| Str)}, that is, \nickel{Str}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=JavaScript]
+let Date = {day | Num, month | Num, year | Num} in
+let DateWeek = {dayOfWeek | Num, week | Num, year | Num} in
+let appendDate | (List (Date @| Str)
+                 @| List (DateWeek @| Str))
+                 -> List
+  = fun list =>
+  lists.cons list "01/01/2021"
+}
+\end{lstlisting}
+\caption{Adding an element to a union of two arrays}
+\label{fig:addElem}
+\end{figure}
 
@@
 let Animal =
-  { species: Str, breed: Str, name: Str } in
-let Pet = { owner: Str, name: Str } in
+  { species | Str, breed | Str, name | Str } in
+let Pet = { owner | Str, name | Str } in
 let myDog | Animal @& Pet =

### 2021-06-10 paper.tex
Section 5.2

@@
 
-\subsection{A conductive semantics}
+\subsection{A coinductive semantics}
+\label{sec:coinductive-sem}
 
@@
 contract composed of nested unions, intersections and higher-order contracts
-appears to require complex book-keeping in order to correctly raise blame.
-
-An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
-these nested contracts using the De Morgan law of
-Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
-intersection as much as possible.
-This transformation flattens arbitrarily nested unions and
-intersections to a union of intersections, making blame resolution more
-manageable.\unsure{(Yann) find better (more precise) wording}
-
-\begin{figure}[h]
-\begin{lstlisting}[frame=none, numbers=none, language=nickel]
-A @& (B @| C) (*@ $\simeq$ @*) (A @& B) @| (A @& C)
-\end{lstlisting}
-\caption{De Morgan law for contracts}
-\label{fig:de-morgan-law}
-\end{figure}
-
-To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
+appears to require complex book-keeping in order to correctly raise
+blame.
+
+All of this still holds true of the coinductive semantics described in
+Section~\ref{sec:coinductive-sem}. Nevertheless,
+\cite{KeilThiemannUnionIntersection} gives an algorithmic system which
+is complete for their coinductive semantics. In a remarkable technical
+tour de force, their algorithmic system is compatible with
+user-defined contract (see Section~\ref{sec:flat-and-inter}).
+
+A key aspect of the approach of~\cite{KeilThiemannUnionIntersection}
+is to rewrite nested union and intersection contracts into a
+disjunctive normal form using the De Morgan's law \nickel{A @& (B @|
+  C)}$\simeq$\nickel{(A @& B) @| (A @& C)}. The goal is to be able to
+delay the choice of branch in intersections as much as possible.
+
+To implement contract verification, \cite{KeilThiemannUnionIntersection} resorts to
 specific reduction rules for unions and intersections which perform this
-rewriting on the fly. The result is that the reduction rules sometimes feel
-ad-hoc, and make us question the composability and extensibility of the system.
-In the words of Williams, Morris, and Wadler (see
-Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
+rewriting on the fly. This aspect is critiqued
+in~\cite{RootCauseOfBlame}:
+``the monitoring semantics for
 contracts of intersection and union types given by Keil and Thiemann are not
 uniform. (\ldots) If uniformity helps composition, then special cases can hinder
-composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
-that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
-potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
-re-evaluated in each branch of the right hand side of
-Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
-mechanism.
-\resolved{(Yann) Does
-it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
-development causes to recheck the contract $A$ that has been duplicated?}
-\resolved{Arnaud says: this whole
-section is too abstract give concrete code and explain what goes
-wrong. It's fine to give code in Nickel syntax}
-
-One important advantage of~\cite{KeilThiemannUnionIntersection} is to handle
-user-defined contracts. We explained in Section~\ref{sec:issues-sem} why a
-direct approach for higher-order intersection contracts (as done
-in~\cite{RootCauseOfBlame}) is not trivial to adapt for user-defined contracts.
-\cite{KeilThiemannUnionIntersection} provides two operational semantics of the
-contract system: a non-deterministic one, more intuitive and proof-friendly, and
-a deterministic one for actual implementation. The non-deterministic semantics
-cleanly handle intersections with user-defined contracts: as each component is
-checked in a separate thread, there is no direct interaction.
-
-Let us recast the problematic example of Figure~\ref{fig:inter-flat-contracts}
-in Figure~\ref{fig:inter-flat-contracts-2}). The execution now splits the
-intersection into two sub-contracts that are checked separately. If one
-component fails, a corresponding flag is set in a global state. In particular,
-the two threads don't interact directly, and don't raise blame by themselves,
-leading this example to run as expected.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel,title={Original program}]
-let C = contracts.fromPred (fun f =>
-  f 0 == 0) in
-let g | (Str -> Str) @& C
-      = fun x => x
-in g 0
-\end{lstlisting}
-\begin{lstlisting}[numbers=none,title={First component}]
-(g | Str -> Str) 0
-\end{lstlisting}
-\begin{lstlisting}[numbers=none,title={Second component}]
-(g | C) 0
-\end{lstlisting}
-\caption{Intersection and user defined contracts}
-\label{fig:inter-flat-contracts-2}
-\end{figure}
-
-In return, the complexity is offloaded to a separate procedure that must decide
-when to raise blame. This complexity is salient in the deterministic, sequential
-version~\cite[Section 5.1]{KeilThiemannUnionIntersection}. There,
-the evaluation procedure has to deal again with the problem of eliding contracts
-that are not \emph{compatible}~\cite[Section 5.1]{KeilThiemannUnionIntersection} (contracts are incompatible
-if they are coming from a different branch and should be ignored). In practice,
-the evaluation procedure maintains a global environment of contracts that is
-extended each time a non-trivial contract (union, intersection or higher-order)
-is opened. Then, any contract application must check the presence of the
-corresponding label in the environment. What's more, each time a function with a contract attached is applied,
-the whole \emph{context} (that is, the continuation of the application)
-must be checked for compatibility with respect to each label of the contract
-environment. Checking for compatibility involves searching the context for
-contract applications.
-
-Keil and Thiemann realize the tour de force of combining union and intersection
-contracts with user-defined contracts. The price to pay, though, is that
-implementing their solution in a reasonable way is far from obvious. Not only
-when starting from the more abstract, non-deterministic formulation, but even
-when starting from the algorithmic version of their system.
-\todo{(Yann) TODO: see if there's a low hanging complexity
-    argument we can make (like it's quadratic in the number of open contracts or
-whatever)}
+composition.''
+
+A cost of this approach is that the De Morgan's law
+\nickel{A @& (B @| C)}$\simeq$\nickel{(A @& B) @| (A @& C)} duplicates
+contract \nickel{A}, which will cause some contracts to be checked
+several times. This can be an issue with user-defined contracts which
+may include costly tests.
+
+Efficiency is also affected another way: each time a function with a
+contract attached is applied, the whole \emph{context} must be
+traverse to check for a compatibility property.
+
+The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on
+balance, a technically impressive realization of the coinductive
+semantics, though it is fairly complex and probably difficult to
+implement efficiently.
 

### 2021-06-10 paper.tex
Typo

@@
 contract attached is applied, the whole \emph{context} must be
-traverse to check for a compatibility property.
+traversed to check for a compatibility property.
 

### 2021-06-10 paper.tex
Section 5.3

@@
 \label{sec:keil-thiemann}
-\unsure{Subsection title?}
 
@@
 \label{sec:will-morr-wadl}
-\unsure{Subsection title?}
-
-As mentioned in Section~\ref{sec:issues-sem}, Williams, Morris, and Wadler,
-downright exclude user-defined contracts from the picture. This choice gives
-them more freedom in the quest of a more uniform operational semantics. While
-this can be a sensible choice (in a gradual typing context e.g.), this is a show
-stopper for our initial use-case of the Nickel configuration language, the main
-motivation being to use contracts as customizable validation functions.
-
-On the semantic side, one of the so-called sound monitoring properties
-linking the operational semantics to the denotational semantics (supposed to
-capture the intuitive interpretation) is at odds with the given interpretation.
-% On top of that, a fundamental problem arises on said paper,
-% symptom of the difficulty of establishing good enough, and yet
-% simple semantics for union and intersection dynamic checks.
-% One of their main results, is the fact that their calculus obeys
-% all of their expected monitoring properties for higher ordered, union
-% and intersection contracts.
-
-For a contract $A$, $\llbracket A \rrbracket^+$ defines the set of terms that
-satisfy $A$. For example,
-
-\begin{lstlisting}[language=nickel,frame=none,numbers=none]
-    if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
-\end{lstlisting}
-
-For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
-contract can be violated in two ways: by the function returning a value that is
-not a string, or by the caller providing an argument that is not a number.  The
-former is embodied by the same notion as before: $f \in \llbracket Num \to Str
-\rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
-important contribution of~\cite{KeilThiemannUnionIntersection}, upon
-which~\cite{RootCauseOfBlame} builds, is to extend the notion of contract
-satisfaction to \emph{contexts}, embodying the caller's
-obligations. A context $K$ is a term with a hole to be filled with a term (Figure~\ref{fig:valid-contexts} has examples). Then, $K \in \llbracket Num \to
-Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
-if it only applies the hole to arguments satisfying the \nickel{Num} contract.
-
-These interpretations $\llbracket . \rrbracket^+$ and $\llbracket .
-\rrbracket^-$ then need to be extended to the setting of unions and
-intersections. In particular, one of the definitions of~\cite{RootCauseOfBlame}
-is the following:
-
-\begin{figure}[h]
+
+Another realization of the coinductive semantics described in
+Section~\ref{sec:coinductive-sem}, is~\cite{RootCauseOfBlame}, which
+aims at simplifying the algorithmic system proposed
+by~\cite{KeilThiemannUnionIntersection} and described in
+Section~\ref{sec:keil-thiemann}.
+
+A key ingredient of~\cite{RootCauseOfBlame} is disallow user-defined
+contracts. This choice gives the authors more freedom in the quest of
+a more uniform operational semantics. This is sensible trade-off in
+the context of gradual typing à la TypeScript: the problem is to match
+contract with static types, and user-defined contracts don't have a
+static type equivalent. On the other hand, the cost would probably not
+be worth it for a configuration language like Nickel.
+
+% Some of this we may want to move to Section 5.1
+%
+% For a contract $A$, $\llbracket A \rrbracket^+$ defines
+% the set of terms that satisfy $A$. For example,
+
+% \begin{lstlisting}[language=nickel,frame=none,numbers=none]
+%     if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
+% \end{lstlisting}
+
+% For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
+% contract can be violated in two ways: by the function returning a value that is
+% not a string, or by the caller providing an argument that is not a number.  The
+% former is embodied by the same notion as before: $f \in \llbracket Num \to Str
+% \rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
+% important contribution of~\cite{KeilThiemannUnionIntersection}, upon
+% which~\cite{RootCauseOfBlame} builds, is to extend the notion of contract
+% satisfaction to \emph{contexts}, embodying the caller's
+% obligations. A context $K$ is a term with a hole to be filled with a term (Figure~\ref{fig:valid-contexts} has examples). Then, $K \in \llbracket Num \to
+% Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
+% if it only applies the hole to arguments satisfying the \nickel{Num} contract.
+
+As a means of proving the correctness of their simplified system, the
+authors introduce what they call \emph{sound monitoring
+  properties}. Here is the sound monitoring property for contexts of
+intersection contracts:
+
 $$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
-\caption{Sound monitoring property for intersection from~\cite{RootCauseOfBlame}}
-\label{fig:wmw-semantics}
-\end{figure}
 
@@
 
-Although valid, this interpretation is weaker than what~\cite{RootCauseOfBlame}
-actually describes it to be and than the semantics given
-in~\cite{KeilThiemannUnionIntersection}. Consider the two contexts presented on
+Although sound, this interpretation is weaker than what the
+coinductive semantics permit. Consider the two contexts presented on
 Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
@@
 Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
-that doesn't satisfy \nickel{Num -> Num} nor \nickel{Bool -> Bool}.  According
-to the semantics of~\cite{RootCauseOfBlame}, it thus doesn't satisfy \nickel{Num
--> Num @& Bool -> Bool}. Yet, our intuition tells us that this example is a
-perfectly valid context for this contract: an overloaded function satisfying
-both \nickel{Num -> Num} and \nickel{Bool -> Bool} can safely be applied to a
-number and to a boolean in a row. And indeed, the denotational semantics given
-in~\cite{KeilThiemannUnionIntersection} does accept this context as valid.
+that doesn't satisfy \nickel{Num -> Num} nor \hbox{\nickel{Bool -> Bool}.}  According
+to the sound monitoring property of intersection, Figure~\ref{fig:invalid-context} thus doesn't satisfy \nickel{Num
+-> Num @& Bool -> Bool}.
 
@@
 
-The issue is not as bad as it sounds: the system presented
-in~\cite{RootCauseOfBlame} actually corresponds to the wished, stronger
-semantics in practice. The specification is just not exactly the right one, but
-this can probably be fixed along the lines
-of~\cite{KeilThiemannUnionIntersection}, where the notion of context used to
-define satisfaction is carefully restricted (\emph{elimination contexts}). While
-this kind of oversight doesn't lower the quality of the work
-of~\cite{RootCauseOfBlame} by any measure, it illustrates that contract
-satisfaction in presence of unions, intersections and functions is a tricky
-notion. This difficulty contrasts with the standard framework of higher-order
-contracts where satisfaction is much more straightforward.
-% Next, we can combine these two contexts into one, that essentially
-% has one single missing value, that will be used independently in both contexts
-% from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
-% This context is not a valid \nickel{Num -> Num} context, since it will apply
-% the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
-% In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
-% Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
-% shown on Figure \ref{fig:wmw-semantics}, it is not a valid
-% \nickel{Num -> Num @& Bool -> Bool} context.
-
-% However, both our intuition, as the most common use case of intersections
-% (that is, overloading), tells us that this should work.
-% While this problem does not invalidate their work (in fact, the system they present
-% handles these cases correctly), it shows that existing research is still not on
-% good shape to handle the growing usecases being developed for
-% unions and intersections around dynamic languages.
-
-% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$
-
-% Keil and Thiemann did noticed this problem, and their contract satisfaction
-% rules only apply to elimination context, a particular kind of context
-% where the hole is either being applied to an argument
-% ($\square N$) or is used as an argument for a primitive
-% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).
-
-% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
-\resolved{(Yann) Maybe adding a concrete example of this (it can be the same term but with
-    concrete types like Number and String and simple contexts like application
-    to a dumb argument) is sufficient to make the point}
+The consequence is that~\cite{RootCauseOfBlame} doesn't prove their
+system complete for the coinductive semantics. It's probably just an
+oversight in the proof: we believe their system to be indeed complete
+for the coinductive semantics. But it does speak to the intrinsic
+complexity of union and intersection contracts: it is remarkably easy
+to get details wrong. This difficulty contrasts with the standard
+framework of higher-order contracts where satisfaction is much more
+straightforward.
 

### 2021-06-10 paper.tex
Small stylistic changes
Section 5.4

@@
                  @| List (DateWeek @| Str))
-                 -> List
-  = fun list =>
-  lists.cons list "01/01/2021"
+                 -> List =
+  fun list => lists.cons list "01/01/2021"
 }
@@
 languages, Racket is probably the one with the most comprehensive contract
-system\cite{RacketContracts}. Regarding union and intersections, Racket provides
-the \racket{and/c} and \racket{or/c} combinators for contract.  These
-combinators correspond to the naive interpretation of unions and intersections
-(as in Figure~\ref{fig:naive-impl} for intersections): \racket{and/c} (resp.  \racket{or/c})
-simply checks sequentially that all (resp. one) contracts pass.  In consequence,
-unions or intersections of higher-order contracts don't match intuition, and in
-particular intersections don't model overloading. As in
+system~\cite{RacketContracts}. Regarding union and intersections, Racket provides
+the \racket{and/c} and \racket{or/c} combinators for contract.
+
+The
+intersection combinator \racket{and/c} corresponds to the naive
+interpretation of intersection described in
+Section~\ref{sec:flat-and-inter}: applying contract \racket{(and/c A
+  B)} is like applying contract \racket{A} then applying contract \racket{B}. In
+particular \racket{and/c} doesn't model overloading. As in
 Figure~\ref{fig:naive-impl}, the example given in
@@
 both \racket{number?} and \racket{string?}.
-
-% Among those contracts, they provide \racket{and/c} and \racket{or/c} contracts.
-% These contracts are lightweight versions of intersections and unions, they simply
-% check that every contract is valid, in the case of \racket{and/c},
-% and that at least one contract
-% is valid, in the case of \racket{or/c}.
-
-% So, for instance, the usual usecase of intersection to encode bounded overloading
-% does not work on Racket. Take a look at the next example, the function
-% \racket{overload} should accept a call like \racket{(overload 3)}, but it fails
-% since 3 is not a valid \racket{string?}.
 
@@
 
-When higher-order contracts are combined using \racket{or/c}, Racket imposes
+The union combinator \racket{or/c} is similarly simple. It must be
+able to decide immediately which branch holds: \racket{or/c} is a
+simple Boolean disjunction. For instance, when higher-order contracts are combined using \racket{or/c}, Racket imposes
 that contracts must be distinguishable by their arity. Doing so, there is at most
@@
 hand, the program of Figure~\ref{code:racket:or/c:non-working} is rejected.
-
-% that could (potentially) apply to a function, these have to be
-% differentiated by first order markers.
-% For instance, the code on Figure~\ref{code:racket:or/c:working}
-% would correctly allow only to call
-% \racket{united} with a \racket{number?},
-% and it can make this distinction since it knows that
-% \racket{(lambda (x) x)} can not be a function that takes 2 parameters.
-% However, the code on Figure~\ref{code:racket:or/c:non-working} would fail
-% before running, since the \racket{or/c} contract is unable to know
-% which of the two function contracts to use.
-% Ideally, this last example should just behave as wrapping
-% the function with the \racket{(-> even? even?)} contract.
 
@@
 
-Racket does provide the \racket{case->} combinator which(partially) supports
-overloading. As for the \racket{or/c}, the candidate contracts must have
-distinct arities to avoid ambiguity. An example is provided in
+To compensate for the fact that \racket{and/c} doesn't support
+overloading,
+Racket provides a second intersection-like combinator: \racket{case->}. As for the \racket{or/c}, the candidate contracts must have
+distinct arities to avoid ambiguity, unlike \racket{or/c} the
+alternative chosen when the function is called rather than when the
+contract is applied to the function. An example is provided in
 Figure~\ref{fig:racket-case-fun}. The resulting possibilities are similar
 to static overloading, where one function can take additional parameters for
-example (e.g. as supported for Java methods). Unfortunately, it excludes the
+example (e.g. as supported for Java methods). On the other hand, it excludes the
 overloading of generic operations with fixed arity such as equality, comparison,
 arithmetic operators, and so on.
-
-% Racket does provide an alternative for overloading methods, called the
-% \racket{case->} contract, however, this overloading of methods only
-% works over contracts of functions with different amount of parameters.
-% It's a good alternative to allow limited overloading of functions.
-% As an example, consider the following non failing piece of code, and
-% try to understand how Racket can decide, at each application,
-% which function contract to use.
 
@@
 
-In conclusion, Racket also seems to avoid the difficulties of general unions and
-intersections. They make the pragmatic choice of a simple semantics that work
-for non higher-order contracts and a limited support for overloading.
-% Even if Racket's intent is not to implement union and intersection
-% dynamic checks, they do provide contracts that are, in some sense,
-% simplified versions of those.
-% In our eyes, Racket has accepted the difficulty of correctly and practically
-% combine union, intersection, and higher order contracts, and has choose
-% to prioritise the highe order ones.
+In conclusion, Racket, a programming language with a large user base,
+avoids the difficulties of general unions and intersections (Section~\ref{sec:issues-sem}). They make
+the pragmatic choice of a simple semantics with limited support for
+higher-order contracts in intersections and unions.
 
@@
 % LocalWords:  TypeScript serializable serialization coinductively
-% LocalWords:  coinduction coinductive
+% LocalWords:  coinduction coinductive combinator combinators arities
+% LocalWords:  disjunction

### 2021-06-10 paper.tex
s/Structure/Contributions/

@@
 
-\subsection{Structure of this paper}
-\unsure{Make this section into a contributions section?
-Or maybe just kill it?}
-Section~\ref{sec:intro} introduced the perspective of designing a configuration
-language. We presented the concept of contracts and detailed why contracts are
-particularly fitted for a configuration language, and more generally for data
-validation or program correctness in a dynamically typed language. The rest of
-the paper is structured as follows:
+\subsection{Contributions}
+Our contributions are as follows
 \begin{itemize}
-    \item Section~\ref{sec:feat-lang} outlines various common features of
-        programming languages that may interact in a bad way with union and
-        intersection contracts.
-    \item Section~\ref{sec:union-inter} introduces unions and intersections and
-        lists the possible motivations for adding them to a language.
-    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
-        encounter when trying to implement union and intersection contracts in
-        practice.
-    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
-        existing attempts to solve said difficulties.
-    \item Section~\ref{sec:related-work} provides some related work.
+\item We describe the fundamental difficulties caused by presence of
+  union and intersection contracts in a language, which are kept
+  implicit in the literature (Section~\ref{sec:issues-sem})
+\item We survey the various trade-offs which appear in implemented
+  languages and in the literature to work around these difficulties
+  (Section~\ref{sec:issues-literature})
 \end{itemize}
-
-Section~\ref{sec:conclusion} concludes the paper.
 

### 2021-06-10 paper.tex
Conclusion

@@
 
-The  design and implementation of programming languages is a challenging task. It
-is a delicate exercise to balance complexity, expressivity, efficiency, and user
-experience all together. Unfortunately for designers, a language cannot usually
-be reduced to a simple sum of small independent features. They are rather
-composed of a lot of moving parts that interact in unexpected ways. The more a
-language grows, the more any new addition has to be weighted carefully, as it
-has the potential to interact with any of the existing features.
-
-Part of recent developments revolves around improving safety of languages,
-ideally with little to none effort required by the programmers.  Contracts are
-crucial to this endeavour, they are fundamental to allow a safe interaction
-between typed and untyped code in gradually typed languages; but they are also
-interesting in themselves in a fully dynamically typed language (and even in a
-fully statically typed language). Contracts are a precious idiom for data
-validation or to enforce runtime pre- and post-condition that would be hard to
-ensure statically.
-
-% In spirit, contracts are similar to
-% assertions, but they are part of a principled framework, they are composable,
-% they have good error reporting and they are integrated as a language feature. In
-% particular, contracts natively handles higher-order functions and built-in data
-% structures. This makes them an great candidate for integration in the Nickel
-% configuration language.
-
-% This is only the beginning of the story. A language designer then has to choose
-% what contract combinators to provide among a number of possibilities: there is
-% the standard package including higher-order function contracts, record contracts
-% and list contracts.  Then, one could add polymorphic
-% contracts~\cite{BlameForAll}, dependent function
-% contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
-% intersection contracts are really appealing. Unions have a intuitive
-% interpretation as modeling alternatives and thus appear implicitly or explicitly
-% in a lot of data structures.
-
-When trying to enforce properties on dynamic languages, unions appear naturally,
-since they allow to express that a variable is in an unknown state, but within
-some bounded set of alternatives.  Sadly, the implementation of union and
-intersection contracts is not as simple as it first appears. In conjunction with
-higher-order contracts, they introduce side-effects in the language. These
-behaviors were, for most of them, already known: we do not claim their
-discovery. However, as language implementers, we have made an actual attempt at
-including them in a concrete language. We exposed what, in our opinion, are the
-most important barriers and their impact on the rest of the language. Even in a
-pure setting, union and intersection contracts break the equivalence between
-call-by-name and call-by-value, and can consequently prevent optimizations.
-There are discrepancies between the expected and actual semantics that make some
-behaviors unintuitive. Their implementation is a non-trivial change that
-contaminates the evaluation of all other contracts, and the performance penalty
-is not obvious to assess.
+Despite the fact that union and intersection of dynamic properties may
+at first appear like an easy task, as soon as they are combined with
+higher-order contracts for increased accuracy of error messages, they
+really aren't.
+
+The problem of union and intersection contracts is that they are not
+orthogonal to apparently independent features of programming
+languages. The mere presence of intersection and union contract
+induces computational effects, and can make it quite difficult to
+perform simple program optimizations such as inlining.
+
+Designing a language with union and intersection contracts means
+making difficult choices: some features of union and intersection
+contracts and of the rest of the language must be abandoned. Various
+trade-offs can be made, but it is worth mentioning that implementing a
+system with union and intersection contracts appears to be pretty
+complex a task when unions and intersections are fairly complete.
+
+It's hard not to have sympathy for the minimalist end of this
+spectrum, where, like Racket, a language only has a very simple notion
+of unions and intersections. Many applications of unions and
+intersections are not possible in such a context, but the presence of
+union and intersection contracts doesn't interact with the rest of the
+language. It's probably more manageable.
 
@@
 in an actual language. We hope that our attempt may serve as a cautionary tale:
-for union and intersection contracts, these gaps might be quite large in places.
+for union and intersection contracts, these gaps may be larger than
+they appear.
 

### 2021-06-10 paper.tex
s/Purity/Referential transparency/
Mild rework Section 2.2

@@
 
-\subsection{Purity}
+\subsection{Referential transparency}
 \label{sec:optimizations}
-
-\unsure{I (Arnaud) changed the title of this subsection. Maybe it's
-  not the best one, but if it is, then we ought to reorder the section
-  a little bit to fit the modified narrative.}
 
@@
 by the compiler or the interpreter. Even more so for functional languages, whose
-execution model is often far removed from actual processors, causing naive
-execution to exhibit unacceptable slowdowns.
-
-One crucial optimization is
-inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
-to make heavy use of functions, and a function call is a costly operation from
-the point of view of low-level execution. Inlining eliminates a function
-application by directly substituting the function for its definition at compile
-time (or before execution, for interpreted language). This is especially
+execution model is often far removed from the hardware, causing naive execution
+to exhibit unacceptable slowdowns.
+
+One such important optimization is inlining (Figure
+\ref{fig:optimizations-inlining-ex}). Functional programs tend to make heavy use
+of functions, and a function call is not a free operation: it usually involves a
+number of low-level operations such as saving and storing registers, pushing a new stack
+frame and jumping to and back from the function's body. Inlining eliminates a
+function call by directly substituting the function for its definition at
+compile time (or before execution, for interpreted language). This is especially
 efficient for small functions that are called repeatedly.
@@
 While inlining expands an expression by substituting a definition for its value,
-an opposite transformation is sometimes beneficial, for example when a composite
-expression is repeated several times. This causes the same expression to be
-wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
-consists in introducing a new identifier for this expression and using the
-identifier in place of the original occurrences
-(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.
+an opposite transformation can be beneficial when a composite expression occurs
+at several places. In this case, the same expression is wastefully recomputed at
+each occurrence. Common subexpression elimination (CSE) consists in storing the
+expression in a variable that is then used in place of the original occurrences
+(Figure \ref{fig:optimizations-cse-ex}), thus evaluating the expression once and
+for all.
 
@@
 
-Beyond CSE, optimizations such as loop-invariant code motion or
-let-floating~\cite{letFloating} apply the same principle of extracting out an invariant
-expression to avoid its recomputation respectively across loop iterations and
-function calls.
+Beyond CSE, other optimizations such as loop-invariant code motion or
+let-floating~\cite{letFloating} apply the same principle of extracting out an
+invariant expression to avoid recomputing it (respectively across loop
+iterations and function calls).
 
@@
 \begin{lstlisting}[language=nickel,title={Optimized program}]
-let g' = g y
+let g' = g y in
 let f' = fun x => g' (x + 1)
@@
 The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
-called. This may be costly, in particular in the presence of contracts: if
-\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
-on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
-to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
-Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
-let-floating transformation could indeed do (given \lstinline+g+ is pure, as
-explained below).
-
-The soundness of all these optimizations is tied to the invariance of the
-semantics of programs with respect to specific substitutions. Inlining requires
-that one can replace the application of a function by its body, which is
-basically $\beta$-reduction: as long as the arguments are evaluated following
-the language's strategy, it is a valid transformation. On the other hand,
-a CSE-like transformation on a term $M$ requires an expansion property of the form
+called. This may be costly, in particular in the presence of contracts: if the
+first argument of \lstinline+g+ must be a list with elements of a specific kind
+and if that precondition is enforced by a function contract (say \nickel{g |
+List Odd -> Even -> List Odd}), the additional cost is linear in the size of
+\lstinline+y+. A sensible thing to do is to factor \lstinline+g y+ out of
+\lstinline+f+ as in Figure~\ref{fig:optimizations-let-floating-ex}, which is
+something a let-floating transformation could indeed do (given \lstinline+g+ is
+pure, as detailed below).
+
+The soundness of these optimizations is tied to the validity of specific program
+equivalences. Inlining requires that one can replace the application of a
+function by its body, which is basically $\beta$-reduction: as long as the
+arguments are evaluated following the language's strategy, this is usually a
+valid transformation. However, Section~\ref{sec:issues-sem} exposes that the
+question of inlining a function with a contract attached is more subtle.
+
+A CSE-like transformation on a term $M$ requires on the other hand an
+equivalence of the form:
 
 \begin{equation}\label{eq:cbn-expansion}
-M[x/N] \simeq let~x~=~N~in~M
+M[N/x] \simeq let~x~=~N~in~M
 \end{equation}
 
-Here, $M[x/N]$ is the substitution of $x$ for $N$ in the term $M$.
-\ref{eq:cbn-expansion} has an evident call-by-name flavour, and consequently
-fails in general for a call-by-value language with side effects.  Indeed, define
-the following function $f$:
-
-\begin{lstlisting}[language=Nickel]
-let f x = print "hi";(x+1) in
-\end{lstlisting}
-
-Then:
-
-\begin{lstlisting}[language=Nickel]
+$M[N/x]$ stands for the substitution of $x$ for the term $N$ in the term $M$.
+This equation clearly fails in presence of side-effects, as demonstrated in
+Figure~\ref{fig:optim-invalid-cse}. In that example, \nickel{(f 1, f 1)} prints
+\nickel{"hi"} two times while \nickel{let y = f 1 in (y,y)} only prints it once.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=Nickel,title={Effectful function}]
+let f x = print "hi";(x+1)
+\end{lstlisting}
+\begin{lstlisting}[language=Nickel,title={Invalid expansion}]
 (f 1,f 1) (*@ $\not \simeq$ @*) let y = f 1 in (y,y)
 \end{lstlisting}
-
-In the example above, the first term prints \nickel{"hi"} two times while the
-second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
-\emph{pure} terms, that are terms without side-effects. In a pure language, or a
-language with effect tracking like PureScript\resolved{Cite a different
-    language, or remove citation. (ADDED) tbh PureScript is still
+\caption{Counter-example to (\ref{eq:cbn-expansion}) in presence of
+side-effects}
+\label{fig:optim-invalid-cse}
+\end{figure}
+
+However, (\ref{eq:cbn-expansion}) does hold for \emph{pure} terms, that are
+terms without side-effects. In a pure language, all these transformations are
+valid.  In impure languages, the situation varies: in some case a large subset
+of pure terms can be identified (in languages with effects tracking such as
+PureScript) to be safely transformed.  Otherwise, the compiler must stay
+conservative and only apply CSE to expressions it can prove are without
+side-effects (arithmetic expressions, for example). \resolved{Cite a
+different language, or remove citation.  (ADDED) tbh PureScript is still
 statically-typed, compiled lang, but who knows: the "script" in the name and the
-web orientation may still be more appealing to the audience of DLS}, pure terms
-can be safely identified for applying optimizations.
-
-Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
-unions and intersections can make \ref{eq:cbn-expansion} unsound \emph{even in a
-pure setting}, preventing the optimization of
+web orientation may still be more appealing to the audience of DLS}
+
+Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
+unions and intersections make (\ref{eq:cbn-expansion}) unsound \emph{even in a
+pure setting}, preventing the kind optimization of
 Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.

### 2021-06-10 paper.tex
Typo
Remove unnecessary paragraph

@@
 Although sound, this interpretation is weaker than what the
-coinductive semantics permit. Consider the two contexts presented on
+coinductive semantics permits. Consider the two contexts presented on
 Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
@@
 
-The implementation of casts relies on abstract interpretation, leading to a
-complex specification, whose implementation is not straightforward. It is not clear
-either how much these casts rely on static type information to function
-correctly, possibly hampering their transformation to more general standalone
-contracts.
-
 \section{Conclusion}

### 2021-06-10 paper.tex
Typo

@@
 constraints are somehow different, resulting in the absence of first-class
-contracts. As for any gradual type system, they do have to implement casts -
-that are very close to contracts - for unions and intersections, but these casts
+contracts. As for any gradual type system, they do have to implement
+casts~--~that are very close to contracts~--~for unions and
+intersections, but these casts
 are neither visible nor available to the programmer. Obviously, no contracts

### 2021-06-10 paper.tex
Typo
Remove unnecessary sentence

@@
 view, and it is thus not obvious how to extend the shared state approach to
-handle user-defined contracts. The authors of~\cite{RootCauseOfBlame}
-acknowledge this fact: they decide to simply drop user-defined contracts
-altogether.
+handle user-defined contracts.
 
@@
 
-A key ingredient of~\cite{RootCauseOfBlame} is disallow user-defined
+A key ingredient of~\cite{RootCauseOfBlame} is to disallow user-defined
 contracts. This choice gives the authors more freedom in the quest of
@@
 the context of gradual typing à la TypeScript: the problem is to match
-contract with static types, and user-defined contracts don't have a
+contracts with static types, and user-defined contracts don't have a
 static type equivalent. On the other hand, the cost would probably not

### 2021-06-11 paper.tex
Pre-arxiv submission cleaning

@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
+\documentclass[sigplan,screen,10pt]{acmart}
 %\usepackage{amssymb}
@@
 \usepackage{hyperref}
+\usepackage{microtype}
+\usepackage{pbalance}
 
 \definecolor{OliveGreen}{RGB}{128,128,0}
-%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%
-
-  % TOGGLE ME to turn off all the commentary:
-  \InputIfFileExists{no-editing-marks}{
-    \def\noeditingmarks{}
-  }
-
-  \usepackage{xargs}
-  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
-  % ^^ Need for pgfsyspdfmark apparently?
-  \ifx\noeditingmarks\undefined
-      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
-      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
-      \newenvironment{alt}{\color{red}}{}
-
-      \newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\info}[2][1=]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,#1]{#2}}
-      \newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
-      \newcommandx{\inconsistent}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\critical}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommand{\improvement}[1]{\todo[linecolor=pink,backgroundcolor=pink!25,bordercolor=pink]{#1}}
-      \newcommandx{\resolved}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}} % use this to mark a resolved question
-  \else
-  %    \newcommand{\Red}[1]{#1}
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{#1}
-      \newcommand{\note}[1]{}
-      \newenvironment{alt}{}{}
-  %    \renewcommand\todo[2]{}
-      \newcommand{\unsure}[2][1=]{}
-      \newcommand{\info}[2][1=]{}
-      \newcommand{\change}[2]{}
-      \newcommand{\inconsistent}[2]{}
-      \newcommand{\critical}[2]{}
-      \newcommand{\improvement}[1]{}
-      \newcommand{\resolved}[2]{}
-  \fi
-
-%%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%
 
@@
 % Copied from https://hal.inria.fr/hal-01140459/file/racket.inc.tex
-\lstdefinelanguage{racket} {
+\lstdefinelanguage{Racket} {
   morekeywords=[1]{define, define-syntax, define-macro, lambda, define-stream, stream-lambda},
@@
 \lstset{
-  language={nickel},
+  language={Nickel},
   basicstyle=\small\ttfamily, % Global Code Style
@@
   breaklines=true, % wrap lines if they don't fit
-  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
-  frameround=tttt, % make the frame round at all four corners
-  framesep=4pt, % quarter circle size of the round corners
-  numbers=left, % show line numbers at the left
-  numberstyle=\tiny\ttfamily, % style of the line numbers
+  % frame=trbl, % draw a frame at the top, right, left and bottom of the listing
+    % frameround=tttt, % make the frame round at all four corners
+  % framesep=4pt, % quarter circle size of the round corners
+  numbers=none, % show line numbers at the left
+  % numberstyle=\tiny\ttfamily, % style of the line numbers
   escapeinside={(*@}{@*)} % escape sequence to insert latex
@@
 \lstset{
-  escapeinside={(*@}{@*)} % escape sequence to insert latex
+  escapeinside={(*@}{@*)}, % escape sequence to insert latex
+  numbers=none, % for the camera-ready version: numbers overlap with margins
+  % Avoid overlap with margins
+  % framexleftmargin = -2pt,
+  % framexrightmargin = -2pt,
+  % framextopmargin = -2pt,
+  % framexbottommargin = -2pt
 }
+
 \newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
@@
 
-\bibliographystyle{ACM-Reference-Format}
-
 \begin{abstract}
-Union and intersection types are a staple of gradually typed language such
+Union and intersection types are a staple of gradually typed languages such
 as TypeScript. While it's long been recognized that union and
@@
 
-\title{Union and intersection contracts are hard, actually}
+\title{Union and Intersection Contracts Are Hard, Actually}
 \author{Teodoro Freund}
@@
 }
-\email{tfreund95@gmail.com}
+\email{tfreund@dc.uba.ar}
 
@@
 \email{arnaud.spiwack@tweag.io}
+
+%%% The following is specific to DLS '21 and the paper
+%%% 'Union and Intersection Contracts Are Hard, Actually'
+%%% by Teodoro Freund, Yann Hamdaoui, and Arnaud Spiwack.
+\setcopyright{acmlicensed}
+\acmPrice{15.00}
+\acmDOI{10.1145/3486602.3486767}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{splashws21dlsmain-p31-p}
+\acmISBN{978-1-4503-9105-4/21/10}
+\acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
+\acmBooktitle{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
+
 \maketitle
@@
 statically. So unions are really only worth it in gradually typed
-language where they formalise existing dynamically typed patterns. On
-the other hand, surely, for dynamic tests, unions are really easy: it
-is simply the Boolean disjunctions of two tests.
-
+language where they formalise existing dynamically typed patterns.
+
+Surely, for dynamic tests, unions ought to be really easy: they
+are simply the Boolean disjunction of two dynamic tests, that fail
+whenever one of those tests fail.
 Unfortunately, as we document in this article, as soon as you extend
@@
 
-\subsection{Configuration languages}
+\subsection{Configuration Languages}
 
@@
 illustration and motivation. At its core, Nickel is the JSON data
-model, augmented with abstraction mechanisms, and its conformed of:
+model, augmented with abstraction mechanisms, and it comprises:
 
 \begin{itemize}
-  \item dictionaries, written as:
+  \item dictionaries, written as
+    \footnote{Note that, unlike JSON, Nickel assigns values in
+    dictionaries using \nickel{=}, to keep \nickel{:} for type
+annotations.}:
 \begin{lstlisting}[frame=none,numbers=none,language=nickel]
@@
   \item arrays:
-\begin{lstlisting}[frame=none,numbers=none,language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
 [x1, x2, ..., xn]
@@
   \item functions:
-\begin{lstlisting}[frame=none,numbers=none,language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
 fun arg1 ... argn => body
@@
   \item and let-definitions:
-\begin{lstlisting}[frame=none,numbers=none,language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
 let id = value in exp
@@
 \end{itemize}
-
-% In this article, we will use the Nickel language~\cite{NickelRepo} as
-% illustration and motivation. At its core, Nickel is the JSON data
-% model, that is dictionaries (written \nickel{\{field1 = value1, ...,
-%   fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
-% with functions (\nickel{fun arg1 ... argn => body}) and
-% let-definitions (\nickel{let id = value in exp}).
-\resolved{Arnaud: this
-  paragraph looks kind of bad, maybe it'll be better to put each
-  syntax entry in an enumeration, we will see if there is still room
-  for this
-  Teo: Tried to enumerate the constructs, left the original commented}
 
@@
 \subsection{Contracts}
+\label{sub-sec:contracts}
 
@@
 \nickel{let id | C = value in exp} stands for \nickel{let id = (value
-  | C) in exp}.\unsure{This point used to be made, do we want to
-  reinsert it? However, Nickel programs are of a special kind: they are usually
-simple, terminating programs that run on fixed inputs. Correctness errors will
-show up at evaluation anyway, or will not matter (dead code).}
+  | C) in exp}.
 
@@
 are \emph{lazy}, that is, the values are only evaluated when
-explicitly required. For instance, when writing
+explicitly required. For instance, when evaluating the expression
 \hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
-evaluated.
+evaluated, even if \nickel{nixpkgs} contained a
+\nickel{world} package as well.
 
@@
 accurately. Contracts can respect laziness of dictionaries, and they
-can be use to add schema validation to functions as well (in fact
+can be used to add schema validation to functions as well (in fact
 functions were the original motivation for contracts).
@@
 There is no Boolean function which can check that a value has type
-\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str}
+\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str} \\
 checks for each call of the function whether
@@
   \begin{subfigure}[b]{0.48\linewidth}
-    \begin{lstlisting}[language=nickel]
+    \begin{lstlisting}[language=Nickel]
 let catHosts = fun last =>
   let hosts = ["foo.com", "bar.org"] in
-  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in
+  lists.fold (fun val acc =>
+    val ++ "," ++ acc) hosts last in
 
@@
     \begin{lstlisting}[language=nickel]
-let catHosts | Str -> Str = fun last =>
-  [(*@{\ldots}@*)]
-\end{lstlisting}
-\vspace{10ex}
+$let$ catHosts | Str -> Str $= fun last =>
+  let hosts = ["foo.com", "bar.org"] in
+  lists.fold (fun val acc =>
+    val ++ "," ++ acc) hosts last in
+
+let makeHost = fun server ext =>
+  server ++ "." ++ ext in
+
+catHosts (makeHost "google")$
+\end{lstlisting}
     \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
@@
 
-Unfortunately, the delayed check of contract, while essential to
+As we shall show, the delayed check of contract, while essential to
 ensuring that schema validation doesn't affect performance (or, indeed,
@@
 appreciated sibling, intersection contracts) quite problematic.
+While usual contracts require only one witness to show the invalidation
+of a contract, the introduction of unions makes
+the number of witnesses not bounded.
 
@@
 \begin{itemize}
-\item We describe the fundamental difficulties caused by presence of
+\item We describe the fundamental difficulties caused by pre-sence of
   union and intersection contracts in a language, which are kept
@@
 \item We survey the various trade-offs which appear in implemented
-  languages and in the literature to work around these difficulties
+  languages and in the academic literature to work around these difficulties
   (Section~\ref{sec:issues-literature})
@@
 
-\section{A typology of language features}
+\section{A Typology of Language Features}
 \label{sec:feat-lang}
@@
 
-\subsection{User-defined contracts}
+\subsection{User-Defined Contracts}
 \label{sec:flat-contracts}
@@
   \begin{center}
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let Port = contracts.fromPred (fun p =>
@@
 This type of contracts are present in many different languages,
-for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
+for instance, the Eiffel programming language~\cite{meyer1987eiffel}, the precursor
 of the Design by Contract philosophy, makes it possible to assert
 these kinds of expression as pre- and post-conditions on
-functions and as invariants on classes\cite{EiffelDesignByContract}.
+functions and as invariants on classes~\cite{EiffelDesignByContract}.
 
 The Racket programming language also has a system to work with
-contracts, powerful enough to define flat contracts, and
+contracts, powerful enough to define user-defined contracts, and
 to compose them with other kinds of dynamic checks,
 like higher order contracts or a lightweight take on union
-and intersection contracts\cite{RacketContracts}.
-
-\subsection{Referential transparency}
+and intersection contracts~\cite{RacketContracts}.
+
+\subsection{Referential Transparency}
 \label{sec:optimizations}
@@
   \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
+\begin{lstlisting}[language=Nickel,title={Source program}]
 let elem = fun elt =>
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
+\begin{lstlisting}[language=Nickel,title={Optimized program}]
 let subList = fun l1 l2 =>
-  lists.any (fun x => x == (lists.head l1)) l2
+  lists.any (fun x => x == (lists.head l1))
+            l2
   && subList (list.tail l1) l2
@@
   \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
+\begin{lstlisting}[language=Nickel,title={Source program}]
 let elemAtOrLast = fun index list =>
   if index > lists.length list - 1 then
-    lists.elemAt (lists.length list - 1) list
+    lists.elemAt (lists.length list - 1)
+                 list
   else
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
+\begin{lstlisting}[language=Nickel,title={Optimized program}]
 let elemAtOrLast = fun index list =>
@@
   \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
+\begin{lstlisting}[language=Nickel,title={Source program}]
 let f = fun x => g y (x + 1)
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
+\begin{lstlisting}[language=Nickel,title={Optimized program}]
 let g' = g y in
@@
 conservative and only apply CSE to expressions it can prove are without
-side-effects (arithmetic expressions, for example). \resolved{Cite a
-different language, or remove citation.  (ADDED) tbh PureScript is still
-statically-typed, compiled lang, but who knows: the "script" in the name and the
-web orientation may still be more appealing to the audience of DLS}
+side-effects (arithmetic expressions, for example).
 
 Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
-unions and intersections make (\ref{eq:cbn-expansion}) unsound \emph{even in a
-pure setting}, preventing the kind optimization of
+union and intersection contracts \emph{breaks referential transparency} and make
+(\ref{eq:cbn-expansion}) \emph{unsound}\footnote{(\ref{eq:cbn-expansion}) holds
+  in Nickel despite non-termination and contract-checking errors
+  because of lazy evaluation. In strict languages (\ref{eq:cbn-expansion}) needs to be restricted, nevertheless
+  union and intersection contracts make it worse.}, preventing the kind of optimization of
 Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.
 
-\section{Union \& intersection}
+\section{Union \& Intersection}
 \label{sec:union-inter}
 
-Let us now consider union and intersection contracts, before we
-explain, in Section~\ref{sec:issues-sem} how they can compromise the
-properties that we described in Section~\ref{sec:feat-lang}.
+Let us now consider union and intersection contracts, before we explain in
+Section~\ref{sec:issues-sem} how they can compromise the properties that we
+described in Section~\ref{sec:feat-lang}.
 
@@
 
-\paragraph{In gradually typed systems}
+\paragraph{In Gradually Typed Systems}
 
@@
 capture, in static types, as many programming patterns as possible
-from the underlying dynamically typed language (JavaScript for
+from the underlying dynamically typed language (Java\-Script for
 TypeScript and Python for MyPy). One such pattern is
@@
 
-% Arnaud: I'm not deleting this quite yet because we may want to use
-% these examples, I just didn't manage to make them fit in the flow of
-% the narrative.
-%
-% Unions are not only useful to model the outcome of alternative
-% branches of computations, but they can be used to
-% formalize very common patterns of data types:
-% \begin{itemize}
-%     \item To represent nullable, or optional, values.
-% \begin{lstlisting}[language=nickel]
-% optionalValue | Str @| Null
-% \end{lstlisting}
-
-%     \item For plain enumerations of literals.
-% \begin{lstlisting}[language=nickel]
-% protocol | "Http" @| "Ftp" @| "Sftp"
-% \end{lstlisting}
-
-%     \item To support different representations of the same data.
-% \begin{lstlisting}[language=nickel]
-% person | {name : Str, age : Int}
-%   @| {name : Str, dateOfBirth : Date}
-% \end{lstlisting}
-
-% \end{itemize}
-% All of which are common in configurations.
-
-\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
-is no value which lets me complete the example, so this example falls
-a bit flat.
-Teo: Improved the example, also added another one commented that is, for me,
-more realistic, but I think the one uncommented is simpler.}
-
-% \begin{figure}[h]
-% \begin{lstlisting}[language=JavaScript]
-% function addElem(values: Array<{ Nix: string }> | Array<{ Nickel: number }>) {
-%     values.push({Nix: "Packages", Nickel: 32})
-% }
-% \end{lstlisting}
-% \caption{Adding an element to a union of two arrays}
-% \label{fig:addElem}
-% \end{figure}
-
 \subsection{Intersections}
@@
 Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
-function \nickel{appendDate} append an element to \nickel{list} whose type is
-only known to be in union of two lists, each using a different representation.
-Both alternatives support falling back to a simple string for unparsed dates.
-Faced with these two possibilities, \nickel{appendDate} can only append a
-value which fits both types: this is precisely the intersection
-\nickel{(Date @| Str) @& (DateWeek @| Str)}, that is, \nickel{Str}.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=JavaScript]
+function \nickel{appendDate} appends an element to \nickel{list}, whose type is
+only known to be the union of two lists, each using a different representation.
+Because the return type is the same as the input type, \nickel{appendDate} must
+preserve this representation (\nickel{Date} and \nickel{DateWeek}
+cannot be mixed in a same list).  Both alternatives support falling back to
+a simple string for unparsed dates.  Faced with these two possibilities,
+\nickel{appendDate} can only append a value which fits both types: this is
+precisely the intersection \nickel{(Date @| Str) @& (DateWeek @| Str)}, that
+is, \nickel{Str}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=Nickel]
 let Date = {day | Num, month | Num, year | Num} in
@@
                  @| List (DateWeek @| Str))
-                 -> List =
-  fun list => lists.cons list "01/01/2021"
-}
+                 -> (List (Date @| Str)
+                 @| List (DateWeek @| Str)) =
+  fun list => lists.cons "01/01/2021" list
 \end{lstlisting}
@@
 
-Yet, intersections are useful in their own right. They can be used to
+Yet, intersection are useful in their own right: they can be used to
 combine dictionaries in the style of object-oriented multiple inheritance. For instance, in
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let Animal =
@@
 \end{figure}
-
-\unsure{Maybe it's better to give it in Typescript syntax?
-left it commented just in case}
-% \begin{figure}[h]
-% \begin{lstlisting}[language=JavaScript]
-% type Animal = {
-%   species: string,
-%   breed:   string,
-%   name:    string}
-
-% type Pet = {owner: string, name: string}
-
-% let myDog : Animal & Pet = {
-%   species : "Canis Lupus",
-%   breed   : "Australian Cattle Dog",
-%   owner   : "Anonymous Author",
-%   name    : "Juno" }
-% \end{lstlisting}
-% \caption{An animal that is also a pet}
-% \label{fig:intersection-record}
-% \end{figure}
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let duplicate
@@
 
-However appealing union and intersection contracts may be, they happen
-to be fundamentally incompatible with the desirable language
-features from Section~\ref{sec:feat-lang}. At least in their
-full-blown form: in Section~\ref{sec:issues-literature} we will
-discuss pragmatic restrictions of union and intersection contracts to
-recover some or all of the features.
-
-\subsection{Union contracts as a side-effect}
-
-In \unsure{Maybe specify its Nickel without Union or inter}Nickel,
-the failure of a function contract can always be traced back to a
+However appealing union and intersection contracts may be, they happen to be
+either hard to combine or even fundamentally incompatible with the desirable
+language features from Section~\ref{sec:feat-lang}. At least in their full-blown
+form: in Section~\ref{sec:issues-literature} we will discuss pragmatic
+restrictions of union and intersection contracts to recover some or all of the
+features.
+
+\subsection{Union Contracts as a Side-Effect}
+
+In Nickel, the failure of a function contract can always be traced back to a
 single call. For example, take the function \nickel{f} with a simple contract
 attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
-error blaming \nickel{f} because the return value of the second call \nickel{f 5}
-violates the \nickel{Positive} contract. The first call to \nickel{f} does
+error blaming \nickel{f} because the return value of the second call \nickel{f
+5} violates the \nickel{Positive} contract. The first call to \nickel{f} does
 not matter, and \nickel{f 5} is a single and independent witness of the contract
@@
 a \nickel{fail} primitive to abort the execution. In practice, applying a
-function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
+function contract to \nickel{f} replaces it with an \nickel{f'} that performs the
 additional checks. Thus, since the core language is pure (albeit
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let f | Positive -> Positive
@@
 contract. \nickel{f} is violating this contract once again, as it neither maps
-positive numbers to positive numbers nor positive numbers to negative numbers.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+all positive numbers to positive numbers nor to nonpositive numbers.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=Nickel]
 let f | (Positive -> Positive)
@@
 
-This behavior indicates that union contracts introduce side-effects. The result
+This behavior shows that union contracts introduce side-effects. The result
 of \nickel{f 5} now depends on the previous execution and more specifically on
-any prior call to \nickel{f}. This behavior of union contracts break the
-property~\ref{eq:cbn-expansion} introduced in Section~\ref{sec:optimizations},
-that is required to perform CSE-like optimizations in all generality. The
-candidate example of Figure~\ref{fig:optimizations-let-floating-ex} in
-Section~\ref{sec:optimizations} can't be optimized in general.
+any prior call to \nickel{f}. This behavior of union contracts breaks
+referential transparency, as well as the property~\ref{eq:cbn-expansion}
+introduced in Section~\ref{sec:optimizations}, that is required to perform
+CSE-like optimizations.
 
@@
 an original program and an optimized version where the common subexpression
-\nickel{f 1} has been eliminated. While equivalent in a pure language without
-contracts, these two programs behave differently because of unions:
+\nickel{f 1} has been eliminated. While equivalent in a pure language with only
+non-termination or plain higher-order contracts, these two programs behave
+differently because of unions:
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=Original]
+\begin{lstlisting}[language=Nickel, title=Original]
 let f | Num -> (Bool -> Num @| Bool -> Str)
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel, title=Optimized]
+\begin{lstlisting}[language=Nickel, title=Optimized]
 let f | Num -> (Bool -> Num @| Bool -> Str)
@@
 \end{lstlisting}
-\caption{Equivalent programs, with inlining or CSE applied}
+\caption{Equivalent programs with CSE applied}
 \label{fig:optimized-programs}
@@
 
-\unsure{These arguments should go in the purity section}
 To sum up, the addition of union contracts introduce side-effects in a pure
@@
 
-\subsection{Intersection with user-defined contracts}
+\subsection{Intersection with User-Defined Contracts}
 \label{sec:flat-and-inter}
 
-A natural - but naive - implementation of intersection contracts could be the
+A natural --- but naive --- implementation of intersection contracts could be the
 following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Naive
+\begin{lstlisting}[language=Nickel,frame=none,numbers=none,title={Naive
 decomposition}]
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Exchange law}]
+\begin{lstlisting}[language=Nickel,frame=none,numbers=none,title={Exchange law}]
 (A -> B) @& (C -> D) (*@ $\simeq$ @*) (A @& C) -> (B @& D)
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Overloaded identity}]
+\begin{lstlisting}[language=Nickel,title={Overloaded identity}]
 let g | Num -> Num @& Str -> Str
@@
 
-\begin{lstlisting}[language=nickel,frame=none,numbers=none]
+\begin{lstlisting}[language=Nickel,frame=none,numbers=none]
 x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
@@
 
-Shared state is represented by the label \nickel{l}. This is in essence the
+Shared state is represented by the label \nickel{l}. Such a shared state is in essence the
 approach proposed by Williams, Morris, and Wadler in~\cite{RootCauseOfBlame}.
@@
 % \begin{figure}[h]
-% \begin{lstlisting}[language=nickel, title=Implementation]
+% \begin{lstlisting}[language=Nickel, title=Implementation]
 % x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
@@
 
-However, their approach has a major drawback: it is not straightforwardly
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let C = contracts.fromPred (fun f =>
+  f 0 == 0) in
+let g | (Str -> Str) @& C
+      = fun x => x
+in g 0
+\end{lstlisting}
+\caption{Intersection and user defined contracts}
+\label{fig:inter-flat-contracts}
+\end{figure}
+
+However, this shared-state approach has a major drawback: it isn't straightforwardly
 compatible with user-defined contracts (introduced in Section
@@
 
-\begin{lstlisting}[language=nickel,frame=none,numbers=none,title={Stateful
+\begin{lstlisting}[language=Nickel,frame=none,numbers=none,title={Stateful
 decomposition}]
@@
 \end{lstlisting}
-
+%
 where \nickel{l} represents the shared state.  At this point,
@@
 
-\begin{lstlisting}[language=nickel,frame=none,numbers=none]
+\begin{lstlisting}[language=Nickel,frame=none,numbers=none]
 ((fun x => x) | (Str -> Str)[l]) 0 == 0.
@@
 
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let c = contracts.fromPred (fun f =>
-  f 0 == 0) in
-let g | (Str -> Str) @& c
-      = fun x => x
-in g 0
-\end{lstlisting}
-\caption{intersection and user defined contracts}
-\label{fig:inter-flat-contracts}
-\end{figure}
-
 Once again, intersection contracts introduce side-effects in the picture. What's
@@
 way, while they are an important feature for validation.
-\unsure{(Yann) Should we talk about thiemann here, who do support user-defined
-contracts?}
-
-\section{Pragmatic trade-offs}
+
+\section{Pragmatic Trade-Offs}
 \label{sec:issues-literature}
@@
 
-\subsection{A coinductive semantics}
+\subsection{A Coinductive Semantics}
 \label{sec:coinductive-sem}
@@
 In order to give a precise definition to what values ought to satisfy
-union and intersection contracts, \cite{KeilThiemannUnionIntersection}
+union and intersection contracts,
+\citeauthor{KeilThiemannUnionIntersection}~\cite{KeilThiemannUnionIntersection}
 give a coinductively defined semantics inspired by union and
-intersection type systems. The key innovation of
-\cite{KeilThiemannUnionIntersection} is recognizing that giving a
+intersection type systems.
+The key innovation of their work is recognizing that giving a
 semantics to higher-order contracts requires defining not only what
@@
 
-Concretely, given a contract $C$, \cite{KeilThiemannUnionIntersection}
+Concretely, given a contract $C$, \citeauthor{KeilThiemannUnionIntersection}
 introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
-\rrbracket^-$ of values and contexts, respectively, satisfying the
-contract. They are defined by mutual induction and coinduction.
-
-This semantics has limited support for overloading. Consider the
-example in Figure~\ref{fig:intersection-distribution}: it could
-evaluate to the pair \nickel{(1, 1)}. But the coinductive semantics
-of~\cite{KeilThiemannUnionIntersection} rejects it as a contract
-violation. This can be phrased pithily as the fact that the
-coinductive semantics doesn't validate the property \nickel{A -> B @&
-  A -> C}$\simeq$\nickel{A -> (B @& C)}.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\rrbracket^-$ of values and contexts, respectively, satisfying the contract.
+They are defined by mutual induction and coinduction.
+
+This semantics has limited support for overloading. Consider the example in
+Figure~\ref{fig:intersection-distribution}: it could evaluate to the pair
+\nickel{(1, 1)}. But \citeauthor{KeilThiemannUnionIntersection}'s coinductive
+semantics rejects it as a contract violation.  This can be phrased pithily as
+the fact that the coinductive semantics doesn't validate the property
+\nickel{A -> B @& A -> C}$\simeq$\nickel{A -> (B @& C)}.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=Nickel]
 let f = fun x y => x in
@@
 \end{figure}
+
+A solution, for the programmer, is to use an uncurried function
+\nickel{fun (x, y) => x}. So one way to think about this limitation is
+that currying function is not fully supported.
 
@@
 
-\subsection{A first realization}
+\subsection{A First Realization}
 \label{sec:keil-thiemann}
@@
 Section~\ref{sec:coinductive-sem}.  Nevertheless,
-\cite{KeilThiemannUnionIntersection} gives an algorithmic system which
-is complete for their coinductive semantics. In a remarkable technical
-tour de force, their algorithmic system is compatible with
-user-defined contract (see Section~\ref{sec:flat-and-inter}).
-
-A key aspect of the approach of~\cite{KeilThiemannUnionIntersection}
+\citeauthor{KeilThiemannUnionIntersection} give an algorithmic system which is
+complete for their coinductive semantics. In a remarkable technical tour de
+force, their algorithmic system allows for user-defined contracts (see
+Section~\ref{sec:flat-and-inter}).
+
+A key aspect of the approach of \citeauthor{KeilThiemannUnionIntersection}
 is to rewrite nested union and intersection contracts into a
@@
 
-To implement contract verification, \cite{KeilThiemannUnionIntersection} resorts to
-specific reduction rules for unions and intersections which perform this
-rewriting on the fly. This aspect is critiqued
-in~\cite{RootCauseOfBlame}:
-``the monitoring semantics for
+To implement contract verification, \citeauthor{KeilThiemannUnionIntersection}
+resort to specific reduction rules for unions and intersections which perform
+this rewriting on the fly. This aspect is critiqued in
+\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}: ``the monitoring semantics for
 contracts of intersection and union types given by Keil and Thiemann are not
@@
 
-The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on
-balance, a technically impressive realization of the coinductive
-semantics, though it is fairly complex and probably difficult to
-implement efficiently.
-
-\subsection{Monitoring properties}
+The algorithmic system of \citeauthor{KeilThiemannUnionIntersection}
+is, on balance, a technically impressive realization of the coinductive
+semantics that supports user-defined contracts, though it is fairly complex and
+probably difficult to implement efficiently.
+
+\subsection{Monitoring Properties}
 \label{sec:will-morr-wadl}
@@
 Another realization of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem}, is~\cite{RootCauseOfBlame}, which
-aims at simplifying the algorithmic system proposed
-by~\cite{KeilThiemannUnionIntersection} and described in
+Section~\ref{sec:coinductive-sem} is given in
+\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}, which aims at simplifying
+the algorithmic system proposed by Keil and Thiemann and described in
 Section~\ref{sec:keil-thiemann}.
 
-A key ingredient of~\cite{RootCauseOfBlame} is to disallow user-defined
-contracts. This choice gives the authors more freedom in the quest of
-a more uniform operational semantics. This is sensible trade-off in
-the context of gradual typing à la TypeScript: the problem is to match
-contracts with static types, and user-defined contracts don't have a
-static type equivalent. On the other hand, the cost would probably not
-be worth it for a configuration language like Nickel.
+A key ingredient of \citeauthor{RootCauseOfBlame} is to disallow user-defined contracts. This
+choice gives the authors more freedom in the quest of a more uniform operational
+semantics. This is sensible trade-off in the context of gradual typing à la
+TypeScript: the problem is to match contracts with static types, and
+user-defined contracts don't have a static type equivalent. On the other hand,
+the cost would probably not be worth it for a configuration language like
+Nickel.
 
@@
 
-% \begin{lstlisting}[language=nickel,frame=none,numbers=none]
+% \begin{lstlisting}[language=Nickel,frame=none,numbers=none]
 %     if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel, title=\nickel{Num -> Num} context]
+\begin{lstlisting}[language=Nickel, title=\nickel{Num -> Num} context]
 (*@$\square$@*) 3
 \end{lstlisting}
-\begin{lstlisting}[language=nickel, title=\nickel{Bool -> Bool} context]
+\begin{lstlisting}[language=Nickel, title=\nickel{Bool -> Bool} context]
 (*@$\square$@*) true
@@
 
-Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
+Now, combining these two contexts as in Figure~\ref{fig:invalid-context} gives a context
 that doesn't satisfy \nickel{Num -> Num} nor \hbox{\nickel{Bool -> Bool}.}  According
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let f = (*@$\square$@*) in
@@
 
-The consequence is that~\cite{RootCauseOfBlame} doesn't prove their
-system complete for the coinductive semantics. It's probably just an
-oversight in the proof: we believe their system to be indeed complete
-for the coinductive semantics. But it does speak to the intrinsic
-complexity of union and intersection contracts: it is remarkably easy
-to get details wrong. This difficulty contrasts with the standard
-framework of higher-order contracts where satisfaction is much more
-straightforward.
+The consequence is that \citeauthor{RootCauseOfBlame} don't prove their system
+complete for the coinductive semantics.  It's probably just an oversight in the
+proof: we believe their system to be indeed complete for the coinductive
+semantics. But it does speak to the intrinsic complexity of union and
+intersection contracts: it is remarkably easy to get details wrong.  This
+difficulty contrasts with the standard framework of higher-order contracts where
+satisfaction is much more straightforward.
 
@@
 languages, Racket is probably the one with the most comprehensive contract
-system~\cite{RacketContracts}. Regarding union and intersections, Racket provides
+system~\cite{RacketContracts}. Regarding union and intersection, Racket provides
 the \racket{and/c} and \racket{or/c} combinators for contract.
@@
 The
-intersection combinator \racket{and/c} corresponds to the naive
+\racket{and/c} combinator corresponds to the naive
 interpretation of intersection described in
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=racket]
+\begin{lstlisting}[language=Racket]
 (define/contract overload
@@
 
-\begin{lstlisting}[language=racket]
+\begin{lstlisting}[language=Racket]
 (define/contract united
@@
 \end{figure}
-
-\unsure{This section is overall a bit hard to follow. I
-  think I've figured out what it means, but it needs more love. Also
-  give an example which doesn't work and one that does for or/c.}
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=racket]
+\begin{lstlisting}[language=Racket]
 (define/contract overcase
@@
 
-\section{(More) Related work}
+\section{(More) Related Work}
 \label{sec:related-work}
-\info{Including a mention of statically typed systems with union}
-
-\unsure{WARNING: what is below is being reworked by Yann}
-
-\subsection{Higher-order contracts}
+
+\subsection{Higher-Order Contracts}
 
 Enforcing pre- and post-conditions at runtime is a widely established practice.
-In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
+In their foundational paper~\cite{FindlerFelleisenHOContracts}, Findler and
 Felleisen introduce higher-order contracts, a principled approach to run-time
@@
 typed terms, dynamically typed terms, and first-class contracts. Higher-order
-contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
-and~\cite{RootCauseOfBlame} that this paper explorted extensively.
-
-\subsection{Unions and intersections in gradual typing}
-
-Castagna et. al \cite{gradualCastagna} introduce a gradual type system based on
-so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types
-feature unions, intersections, negation types together with a notion of
-subtyping. This work adheres to the \emph{static first school} (see
-~\cite{practiceTheoryGrad}) of gradual typing: the reason for gradual types is
-to allow for a less precise type information. It follows that their goals and
-constraints are somehow different, resulting in the absence of first-class
-contracts. As for any gradual type system, they do have to implement
-casts~--~that are very close to contracts~--~for unions and
-intersections, but these casts
-are neither visible nor available to the programmer. Obviously, no contracts
-mean no user-defined contracts as well.
+contracts are the basis of the work of
+\citeauthor{KeilThiemannUnionIntersection}~\cite{KeilThiemannUnionIntersection}
+and \citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame} that this paper
+explored extensively.
+
+\subsection{Unions and Intersections in Gradual Typing}
+
+Castagna and Lanvin~\cite{CastagnaLanvinGradualUnionIntersection} introduce a
+gradual type system based on set-theoretic
+types~\cite{SetTheoreticTypes}.  Set-theoretic types feature unions,
+intersections, negation types together with a notion of subtyping. This work
+adheres to the \emph{static first school} (see~\cite{practiceTheoryGrad}) of
+gradual typing: the reason for gradual types is to allow for a less precise type
+information. It follows that their goals and constraints are somehow different,
+resulting in the absence of first-class contracts. As for any gradual type
+system, they do have to implement casts~---~that are very close to
+contracts~---~for unions and intersections, but these casts are neither visible
+nor available to the programmer. Obviously, no contracts mean no user-defined
+contracts as well.
+
+Castagna and Lanvin use abstract interpretation to derive the semantics of
+unions and intersections. In their words: ``the resulting definitions are
+quite technical and barely intuitive but they have the properties we seek for
+[\ldots]''.  This makes any comparison with the coinductive semantics of
+\citeauthor{KeilThiemannUnionIntersection} rather difficult. Castagna et
+al.~\cite{gradualCastagna} build on Castagna and Lanvin to add polymorphism in
+the picture. However, it is at the price of restricting the union and
+intersection part: it is not possible to assign intersection types to a function
+anymore.
+
+\subsection{Context Sensitive Contracts}
+
+The work by Dimoulas et al.~\cite{dimoulasESOP,dimoulasPOPL} presents
+a complete study on systems that may execute (and potentially invalidate)
+contracts inside other contracts, in particular through the study
+of dependent function contracts.
+Their ideas have the potential to help solve the problems outlined
+on Sections \ref{sec:flat-and-inter} and \ref{sec:keil-thiemann},
+however, it isn't clear how to leverage their techniques
+for intersections.
+
+% As a first obstacle, it's worth noting that their system is intended
+% to get stuck upon the checking of a contract in a context that shouldn't
+% be checked in, while to keep the semantics of unions and intersections
+% valid, we'd expect for that check to get skipped, and for execution
+% to continue.
+
+% > Also worth noting is that Dimoulas's work deals extensively with these kinds
+% > of contracts and there is no mention of it in this paper. His POPL 2011 and
+% > ESOP 2012 papers are relevant to such contracts.
+
+% Regarding Dimoulas's papers, their work presents a solution
+% for contracts that may, arbitrarily, break other contracts in their
+% evaluation. Although this looks related to this problem with
+% intersection and user-defined contracts, it is not clear how to
+% leverage Dimoulas's techniques for intersection in practice.
+
+% It's worth noting, for instance, that Dimoulas's systems behave by getting stuck
+% in the case an undesired event happens (a contract fails because of external,
+% semantically unrelated, values flow in). With unions and intersections as
+% developed by Keil and Thiemann, the expected outcome is rather to not execute a
+% contract under the wrong context.
 
@@
 
-Designing a language with union and intersection contracts means
+Designing a language with union and intersection contracts necessarily
+means
 making difficult choices: some features of union and intersection
@@
 union and intersection contracts doesn't interact with the rest of the
-language. It's probably more manageable.
+language. It's probably more manageable, and it's the approach that
+we are currently considering for Nickel.
 

### 2021-06-11 paper.tex
Pre-arxiv submission cleaning

--- 2021-06-10 paper.tex
@@
-\documentclass[sigplan,10pt,review,anonymous]{acmart}
+\documentclass[sigplan,10pt]{acmart}
 %\usepackage{amssymb}
@@
 \definecolor{OliveGreen}{RGB}{128,128,0}
-%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%
-
-  % TOGGLE ME to turn off all the commentary:
-  \InputIfFileExists{no-editing-marks}{
-    \def\noeditingmarks{}
-  }
-
-  \usepackage{xargs}
-  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
-  % ^^ Need for pgfsyspdfmark apparently?
-  \ifx\noeditingmarks\undefined
-      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
-      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
-      \newenvironment{alt}{\color{red}}{}
-
-      \newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\info}[2][1=]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,#1]{#2}}
-      \newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
-      \newcommandx{\inconsistent}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommandx{\critical}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
-      \newcommand{\improvement}[1]{\todo[linecolor=pink,backgroundcolor=pink!25,bordercolor=pink]{#1}}
-      \newcommandx{\resolved}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}} % use this to mark a resolved question
-  \else
-  %    \newcommand{\Red}[1]{#1}
-      \newcommand{\Red}[1]{{\color{red}{#1}}}
-      \newcommand{\newaudit}[1]{#1}
-      \newcommand{\note}[1]{}
-      \newenvironment{alt}{}{}
-  %    \renewcommand\todo[2]{}
-      \newcommand{\unsure}[2][1=]{}
-      \newcommand{\info}[2][1=]{}
-      \newcommand{\change}[2]{}
-      \newcommand{\inconsistent}[2]{}
-      \newcommand{\critical}[2]{}
-      \newcommand{\improvement}[1]{}
-      \newcommand{\resolved}[2]{}
-  \fi
-
-%%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%
 
@@
 }
-\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
-\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
-\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
+\newcommand{\nickel}[1]{\lstinline[language=Nickel]{#1}}
+\newcommand{\racket}[1]{\lstinline[language=Racket]{#1}}
 \newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
@@
   \item dictionaries, written as:
-\begin{lstlisting}[frame=none,numbers=none,language=nickel]
+\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
 {field1 = value1, ..., fieldn = valuen}
@@
   \begin{subfigure}[b]{0.48\linewidth}
-    \begin{lstlisting}[language=nickel]
+    \begin{lstlisting}[language=Nickel]
 let catHosts = fun last =>
@@
   \begin{subfigure}[b]{0.48\linewidth}
-    \begin{lstlisting}[language=nickel]
+    \begin{lstlisting}[language=Nickel]
 let catHosts | Str -> Str = fun last =>
@@
 \end{lstlisting}
-\begin{lstlisting}[language=nickel,title={Optimized program}]
+\begin{lstlisting}[language=Nickel,title={Optimized program}]
 let subList = fun l1 l2 =>
@@
   \begin{center}
-\begin{lstlisting}[language=nickel,title={Source program}]
+\begin{lstlisting}[language=Nickel,title={Source program}]
 let elemAtOrLast = fun index list =>
@@
 conservative and only apply CSE to expressions it can prove are without
-side-effects (arithmetic expressions, for example). \resolved{Cite a
-different language, or remove citation.  (ADDED) tbh PureScript is still
-statically-typed, compiled lang, but who knows: the "script" in the name and the
-web orientation may still be more appealing to the audience of DLS}
+side-effects (arithmetic expressions, for example).
 
@@
 
-In \unsure{Maybe specify its Nickel without Union or inter}Nickel,
-the failure of a function contract can always be traced back to a
+In Nickel, the failure of a function contract can always be traced back to a
 single call. For example, take the function \nickel{f} with a simple contract
 attached of Figure~\ref{fig:pos-to-pos}. The whole program fails with a contract
-error blaming \nickel{f} because the return value of the second call \nickel{f 5}
-violates the \nickel{Positive} contract. The first call to \nickel{f} does
+error blaming \nickel{f} because the return value of the second call \nickel{f
+5} violates the \nickel{Positive} contract. The first call to \nickel{f} does
 not matter, and \nickel{f 5} is a single and independent witness of the contract
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let f | (Positive -> Positive)
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let c = contracts.fromPred (fun f =>
@@
 way, while they are an important feature for validation.
-\unsure{(Yann) Should we talk about thiemann here, who do support user-defined
-contracts?}
 
@@
 \begin{figure}[h]
-\begin{lstlisting}[language=nickel]
+\begin{lstlisting}[language=Nickel]
 let f = fun x y => x in
@@
 \label{sec:related-work}
-\info{Including a mention of statically typed systems with union}
-
-\unsure{WARNING: what is below is being reworked by Yann}
 

### 2021-07-19 paper.tex
Add arxiv-specific settings, clean junk files
Implement minor corrections of reviewer 1

--- 2021-06-10 paper.tex
@@
   let hosts = ["foo.com", "bar.org"] in
-  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in
+  lists.fold (fun val acc =>
+    val ++ "," ++ acc) hosts last in
 
@@
 let catHosts | Str -> Str = fun last =>
-  [(*@{\ldots}@*)]
-\end{lstlisting}
-\vspace{10ex}
+$  let hosts = ["foo.com", "bar.org"] in
+  lists.fold (fun val acc =>
+    val ++ "," ++ acc) hosts last in
+
+let makeHost = fun server ext =>
+  server ++ "." ++ ext in
+
+catHosts (makeHost "google")$
+\end{lstlisting}
     \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
@@
 This type of contracts are present in many different languages,
-for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
+for instance, the Eiffel programming language~\cite{meyer1987eiffel}, the precursor
 of the Design by Contract philosophy, makes it possible to assert
 these kinds of expression as pre- and post-conditions on
-functions and as invariants on classes\cite{EiffelDesignByContract}.
+functions and as invariants on classes~\cite{EiffelDesignByContract}.
 
@@
 like higher order contracts or a lightweight take on union
-and intersection contracts\cite{RacketContracts}.
+and intersection contracts~\cite{RacketContracts}.
 
@@
 
-Let us now consider union and intersection contracts, before we
-explain, in Section~\ref{sec:issues-sem} how they can compromise the
-properties that we described in Section~\ref{sec:feat-lang}.
+Let us now consider union and intersection contracts, before we explain in
+Section~\ref{sec:issues-sem} how they can compromise the properties that we
+described in Section~\ref{sec:feat-lang}.
 
@@
                  -> List =
-  fun list => lists.cons list "01/01/2021"
-}
+  fun list => lists.cons "01/01/2021" list
 \end{lstlisting}
@@
 
-A natural - but naive - implementation of intersection contracts could be the
+A natural --- but naive --- implementation of intersection contracts could be the
 following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
@@
 tour de force, their algorithmic system is compatible with
-user-defined contract (see Section~\ref{sec:flat-and-inter}).
+user-defined contracts (see Section~\ref{sec:flat-and-inter}).
 
@@
 Another realization of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem}, is~\cite{RootCauseOfBlame}, which
+Section~\ref{sec:coinductive-sem} is given in \cite{RootCauseOfBlame}, which
 aims at simplifying the algorithmic system proposed
@@
 Enforcing pre- and post-conditions at runtime is a widely established practice.
-In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
+In their foundational paper~\cite{FindlerFelleisenHOContracts}, Findler and
 Felleisen introduce higher-order contracts, a principled approach to run-time
@@
 contracts. As for any gradual type system, they do have to implement
-casts~--~that are very close to contracts~--~for unions and
+casts~---~that are very close to contracts~---~for unions and
 intersections, but these casts

### 2021-07-19 paper.tex
Fix typos reported by reviewer 3

@@
 \begin{lstlisting}[language=nickel]
-let c = contracts.fromPred (fun f =>
+let C = contracts.fromPred (fun f =>
   f 0 == 0) in
-let g | (Str -> Str) @& c
+let g | (Str -> Str) @& C
       = fun x => x
@@
 \end{lstlisting}
-\caption{intersection and user defined contracts}
+\caption{Intersection and user defined contracts}
 \label{fig:inter-flat-contracts}
@@
 typed terms, dynamically typed terms, and first-class contracts. Higher-order
-contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
-and~\cite{RootCauseOfBlame} that this paper explorted extensively.
+contracts are the basis of the work of \cite{KeilThiemannUnionIntersection}
+and \cite{RootCauseOfBlame} that this paper explored extensively.
 
@@
 
-Castagna et. al \cite{gradualCastagna} introduce a gradual type system based on
+Castagna et al.~\cite{gradualCastagna} introduce a gradual type system based on
 so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types

### 2021-07-24 paper.tex
Dimoulas papers, reformatting
Revision: review of Castagna et al. (draft)

@@
 
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let C = contracts.fromPred (fun f =>
+  f 0 == 0) in
+let g | (Str -> Str) @& C
+      = fun x => x
+in g 0
+\end{lstlisting}
+\caption{Intersection and user defined contracts}
+\label{fig:inter-flat-contracts}
+\end{figure}
+
 However, their approach has a major drawback: it is not straightforwardly
@@
 handle user-defined contracts.
-
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let C = contracts.fromPred (fun f =>
-  f 0 == 0) in
-let g | (Str -> Str) @& C
-      = fun x => x
-in g 0
-\end{lstlisting}
-\caption{Intersection and user defined contracts}
-\label{fig:inter-flat-contracts}
-\end{figure}
 
@@
 
-Castagna et al.~\cite{gradualCastagna} introduce a gradual type system based on
-so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types
-feature unions, intersections, negation types together with a notion of
-subtyping. This work adheres to the \emph{static first school} (see
-~\cite{practiceTheoryGrad}) of gradual typing: the reason for gradual types is
-to allow for a less precise type information. It follows that their goals and
-constraints are somehow different, resulting in the absence of first-class
-contracts. As for any gradual type system, they do have to implement
-casts~---~that are very close to contracts~---~for unions and
-intersections, but these casts
-are neither visible nor available to the programmer. Obviously, no contracts
-mean no user-defined contracts as well.
+Castagna and Lanvin~\cite{CastagnaLanvinGradualUnionIntersection} introduce a
+gradual type system based on so-called set-theoretic
+types~\cite{SetTheoreticTypes}.  Set-theoretic types feature unions,
+intersections, negation types together with a notion of subtyping. This work
+adheres to the \emph{static first school} (see ~\cite{practiceTheoryGrad}) of
+gradual typing: the reason for gradual types is to allow for a less precise type
+information. It follows that their goals and constraints are somehow different,
+resulting in the absence of first-class contracts. As for any gradual type
+system, they do have to implement casts~---~that are very close to
+contracts~---~for unions and intersections, but these casts are neither visible
+nor available to the programmer. Obviously, no contracts mean no user-defined
+contracts as well.
+
+Castagna and Lanvin use abstract interpretation to derive the semantics of
+unions and intersections. In their own words, "the resulting definitions are
+quite technical and barely intuitive but they have the properties we seek for
+[\ldots]".  This makes any comparison with the co-inductive semantics of Kiel
+and Thiemann rather difficult. Castagna et al.~\cite{gradualCastagna} build on
+Castagna and Lanvin to add polymorphism in the picture. However, it is at the
+price of restricting the union and intersection part: it is not possible to
+assign intersection types to a function anymore.
 

### 2021-07-25 paper.tex
Revision: side-effects

@@
 
+Higher-order contracts described in Section~\ref{sub-sec:contracts} already
+introduce side-effects: semantically, the \nickel{blame} operator signaling
+contract violation is equivalent to non-termination (error reporting aside).
+But non-termination is a special kind of effect, if only because it preserves
+referential transparency. In a lazy language like Nickel,
+(\ref{eq:cbn-expansion}) still holds for all terms in the presence of
+non-termination. In a strict language (call-by-value), (\ref{eq:cbn-expansion})
+fails in all generality because the transformation may cause a previously
+unevaluated term to be suddenly evaluated.  However, (\ref{eq:cbn-expansion})
+holds given slight restrictions on the context $M$: for example, it holds as
+long as at least one occurrence of $x$ in $M$ is in scope (i.e. not inside a
+function). Beside, many languages out there already feature non-termination. For
+such langauges, adding contracts does not change what optimizations are valid.
+
 Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
-unions and intersections make (\ref{eq:cbn-expansion}) unsound \emph{even in a
-pure setting}, preventing the kind optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.
+unions and intersections \emph{breaks referential transparency}. Unions and
+intersections make (\ref{eq:cbn-expansion}) unsound even in a lazy language,
+preventing the kind optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} to fire in general. In a strict
+language, the restrictions to (\ref{eq:cbn-expansion}) outlined above are not
+sufficient anymore. If the effect of plain higher-order contracts behaves as
+non-termination, the effect of union and intersection contracts rather behaves
+as shared state.
 

### 2021-07-25 paper.tex
Revision: user-defined contracts compatibility

@@
 
-However appealing union and intersection contracts may be, they happen
-to be fundamentally incompatible with the desirable language
-features from Section~\ref{sec:feat-lang}. At least in their
-full-blown form: in Section~\ref{sec:issues-literature} we will
-discuss pragmatic restrictions of union and intersection contracts to
-recover some or all of the features.
+However appealing union and intersection contracts may be, they happen to be
+either hard to combine or even fundamentally incompatible with the desirable
+language features from Section~\ref{sec:feat-lang}. At least in their full-blown
+form: in Section~\ref{sec:issues-literature} we will discuss pragmatic
+restrictions of union and intersection contracts to recover some or all of the
+features.
 
@@
 
-The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on
-balance, a technically impressive realization of the coinductive
-semantics, though it is fairly complex and probably difficult to
+The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on balance, a
+technically impressive realization of the coinductive semantics that supports
+user-defined contracts, though it is fairly complex and probably difficult to
 implement efficiently.

### 2021-07-26 paper.tex
Revision: side-effects (part 2)

@@
 of \nickel{f 5} now depends on the previous execution and more specifically on
-any prior call to \nickel{f}. This behavior of union contracts break the
-property~\ref{eq:cbn-expansion} introduced in Section~\ref{sec:optimizations},
-that is required to perform CSE-like optimizations in all generality. The
-candidate example of Figure~\ref{fig:optimizations-let-floating-ex} in
-Section~\ref{sec:optimizations} can't be optimized in general.
+any prior call to \nickel{f}. This behavior of union contracts breaks
+referential transparency, as well as the property~\ref{eq:cbn-expansion}
+introduced in Section~\ref{sec:optimizations}, that is required to perform
+CSE-like optimizations.
 

### 2021-07-26 paper.tex
Hardcode no-editing-marks for usage with git-latexdiff

--- 2021-07-25 paper.tex
@@
   % TOGGLE ME to turn off all the commentary:
-  \InputIfFileExists{no-editing-marks}{
   \def\noeditingmarks{}
-  }
 
@@
 \subsection{Contracts}
-\label{sub-sec:contracts}
 
@@
 accurately. Contracts can respect laziness of dictionaries, and they
-can be used to add schema validation to functions as well (in fact
+can be use to add schema validation to functions as well (in fact
 functions were the original motivation for contracts).
@@
   let hosts = ["foo.com", "bar.org"] in
-  lists.fold (fun val acc =>
-    val ++ "," ++ acc) hosts last in
+  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in
 
@@
 let catHosts | Str -> Str = fun last =>
-$  let hosts = ["foo.com", "bar.org"] in
-  lists.fold (fun val acc =>
-    val ++ "," ++ acc) hosts last in
-
-let makeHost = fun server ext =>
-  server ++ "." ++ ext in
-
-catHosts (makeHost "google")$
-\end{lstlisting}
+  [(*@{\ldots}@*)]
+\end{lstlisting}
+\vspace{10ex}
     \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
@@
 This type of contracts are present in many different languages,
-for instance, the Eiffel programming language~\cite{meyer1987eiffel}, the precursor
+for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
 of the Design by Contract philosophy, makes it possible to assert
 these kinds of expression as pre- and post-conditions on
-functions and as invariants on classes~\cite{EiffelDesignByContract}.
+functions and as invariants on classes\cite{EiffelDesignByContract}.
 
@@
 like higher order contracts or a lightweight take on union
-and intersection contracts~\cite{RacketContracts}.
+and intersection contracts\cite{RacketContracts}.
 
@@
 
-Higher-order contracts described in Section~\ref{sub-sec:contracts} already
-introduce side-effects: semantically, the \nickel{blame} operator signaling
-contract violation is equivalent to non-termination (error reporting aside).
-But non-termination is a special kind of effect, if only because it preserves
-referential transparency. In a lazy language like Nickel,
-(\ref{eq:cbn-expansion}) still holds for all terms in the presence of
-non-termination. In a strict language (call-by-value), (\ref{eq:cbn-expansion})
-fails in all generality because the transformation may cause a previously
-unevaluated term to be suddenly evaluated.  However, (\ref{eq:cbn-expansion})
-holds given slight restrictions on the context $M$: for example, it holds as
-long as at least one occurrence of $x$ in $M$ is in scope (i.e. not inside a
-function). Beside, many languages out there already feature non-termination. For
-such langauges, adding contracts does not change what optimizations are valid.
-
 Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
-unions and intersections \emph{breaks referential transparency}. Unions and
-intersections make (\ref{eq:cbn-expansion}) unsound even in a lazy language,
-preventing the kind optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} to fire in general. In a strict
-language, the restrictions to (\ref{eq:cbn-expansion}) outlined above are not
-sufficient anymore. If the effect of plain higher-order contracts behaves as
-non-termination, the effect of union and intersection contracts rather behaves
-as shared state.
+unions and intersections make (\ref{eq:cbn-expansion}) unsound \emph{even in a
+pure setting}, preventing the kind optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.
 
@@
 
-Let us now consider union and intersection contracts, before we explain in
-Section~\ref{sec:issues-sem} how they can compromise the properties that we
-described in Section~\ref{sec:feat-lang}.
+Let us now consider union and intersection contracts, before we
+explain, in Section~\ref{sec:issues-sem} how they can compromise the
+properties that we described in Section~\ref{sec:feat-lang}.
 
@@
                  -> List =
-  fun list => lists.cons "01/01/2021" list
+  fun list => lists.cons list "01/01/2021"
+}
 \end{lstlisting}
@@
 \end{lstlisting}
-\caption{Equivalent programs with CSE applied}
+\caption{Equivalent programs, with inlining or CSE applied}
 \label{fig:optimized-programs}
@@
 
-A natural --- but naive --- implementation of intersection contracts could be the
+A natural - but naive - implementation of intersection contracts could be the
 following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
@@
 
-\begin{figure}[h]
-\begin{lstlisting}[language=nickel]
-let C = contracts.fromPred (fun f =>
-  f 0 == 0) in
-let g | (Str -> Str) @& C
-      = fun x => x
-in g 0
-\end{lstlisting}
-\caption{Intersection and user defined contracts}
-\label{fig:inter-flat-contracts}
-\end{figure}
-
 However, their approach has a major drawback: it is not straightforwardly
@@
 handle user-defined contracts.
+
+\begin{figure}[h]
+\begin{lstlisting}[language=nickel]
+let c = contracts.fromPred (fun f =>
+  f 0 == 0) in
+let g | (Str -> Str) @& c
+      = fun x => x
+in g 0
+\end{lstlisting}
+\caption{intersection and user defined contracts}
+\label{fig:inter-flat-contracts}
+\end{figure}
 
@@
 tour de force, their algorithmic system is compatible with
-user-defined contracts (see Section~\ref{sec:flat-and-inter}).
+user-defined contract (see Section~\ref{sec:flat-and-inter}).
 
@@
 Another realization of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem} is given in \cite{RootCauseOfBlame}, which
+Section~\ref{sec:coinductive-sem}, is~\cite{RootCauseOfBlame}, which
 aims at simplifying the algorithmic system proposed
@@
 
-Now, combining these two contexts as in Figure~\ref{fig:invalid-context} gives a context
+Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
 that doesn't satisfy \nickel{Num -> Num} nor \hbox{\nickel{Bool -> Bool}.}  According
@@
 Enforcing pre- and post-conditions at runtime is a widely established practice.
-In their foundational paper~\cite{FindlerFelleisenHOContracts}, Findler and
+In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
 Felleisen introduce higher-order contracts, a principled approach to run-time
@@
 typed terms, dynamically typed terms, and first-class contracts. Higher-order
-contracts are the basis of the work of \cite{KeilThiemannUnionIntersection}
-and \cite{RootCauseOfBlame} that this paper explored extensively.
+contracts are the basis of the work~\cite{KeilThiemannUnionIntersection}
+and~\cite{RootCauseOfBlame} that this paper explorted extensively.
 
@@
 
-Castagna and Lanvin~\cite{CastagnaLanvinGradualUnionIntersection} introduce a
-gradual type system based on so-called set-theoretic
-types~\cite{SetTheoreticTypes}.  Set-theoretic types feature unions,
-intersections, negation types together with a notion of subtyping. This work
-adheres to the \emph{static first school} (see ~\cite{practiceTheoryGrad}) of
-gradual typing: the reason for gradual types is to allow for a less precise type
-information. It follows that their goals and constraints are somehow different,
-resulting in the absence of first-class contracts. As for any gradual type
-system, they do have to implement casts~---~that are very close to
-contracts~---~for unions and intersections, but these casts are neither visible
-nor available to the programmer. Obviously, no contracts mean no user-defined
-contracts as well.
-
-Castagna and Lanvin use abstract interpretation to derive the semantics of
-unions and intersections. In their own words, "the resulting definitions are
-quite technical and barely intuitive but they have the properties we seek for
-[\ldots]".  This makes any comparison with the co-inductive semantics of Kiel
-and Thiemann rather difficult. Castagna et al.~\cite{gradualCastagna} build on
-Castagna and Lanvin to add polymorphism in the picture. However, it is at the
-price of restricting the union and intersection part: it is not possible to
-assign intersection types to a function anymore.
+Castagna et. al \cite{gradualCastagna} introduce a gradual type system based on
+so-called set-theoretic types~\cite{SetTheoreticTypes}.  Set-theoretic types
+feature unions, intersections, negation types together with a notion of
+subtyping. This work adheres to the \emph{static first school} (see
+~\cite{practiceTheoryGrad}) of gradual typing: the reason for gradual types is
+to allow for a less precise type information. It follows that their goals and
+constraints are somehow different, resulting in the absence of first-class
+contracts. As for any gradual type system, they do have to implement
+casts~--~that are very close to contracts~--~for unions and
+intersections, but these casts
+are neither visible nor available to the programmer. Obviously, no contracts
+mean no user-defined contracts as well.
 

### 2021-07-26 paper.tex
Hardcode no-editing-marks to make it work with git-latexdiff
Revision: small precisions

--- 2021-07-26 paper.tex
@@
 long as at least one occurrence of $x$ in $M$ is in scope (i.e. not inside a
-function). Beside, many languages out there already feature non-termination. For
-such langauges, adding contracts does not change what optimizations are valid.
+function). Beside, many languages out there (including Nickel) already feature
+non-termination. For such languages, adding contracts does not change what
+optimizations are valid.
 

### 2021-07-26 paper.tex
Smallfix

@@
 \begin{figure}[h]
-\begin{lstlisting}[language=JavaScript]
+\begin{lstlisting}[language=Nickel]
 let Date = {day | Num, month | Num, year | Num} in

### 2021-07-26 paper.tex
Fix intersection example on lists

@@
 Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
-function \nickel{appendDate} append an element to \nickel{list} whose type is
+function \nickel{appendDate} appends an element to \nickel{list} whose type is
 only known to be in union of two lists, each using a different representation.
-Both alternatives support falling back to a simple string for unparsed dates.
-Faced with these two possibilities, \nickel{appendDate} can only append a
-value which fits both types: this is precisely the intersection
-\nickel{(Date @| Str) @& (DateWeek @| Str)}, that is, \nickel{Str}.
+Because the return type is the same as the input type, \nickel{appendDate} must
+preserve this representation (\nickel{Date} and \nickel{DateWeek}
+cannot be mixed in a same list).  Both alternatives support falling back to
+a simple string for unparsed dates.  Faced with these two possibilities,
+\nickel{appendDate} can only append a value which fits both types: this is
+precisely the intersection \nickel{(Date @| Str) @& (DateWeek @| Str)}, that
+is, \nickel{Str}.
 
@@
                  @| List (DateWeek @| Str))
-                 -> List =
+                 -> (List (Date @| Str)
+                 @| List (DateWeek @| Str)) =
   fun list => lists.cons "01/01/2021" list

### 2021-07-27 paper.tex
Revision: user defined contracts and currying

@@
 intersections make (\ref{eq:cbn-expansion}) unsound even in a lazy language,
-preventing the kind optimization of
+preventing the kind of optimization of
 Figure~\ref{fig:optimizations-let-floating-ex} to fire in general. In a strict
@@
 Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
-function \nickel{appendDate} appends an element to \nickel{list} whose type is
-only known to be in union of two lists, each using a different representation.
+function \nickel{appendDate} appends an element to \nickel{list}, whose type is
+only known to be the union of two lists, each using a different representation.
 Because the return type is the same as the input type, \nickel{appendDate} must
@@
 In order to give a precise definition to what values ought to satisfy
-union and intersection contracts, \cite{KeilThiemannUnionIntersection}
+union and intersection contracts,
+Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
 give a coinductively defined semantics inspired by union and
-intersection type systems. The key innovation of
-\cite{KeilThiemannUnionIntersection} is recognizing that giving a
+intersection type systems.
+The key innovation of their work is recognizing that giving a
 semantics to higher-order contracts requires defining not only what
@@
 
-Concretely, given a contract $C$, \cite{KeilThiemannUnionIntersection}
+Concretely, given a contract $C$,
+Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
 introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
@@
 evaluate to the pair \nickel{(1, 1)}. But the coinductive semantics
-of~\cite{KeilThiemannUnionIntersection} rejects it as a contract
-violation. This can be phrased pithily as the fact that the
+of this work rejects it as a contract violation.
+This can be phrased pithily as the fact that the
 coinductive semantics doesn't validate the property \nickel{A -> B @&
@@
 All of this still holds true of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem}. Nevertheless,
-\cite{KeilThiemannUnionIntersection} gives an algorithmic system which
+Section~\ref{sec:coinductive-sem}.
+Nevertheless, Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
+give an algorithmic system which
 is complete for their coinductive semantics. In a remarkable technical
@@
 
-To implement contract verification, \cite{KeilThiemannUnionIntersection} resorts to
+To implement contract verification,
+Keil and Thiemann~\cite{KeilThiemannUnionIntersection} resort to
 specific reduction rules for unions and intersections which perform this
@@
 
-The consequence is that~\cite{RootCauseOfBlame} doesn't prove their
-system complete for the coinductive semantics. It's probably just an
+The consequence is that Williams et al.~\cite{RootCauseOfBlame} don't
+prove their system complete for the coinductive semantics.
+It's probably just an
 oversight in the proof: we believe their system to be indeed complete
@@
 intersections, negation types together with a notion of subtyping. This work
-adheres to the \emph{static first school} (see ~\cite{practiceTheoryGrad}) of
+adheres to the \emph{static first school} (see~\cite{practiceTheoryGrad}) of
 gradual typing: the reason for gradual types is to allow for a less precise type

### 2021-07-29 paper.tex
A note on uncurrying as a work around for K&T overloading limitation

@@
 example in Figure~\ref{fig:intersection-distribution}: it could
-evaluate to the pair \nickel{(1, 1)}. But the coinductive semantics
-of this work rejects it as a contract violation.
+evaluate to the pair \nickel{(1, 1)}. But
+\citeauthor{KeilThiemannUnionIntersection}'s coinductive semantics
+rejects it as a contract violation.
 This can be phrased pithily as the fact that the

### 2021-07-29 paper.tex
Make citations uniform

@@
 union and intersection contracts,
-Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
+\citeauthor{KeilThiemannUnionIntersection}~\cite{KeilThiemannUnionIntersection}
 give a coinductively defined semantics inspired by union and
@@
 
-Concretely, given a contract $C$,
-Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
+Concretely, given a contract $C$, \citeauthor{KeilThiemannUnionIntersection}
 introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
-\rrbracket^-$ of values and contexts, respectively, satisfying the
-contract. They are defined by mutual induction and coinduction.
-
-This semantics has limited support for overloading. Consider the
-example in Figure~\ref{fig:intersection-distribution}: it could
-evaluate to the pair \nickel{(1, 1)}. But
-\citeauthor{KeilThiemannUnionIntersection}'s coinductive semantics
-rejects it as a contract violation.
-This can be phrased pithily as the fact that the
-coinductive semantics doesn't validate the property \nickel{A -> B @&
-  A -> C}$\simeq$\nickel{A -> (B @& C)}.
+\rrbracket^-$ of values and contexts, respectively, satisfying the contract.
+They are defined by mutual induction and coinduction.
+
+This semantics has limited support for overloading. Consider the example in
+Figure~\ref{fig:intersection-distribution}: it could evaluate to the pair
+\nickel{(1, 1)}. But \citeauthor{KeilThiemannUnionIntersection}'s coinductive
+semantics rejects it as a contract violation.  This can be phrased pithily as
+the fact that the coinductive semantics doesn't validate the property
+\nickel{A -> B @& A -> C}$\simeq$\nickel{A -> (B @& C)}.
 
@@
 All of this still holds true of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem}.
-Nevertheless, Keil and Thiemann~\cite{KeilThiemannUnionIntersection}
-give an algorithmic system which
-is complete for their coinductive semantics. In a remarkable technical
-tour de force, their algorithmic system is compatible with
-user-defined contracts (see Section~\ref{sec:flat-and-inter}).
-
-A key aspect of the approach of~\cite{KeilThiemannUnionIntersection}
+Section~\ref{sec:coinductive-sem}.  Nevertheless,
+\citeauthor{KeilThiemannUnionIntersection} give an algorithmic system which is
+complete for their coinductive semantics. In a remarkable technical tour de
+force, their algorithmic system is compatible with user-defined contracts (see
+Section~\ref{sec:flat-and-inter}).
+
+A key aspect of the approach of \citeauthor{KeilThiemannUnionIntersection}
 is to rewrite nested union and intersection contracts into a
@@
 
-To implement contract verification,
-Keil and Thiemann~\cite{KeilThiemannUnionIntersection} resort to
-specific reduction rules for unions and intersections which perform this
-rewriting on the fly. This aspect is critiqued
-in~\cite{RootCauseOfBlame}:
-``the monitoring semantics for
+To implement contract verification, \citeauthor{KeilThiemannUnionIntersection}
+resort to specific reduction rules for unions and intersections which perform
+this rewriting on the fly. This aspect is critiqued in
+\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}: ``the monitoring semantics for
 contracts of intersection and union types given by Keil and Thiemann are not
@@
 
-The algorithmic system of~\cite{KeilThiemannUnionIntersection} is, on balance, a
-technically impressive realization of the coinductive semantics that supports
-user-defined contracts, though it is fairly complex and probably difficult to
-implement efficiently.
+The algorithmic system of \citeauthor{KeilThiemannUnionIntersection}
+is, on balance, a technically impressive realization of the coinductive
+semantics that supports user-defined contracts, though it is fairly complex and
+probably difficult to implement efficiently.
 
@@
 Another realization of the coinductive semantics described in
-Section~\ref{sec:coinductive-sem} is given in \cite{RootCauseOfBlame}, which
-aims at simplifying the algorithmic system proposed
-by~\cite{KeilThiemannUnionIntersection} and described in
+Section~\ref{sec:coinductive-sem} is given in
+\citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}, which aims at simplifying
+the algorithmic system proposed by Keil and Thiemann and described in
 Section~\ref{sec:keil-thiemann}.
 
-A key ingredient of~\cite{RootCauseOfBlame} is to disallow user-defined
-contracts. This choice gives the authors more freedom in the quest of
-a more uniform operational semantics. This is sensible trade-off in
-the context of gradual typing à la TypeScript: the problem is to match
-contracts with static types, and user-defined contracts don't have a
-static type equivalent. On the other hand, the cost would probably not
-be worth it for a configuration language like Nickel.
+A key ingredient of \citeauthor{RootCauseOfBlame} is to disallow user-defined contracts. This
+choice gives the authors more freedom in the quest of a more uniform operational
+semantics. This is sensible trade-off in the context of gradual typing à la
+TypeScript: the problem is to match contracts with static types, and
+user-defined contracts don't have a static type equivalent. On the other hand,
+the cost would probably not be worth it for a configuration language like
+Nickel.
 
@@
 
-The consequence is that Williams et al.~\cite{RootCauseOfBlame} don't
-prove their system complete for the coinductive semantics.
-It's probably just an
-oversight in the proof: we believe their system to be indeed complete
-for the coinductive semantics. But it does speak to the intrinsic
-complexity of union and intersection contracts: it is remarkably easy
-to get details wrong. This difficulty contrasts with the standard
-framework of higher-order contracts where satisfaction is much more
-straightforward.
+The consequence is that \citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}
+don't prove their system complete for the coinductive semantics.  It's probably
+just an oversight in the proof: we believe their system to be indeed complete
+for the coinductive semantics. But it does speak to the intrinsic complexity of
+union and intersection contracts: it is remarkably easy to get details wrong.
+This difficulty contrasts with the standard framework of higher-order contracts
+where satisfaction is much more straightforward.
 
@@
 quite technical and barely intuitive but they have the properties we seek for
-[\ldots]".  This makes any comparison with the co-inductive semantics of Kiel
-and Thiemann rather difficult. Castagna et al.~\cite{gradualCastagna} build on
-Castagna and Lanvin to add polymorphism in the picture. However, it is at the
-price of restricting the union and intersection part: it is not possible to
-assign intersection types to a function anymore.
+[\ldots]".  This makes any comparison with the coinductive semantics of
+\citeauthor{KeilThiemannUnionIntersection} rather difficult. Castagna et
+al.~\cite{gradualCastagna} build on Castagna and Lanvin to add polymorphism in
+the picture. However, it is at the price of restricting the union and
+intersection part: it is not possible to assign intersection types to a function
+anymore.
 

### 2021-07-29 paper.tex
More citation fixes

@@
 
-The consequence is that \citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame}
-don't prove their system complete for the coinductive semantics.  It's probably
-just an oversight in the proof: we believe their system to be indeed complete
-for the coinductive semantics. But it does speak to the intrinsic complexity of
-union and intersection contracts: it is remarkably easy to get details wrong.
-This difficulty contrasts with the standard framework of higher-order contracts
-where satisfaction is much more straightforward.
+The consequence is that \citeauthor{RootCauseOfBlame} don't prove their system
+complete for the coinductive semantics.  It's probably just an oversight in the
+proof: we believe their system to be indeed complete for the coinductive
+semantics. But it does speak to the intrinsic complexity of union and
+intersection contracts: it is remarkably easy to get details wrong.  This
+difficulty contrasts with the standard framework of higher-order contracts where
+satisfaction is much more straightforward.
 
@@
 typed terms, dynamically typed terms, and first-class contracts. Higher-order
-contracts are the basis of the work of \cite{KeilThiemannUnionIntersection}
-and \cite{RootCauseOfBlame} that this paper explored extensively.
+contracts are the basis of the work of
+\citeauthor{KeilThiemannUnionIntersection}~\cite{KeilThiemannUnionIntersection}
+and \citeauthor{RootCauseOfBlame}~\cite{RootCauseOfBlame} that this paper
+explored extensively.
 

### 2021-09-09 paper.tex
Non review version
Arjun notes and pass

@@
 
+\acmConference[DLS 2021]{Dynamic Languages Symposium}{October
+17--22}{Chicago, Illinois, United States}
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \begin{abstract}
-Union and intersection types are a staple of gradually typed language such
+Union and intersection types are a staple of gradually typed languages such
 as TypeScript. While it's long been recognized that union and
@@
 illustration and motivation. At its core, Nickel is the JSON data
-model, augmented with abstraction mechanisms, and its conformed of:
+model, augmented with abstraction mechanisms, and it's conformed of:
 
 \begin{itemize}
-  \item dictionaries, written as:
+  \item dictionaries, written as
+    \footnote{Note that, unlike JSON, Nickel assigns values in
+    dictionaries using \nickel{=}, and not \nickel{:}.}:
 \begin{lstlisting}[frame=none,numbers=none,language=nickel]
@@
 \hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
-evaluated.
+evaluated, even if \nickel{nixpkgs} contained a
+\nickel{world} package as well.
 
@@
     \begin{lstlisting}[language=nickel]
-let catHosts | Str -> Str = fun last =>
-$  let hosts = ["foo.com", "bar.org"] in
+$let$ catHosts | Str -> Str $= fun last =>
+  let hosts = ["foo.com", "bar.org"] in
   lists.fold (fun val acc =>
@@
 appreciated sibling, intersection contracts) quite problematic.
+While usual contracts require only one witness to show the invalidation
+of a contract, the introduction of unions makes
+the number of witnesses not constantly bounded.
 
@@
 The Racket programming language also has a system to work with
-contracts, powerful enough to define flat contracts, and
+contracts, powerful enough to define user-defined contracts, and
 to compose them with other kinds of dynamic checks,
@@
 let subList = fun l1 l2 =>
-  lists.any (fun x => x == (lists.head l1)) l2
+  lists.any (fun x => x == (lists.head l1)) 
+            l2
   && subList (list.tail l1) l2
@@
   if index > lists.length list - 1 then
-    lists.elemAt (lists.length list - 1) list
+    lists.elemAt (lists.length list - 1) 
+                 list
   else
@@
 Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
-unions and intersections \emph{breaks referential transparency}. Unions and
+union and intersection contracts \emph{breaks referential transparency}. Unions and
 intersections make (\ref{eq:cbn-expansion}) unsound even in a lazy language,
@@
 as shared state.
+
+% TODO Will this be demostrated?
+% yes, I think is better untold, and the paragraph starts by
+% referencing section 5
 
@@
 contract. \nickel{f} is violating this contract once again, as it neither maps
-positive numbers to positive numbers nor positive numbers to negative numbers.
+positive numbers to positive numbers nor positive numbers to non positive numbers.
 

### 2021-09-10 paper.tex
Small change in footnote
small text on nickel outcome

@@
     \footnote{Note that, unlike JSON, Nickel assigns values in
-    dictionaries using \nickel{=}, and not \nickel{:}.}:
+    dictionaries using \nickel{=}, to keep \nickel{:} for type
+annotations.}:
 \begin{lstlisting}[frame=none,numbers=none,language=nickel]
@@
 union and intersection contracts doesn't interact with the rest of the
-language. It's probably more manageable.
+language. It's probably more manageable, and it's the approach taken
+by Nickel to be able to logically combine contracts.
 

### 2021-09-14 paper.tex
Cut some text on effects, purity and CSE

@@
 
-Higher-order contracts described in Section~\ref{sub-sec:contracts} already
-introduce side-effects: semantically, the \nickel{blame} operator signaling
-contract violation is equivalent to non-termination (error reporting aside).
-But non-termination is a special kind of effect, if only because it preserves
-referential transparency. In a lazy language like Nickel,
-(\ref{eq:cbn-expansion}) still holds for all terms in the presence of
-non-termination. In a strict language (call-by-value), (\ref{eq:cbn-expansion})
-fails in all generality because the transformation may cause a previously
-unevaluated term to be suddenly evaluated.  However, (\ref{eq:cbn-expansion})
-holds given slight restrictions on the context $M$: for example, it holds as
-long as at least one occurrence of $x$ in $M$ is in scope (i.e. not inside a
-function). Beside, many languages out there (including Nickel) already feature
-non-termination. For such languages, adding contracts does not change what
-optimizations are valid.
-
 Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
-union and intersection contracts \emph{breaks referential transparency}. Unions and
-intersections make (\ref{eq:cbn-expansion}) unsound even in a lazy language,
-preventing the kind of optimization of
-Figure~\ref{fig:optimizations-let-floating-ex} to fire in general. In a strict
-language, the restrictions to (\ref{eq:cbn-expansion}) outlined above are not
-sufficient anymore. If the effect of plain higher-order contracts behaves as
-non-termination, the effect of union and intersection contracts rather behaves
-as shared state.
-
-% TODO Will this be demostrated?
-% yes, I think is better untold, and the paragraph starts by
-% referencing section 5
+union and intersection contracts \emph{breaks referential transparency} and make
+(\ref{eq:cbn-expansion}) \emph{unsound}\footnote{(\ref{eq:cbn-expansion}) holds
+  even in a technically impure language like Nickel with non-termination and
+  higher-order contracts because of lazy evaluation. This is not the case in a strict language,
+  although in that case, (\ref{eq:cbn-expansion}) still holds more often without
+  unions and intersections than with.}, preventing the kind of optimization of
+Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.
 

### 2021-09-15 paper.tex

@@
 
+\subsection{Context sensitive contracts}
+
+The work by Dimoulas et al.~\cite{dimoulasESOP,dimoulasPOPL} presents
+a complete study on systems that may execute (and potentially invalidate)
+contracts inside other contracts, in particular through the study
+of dependent function contracts.
+Their ideas have the potential to help solve the problems outlined
+on Sections \ref{sec:flat-and-inter} and \ref{sec:keil-thiemann},
+however, we consider it's not clear how to leverage their techniques
+for intersections.
+
+As a first obstacle, it's worth noting that their system is intended
+to get stuck upon the checking of a contract in a context that shouldn't
+be checked in, while to keep the semantics of unions and intersections
+valid, we'd expect for that check to get skipped, and for execution
+to continue.
+
+% > Also worth noting is that Dimoulas's work deals extensively with these kinds
+% > of contracts and there is no mention of it in this paper. His POPL 2011 and
+% > ESOP 2012 papers are relevant to such contracts.
+
+% Regarding Dimoulas's papers, their work presents a solution
+% for contracts that may, arbitrarily, break other contracts in their
+% evaluation. Although this looks related to this problem with
+% intersection and user-defined contracts, it is not clear how to
+% leverage Dimoulas's techniques for intersection in practice.
+
+% It's worth noting, for instance, that Dimoulas's systems behave by getting stuck
+% in the case an undesired event happens (a contract fails because of external,
+% semantically unrelated, values flow in). With unions and intersections as
+% developed by Keil and Thiemann, the expected outcome is rather to not execute a
+% contract under the wrong context.
+
 \section{Conclusion}

### 2021-09-15 paper.tex
Update paper.tex
Co-authored-by: Arnaud Spiwack <arnaud.spiwack@tweag.io>

@@
 (\ref{eq:cbn-expansion}) \emph{unsound}\footnote{(\ref{eq:cbn-expansion}) holds
-  even in a technically impure language like Nickel with non-termination and
-  higher-order contracts because of lazy evaluation. This is not the case in a strict language,
-  although in that case, (\ref{eq:cbn-expansion}) still holds more often without
-  unions and intersections than with.}, preventing the kind of optimization of
+  in Nickel despite non-termination and contract-checking errors
+  because of lazy evaluation. In strict languages (\ref{eq:cbn-expansion}) needs to be restricted, nevertheless
+  union and intersection contracts make it worse.}, preventing the kind of optimization of
 Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.

### 2021-09-15 paper.tex
Update paper.tex
Co-authored-by: Arnaud Spiwack <arnaud.spiwack@tweag.io>

@@
 contract. \nickel{f} is violating this contract once again, as it neither maps
-positive numbers to positive numbers nor positive numbers to non positive numbers.
+positive numbers to positive numbers nor positive numbers to nonpositive numbers.
 

### 2021-09-15 paper.tex
Update paper.tex
Co-authored-by: Arnaud Spiwack <arnaud.spiwack@tweag.io>

@@
 illustration and motivation. At its core, Nickel is the JSON data
-model, augmented with abstraction mechanisms, and it's conformed of:
+model, augmented with abstraction mechanisms, and it comprises:
 
@@
 union and intersection contracts doesn't interact with the rest of the
-language. It's probably more manageable, and it's the approach taken
-by Nickel to be able to logically combine contracts.
+language. It's probably more manageable, and it's the approach that
+we are currently considering for Nickel.
 

### 2021-09-15 paper.tex
Use microtype package
Fill-in acm information [INCOMPLETE]
Remove duplicate command, move bibliographystyle

@@
-\documentclass[sigplan,10pt]{acmart}
+\documentclass[sigplan,screen,10pt]{acmart}
 %\usepackage{amssymb}
@@
 \usepackage{hyperref}
+\usepackage{microtype}
 
@@
 
-\acmConference[DLS 2021]{Dynamic Languages Symposium}{October
-17--22}{Chicago, Illinois, United States}
-
-\bibliographystyle{ACM-Reference-Format}
-
 \begin{abstract}
@@
 \email{arnaud.spiwack@tweag.io}
+
+%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
+%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
+%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
+%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
+%%%
+%%% The following is specific to DLS '21 and the paper
+%%% 'Union and Intersection Contracts Are Hard, Actually'
+%%% by Teodoro Freund, Yann Hamdaoui, and Arnaud Spiwack.
+%%%
+\setcopyright{ACMUNKNOWN}
+\acmPrice{15.00}
+\acmDOI{10.1145/3486602.3486767}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{splashws21dlsmain-p31-p}
+\acmISBN{978-1-4503-9105-4/21/10}
+\acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
+\acmBooktitle{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
 

### 2021-09-16 paper.tex
Capitalize headings
Use pbalance for balancing, update copyright information

@@
 \usepackage{microtype}
+\usepackage{pbalance}
 
@@
 
-%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
-%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
-%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
-%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
-%%%
 %%% The following is specific to DLS '21 and the paper
@@
 %%% by Teodoro Freund, Yann Hamdaoui, and Arnaud Spiwack.
-%%%
-\setcopyright{ACMUNKNOWN}
+\copyrightyear{2021}
+\acmYear{2021}
+\setcopyright{acmlicensed}\acmConference[DLS '21]{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
+\acmBooktitle{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
 \acmPrice{15.00}
 \acmDOI{10.1145/3486602.3486767}
-\acmYear{2021}
-\copyrightyear{2021}
-\acmSubmissionID{splashws21dlsmain-p31-p}
 \acmISBN{978-1-4503-9105-4/21/10}
+
 \acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
@@
 
-\section{A typology of language features}
+\section{A Typology Of Language Features}
 \label{sec:feat-lang}
@@
 
-\subsection{Referential transparency}
+\subsection{Referential Transparency}
 \label{sec:optimizations}
@@
 
-\section{Union \& intersection}
+\section{Union \& Intersection}
 \label{sec:union-inter}
@@
 
-\subsection{Union contracts as a side-effect}
+\subsection{Union Contracts as a Side-Effect}
 
@@
 
-\subsection{Intersection with user-defined contracts}
+\subsection{Intersection With User-Defined Contracts}
 \label{sec:flat-and-inter}
@@
 
-\section{Pragmatic trade-offs}
+\section{Pragmatic Trade-Offs}
 \label{sec:issues-literature}
@@
 
-\subsection{Monitoring properties}
+\subsection{Monitoring Properties}
 \label{sec:will-morr-wadl}
@@
 
-\section{(More) Related work}
+\section{(More) Related Work}
 \label{sec:related-work}
@@
 
-\subsection{Higher-order contracts}
+\subsection{Higher-Order Contracts}
 
@@
 
-\subsection{Unions and intersections in gradual typing}
+\subsection{Unions and Intersections in Gradual Typing}
 
@@
 
-\subsection{Context sensitive contracts}
+\subsection{Context Sensitive Contracts}
 

### 2021-09-16 paper.tex
Remove line numbers in listings
ACM author instructions require that nothing overlaps with margins,
which line numbers did.

@@
   framesep=4pt, % quarter circle size of the round corners
-  numbers=left, % show line numbers at the left
-  numberstyle=\tiny\ttfamily, % style of the line numbers
+  numbers=none, % show line numbers at the left
+  % numberstyle=\tiny\ttfamily, % style of the line numbers
   escapeinside={(*@}{@*)} % escape sequence to insert latex
@@
 \lstset{
-  escapeinside={(*@}{@*)} % escape sequence to insert latex
+  escapeinside={(*@}{@*)}, % escape sequence to insert latex
+  numbers=none % for the camera-ready version: numbers overlap with margins
 }

### 2021-09-16 paper.tex
Fix duplicate commands

@@
 \acmYear{2021}
-\setcopyright{acmlicensed}\acmConference[DLS '21]{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
+\setcopyright{acmlicensed}
+\acmConference[DLS '21]{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
 \acmBooktitle{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
@@
 \acmISBN{978-1-4503-9105-4/21/10}
-
-\acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
-\acmBooktitle{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
 

### 2021-09-16 paper.tex
Fix ACM metadata

--- 2021-09-16 paper.tex
@@
 %%% by Teodoro Freund, Yann Hamdaoui, and Arnaud Spiwack.
-\copyrightyear{2021}
-\acmYear{2021}
-\setcopyright{acmlicensed}\acmConference[DLS '21]{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
-\acmBooktitle{Proceeding of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}
+\setcopyright{acmlicensed}
 \acmPrice{15.00}
 \acmDOI{10.1145/3486602.3486767}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{splashws21dlsmain-p31-p}
 \acmISBN{978-1-4503-9105-4/21/10}
-
 \acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}

### 2021-09-16 paper.tex
Add negative margins to listings to avoid reaching into margins

@@
   escapeinside={(*@}{@*)}, % escape sequence to insert latex
-  numbers=none % for the camera-ready version: numbers overlap with margins
+  numbers=none, % for the camera-ready version: numbers overlap with margins
+  % Avoid overlap with margins
+  framexleftmargin = -2pt,
+  framexrightmargin = -2pt,
+  framextopmargin = -2pt,
+  framexbottommargin = -2pt
 }
+
 \newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}

### 2021-09-16 paper.tex
Fix some overfull hboxes

@@
 are \emph{lazy}, that is, the values are only evaluated when
-explicitly required. For instance, when writing
+explicitly required. For instance, when evaluating the expression
 \hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
@@
 contract. \nickel{f} is violating this contract once again, as it neither maps
-positive numbers to positive numbers nor positive numbers to nonpositive numbers.
+all positive numbers to positive numbers nor to nonpositive numbers.
 

### 2021-09-16 paper.tex
More overfull hbox + typography

@@
 
-This behavior indicates that union contracts introduce side-effects. The result
+This behavior shows that union contracts introduce side-effects. The result
 of \nickel{f 5} now depends on the previous execution and more specifically on
@@
 
-However, their approach has a major drawback: it is not straightforwardly
+However, this shared-state approach has a major drawback: it isn't straightforwardly
 compatible with user-defined contracts (introduced in Section
@@
 complete for their coinductive semantics. In a remarkable technical tour de
-force, their algorithmic system is compatible with user-defined contracts (see
+force, their algorithmic system allows for user-defined contracts (see
 Section~\ref{sec:flat-and-inter}).
@@
 Castagna and Lanvin use abstract interpretation to derive the semantics of
-unions and intersections. In their own words, "the resulting definitions are
+unions and intersections. In their words: ``the resulting definitions are
 quite technical and barely intuitive but they have the properties we seek for
-[\ldots]".  This makes any comparison with the coinductive semantics of
+[\ldots]''.  This makes any comparison with the coinductive semantics of
 \citeauthor{KeilThiemannUnionIntersection} rather difficult. Castagna et

### 2021-09-16 paper.tex
Final overfull hboxes

@@
 Castagna and Lanvin~\cite{CastagnaLanvinGradualUnionIntersection} introduce a
-gradual type system based on so-called set-theoretic
+gradual type system based on set-theoretic
 types~\cite{SetTheoreticTypes}.  Set-theoretic types feature unions,

### 2021-09-16 paper.tex
Update paper.tex
Co-authored-by: Arnaud Spiwack <arnaud.spiwack@tweag.io>

@@
 on Sections \ref{sec:flat-and-inter} and \ref{sec:keil-thiemann},
-however, we consider it's not clear how to leverage their techniques
+however, it isn't clear how to leverage their techniques
 for intersections.

### 2021-09-16 paper.tex
small fix + removed paragraph

@@
 of a contract, the introduction of unions makes
-the number of witnesses not constantly bounded.
+the number of witnesses not bounded.
 
@@
 
-As a first obstacle, it's worth noting that their system is intended
-to get stuck upon the checking of a contract in a context that shouldn't
-be checked in, while to keep the semantics of unions and intersections
-valid, we'd expect for that check to get skipped, and for execution
-to continue.
+% As a first obstacle, it's worth noting that their system is intended
+% to get stuck upon the checking of a contract in a context that shouldn't
+% be checked in, while to keep the semantics of unions and intersections
+% valid, we'd expect for that check to get skipped, and for execution
+% to continue.
 

### 2021-09-28 paper.tex
Implement corrections for camera-ready version

@@
   breaklines=true, % wrap lines if they don't fit
-  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
-    frameround=tttt, % make the frame round at all four corners
-  framesep=4pt, % quarter circle size of the round corners
+  % frame=trbl, % draw a frame at the top, right, left and bottom of the listing
+    % frameround=tttt, % make the frame round at all four corners
+  % framesep=4pt, % quarter circle size of the round corners
   numbers=none, % show line numbers at the left
@@
   % Avoid overlap with margins
-  framexleftmargin = -2pt,
-  framexrightmargin = -2pt,
-  framextopmargin = -2pt,
-  framexbottommargin = -2pt
+  % framexleftmargin = -2pt,
+  % framexrightmargin = -2pt,
+  % framextopmargin = -2pt,
+  % framexbottommargin = -2pt
 }
@@
 
-\section{A Typology Of Language Features}
+\section{A Typology of Language Features}
 \label{sec:feat-lang}
@@
 
-\subsection{Intersection With User-Defined Contracts}
+\subsection{Intersection with User-Defined Contracts}
 \label{sec:flat-and-inter}
